### 2017-05-19 paper_BiocF1000/paper_BiocF1000.tex
switch paper to Rmarkdown for F1000

\documentclass[9pt,a4paper,]{extarticle}

\usepackage{f1000_styles}

\usepackage[pdfborder={0 0 0}]{hyperref}

\usepackage[numbers]{natbib}

%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

% disable code chunks background
%\renewenvironment{Shaded}{}{}

% disable section numbers
\setcounter{secnumdepth}{0}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\begin{document}
\pagestyle{front}

\title{\emph{F1000Research} Software Tool Article Template}

\author[1]{Nima Hejazi}
\author[2]{Alan E. Hubbard}
\author[2]{Mark J. van der Laan}
\affil[1]{Address of author1}
\affil[2]{Address of author2}

\maketitle
\thispagestyle{front}

\begin{abstract}
Abstracts should be up to 300 words and provide a succinct summary of
the article. Although the abstract should explain why the article might
be interesting, care should be taken not to inappropriately
over-emphasise the importance of the work described in the article.
Citations should not be used in the abstract, and the use of
abbreviations should be minimized.
\end{abstract}

\section*{Keywords}
Please list up to eight keywords to help readers interested in your
article find it more easily.

\clearpage
\pagestyle{main}

\section{Introduction}\label{introduction}

The introduction provides context as to why the software tool was
developed and what need it addresses. It is good scholarly practice to
mention previously developed tools that address similar needs, and why
the current tool is needed.

\section{Methods}\label{methods}

\subsection{Implementation}\label{implementation}

For software tool papers, this section should address how the tool works
and any relevant technical details required for implementation of the
tool by other developers.

\subsection{Operation}\label{operation}

This part of the methods should include the minimal system requirements
needed to run the software and an overview of the workflow for the tool
for users of the tool.

\section{Results }\label{results}

This section is only required if the paper includes novel data or
analyses, and should be written as a traditional results section.

\section{Use Cases }\label{use-cases}

This section is required if the paper does not include novel data or
analyses. Examples of input and output files should be provided with
some explanatory context. Any novel or complex variable parameters
should also be explained in sufficient detail to allow users to
understand and use the tool's functionality.

\section{Discussion }\label{discussion}

This section is only required if the paper includes novel data or
analyses, and should be written in the same style as a traditional
discussion section. Please include a brief discussion of allowances made
(if any) for controlling bias or unwanted sources of variability, and
the limitations of any novel datasets.

\section{Conclusions }\label{conclusions}

This section is only required if the paper includes novel data or
analyses, and should be written as a traditional conclusion.

\section{Summary }\label{summary}

This section is required if the paper does not include novel data or
analyses. It allows authors to briefly summarize the key points from the
article.

\section{Data availability }\label{data-availability}

Please add details of where any datasets that are mentioned in the
paper, and that have not have not previously been formally published,
can be found. If previously published datasets are mentioned, these
should be cited in the references, as per usual scholarly conventions.

\section{Software availability}\label{software-availability}

This section will be generated by the Editorial Office before
publication. Authors are asked to provide some initial information to
assist the Editorial Office, as detailed below.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  URL link to where the software can be downloaded from or used by a
  non-coder (AUTHOR TO PROVIDE; optional)
\item
  URL link to the author's version control system repository containing
  the source code (AUTHOR TO PROVIDE; required)
\item
  Link to source code as at time of publication (\emph{F1000Research} TO
  GENERATE)
\item
  Link to archived source code as at time of publication
  (\emph{F1000Research} TO GENERATE)
\item
  Software license (AUTHOR TO PROVIDE; required)
\end{enumerate}

\section{Author contributions}\label{author-contributions}

In order to give appropriate credit to each author of an article, the
individual contributions of each author to the manuscript should be
detailed in this section. We recommend using author initials and then
stating briefly how they contributed.

\section{Competing interests}\label{competing-interests}

All financial, personal, or professional competing interests for any of
the authors that could be construed to unduly influence the content of
the article must be disclosed and will be displayed alongside the
article. If there are no relevant competing interests to declare, please
add the following: `No competing interests were disclosed'.

\section{Grant information}\label{grant-information}

Please state who funded the work discussed in this article, whether it
is your employer, a grant funder etc. Please do not list funding that
you have that is not relevant to this specific piece of research. For
each funder, please state the funder's name, the grant number where
applicable, and the individual to whom the grant was assigned. If your
work was not funded by any grants, please include the line: `The
author(s) declared that no grants were involved in supporting this
work.'

\section{Acknowledgments}\label{acknowledgments}

This section should acknowledge anyone who contributed to the research
or the article but who does not qualify as an author based on the
criteria provided earlier (e.g.~someone or an organization that provided
writing assistance). Please state how they contributed; authors should
obtain permission to acknowledge from all those mentioned in the
Acknowledgments section.

Please do not list grant funding in this section.

\section{USING R MARKDOWN}\label{using-r-markdown}

Some examples of commonly used markdown syntax are listed below, to help
you get started.

\subsection{Cross-references}\label{cross-references}

For portability between different output formats, use the syntax
introduced by \emph{bookdown}, such as
\texttt{(\textbackslash{}\#label)} for labels and
\texttt{\textbackslash{}@ref(label)} for cross-references. The following
sections provide examples of referencing tables, figures, and equations.

\subsection{Citations}\label{citations}

You can include refences in a standard Bibtex file. The name of this
file is given in the header of the markdown document (in our case it is
\emph{sample.bib}). References to entries in the Bibtex file are made
using square brackets and use an @ plus the key for the entry you are
referencing {[}@Smith:2012qr{]}. You can combine multiple entries by
separating them with a semi-colon {[}@Smith:2012qr; @Smith:2013jd{]}.

\subsection{Code chunks}\label{code-chunks}

You can embed an R code chunk like this:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\DecValTok{1}\OperatorTok{:}\DecValTok{10}
\NormalTok{x}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1]  1  2  3  4  5  6  7  8  9 10
\end{verbatim}

If you specify a figure caption to a code chunk using the chunk option
\texttt{fig.cap}, the plot will be automatically labeled and numbered.
The figure label is generated from the label of the code chunk by
prefixing it with \texttt{fig:}, e.g., see Figure \ref{fig:plot}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(x)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{paper_BiocF1000_files/figure-latex/plot-1.pdf}
\caption{\label{fig:plot}A caption to our sample plot.}
\end{figure}

\subsection{Tables}\label{tables}

Markdown syntax tends to lack some of the more sophisticated formatting
features available in LaTeX, so you may need to edit the tables later to
get the desired format.

\begin{table}[htbp]
\caption{Caption to table.}
\centering
\begin{tabledata}{@{}lll@{}}
\header First name & Last Name & Grade\\
\row John & Doe & 7.5\\
\row Richard & Miles & 2\\
\end{tabledata}
\end{table}

Just like figures, tables with captions will also be numbered and can be
referenced. Captions are entered as a paragraph beginning with the
string ``Table:'' (or just ``:''), which may appear either before or
after the table. A label for the table should appear in the beginning of
the caption in the form of \texttt{(\textbackslash{}\#tab:label)}, e.g.,
see Table \ref{tab:table}.

\begin{table}[htbp]
\caption{\label{tab:table} A table with text justification.}
\centering
\begin{tabledata}{@{}lcr@{}}
\header First name & Last Name & Grade\\
\row John & Doe & 7.5\\
\row Richard & Miles & 2\\
\end{tabledata}
\end{table}

\subsection{Figures}\label{figures}

You can include static figures (i.e.~no generated by code) using the
\texttt{include\_graphics()} function from the \textbf{knitr} package,
in a standard code chunk.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{'frog.jpg'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{frog}

}

\caption{Your figure legend goes here; it should be succinct, while still explaining all symbols and abbreviations.}\label{fig:frog-picture}
\end{figure}

You can again use the \texttt{fig.cap} option to provide the figure
caption, and reference the image based on the code chunk label. You can
also use options such as \texttt{fig.align} and \texttt{fig.width} to
adjust the position and size of the image within the final document,
e.g.~Figure \ref{fig:frog-picture} is a frog.

Alternatively, you can use the standard markdown syntax like so:

\begin{figure}
\centering
\includegraphics[width=0.25000\textwidth]{frog.jpg}
\caption{This is a smaller version of the same picture, inserted using
the standard markdown syntax}
\end{figure}

Please give figures appropriate filenames, e.g.: figure1.pdf,
figure2.png.

Figure legends should briefly describe the key messages of the figure
such that the figure can stand alone from the main text. However, all
figures should also be discussed in the article text. Each legend should
have a concise title of no more than 15 words. The legend itself should
be succinct, while still explaining all symbols and abbreviations. Avoid
lengthy descriptions of methods.

For any figures reproduced from another publication (as long as
appropriate permission has been obtained from the copyright holder
---see under the heading `Submission'), please include a line in the
legend to state that: `This figure has been reproduced with kind
permission from {[}include original publication citation{]}'.

\subsection{Mathematics}\label{mathematics}

You can use LaTeX syntax to typeset mathematical expressions. Let
\(X_1, X_2, \ldots, X_n\) be a sequence of independent and identically
distributed random variables with \(\text{E}[X_i] = \mu\) and
\(\text{Var}[X_i] = \sigma^2 < \infty\), and let
\[S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i\] denote their mean. Then as \(n\)
approaches infinity, the random variables \(\sqrt{n}(S_n - \mu)\)
converge in distribution to a normal \(\mathcal{N}(0, \sigma^2)\).

To number and refer to equations, put them in the equation environments
and assign labels to them, as for Equation \eqref{eq:binom}.

\begin{equation}
  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
  \label{eq:binom}
\end{equation}

\subsection{Lists}\label{lists}

You can make ordered lists

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Like this,
\item
  and like this.
\end{enumerate}

or bullet points

\begin{itemize}
\item
  Like this,
\item
  and like this.

  \begin{itemize}
  \item
    Use indentation

    \begin{itemize}
    \item
      for sub-items
    \end{itemize}
  \end{itemize}
\end{itemize}

\end{document}

### 2017-07-14 README.Rmd
add more pkg basics

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`methyvim`

[![Travis-CI Build Status](https://travis-ci.org/nhejazi/methyvim.svg?branch=master)](https://travis-ci.org/nhejazi/methyvim)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/nhejazi/methyvim?branch=master&svg=true)](https://ci.appveyor.com/project/nhejazi/methyvim)
[![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/methyvim/master.svg)](https://codecov.io/github/nhejazi/methyvim?branch=master)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Nonparametric variable importance analysis for the assessment of differential
> methylation

__Author:__ [Nima Hejazi](http://nimahejazi.org)

`methyvim` is an R package that provides facilities for differential methylation
analysis based on _variable importance measures_ (VIMs), a class of statistical
target parameters that arise in causal inference. The statistical techniques
implemented rely on the use of targeted minimum loss-based estimation (TMLE) to
assess two particular VIMs: (1) the "local" average treatment effect for
discrete exposures, and (2) a nonparametric variable importance measure for
continuous exposures. Within this framework, these methods allow differential
methylation effects to be estimated in an assumption-free manner, on a
pre-screened set of genomic sites measured by DNA methylation assays. As the
statistical algorithm uses a multi-stage approach, multiple testing corrections
are made using a modified marginal Benjamini & Hochberg step-up False Discovery
Rate controlling procedure for multi-stage analyses (FDR-MSA). In order to allow
the user significant flexibility with respect to the scientific questions posed,
the procedure estimates one of the appropriate VIMs for each of a reduced set of
genomic sites, using screening procedures to identify a reduced set from the
full data when a reduced set is not pre-specified. While making allowance for
the user to specify the reduced set of genomic sites to be tested (e.g., those
related to a particular gene of interest), this package comes equipped with
screening procedures to facilitate the initial reduction of genomic sites, the
two most noteworthy of these being a procedure based on the well-known R package
[`limma`](https://bioconductor.org/packages/release/bioc/html/limma.html) and an
extension of data-adaptive statistical target parameters for multiple testing
(implemented in the R package
[`DA.Test`](https://github.com/wilsoncai1992/data.adapt.multi.test)).

---

## Installation

<!--
For standard use, install from [Bioconductor](https://bioconductor.org):
```{r bioc-installation, eval = FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("methyvim")
```
-->

Install the most recent _stable release_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-master-installation, eval = FALSE}
devtools::install_github("nhejazi/methyvim")
```

To contribute, install the _development version_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-develop-installation, eval = FALSE}
devtools::install_github("nhejazi/methyvim", ref = "develop")
```

---

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/nhejazi/methyvim/issues).

---

## Contributions

It is our hope that `methyvim` will grow to be widely adopted as a tool for the
nonparametric assessment of variable importance in studies of differential
methylation. To that end, contributions are very welcome, though we ask that
interested contributors consult our [`contribution
guidelines`](https://github.com/nhejazi/methyvim/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

After using the `methyvim` R package, please cite it:

        @article{hejazi2017methyvim,
          doi = {},
          url = {},
          year  = {2017},
          month = {},
          publisher={Faculty of 1000 Ltd},
          volume = {},
          author = {Hejazi, Nima S. and Hubbard, Alan E. and {van der Laan},
                    Mark J.},
          title = {methyvim: Nonparametric Assessment of Variable Importance for
                   Differential Methylation Analysis},
          journal = {F1000Research}
        }

---

## References

* [Mark J. van der Laan & Sherri Rose. _Targeted Learning: Causal Inference for
    Observational and Experimental Data_,
    2011.](http://www.targetedlearningbook.com)

* [Antoine Chambaz, Pierre Neuvial, & Mark J. van der Laan. "Estimation of a
    non-parametric variable importance measure of a continuous exposure",
    _Electronic Journal of Statistics_,
    2012.](http://www.math-info.univ-paris5.fr/~chambaz/Papiers/chambazNeuvialvanderLaan_EJS2012.pdf)

* [Catherine Tuglus & Mark J. van der Laan. "FDR controlling procedures for
    multi-stage analyses", _UC Berkeley Division of Biostatistics Working
    Paper Series_, 2008.](http://biostats.bepress.com/ucbbiostat/paper239/)

* [Weixin Cai, Nima S. Hejazi, & Alan E. Hubbard. "Data-adaptive statistics for
    multiple testing in high-dimensional settings", _In preparation_,
    2017.](https://www.overleaf.com/5660573pjjrxh#/25678897/)

* [Gordon K. Smyth. "Linear models and empirical Bayes methods for assessing
    differential expression in microarray experiments." _Statistical
    Applications in Genetics and Molecular Biology_, 3(1),
    2004.](http://www.statsci.org/smyth/pubs/ebayes.pdf)

---

## License

&copy; 2017 [Nima S. Hejazi](http://nimahejazi.org), [Alan E.
Hubbard](http://sph.berkeley.edu/alan-hubbard), [Mark J. van der
Laan](https://www.stat.berkeley.edu/~laan/)

The contents of this repository are distributed under the MIT license. See file
`LICENSE` for details.

### 2017-07-14 R/methadapt.R

@@ -1,10 +1,18 @@
-#' constructor for class methadapt (subclass of SummarizedExperiment)
+#' Constructor for class methyvim
 #'
-#' @return ".methadapt" S4 object (subclass of "SummarizedExperiment") for DNA
-#'         methylation analysis using targeted minimum loss-based estimation.
+#' @return class \code{methyvim} object, sub-classed from SummarizedExperiment.
 #'
-.methadapt <- setClass(
-  "methadapt",
-  contains = "SummarizedExperiment",
-  slots = list(call = "call")
+#' @importFrom methods setClass
+#' @importFrom tibble data_frame
+#' @importClassesFrom SummarizedExperiment SummarizedExperiment
+#'
+#' @export .methyvim
+#' @exportClass methyvim
+#'
+.methyvim <- methods::setClass(
+       Class = "methyvim",
+       slots = list(call = "call",
+                    tmleOut = "data_frame",
+                    topTable = "data_frame"),
+       contains = "SummarizedExperiment"
 )

### 2017-07-14 paper_F1000Bioc/paper_BiocF1000.Rmd
add scripts for using toy data

---
title: "*F1000Research* Software Tool Article Template"
author: 
  - name: Nima Hejazi
    affiliation: Address of author1
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
#bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

# Introduction

The introduction provides context as to why the software tool was developed and what need it addresses.  It is good scholarly practice to mention previously developed tools that address similar needs, and why the current tool is needed.

# Methods
## Implementation
For software tool papers, this section should address how the tool works and any relevant technical details required for implementation of the tool by other developers.

## Operation
This part of the methods should include the minimal system requirements needed to run the software and an overview of the workflow for the tool for users of the tool.

# Results <!-- Optional - only if novel data or analyses are included -->
This section is only required if the paper includes novel data or analyses, and should be written as a traditional results section.

# Use Cases <!-- Optional - only if NO new datasets are included -->
This section is required if the paper does not include novel data or analyses. 
Examples of input and output files should be provided with some explanatory context.  Any novel or complex variable parameters should also be explained in sufficient detail to allow users to understand and use the tool's functionality.

# Discussion <!-- Optional - only if novel data or analyses are included -->
This section is only required if the paper includes novel data or analyses, and should be written in the same style as a traditional discussion section.
Please include a brief discussion of allowances made (if any) for controlling bias or unwanted sources of variability, and the limitations of any novel datasets.

# Conclusions <!-- Optional - only if novel data or analyses are included -->
This section is only required if the paper includes novel data or analyses, and should be written as a traditional conclusion.

# Summary <!-- Optional - only if NO new datasets are included -->
This section is required if the paper does not include novel data or analyses.  It allows authors to briefly summarize the key points from the article.

# Data availability <!-- Optional - only if novel data or analyses are included -->
Please add details of where any datasets that are mentioned in the paper, and that have not have not previously been formally published, can be found.  If previously published datasets are mentioned, these should be cited in the references, as per usual scholarly conventions.

# Software availability
This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.

1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
5. Software license (AUTHOR TO PROVIDE; required)

# Author contributions
In order to give appropriate credit to each author of an article, the individual
contributions of each author to the manuscript should be detailed in this section. We
recommend using author initials and then stating briefly how they contributed.

# Competing interests
All financial, personal, or professional competing interests for any of the authors that
could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article. If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.

# Grant information
Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc. Please do not list funding that you have that is not relevant to this
specific piece of research. For each funder, please state the funder’s name, the grant
number where applicable, and the individual to whom the grant was assigned.
If your work was not funded by any grants, please include the line: 'The author(s)
declared that no grants were involved in supporting this work.'

# Acknowledgments
This section should acknowledge anyone who contributed to the research or the
article but who does not qualify as an author based on the criteria provided earlier
(e.g. someone or an organization that provided writing assistance). Please state how
they contributed; authors should obtain permission to acknowledge from all those
mentioned in the Acknowledgments section.

Please do not list grant funding in this section.

# USING R MARKDOWN
Some examples of commonly used markdown syntax are listed below, to help you get started.

## Cross-references
For portability between different output formats, use the syntax introduced by *bookdown*, such as `(\#label)` for labels and `\@ref(label)` for cross-references. The following sections provide examples of referencing tables, figures, and equations.

## Citations

You can include refences in a standard Bibtex file.  The name of this file is given in the header of the markdown document (in our case it is *sample.bib*).  References to entries in the Bibtex file are made using square brackets and use an @ plus the key for the entry you are referencing [@Smith:2012qr].  You can combine multiple entries by separating them with a semi-colon [@Smith:2012qr; @Smith:2013jd].

## Code chunks
You can embed an R code chunk like this:
```{r firstCodeChunk}
x <- 1:10
x
```

If you specify a figure caption to a code chunk using the chunk option `fig.cap`, the plot will be automatically labeled and numbered. The figure label is generated from the label of the code chunk by prefixing it with `fig:`, e.g., see Figure \@ref(fig:plot).

```{r plot, fig.width=4, fig.height=4, fig.cap='A caption to our sample plot.'}
plot(x)
```

## Tables

Markdown syntax tends to lack some of the more sophisticated formatting features available in LaTeX, so you may need to edit the tables later to get the desired format.

| First name  | Last Name | Grade |
| ----------- | --------- | ----- |
| John        | Doe       |   7.5 |
| Richard     | Miles     |     2 |

Table: Caption to table.

Just like figures, tables with captions will also be numbered and can be referenced. Captions are entered as a paragraph beginning with the string "Table:" (or just ":"), which may appear either before or after the table. A label for the table should appear in the beginning of the caption in the form of `(\#tab:label)`, e.g., see Table \@ref(tab:table).

: (\#tab:table) A table with text justification.

| First name  | Last Name | Grade |
| ----------- | :-------: | ----: |
| John        | Doe       |   7.5 |
| Richard     | Miles     |     2 |

## Figures

You can include static figures (i.e. no generated by code) using the `include_graphics()` function from the **knitr** package, in a standard code chunk.

```{r frog-picture, out.width='50%', fig.align='center', fig.cap = 'Your figure legend goes here; it should be succinct, while still explaining all symbols and abbreviations.'}
knitr::include_graphics('frog.jpg')
```

You can again use the `fig.cap` option to provide the figure caption, and reference the image based on the code chunk label.  You can also use options such as `fig.align` and `fig.width` to adjust the position and size of the image within the final document, e.g. Figure \@ref(fig:frog-picture) is a frog.

Alternatively, you can use the standard markdown syntax like so:

![This is a smaller version of the same picture, inserted using the standard markdown syntax](frog.jpg){width=25%}

Please give figures appropriate filenames, e.g.: figure1.pdf, figure2.png.

Figure legends should briefly describe the key messages of the figure such that the figure can stand alone from the main text. However, all figures should also be discussed in the article text. Each legend should have a concise title of no more than 15 words. The legend itself should be succinct, while still explaining all symbols and abbreviations. Avoid lengthy descriptions of methods.

For any figures reproduced from another publication (as long as appropriate permission has been obtained from the copyright holder —see under the heading 'Submission'), please include a line in the legend to state that: 'This figure has been reproduced with kind permission from [include original publication citation]'.

## Mathematics

You can use LaTeX syntax to typeset mathematical expressions.  Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i$$
denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

To number and refer to equations, put them in the equation environments and assign labels to them, as for Equation \@ref(eq:binom).

\begin{equation}
  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
  (\#eq:binom)
\end{equation}

## Lists

You can make ordered lists

1. Like this,
2. and like this.

or bullet points

* Like this,
* and like this.
    + Use indentation
        - for sub-items

### 2017-07-14 R/methyvim_tmle_ate.R

@@ -1,0 +1,187 @@
+methytmle_disc <- function(sumExp,
+                           clusters,
+                           outcomeVar,
+                           targetSites,
+                           txCutoff = "50%",
+                           type = "exposure",
+                           numberSites = NULL,
+                           positivityMin = 10,
+                           covarPerSubj = 20,
+                           parallel = TRUE,
+                           family = "binomial",
+                           g_lib = list("SL.mean", "SL.glm", "SL.bayesglm"),
+                           Q_lib = list("SL.mean", "SL.glm", "SL.knn", "SL.gam")
+                          ) {
+
+  # ============================================================================
+  # check inputs are of the correct type before proceeding
+  # ============================================================================
+  type <- match.arg(type)
+  family <- match.arg(family)
+
+  # ============================================================================
+  # catch input and return in output object for user convenience
+  # ============================================================================
+  call <- match.call(expand.dots = TRUE)
+
+  #=============================================================================
+  # set up parallelization based on input
+  # ============================================================================
+  if (class(parallel) == "numeric") doParallel::registerDoParallel(parallel)
+  if (class(parallel) == "logical") {
+     nCores <- parallel::detectCores()
+     if (nCores > 1) {
+        doParallel::registerDoParallel(nCores)
+     } else {
+        warning("option 'parallel' is set to TRUE but only 1 core detected.")
+     }
+     if (parallel == FALSE) {
+        warning("parallelization set to FALSE: manually abort procedure.")
+     }
+  }
+
+  #=============================================================================
+  # heuristics for getting the value of the outcome variable
+  # ============================================================================
+  if (length(outcomeVar) == 1) {
+    # assume referring to column of design matrix if a scalar
+    outcomeValues <- as.data.frame(colData(sumExp))[, outcomeVar]
+  } else if (length(outcomeVar) > 1) {
+    # assume actual outcome data if a vector
+    outcomeValues <- outcomeVar
+  } else {
[…]

### 2017-07-14 R/methyvim_tmle_npvi.R

@@ -1,0 +1,193 @@
+methytmle_cont <- function(sumExp,
+                           clusters,
+                           outcomeVar,
+                           targetSites,
+                           type = "exposure",
+                           confid_level = 0.95,
+                           nullMode = "support",
+                           nullSupport = 0.20,
+                           nullRange = c(-2, 2),
+                           numberSites = NULL,
+                           parallel = TRUE,
+                           cvControl = 2,
+                           nMax = 30,
+                           ...
+                          ) {
+
+  # ============================================================================
+  # check inputs are of the correct type before proceeding
+  # ============================================================================
+  type <- match.arg(type)
+
+  # ============================================================================
+  # catch input and return in output object for user convenience
+  # ============================================================================
+  call <- match.call(expand.dots = TRUE)
+
+  # ============================================================================
+  # create TMLE-NPVI description list for convenient input later
+  # ============================================================================
+  descr <- list(f = identity, iter = 10, cvControl = cvControl, nMax = nMax,
+                stoppingCriteria = list(mic = 0.001, div = 0.001, psi = 0.01))
+
+  #=============================================================================
+  # set up parallelization based on input
+  # ============================================================================
+  if (class(parallel) == "numeric") doParallel::registerDoParallel(parallel)
+  if (class(parallel) == "logical") {
+     nCores <- parallel::detectCores()
+     if (nCores > 1) {
+        doParallel::registerDoParallel(nCores)
+     } else {
+        warning("option 'parallel' is set to TRUE but only 1 core detected.")
+     }
+     if (parallel == FALSE) {
+        warning("parallelization set to FALSE: manually abort procedure.")
+     }
+  }
+
+  #=============================================================================
+  # heuristics for getting the value of the outcome variable
+  # ============================================================================
+  if (length(outcomeVar) == 1) {
+    # assume referring to column of design matrix if a scalar
[…]

### 2017-07-14 R/shrink_ic.R

@@ -1,0 +1,6 @@
+# procedure to perform variance shrinkage over the influence curve
+# representation of a particular target parameter, as per Hejazi & Hubbard,
+# 2017-2018
+
+# function will check whether return_ic = TRUE, if so, apply limma shrinkage;
+# else, merely error out

### 2017-07-14 debug/1_toyData.R

@@ -1,0 +1,28 @@
+# setup project
+library(here)
+
+# setting project and data directories...
+proj_dir <- here()
+if (substr(Sys.info()["nodename"], 1, 10) == "bluevelvet") {
+  data_mtmle <- normalizePath(here("..", "..", "..", "..",
+                                   "local_ATE-methylation",
+                                   "methytmle_data_fixed.RData"))
+} else {
+  warning("Use Bluevelvet server to analyze this data set (due to size).")
+}
+load(data_mtmle)
+
+
+# extract annotation and M-values
+annot <- methytmleTestData$annot
+batch <- methytmleTestData$batch
+mvals <- methytmleTestData$mvals[, as.character(batch$Sample_Name)]
+
+# extract outcomes of interest
+outcomes <- c("fsiq_i_7y", "vciq_i_7y", "priq_i_7y", "wmiq_i_7y", "psiq_7y",
+              "baattss_7y", "pbdss_7y", "ftdr_7y")
+
+# extract potential confounders
+confounders <- c("educcat", "sppstd_6m", "CD8T.bakulski", "CD4T.bakulski",
+                 "NK.bakulski", "Bcell.bakulski", "Mono.bakulski",
+                 "Gran.bakulski", "nRBC.bakulski")

### 2017-07-14 debug/2_makeGRanges.R

@@ -1,0 +1,22 @@
+library(GenomicRanges)
+
+# build GRanges structure of CpG sites from annotation data
+cpg_gr <- makeGRangesFromDataFrame(annot, ignore.strand = TRUE,
+                                   seqnames.field = "CHR",
+                                   start.field = "MAPINFO",
+                                   end.field = "MAPINFO")
+
+# extract annotation data from grouped object and make GRanges
+cluster_sites <- function(granges, ...) {
+  if(dim(mcols(granges))[2] != 0) {
+    mcols(granges) <- NULL
+  }
+  clusters <- bumphunter::clusterMaker(chr = seqnames(granges),
+                                       pos = start(ranges(granges)),
+                                       assumeSorted = FALSE,
+                                       maxGap = 1000)
+  return(clusters)
+}
+
+# set up and genomic CpG matrix with M-value measurements
+clusters <- cluster_sites(cpg_gr)

### 2017-07-14 debug/3_setupLimma.R

@@ -1,0 +1,26 @@
+# use LIMMA to assess association with neurodevelopment with adjusted models
+designMats_full <- vector("list", length(outcomes))
+designMats_min <- vector("list", length(outcomes))
+keep_cases <- vector("list", length(outcomes))
+
+for (covar in 1:length(outcomes)) {
+  # extract full design matrix
+  covarsNames <- c(outcomes[covar], confounders, "batch")
+  design <- subset(batch, select = covarsNames)
+
+  # remove cases where covariate of interest is missing
+  keep_cases[[covar]] <- complete.cases(design)
+  design <- design[keep_cases[[covar]], ]
+
+ # set up intercept term
+ design <- as.data.frame(cbind(rep(1, nrow(design)), design))
+ colnames(design)[1] <- "intercept"
+ design <- sapply(design, as.numeric)
+
+ # reduce to minimal design matrix
+ design_min <- subset(design, select = c("intercept", outcomes[covar]))
+
+ # store design matrix for later use
+ designMats_full[[covar]] <- design
+ designMats_min[[covar]] <- design_min
+}

### 2017-07-14 debug/4_screenLimma.R

@@ -1,0 +1,16 @@
+# use LIMMA to assess neurocognitive scores using fully adjusted models
+tt_neuro_full <- vector("list", length(outcomes))
+
+for (neuro in 1:length(outcomes)) {
+  # normalization and subject reduction
+  meth <- as.data.frame(mcols(cpg_gr_mval)[, 2:length(mcols(cpg_gr_mval))])
+  meth_complete <- meth[, keep_cases[[neuro]]]
+
+  # fit LIMMA model
+  fit_full <- limma::lmFit(meth_complete, designMats_full[[neuro]])
+  fit_full <- limma::eBayes(fit_full)
+  tt_full <- limma::topTable(fit_full, coef = 2, num = Inf, sort.by = "none")
+
+  # store topTable result for later examination
+  tt_neuro_full[[neuro]] <- tt_full
+}

### 2017-07-14 R/data.R
more minor revisions

@@ -1,11 +1,0 @@
-#' Baseline covariates and Illumina 450K methylation data
-#'
-#' A dataset (list) containing a matrix of baseline/background covariates and #' a matrix of Illumina 450K methylation M-value measures.
-#'
-#' @format A \code{list} containing (1) a \code{data.frame} with N rows and P
-#'         variables of backround/baseline (phenotype) covariates, and (2) a
-#'         \code{granges} object containing annotation information on CpG sites #'         and m-values from Illumina 450K methylation sequencing.
-#' \describe{
-#'   ...some information ought to go here...
-#' }
-"neuro450k"

### 2017-07-14 R/zzz.R

@@ -1,4 +1,5 @@
-packageStartupMessage(
-      "methadapt: Adaptive Estimation of DNA Methylation",
-      "\nVersion: ", utils::packageVersion("methadapt")
-)
+.onAttach <- function(...) {
+  packageStartupMessage("methyvim: Nonparametric Variable Importance for Differential Methylation Analysis")
+  packageStartupMessage("Version: ",
+                        utils::packageDescription("methyvim")$Version)
+}

### 2017-07-31 README.Rmd
add scripts for using toy data
add bib for readme refs

@@
 output: github_document
+bibliography: refs.bib
 ---
@@
 
+---
+
+## Description
+
 `methyvim` is an R package that provides facilities for differential methylation
 analysis based on _variable importance measures_ (VIMs), a class of statistical
-target parameters that arise in causal inference. The statistical techniques
-implemented rely on the use of targeted minimum loss-based estimation (TMLE) to
-assess two particular VIMs: (1) the "local" average treatment effect for
-discrete exposures, and (2) a nonparametric variable importance measure for
-continuous exposures. Within this framework, these methods allow differential
-methylation effects to be estimated in an assumption-free manner, on a
-pre-screened set of genomic sites measured by DNA methylation assays. As the
-statistical algorithm uses a multi-stage approach, multiple testing corrections
-are made using a modified marginal Benjamini & Hochberg step-up False Discovery
-Rate controlling procedure for multi-stage analyses (FDR-MSA). In order to allow
-the user significant flexibility with respect to the scientific questions posed,
-the procedure estimates one of the appropriate VIMs for each of a reduced set of
-genomic sites, using screening procedures to identify a reduced set from the
-full data when a reduced set is not pre-specified. While making allowance for
-the user to specify the reduced set of genomic sites to be tested (e.g., those
-related to a particular gene of interest), this package comes equipped with
-screening procedures to facilitate the initial reduction of genomic sites, the
-two most noteworthy of these being a procedure based on the well-known R package
-[`limma`](https://bioconductor.org/packages/release/bioc/html/limma.html) and an
-extension of data-adaptive statistical target parameters for multiple testing
-(implemented in the R package
-[`DA.Test`](https://github.com/wilsoncai1992/data.adapt.multi.test)).
+target parameters that arise in causal inference.
+
+The statistical methodology implemented computes targeted minimum loss-based
+estimates of well-studied variable importance measures:
+
+1. The average treatment effect (ATE) for discrete exposures/treatments,
+2. A nonparametric variable importance measure (NPVI) for continuous exposures
+[@chambaz2012estimation].
+
+These methods allow differential methylation effects to be quantified in a
+manner that is largely free of assumptions, especially of the variety exploited
+in parametric models. __The statistical algorithm consists in several major
+steps:__
+
+1. Pre-screening of genomic sites is used to isolate a subset of sites for
+which there is evidence of differential methylation. Targeted minimum
+loss-based estimates of VIMs are computed for this subset of sites only (for
+the sake of computational feasibility). Several screening approaches are
+available, adapting core routines from the following R packages:
+[`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html),
+[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi),
+[`adaptest`](https://github.com/nhejazi/adaptest).
+2. Nonparametric VIMs are estimated for the specified parameter, currently using
+routines from the R packages
+[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) to estimate the
+NPVI parameter, or [`tmle`](https://CRAN.R-project.org/package=tmle) to
+estimate the ATE.
+3. Since pre-screening is performed prior to estimating VIMs, we make use of a
+multiple testing correction uniquely suited to such settings, specifically we
+use the modified marginal Benjamini & Hochberg step-up False Discovery Rate
+controlling procedure for multi-stage analyses (FDR-MSA) [@tuglus2009modified].
+
+See @vdl2011targeted for a general description of the framework of targeted
+minimum loss-based estimation.
+
+Note about shrinkage of influence curves, adapting @smyth2004linear.
+
+More exposition here....
 
@@
 
+<!--
 To contribute, install the _development version_ from GitHub via
@@
 ```
+-->
 
@@
 
-After using the `methyvim` R package, please cite it:
+After using the `methyvim` R package, please cite both of the following:
 
@@
             Mark J.},
-          title = {methyvim: Nonparametric Assessment of Variable Importance for
-                   Differential Methylation Analysis},
+          title = {methyvim: Differential methylation analysis with targeted
+            nonparametric variable importance measures},
           journal = {F1000Research}
@@
 
+        @misc{hejazi2017methyvim-r,
+          doi = {},
+          url = {},
+          howpublished = {\url{https://github.com/nhejazi/methyvim}},
+          year  = {2017},
+          author = {Hejazi, Nima S. and Hubbard, Alan E. and {van der Laan},
+            Mark J.},
+          title = {methyvim: Differential methylation analysis with targeted
+            minimum loss-based estimates of variable importance measures}
+        }
+
 ---
-
-## References
-
-* [Mark J. van der Laan & Sherri Rose. _Targeted Learning: Causal Inference for
-    Observational and Experimental Data_,
-    2011.](http://www.targetedlearningbook.com)
-
-* [Antoine Chambaz, Pierre Neuvial, & Mark J. van der Laan. "Estimation of a
-    non-parametric variable importance measure of a continuous exposure",
-    _Electronic Journal of Statistics_,
-    2012.](http://www.math-info.univ-paris5.fr/~chambaz/Papiers/chambazNeuvialvanderLaan_EJS2012.pdf)
-
-* [Catherine Tuglus & Mark J. van der Laan. "FDR controlling procedures for
-    multi-stage analyses", _UC Berkeley Division of Biostatistics Working
-    Paper Series_, 2008.](http://biostats.bepress.com/ucbbiostat/paper239/)
-
-* [Weixin Cai, Nima S. Hejazi, & Alan E. Hubbard. "Data-adaptive statistics for
-    multiple testing in high-dimensional settings", _In preparation_,
-    2017.](https://www.overleaf.com/5660573pjjrxh#/25678897/)
-
-* [Gordon K. Smyth. "Linear models and empirical Bayes methods for assessing
-    differential expression in microarray experiments." _Statistical
-    Applications in Genetics and Molecular Biology_, 3(1),
-    2004.](http://www.statsci.org/smyth/pubs/ebayes.pdf)
 
@@
 `LICENSE` for details.
+
+---
+
+## References

### 2017-08-07 R/methyvim.R
fix edge cases in ATE algo; NPVI algo now working

@@ -1,23 +1,64 @@
-#' Differential Methylation Analysis with Nonparametric Variable Importance
+#' Differential Methylation Statistics with Variable Importance Measures
 #'
-#' Computes the Targeted Minimum Loss-Based Estimate of...
+#' Computes the Targeted Minimum Loss-Based Estimate of a specified statistical
+#' target parameter, formally defined within models from causal inference. The
+#' variable importance measures currently supported are the Average Treatment
+#' Effect (ATE) and a Nonparametric Variable Importance Measure (NPVI, formally
+#' defined by Chambaz, Neuvial, and van der Laan <doi:10.1214/12-EJS703>).
 #'
-#' @param data_grs - ...
+#' @param data_grs An object of class \code{minfi::GenomicRatioSet}, containing
+#'        standard data structures associated with DNA Methylation experiments.
+#'        Consult the documentation for \code{minfi} to construct such objects.
+#' @param var_int Numeric indicating the column index of the variable of
+#'        interest, whether exposure or outcome. If argument \code{vim} is set
+#'        to the ATE, then the variable of interest is treated as an exposure;
+#'        it is treated as an outcome if this is set to be the NPVI.
+#' @param vim Character indicating the variable importance measure to be used
+#'        in the estimation procedure. Currently supported options are the ATE
+#'        for discretized exposures and NPVI for continuous exposures. ATE is
+#'        the appropriate choice when the underlying scientific question is of
+#'        the effect of an exposure on methylation, while NPVI is the parameter
+#'        of choice when the effect of methylation on an outcome is sought.
+#' @param type Character indicating the particular measure of DNA methylation to
+#'        be used as the observed data in the estimation procedure, either Beta
+#'        values or M-values. The data are accessed via \code{minfi::getBeta} or
+#'        \code{minfi::getM}.
+#' @param filter Character indicating...
 #'        ...
-#' @param var_int - ...
+#' @param filter_cutoff Numeric indicating...
 #'        ...
-#' @param min_sites Numeric indicating the minimum number of sites to be kept
-#'        after the application of a filtering procedure.
-#' @param neighbors Numeric indicating the size (in bp) of genomic windows to
-#'        use when assigning CpG sites to groups.
-#' @param parallel c(logical, numeric) - whether to use or number of cores to be
-#'        used in parallelizing the TMLE-based procedure across genomic sites.
-#' @param family (character) - specification of error family; available options
[…]

### 2017-08-07 R/tmle_ate.R

@@ -1,4 +1,18 @@
-methyvim_ate <- function(methy_tmle, catch_inputs) {
+#' Differential Methylation with the Average Treatment Effect
+#'
+#' Computes the Targeted Minimum Loss-Based Estimate of the Average Treatment
+#' Effect (ATE), treating DNA methylation as an outcome (Y) and the indicated
+#' variable of interest (which ought to be binarized) as a treatment/exposure
+#' (A), using the neighbors of a given CpG site as the adjustment set (W).
+#'
+#' @param methy_tmle An object of class \code{methytmle}...
+#'        ...
+#' @param catch_inputs_ate List...
+#'        ...
+#'
+
+methyvim_ate <- function(methy_tmle,
+                         catch_inputs_ate) {
 
   # make sure that the outcome data is of class numeric
   var_of_interest <- as.numeric(colData(methy_tmle)[, catch_inputs$var])
@@ -6,115 +20,153 @@ methyvim_ate <- function(methy_tmle, catch_inputs) {
     var_of_interest <- as.numeric(var_of_interest)
   }
 
-  # find all cases that have no missing values
-  cases_complete <- complete.cases(colData(methy_tmle))
+  # get names of sites to be added to output object
+  cpg_screened_names <- names(methy_tmle[methy_tmle@screen_ind])
+
+  # object of screened CpG site indices to loop over in TMLE procedure
+  methy_tmle_ind <- seq_along(methy_tmle_screened@screen_ind)
 
   #=============================================================================
   # perform TMLE estimation of L-ATE for all genomic (CpG) sites individually
   # ============================================================================
-  methy_vim_out <- foreach::foreach(site = seq_along(methy_tmle@screen_ind),
+  methy_vim_out <- foreach::foreach(i_site = methy_tmle_ind,
                                     .packages = c("tmle", "class", "gtools"),
                                     .combine = rbind) %dopar% {
 
-    # find neighbors of target site
+    ### create message for monitoring
+    message(paste("Beginning estimation for site", i_site))
 
-    # scale outcome variable for use with binomial TMLE
-    a = min(var_of_interest, na.rm = TRUE)
-    b = max(var_of_interest, na.rm = TRUE)
-    y_star <- (var_of_interest - a) / (b - a)
+    ### get target site
+    target_site <- methy_tmle_screened@screen_ind[i_site]
 
[…]

### 2017-08-07 R/tmle_npvi.R

@@ -1,156 +1,202 @@
-methyvim_npvi <- function(sumExp,
-                           clusters,
-                           outcomeVar,
-                           targetSites,
-                           type = "exposure",
-                           confid_level = 0.95,
-                           nullMode = "support",
-                           nullSupport = 0.20,
-                           nullRange = c(-2, 2),
-                           numberSites = NULL,
-                           parallel = TRUE,
-                           cvControl = 2,
-                           nMax = 30,
-                           ...
-                          ) {
+#' Differential Methylation with a Nonparametric Variable Importance Measure
+#'
+#' Computes the Targeted Minimum Loss-Based Estimate of a Nonparametric
+#' Variable Importance Measure (NPVI) of a continous exposure, treating DNA
+#' methylation as said exposure (X) and the indicated variable of interest
+#' as an outcome (Y), using the neighbors of a given CpG site as the adjustment
+#' set (W).
+#'
+#' @param methy_tmle An object of class \code{methytmle}...
+#'        ...
+#' @param catch_inputs_ate List...
+#'        ...
+#'
 
-  # ============================================================================
-  # create TMLE-NPVI description list for convenient input later
-  # ============================================================================
-  descr <- list(f = identity, iter = 10, cvControl = cvControl, nMax = nMax,
-                stoppingCriteria = list(mic = 0.001, div = 0.001, psi = 0.01))
+methyvim_ate <- function(methy_tmle,
+                         catch_inputs_ate) {
 
-  #=============================================================================
-  # heuristics for getting the value of the outcome variable
-  # ============================================================================
-  if (length(outcomeVar) == 1) {
-    # assume referring to column of design matrix if a scalar
-    outcomeValues <- as.data.frame(colData(sumExp))[, outcomeVar]
-  } else if (length(outcomeVar) > 1) {
-    # assume actual outcome data if a vector
-    outcomeValues <- outcomeVar
-    # NOTE: add r2weight here?
-  } else {
-    warning("inappropriate value specified for 'outcomeVar' argument")
+  # extract the variable of interest and ensure numeric
+  var_of_interest <- colData(methy_tmle_screened)[, catch_inputs_npvi$var]
[…]

### 2017-08-07 R/utils.R

@@ -63,28 +63,16 @@ set_parallel <- function(parallel) {
 
 ################################################################################
 
-#' Find correlations between neighboring CpGs
+#' Enforce the Assumption of Positivity to Compute the Average Treatment Effect
 #'
-#' Determine those CpG sites highly correlated with a particular site, towards
-#' the purpose of removing such sites from the baseline covariates included in
-#' TMLEs of a parameter of interest, to avoid practical positivity violations.
+#' Discretizes the numeric columns of an input matrix such that the newly
+#' created levels of each variable individually contain at least a specified
+#' mass when considering each level against levels of the treatment variable.
 #'
-#'
-#'
-#'
-#'
-
-corr_cpg <- function() {
-  message("This method has not been implemented yet.")
-}
-
-################################################################################
-
-#' Discretize a vector
-#'
-#' Discretizes a non-factor input vector and returns the result as numeric.
-#'
-#' @param x A vector containing arbitrary data.
+#' @param A...
+#' @param W...
+#' @param pos_min Numeric indicating...
+#' @param q_init Numeric indicating...
 #'
 #' @return A numeric vector with the data re-coded to based on the quantiles.
 #'
@@ -92,14 +80,31 @@ corr_cpg <- function() {
 #'
 #' @export
 #'
-#' @examples
-#' x <- rnorm(1000)
-#' discrete_by_quantile(x)
 
-discrete_by_quantile <- function(x) {
-  if( class(x) != "factor" ) {
-    as.numeric(gtools::quantcut(x))
-  } else {
-    as.numeric(x)
+force_positivity <- function(A, W, pos_min = 0.1, q_init = 10) {
+  stopifnot(length(A) == nrow(W))
+
+  if (class(W) != "data.frame") W <- as.data.frame(W) # cover use of "ncol"
+  out_w <- NULL # concatenate W columnwise as we discretize each covar below
+
+  for (obs_w in seq_len(ncol(W))) {
+    in_w <- as.numeric(W[, obs_w])
+    discr_w <- as.numeric(as.factor(gtools::quantcut(x = in_w, q = q_init)))
+    check <- sum((table(A, discr_w) / length(A)) < pos_min)
+    next_guess_q <- q_init
+    while (check > 0) {
+      next_guess_q <- (next_guess_q - 1)
+      discr_w <- as.numeric(as.factor(gtools::quantcut(x = in_w,
[…]

### 2017-08-07 sandbox/1_setup_data.R

@@ -1,6 +1,7 @@
 # needed pkgs
 library(here)
 library(minfi)
+library(foreach)
 library(parallel)
 library(doParallel)
 n_cores <- detectCores()
@@ -30,6 +31,7 @@ family = "binomial"  # for logistic fluctuation model with scaled Y
 g_lib = c("SL.mean", "SL.glm", "SL.randomForest")
 Q_lib = c("SL.mean", "SL.randomForest")
 npvi_cutoff = 0.25
+npvi_conf = 0.95
 npvi_descr = list(f = identity, iter = 10, cvControl = 2, nMax = 30,
                   stoppingCriteria = list(mic = 0.001, div = 0.001,
                                           psi = 0.01)
@@ -50,7 +52,7 @@ catch_inputs_ate <- list(data = grs, var = exp_var_int, vim = "ATE",
                         )
 
 # catch inputs for NPVI procedure
-catch_inputs_npvi <- list(data = grs, var = exp_var_int, vim = "NPVI",
+catch_inputs_npvi <- list(data = grs, var = out_var_int, vim = "NPVI",
                           type = type, window = window_bp, corr = corr_max,
                           obs_per_covar = obs_per_covar, par = parallel,
                           filter = filter, filter_cutoff = cutoff,
@@ -59,6 +61,7 @@ catch_inputs_npvi <- list(data = grs, var = exp_var_int, vim = "NPVI",
                                            g_lib = g_lib,
                                            Q_lib = Q_lib,
                                            npvi_cutoff = npvi_cutoff,
+                                           npvi_conf = npvi_conf,
                                            npvi_descr = npvi_descr
                                           )
                          )

### 2017-08-07 sandbox/3_methyvim_ate.R

@@ -18,10 +18,14 @@ if (class(var_of_interest) != "numeric") {
 ## run the ATE procedure
 methy_tmle_ind <- seq_along(methy_tmle_screened@screen_ind)
 sites_rand <- sort(sample(methy_tmle_ind, 50))
+sites <- names(methy_tmle_screened[methy_tmle_screened@screen_ind[sites_rand],])
 
 methy_vim_out <- foreach::foreach(i_site = sites_rand,
                                   .packages = c("tmle", "class", "gtools"),
                                   .combine = rbind) %dopar% {
+
+  ### create message for monitoring
+  message(paste("Beginning estimation for site", i_site))
 
   ### get target site
   target_site <- methy_tmle_screened@screen_ind[i_site]
@@ -68,23 +72,36 @@ methy_vim_out <- foreach::foreach(i_site = sites_rand,
       # if all neighbors are too highly correlated, we'll simply ignore W
       if (length(neighbors_corr_high) == length(only_neighbors)) {
         w_no_corr <- NULL
+        w_in <- as.data.frame(t(rep(1, length(y))))
       } else if (length(neighbors_corr_high) != 0) {
-        w_no_corr <- w[-neighbors_corr_high, ]
-      } else {
-        w_no_corr <- w
+        w_no_corr <- TRUE
+        w_in <- as.data.frame(w[-neighbors_corr_high, ])
       }
-      w <- w_no_corr
+    } else {
+      w_no_corr <- TRUE
+      w_in <- w
     }
 
     # use PAM to reduce W by selecting medoids
-    if (nrow(w) > w_max) {
+    if (!is.null(w_no_corr) & nrow(w_in) > w_max) {
       message("PAM will be used to reduce W but is not yet implemented.")
       # write utility function to perform PAM clustering and select medoids
       # TODO: w <- cluster_w_pam(w)
     }
 
     # strictly enforces the assumption of positivity by discretizing W
-    w_pos <- force_positivity(var_of_interest, t(w), pos_min = 0.15)
+    if (!is.null(w_no_corr)) {
+      w_pos <- force_positivity(var_of_interest, t(w_in), pos_min = 0.15)
+    } else {
+      w_pos <- as.data.frame(t(w_in))
+    }
+
+    # get length of remaining neighbors in the adjustment set
+    if (!is.null(w_no_corr)) {
+      n_neighbors_reduced <- ncol(w_pos)
+    } else {
[…]

### 2017-08-07 sandbox/4_methyvim_npvi.R

@@ -1,96 +1,125 @@
 # screen CpG sites using LIMMA method
 methy_tmle_screened <- limma_screen(methytmle = methy_tmle,
-                                    var_int = catch_inputs$var,
-                                    type = catch_inputs$type,
-                                    cutoff = catch_inputs$cutoff)
+                                    var_int = catch_inputs_npvi$var,
+                                    type = catch_inputs_npvi$type,
+                                    cutoff = catch_inputs_npvi$filter_cutoff)
 
 # NOTE: work out NPVI procedure "by hand"
-## first, randomly generate an outcome since we don't have this in the data
-outcome_int <- rnorm(nrow(colData(methy_tmle)))
-
 # create clusters
-methy_tmle_screened <- cluster_sites(methy_tmle = methy_tmle_screened)
+methy_tmle_screened <- cluster_sites(methy_tmle = methy_tmle_screened,
+                                     window_size = catch_inputs_npvi$window)
 
-# find all cases that have no missing values
-cases_complete <- complete.cases(colData(methy_tmle_screened))
-
-# remove subjects with missingness from the outcome
-if (length(outcome_int) > sum(cases_complete)) {
-  y <- as.numeric(outcome_int[cases_complete])
-} else {
-  y <- as.numeric(outcome_int)
+# extract the variable of interest and ensure numeric
+var_of_interest <- colData(methy_tmle_screened)[, catch_inputs_npvi$var]
+if (class(var_of_interest) != "numeric") {
+  var_of_interest <- as.numeric(var_of_interest)
 }
 
-# run the NPVI procedure
-confid_level = 0.95
-descr <- catch_inputs$npvi_descr
-
 # setup NPVI params
+confid_level <- catch_inputs_npvi$tmle_args$npvi_conf
+descr <- catch_inputs_npvi$tmle_args$npvi_descr
 alpha <- (1 - confid_level)
-samp_size <- length(y)
+samp_size <- length(var_of_interest)
 
-# setup shortened TMLE loop
+# run the NPVI procedure
 methy_tmle_ind <- seq_along(methy_tmle_screened@screen_ind)
 sites_rand <- sort(sample(methy_tmle_ind, 50))
+sites <- names(methy_tmle_screened[methy_tmle_screened@screen_ind[sites_rand],])
 
 methy_vim_out <- foreach::foreach(i_site = sites_rand,
[…]

### 2017-08-07 tests/testthat.R

@@ -1,4 +1,4 @@
 library(testthat)
-library(rrbsTMLE)
+library(methyvim)
 
-test_check("rrbsTMLE")
+#test_check("methyvim")

### 2017-08-07 tests/testthat/test_methyvim_ate.R

@@ -1,0 +1,5 @@
+# sites: 29537, 85038, 142384 (when input in i_site) make good examples for
+# several tests -- specifically:
+# 1. 29537 appears "normal", allowing for regular functionality to be tested.
+# 2. 85038 has no neighbors without a high correlation...
+# 3. 142384 has many neighbors with low correlations...

### 2017-08-07 README.Rmd
clean up readme

@@
 
-> Nonparametric variable importance analysis for the assessment of differential
-> methylation
+> Targeted learning of variable importance measures for the analysis of
+> differential methylation experiments
 
@@
 
+<!--
 Note about shrinkage of influence curves, adapting @smyth2004linear.
@@
 More exposition here....
+-->
 
@@
 
+<!--
 ## Example
@@
 ```
-
----
+-->
 
@@
 
----
-
 ## License

### 2017-08-08 R/check_inputs.R
update documentation

@@ -30,7 +30,7 @@ check_inputs <- function(catch_inputs) {
   }
 
   if (catch_inputs$tmle_type == "glm") {
-    if (catch_inputs$vim == "ATE")
+    if (catch_inputs$vim == "ATE") {
       # set GLM libraries for "tmle" package
       g_lib <- "SL.mean"
       Q_lib <- "SL.glm"

### 2017-08-08 R/screen_adaptest.R

@@ -1,10 +1,0 @@
-#' Screening procedure based on data-adaptive multiple testing
-#'
-#' @importFrom adaptest adaptest
-#'
-#' @export adaptest_screen
-#'
-
-adaptest_screen <- function() {
-  message("This method has not yet been implemented.")
-}

### 2017-08-08 R/utils.R

@@ -1,4 +1,14 @@
-#' Clustering of sites to generate CpG neighborhoods
+#' CpG Neighborhoods from Genomic Distance
+#'
+#' Clustering of CpG sites to define CpG neighborhoods based on distance (bp).
+#'
+#' @param methy_tmle Object of class \code{methytmle} produced from an object of
+#'        class \code{GenomicRatioSet}, but containing extra slots.
+#' @param window_size Numeric giving the number of base pairs used to define
+#'        neighborhoods along the genome. CpG sites within this distance (bp) of
+#'        one another are denoted as neighbors. Chromosome boundaries and other
+#'        biological constraints are respected (see the documentation available
+#'        for \code{bumphunter::clusterMaker} for details).
 #'
 #' @importFrom bumphunter clusterMaker
 #' @importFrom SummarizedExperiment rowRanges
@@ -6,7 +16,6 @@
 #' @importFrom BiocGenerics start
 #' @importFrom IRanges ranges
 #'
-
 cluster_sites <- function(methy_tmle, window_size = 1000) {
   gr <- SummarizedExperiment::rowRanges(methy_tmle)
   pos <- BiocGenerics::start(IRanges::ranges(gr))
@@ -25,9 +34,14 @@ cluster_sites <- function(methy_tmle, window_size = 1000) {
 #' Modified FDR Controlling Procedure for Multi-Stage Analyses (MJ van der Laan
 #' and C Tuglus, 2009, <doi:10.2202/1544-6115.1397>)
 #'
+#' @param pvals Numeric vector containing the p-values that result from any
+#'        chosen statistical hypothesis testing procedure.
+#' @param total_obs Numeric indicating the total number of observations that
+#'        would have been available for testing prior to the selection procedure
+#'        employed in the multi-stage analysis performed.
+#'
 #' @importFrom stats p.adjust
 #'
-
 fdr_msa <- function(pvals, total_obs) {
   pvals_not_tested <- rep(1, total_obs - length(pvals))
   pvals_all <- c(pvals, pvals_not_tested)
@@ -40,11 +54,16 @@ fdr_msa <- function(pvals, total_obs) {
 
 #' Set up parallelization
 #'
+#' @param parallel Numeric or Logical indicating either the number of cores to
+#'        use in parallelized computation or whether parallelization ought to be
+#'        used at all. If \code{TRUE}, all available cores are found via a call
+#'        to \code{parallel::detectCores} and used. If \code{FALSE}, computation
[…]

### 2017-08-11 R/tmle_npvi.R
beginning vignette work

@@ -129,8 +129,12 @@ methyvim_ate <- function(methy_tmle,
 
       # create observed data matrix for input into tmle.npvi
       obs_data_in <- as.data.frame(cbind(var_of_interest, x, t(w_in)))
-      colnames(obs_data_in) <- c("Y", "X",
-                                 paste0("W", 1:(ncol(obs_data_in) - 2)))
+      if (nrow(w_in) > 1) {
+        colnames(obs_data_in) <- c("Y", "X",
+                                   paste0("W", 1:(ncol(obs_data_in) - 2)))
+      } else {
+        colnames(obs_data_in) <- c("Y", "X", "W")
+      }
 
       # compute the NPVI
       out <- tmle.npvi(obs = obs_data_in,
@@ -158,8 +162,7 @@ methyvim_ate <- function(methy_tmle,
 
       # create observed data matrix for input into tmle.npvi
       obs_data_in <- as.data.frame(cbind(var_of_interest, x, w_int))
-      colnames(obs_data_in) <- c("Y", "X",
-                                 paste0("W", 1:(ncol(obs_data_in) - 2)))
+      colnames(obs_data_in) <- c("Y", "X", "W")
 
       # compute the NPVI
       out <- tmle.npvi(obs = obs_data_in,

### 2017-08-11 sandbox/1_setup_data.R

@@ -28,8 +28,8 @@ parallel = TRUE
 return_ic = FALSE
 shrink_ic = FALSE
 family = "binomial"  # for logistic fluctuation model with scaled Y
-g_lib = c("SL.mean", "SL.glm", "SL.randomForest")
-Q_lib = c("SL.mean", "SL.randomForest")
+g_lib = c("SL.mean", "SL.glm", "SL.glmnet", "SL.polymars", "SL.nnet")
+Q_lib = c("SL.mean", "SL.glm", "SL.nnet")
 npvi_cutoff = 0.25
 npvi_conf = 0.95
 npvi_descr = list(f = identity, iter = 10, cvControl = 2, nMax = 30,

### 2017-08-11 sandbox/3_methyvim_ate.R

@@ -17,7 +17,7 @@ if (class(var_of_interest) != "numeric") {
 
 ## run the ATE procedure
 methy_tmle_ind <- seq_along(methy_tmle_screened@screen_ind)
-sites_rand <- sort(sample(methy_tmle_ind, 50))
+sites_rand <- sort(sample(methy_tmle_ind, 25))
 sites <- names(methy_tmle_screened[methy_tmle_screened@screen_ind[sites_rand],])
 
 methy_vim_out <- foreach::foreach(i_site = sites_rand,
@@ -153,5 +153,5 @@ methy_vim_out <- foreach::foreach(i_site = sites_rand,
 methy_vim_out <- as.data.frame(methy_vim_out)
 colnames(methy_vim_out) <- c("lower_CI_ATE", "est_ATE", "upper_CI_ATE", "Var",
                              "pval", "n_neighbors_all", "n_neighbors_w",
-                             "max_corr_w")
+                             "max_corr_all")
 rownames(methy_vim_out) <- sites

### 2017-08-11 sandbox/4_methyvim_npvi.R

@@ -23,7 +23,7 @@ samp_size <- length(var_of_interest)
 
 # run the NPVI procedure
 methy_tmle_ind <- seq_along(methy_tmle_screened@screen_ind)
-sites_rand <- sort(sample(methy_tmle_ind, 50))
+sites_rand <- sort(sample(methy_tmle_ind, 25))
 sites <- names(methy_tmle_screened[methy_tmle_screened@screen_ind[sites_rand],])
 
 methy_vim_out <- foreach::foreach(i_site = sites_rand,
@@ -117,7 +117,13 @@ methy_vim_out <- foreach::foreach(i_site = sites_rand,
 
     # create observed data matrix for input into tmle.npvi
     obs_data_in <- as.data.frame(cbind(var_of_interest, x, t(w_in)))
-    colnames(obs_data_in) <- c("Y", "X", paste0("W", 1:(ncol(obs_data_in) - 2)))
+    if (nrow(w_in) > 1) {
+      colnames(obs_data_in) <- c("Y", "X",
+                                 paste0("W", 1:(ncol(obs_data_in) - 2)))
+    } else {
+      colnames(obs_data_in) <- c("Y", "X", "W")
+    }
+
 
     # compute the NPVI
     out <- tmle.npvi(obs = obs_data_in,
@@ -144,7 +150,7 @@ methy_vim_out <- foreach::foreach(i_site = sites_rand,
 
     # create observed data matrix for input into tmle.npvi
     obs_data_in <- as.data.frame(cbind(var_of_interest, x, w_int))
-    colnames(obs_data_in) <- c("Y", "X", paste0("W", 1:(ncol(obs_data_in) - 2)))
+    colnames(obs_data_in) <- c("Y", "X", "W")
 
     # compute the NPVI
     out <- tmle.npvi(obs = obs_data_in,
@@ -180,5 +186,5 @@ methy_vim_out <- foreach::foreach(i_site = sites_rand,
 }
 methy_vim_out <- as.data.frame(methy_vim_out)
 colnames(methy_vim_out) <- c("lower_CI_ATE", "est_ATE", "upper_CI_ATE", "pval",
-                             "n_neighbors_all", "n_neighbors_w", "max_corr_w")
+                             "n_neighbors_all", "n_neighbors_w", "max_corr_all")
 rownames(methy_vim_out) <- sites

### 2017-08-15 README.Rmd
adds data package note
expand description and add references

@@
 
-## Description
+## What's `methyvim`?
 
@@
 analysis based on _variable importance measures_ (VIMs), a class of statistical
-target parameters that arise in causal inference.
+estimable target parameters that arise in causal inference.
 
@@
 
-1. The average treatment effect (ATE) for discrete exposures/treatments,
-2. A nonparametric variable importance measure (NPVI) for continuous exposures
+1. The _average treatment effect_ (ATE) for discrete exposures/treatments,
+2. A _nonparametric variable importance measure_ (NPVI) for continuous exposures
 [@chambaz2012estimation].
@@
 1. Pre-screening of genomic sites is used to isolate a subset of sites for
-which there is evidence of differential methylation. Targeted minimum
-loss-based estimates of VIMs are computed for this subset of sites only (for
-the sake of computational feasibility). Several screening approaches are
+which there is cursory evidence of differential methylation. For the sake of
+computational feasibility, targeted minimum loss-based estimates of VIMs are
+computed only for this subset of sites. Several screening approaches are
 available, adapting core routines from the following R packages:
 [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html),
-[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi),
-[`adaptest`](https://github.com/nhejazi/adaptest).
+[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi).
 2. Nonparametric VIMs are estimated for the specified parameter, currently using
 routines from the R packages
-[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) to estimate the
-NPVI parameter, or [`tmle`](https://CRAN.R-project.org/package=tmle) to
-estimate the ATE.
+[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) (to estimate the
+NPVI parameter) or [`tmle`](https://CRAN.R-project.org/package=tmle) (to
+estimate the ATE).
 3. Since pre-screening is performed prior to estimating VIMs, we make use of a
-multiple testing correction uniquely suited to such settings, specifically we
-use the modified marginal Benjamini & Hochberg step-up False Discovery Rate
-controlling procedure for multi-stage analyses (FDR-MSA) [@tuglus2009modified].
+multiple testing correction uniquely suited to such settings. Rather generally,
+the Benjamini & Hochberg procedure for controlling the False Discovery Rate
+(FDR) [@benjamini1995controlling] is applied due to multiple testing issues
+inherent to the estimation procedure. Specifically, we apply a modified marginal
+Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
+multi-stage analyses (FDR-MSA) [@tuglus2009modified].
 
-See @vdl2011targeted for a general description of the framework of targeted
-minimum loss-based estimation.
+For a general discussion of the framework of targeted minimum loss-based
+estimation, the many applications of this methodology, and the role the
+framework plays in statistical causal inference, the recommended references are
+@vdl2011targeted and @vdl2017targeted. @hernan2018causal and @pearl2009causality
+may be of interest to those desiring a more general introduction to statistical
+causal inference.
 
@@
 
+## Related
+
+* [R/`methyvimData`](https://github.com/nhejazi/methyvimData) - R package with
+    sample experimental DNA methylation data for use as an example with this
+    analysis package.
+
+---
+
 ## License

### 2017-08-16 README.Rmd
re-structure tmle-ate, other misc changes

@@
 `methyvim` is an R package that provides facilities for differential methylation
-analysis based on _variable importance measures_ (VIMs), a class of statistical
-estimable target parameters that arise in causal inference.
+analysis based on _variable importance measures_ (VIMs), a class of
+statistically estimable target parameters that arise in causal inference.
 
@@
 3. Since pre-screening is performed prior to estimating VIMs, we make use of a
-multiple testing correction uniquely suited to such settings. Rather generally,
-the Benjamini & Hochberg procedure for controlling the False Discovery Rate
-(FDR) [@benjamini1995controlling] is applied due to multiple testing issues
-inherent to the estimation procedure. Specifically, we apply a modified marginal
+multiple testing correction uniquely suited to such settings. Due to the
+multiple testing nature of the estimation problem, a variant of the Benjamini &
+Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
+[@benjamini1995controlling]. Specifically, we apply the modified marginal
 Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
@@
 
-After using the `methyvim` R package, please cite both of the following:
+After using the `methyvim` R package, please cite the following:
 
@@
           volume = {},
-          author = {Hejazi, Nima S. and Hubbard, Alan E. and {van der Laan},
-            Mark J.},
+          author = {Hejazi, Nima S and Hubbard, Alan E and {van der Laan}, Mark
+            J},
           title = {methyvim: Differential methylation analysis with targeted
-            nonparametric variable importance measures},
+            esitmates of nonparametric variable importance measures},
           journal = {F1000Research}
-        }
-
-        @misc{hejazi2017methyvim-r,
-          doi = {},
-          url = {},
-          howpublished = {\url{https://github.com/nhejazi/methyvim}},
-          year  = {2017},
-          author = {Hejazi, Nima S. and Hubbard, Alan E. and {van der Laan},
-            Mark J.},
-          title = {methyvim: Differential methylation analysis with targeted
-            minimum loss-based estimates of variable importance measures}
         }

### 2017-08-16 R/methyvim.R

@@ -24,7 +24,7 @@
 #'        values or M-values. The data are accessed via \code{minfi::getBeta} or
 #'        \code{minfi::getM}.
 #' @param filter Character indicating the model to be implemented when screening
-#'        the \code{data_grs} object for CpG sites. Currently 
+#'        the \code{data_grs} object for CpG sites. Currently
 #'        supported options are limma, npvi, and adaptest.  (?)
 #' @param filter_cutoff Numeric indicating the p-value cutoff that defines which
 #'        sites pass through the \code{filter}.
@@ -71,6 +71,8 @@
 #'         regressions, as well as the original data rotated into influence
 #'         curve space may be returned, if so requested.
 #'
+#' @importFrom BiocParallel bplapply
+#'
 #' @export methyvim
 #'
 
@@ -109,25 +111,25 @@ methyvim <- function(data_grs,
   # ============================================================================
   # catch inputs to pass to downstream functions (for estimation and such)
   # ============================================================================
-  catch_inputs <- list(data = data_grs,
-                       var = var_int,
-                       vim = vim,
-                       type = type,
-                       filter = filter,
-                       filter_cutoff = filter_cutoff,
-                       window = window_bp,
-                       corr = corr_max,
-                       obs_per_covar = obs_per_covar,
-                       par = parallel,
-                       future = future_param,
-                       bppar = bppar_type,
-                       return_ic = return_ic,
-                       shrink_ic = shrink_ic,
-                       tmle_type = tmle_type,
-                       tmle_args = tmle_args)
   # check that inputs satisfy expectations
   ## and clean up check_inputs (e.g., rm NPVI stuff if ATE is specified)
-  catch_inputs <- check_inputs(catch_inputs)
+  check_inputs(data = data_grs,
+               var = var_int,
+               vim = vim,
+               type = type,
+               filter = filter,
+               filter_cutoff = filter_cutoff,
+               window = window_bp,
+               corr = corr_max,
+               obs_per_covar = obs_per_covar,
+               parallel = parallel,
+               future = future_param,
+               bppar = bppar_type,
+               return_ic = return_ic,
+               shrink_ic = shrink_ic,
+               tmle_type = tmle_type,
+               tmle_args = tmle_args
+              )
 
   # ============================================================================
   # invoke S4 class constructor for "methadapt" object
@@ -138,9 +140,9 @@ methyvim <- function(data_grs,
[…]

### 2017-08-16 R/tmle_ate.R

@@ -5,168 +5,149 @@
 #' variable of interest (which ought to be binarized) as a treatment/exposure
 #' (A), using the neighbors of a given CpG site as the adjustment set (W).
 #'
-#' @param methy_tmle An object of class \code{methytmle}...
-#'        ...
-#' @param catch_inputs_ate List...
-#'        ...
+#' @param target_site Numeric ...
+#' @param methy_tmle_screened An object of class \code{methytmle}...
+#' @param i_site Numeric ...
+#' @param type Character ...
+#' @param corr Numeric ...
+#' @param obs_per_covar Numeric ...
+#' @param g_lib Character or vector of characters...
+#' @param Q_lib Character or vector of characters...
+#' @param family Character ...
+#' @param return_ic Logical ...
 #'
+methyvim_ate <- function(target_site,
+                         methy_tmle_screened,
+                         var_of_interest,
+                         type = c("Beta", "Mval"),
+                         corr = 0.80,
+                         obs_per_covar = 20,
+                         g_lib = c("SL.mean", "SL.glm"),
+                         Q_lib = c("SL.mean", "SL.glm"),
+                         family = c("gaussian", "binomial"),
+                         return_ic = FALSE
+                         ) {
+  ### check arguments where possible
+  #type <- match.arg(type)
+  #family <- match.arg(family)
 
-methyvim_ate <- function(methy_tmle,
-                         catch_inputs_ate) {
+  ### get neighboring site
+  in_cluster <- which(methy_tmle_screened@clusters %in%
+                      methy_tmle_screened@clusters[target_site])
 
-  # make sure that the outcome data is of class numeric
-  var_of_interest <- as.numeric(colData(methy_tmle)[, catch_inputs$var])
-  if (class(var_of_interest) != "numeric") {
-    var_of_interest <- as.numeric(var_of_interest)
+  ### remove target site from the set of neighbors
+  only_neighbors <- in_cluster[in_cluster != target_site]
+
+  # get expression measures based on input
+  if (type == "Beta") {
+    expr <- minfi::getBeta(methy_tmle_screened)
+  } else if (type == "Mval") {
+    expr <- minfi::getM(methy_tmle_screened)
   }
 
-  # get names of sites to be added to output object
-  cpg_screened_names <- names(methy_tmle[methy_tmle@screen_ind])
[…]

### 2017-08-16 sandbox/2_methyvim_funs.R

@@ -89,6 +89,36 @@ force_positivity <- function(A, W, pos_min = 0.1, q_init = 10) {
   return(out)
 }
 
+set_parallel <- function(parallel = c(TRUE, FALSE),
+                         future_param = NULL,
+                         bppar_type = NULL) {
+  # invoke a future-based backend
+  doFuture::registerDoFuture()
+
+  if (parallel == TRUE) {
+    if (!is.null(future_param)) {
+      set_future_param <- parse(text = paste0("future", "::", future_param))
+      future::plan(eval(set_future_param))
+    } else {
+      future::plan(future::multiprocess)
+    }
+  } else if (parallel == FALSE) {
+    warning(paste("Sequential evaluation is strongly discouraged.",
+                  "\n Proceed with caution."))
+    future::plan(future::sequential)
+  }
+  if (!is.null(bppar_type)) {
+    bp_type <- eval(parse(text = paste0("BiocParallel", "::",
+                                        bppar_type, "()")))
+  } else {
+    bp_type <- BiocParallel::DoparParam()
+  }
+  # try to use a progress bar is supported in the parallelization plan
+  BiocParallel::bpprogressbar(bp_type) <- TRUE
+  # register the chosen parallelization plan
+  BiocParallel::register(bp_type, default = TRUE)
+}
+
 # advanced vingette discussing internals of the package
 # not writing about force positivity in the
 # treatment (discrete) and W is a matrix of cont measures assoc w each of the
@@ -103,4 +133,4 @@ force_positivity <- function(A, W, pos_min = 0.1, q_init = 10) {
 # now each of the obs data are now replaced with a level of that obs while
 # maintaining the positivity
 
-# somehow this shit is related to cTMLE
+# somehow this is related to cTMLE

### 2017-08-17 R/check_inputs.R
clean up

@@ -8,8 +8,24 @@
 #' @return Options to be passed to various \code{tmle_*} functions.
 #'
 
-check_inputs <- function(catch_inputs) {
-  message("check_inputs is only partially implemented currently.")
+check_inputs <- function(data = data_grs,
+                         var = var_int,
+                         vim = vim,
+                         type = type,
+                         filter = filter,
+                         filter_cutoff = filter_cutoff,
+                         window = window_bp,
+                         corr = corr_max,
+                         obs_per_covar = obs_per_covar,
+                         parallel = parallel,
+                         future = future_param,
+                         bppar = bppar_type,
+                         return_ic = return_ic,
+                         shrink_ic = shrink_ic,
+                         tmle_type = tmle_type,
+                         tmle_args = tmle_args) {
+
+  message("This function is only partially implemented currently.")
   # NOTE: make sure to check combinations of the variable of interest and the
   # stated parameter of interest (i.e., VIM). In particular, we will use the
   # following heuristic to decide between NPVI and ATE: if the variable of
@@ -19,24 +35,24 @@ check_inputs <- function(catch_inputs) {
   # heuristic, but use it to check that the user has specified the ATE vs. NPVI
   # parameters correctly.
 
-  if (catch_inputs$vim == "NPVI" & is.null(catch_inputs$tmle_args$npvi_descr)) {
+  if (vim == "NPVI" & is.null(tmle_args$npvi_descr)) {
     npvi_descr_defaults <- list(f = identity, iter = 10, cvControl = 2,
                                 nMax = 30,
                                 stoppingCriteria = list(mic = 0.001,
                                                         div = 0.001,
                                                         psi = 0.01)
                                )
-    catch_inputs$tmle_args$npvi_descr <- npvi_descr_defaults
+    tmle_args$npvi_descr <- npvi_descr_defaults
   }
 
-  if (catch_inputs$tmle_type == "glm") {
-    if (catch_inputs$vim == "ATE") {
+  if (tmle_type == "glm") {
+    if (vim == "ATE") {
       # set GLM libraries for "tmle" package
       g_lib <- "SL.mean"
       Q_lib <- "SL.glm"
-      catch_inputs$tmle_args$g_lib <- g_lib
-      catch_inputs$tmle_args$Q_lib <- Q_lib
-    } else if (catch_inputs$vim == "NPVI") {
+      tmle_args$g_lib <- g_lib
[…]

### 2017-08-17 R/methyvim.R

@@ -72,6 +72,9 @@
 #'         curve space may be returned, if so requested.
 #'
 #' @importFrom BiocParallel bplapply
+#' @importFrom parallel detectCores
+#' @importFrom doParallel registerDoParallel
+#' @importFrom foreach foreach "%dopar%"
 #'
 #' @export methyvim
 #'
@@ -134,7 +137,7 @@ methyvim <- function(data_grs,
   # ============================================================================
   # invoke S4 class constructor for "methadapt" object
   # ============================================================================
-  methy_tmle <- .methytmle(catch_inputs$data)
+  methy_tmle <- .methytmle(data)
   methy_tmle@call <- call
 
   #=============================================================================
@@ -143,15 +146,8 @@ methyvim <- function(data_grs,
   #set_parallel(parallel = parallel,
   #             future_param = future_param,
   #             bppar_type = bppar_type)
-
-  # ============================================================================
-  # operate on the type of data specified
-  # ============================================================================
-  #if (catch_inputs$type == "Beta") {
-  #  extract_measures <- parse(text = "getBeta(methy_tmle)")
-  #} else if (catch_inputs$type == "Mval") {
-  #  extract_measures <- parse(text = "getM(methy_tmle)")
-  #}
+  n_cores <- parallel::detectCores()
+  doParallel::registerDoParallel(n_cores)
 
   #=============================================================================
   # check if there is missing data in the phenotype-level matrix and drop if so
@@ -167,7 +163,7 @@ methyvim <- function(data_grs,
   #=============================================================================
   # screen sites to produce a subset on which to estimate VIMs
   # ============================================================================
-  if (catch_inputs$filter == "limma") {
+  if (filter == "limma") {
     methy_tmle <- limma_screen(methytmle = methy_tmle,
                                var_int = var_int,
                                type = type)
@@ -183,7 +179,7 @@ methyvim <- function(data_grs,
   #=============================================================================
   # ATE TMLE procedure for targeted differential methylation analysis
   # ============================================================================
-  if (catch_inputs$vim == "ATE") {
+  if (vim == "ATE") {
 
     # make sure that the outcome data is of class numeric
     var_of_interest <- as.numeric(colData(methy_tmle)[, var_int])
@@ -193,9 +189,11 @@ methyvim <- function(data_grs,
 
[…]

### 2017-08-17 R/methyvim.R
minor vignette update

@@ -116,23 +116,23 @@ methyvim <- function(data_grs,
   # ============================================================================
   # check that inputs satisfy expectations
   ## and clean up check_inputs (e.g., rm NPVI stuff if ATE is specified)
-  check_inputs(data = data_grs,
-               var = var_int,
-               vim = vim,
-               type = type,
-               filter = filter,
-               filter_cutoff = filter_cutoff,
-               window = window_bp,
-               corr = corr_max,
-               obs_per_covar = obs_per_covar,
-               parallel = parallel,
-               future = future_param,
-               bppar = bppar_type,
-               return_ic = return_ic,
-               shrink_ic = shrink_ic,
-               tmle_type = tmle_type,
-               tmle_args = tmle_args
-              )
+  #check_inputs(data = data_grs,
+  #             var = var_int,
+  #             vim = vim,
+  #             type = type,
+  #             filter = filter,
+  #             filter_cutoff = filter_cutoff,
+  #             window = window_bp,
+  #             corr = corr_max,
+  #             obs_per_covar = obs_per_covar,
+  #             parallel = parallel,
+  #             future = future_param,
+  #             bppar = bppar_type,
+  #             return_ic = return_ic,
+  #             shrink_ic = shrink_ic,
+  #             tmle_type = tmle_type,
+  #             tmle_args = tmle_args
+  #            )
 
   # ============================================================================
   # invoke S4 class constructor for "methadapt" object
@@ -174,7 +174,7 @@ methyvim <- function(data_grs,
   #=============================================================================
   # cluster sites based on genomic windows
   #=============================================================================
-  methy_tmle <- cluster_sites(methy_tmle = methy_tmle)
+  methy_tmle <- cluster_sites(methytmle = methy_tmle)
 
   #=============================================================================
   # ATE TMLE procedure for targeted differential methylation analysis

### 2017-08-17 R/utils.R

@@ -16,7 +16,7 @@
 #' @importFrom BiocGenerics start
 #' @importFrom IRanges ranges
 #'
-cluster_sites <- function(methy_tmle, window_size = 1000) {
+cluster_sites <- function(methytmle, window_size = 1000) {
   gr <- SummarizedExperiment::rowRanges(methy_tmle)
   pos <- BiocGenerics::start(IRanges::ranges(gr))
   clusters <- bumphunter::clusterMaker(chr = GenomeInfoDb::seqnames(gr),

### 2017-08-17 R/methyvim.R
fix .methytmle error: data -> data_grs

@@ -137,7 +137,7 @@ methyvim <- function(data_grs,
   # ============================================================================
   # invoke S4 class constructor for "methadapt" object
   # ============================================================================
-  methy_tmle <- .methytmle(data)
+  methy_tmle <- .methytmle(data_grs)
   methy_tmle@call <- call
 
   #=============================================================================

### 2017-08-17 R/methyvim.R
fix synatx errors from argument renaming

@@ -196,7 +196,7 @@ methyvim <- function(data_grs,
     methy_tmle_ind <- methy_tmle_ind[seq_len(100)] ## TODO: REMOVE, FOR TESTING
     #methy_vim_out <- BiocParallel::bplapply(X = methy_tmle_ind,
     #                                        FUN = methyvim_ate,
-    #                                        methy_tmle_screened = methy_tmle,
+    #                                        methytmle_screened = methy_tmle,
     #                                        var_of_interest = var_of_interest,
     #                                        type = "Mval",
     #                                        corr = 0.80,
@@ -214,7 +214,7 @@ methyvim <- function(data_grs,
                     length(methy_tmle_ind)))
 
       out <- methyvim_ate(target_site = i_site,
-                          methy_tmle_screened = methy_tmle,
+                          methytmle_screened = methy_tmle,
                           var_of_interest = var_of_interest,
                           type = "Mval",
                           corr = 0.80,

### 2017-08-17 R/tmle_ate.R

@@ -17,7 +17,7 @@
 #' @param return_ic Logical ...
 #'
 methyvim_ate <- function(target_site,
-                         methy_tmle_screened,
+                         methytmle_screened,
                          var_of_interest,
                          type = c("Beta", "Mval"),
                          corr = 0.80,
@@ -32,17 +32,17 @@ methyvim_ate <- function(target_site,
   #family <- match.arg(family)
 
   ### get neighboring site
-  in_cluster <- which(methy_tmle_screened@clusters %in%
-                      methy_tmle_screened@clusters[target_site])
+  in_cluster <- which(methytmle_screened@clusters %in%
+                      methytmle_screened@clusters[target_site])
 
   ### remove target site from the set of neighbors
   only_neighbors <- in_cluster[in_cluster != target_site]
 
   # get expression measures based on input
   if (type == "Beta") {
-    expr <- minfi::getBeta(methy_tmle_screened)
+    expr <- minfi::getBeta(methytmle_screened)
   } else if (type == "Mval") {
-    expr <- minfi::getM(methy_tmle_screened)
+    expr <- minfi::getM(methytmle_screened)
   }
 
   # get measures at the target site and perform scaling

### 2017-08-17 R/utils.R

@@ -17,14 +17,14 @@
 #' @importFrom IRanges ranges
 #'
 cluster_sites <- function(methytmle, window_size = 1000) {
-  gr <- SummarizedExperiment::rowRanges(methy_tmle)
+  gr <- SummarizedExperiment::rowRanges(methytmle)
   pos <- BiocGenerics::start(IRanges::ranges(gr))
   clusters <- bumphunter::clusterMaker(chr = GenomeInfoDb::seqnames(gr),
                                        pos = pos,
                                        assumeSorted = FALSE,
                                        maxGap = window_size)
-  methy_tmle@clusters <- as.numeric(clusters)
-  return(methy_tmle)
+  methytmle@clusters <- as.numeric(clusters)
+  return(methytmle)
 }
 
 ################################################################################

### 2017-08-17 R/methyvim.R
fix function args

@@ -9,6 +9,9 @@
 #' @param data_grs An object of class \code{minfi::GenomicRatioSet}, containing
 #'        standard data structures associated with DNA Methylation experiments.
 #'        Consult the documentation for \code{minfi} to construct such objects.
+#' @param sites_comp Numeric indicating the number of sites for which a variable
+#'        importance measure is to be estimated post-screening. This arugment is
+#'        TEMPORARY, USED ONLY FOR TESTING PURPOSES currently.
 #' @param var_int Numeric indicating the column index of the variable of
 #'        interest, whether exposure or outcome. If argument \code{vim} is set
 #'        to the ATE, then the variable of interest is treated as an exposure;
@@ -80,14 +83,15 @@
 #'
 
 methyvim <- function(data_grs,
+                     sites_comp = 100,
                      var_int = 1,
                      vim = c("ATE", "NPVI"),
                      type = c("Beta", "Mval"),
                      filter = c("limma", "npvi", "adaptest"),
                      filter_cutoff = 0.05,
                      window_bp = 1e3,
-                     corr_max = 0.35,
-                     obs_per_covar = 15,
+                     corr_max = 0.70,
+                     obs_per_covar = 20,
                      parallel = TRUE,
                      future_param = NULL,
                      bppar_type = NULL,
@@ -95,8 +99,7 @@ methyvim <- function(data_grs,
                      shrink_ic = FALSE,
                      tmle_type = c("glm", "super_learning"),
                      tmle_args = list(family = "binomial",
-                                      g_lib = c("SL.mean", "SL.glm",
-                                                "SL.randomForest"),
+                                      g_lib = c("SL.mean", "SL.glm"),
                                       Q_lib = c("SL.mean", "SL.glm"),
                                       npvi_cutoff = 0.25,
                                       npvi_descr = NULL)
@@ -189,11 +192,16 @@ methyvim <- function(data_grs,
 
     # get names of sites to be added to output object
     cpg_screened_names <- names(methy_tmle[methy_tmle@screen_ind])
-    cpg_screened_names <- cpg_screened_names[seq_len(100)] ## TODO: REMOVE, FOR TESTING ONLY
+
+    ## TODO: THIS IS FOR TESTING ONLY
+    cpg_screened_names <- cpg_screened_names[seq_len(sites_comp)]
 
     # object of screened CpG site indices to loop over in TMLE procedure
     methy_tmle_ind <- seq_along(methy_tmle@screen_ind)
[…]

### 2017-08-17 R/methyvim.R
fix argument passing error

@@ -206,12 +206,12 @@ methyvim <- function(data_grs,
     #                                        FUN = methyvim_ate,
     #                                        methytmle_screened = methy_tmle,
     #                                        var_of_interest = var_of_interest,
-    #                                        type = "Mval",
-    #                                        corr = 0.80,
-    #                                        obs_per_covar = 20,
-    #                                        family = "gaussian",
-    #                                        g_lib = c("SL.mean", "SL.glm"),
-    #                                        Q_lib = c("SL.mean", "SL.glm"),
+    #                                        type = type,
+    #                                        corr = corr_max,
+    #                                        obs_per_covar = obs_per_covar,
+    #                                        family = tmle_args$family,
+    #                                        g_lib = tmle_args$g_lib,
+    #                                        Q_lib = tmle_args$Q_lib,
     #                                        return_ic = FALSE
     #                                       )
     methy_vim_out <- foreach::foreach(i_site = methy_tmle_ind,

### 2017-08-17 R/tmle_ate.R

@@ -6,8 +6,7 @@
 #' (A), using the neighbors of a given CpG site as the adjustment set (W).
 #'
 #' @param target_site Numeric ...
-#' @param methy_tmle_screened An object of class \code{methytmle}...
-#' @param i_site Numeric ...
+#' @param methytmle_screened An object of class \code{methytmle}...
 #' @param type Character ...
 #' @param corr Numeric ...
 #' @param obs_per_covar Numeric ...

### 2017-08-17 R/methyvim.R
foreach syntax errors

@@ -83,7 +83,7 @@
 #'
 
 methyvim <- function(data_grs,
-                     sites_comp = 100,
+                     sites_comp = 10,
                      var_int = 1,
                      vim = c("ATE", "NPVI"),
                      type = c("Beta", "Mval"),
@@ -215,7 +215,7 @@ methyvim <- function(data_grs,
     #                                        return_ic = FALSE
     #                                       )
     methy_vim_out <- foreach::foreach(i_site = methy_tmle_ind,
-                                      .packages = c("tmle"),
+                                      .packages = c("tmle", "SuperLearner"),
                                       .combine = rbind) %dopar% {
 
       message(paste("Computing targeted estimate for site", i_site, "of",

### 2017-08-17 R/tmle_ate.R
add missing importFrom tags

@@ -15,6 +15,11 @@
 #' @param family Character ...
 #' @param return_ic Logical ...
 #'
+#' @importFrom minfi getBeta getM
+#' @importFrom tmle tmle
+#'
+#' @export
+#'
 methyvim_ate <- function(target_site,
                          methytmle_screened,
                          var_of_interest,
@@ -107,14 +112,14 @@ methyvim_ate <- function(target_site,
     max_corr_w <- max(cor(y, t(w)))
 
     # compute the ATE
-    out <- tmle(Y = as.numeric(y_star),
-                A = as.numeric(var_of_interest),
-                W = as.data.frame(w_pos),
-                Q.SL.library = Q_lib,
-                g.SL.library = g_lib,
-                family = family,
-                verbose = FALSE
-               )
+    out <- tmle::tmle(Y = as.numeric(y_star),
+                      A = as.numeric(var_of_interest),
+                      W = as.data.frame(w_pos),
+                      Q.SL.library = Q_lib,
+                      g.SL.library = g_lib,
+                      family = family,
+                      verbose = FALSE
+                     )
   } else {
     n_neighbors_total <- 0
     n_neighbors_reduced <- 0
@@ -124,14 +129,14 @@ methyvim_ate <- function(target_site,
     w_int <- as.data.frame(rep(1, length(var_of_interest)))
 
     # compute the ATE
-    out <- tmle(Y = as.numeric(y_star),
-                A = as.numeric(var_of_interest),
-                W = as.data.frame(w_int),
-                Q.SL.library = Q_lib,
-                g.SL.library = g_lib,
-                family = family,
-                verbose = FALSE
-               )
+    out <- tmle::tmle(Y = as.numeric(y_star),
+                      A = as.numeric(var_of_interest),
+                      W = as.data.frame(w_int),
+                      Q.SL.library = Q_lib,
+                      g.SL.library = g_lib,
+                      family = family,
+                      verbose = FALSE
+                     )
   }
 
   # get the influence curve estimates if so requested

### 2017-08-17 R/utils.R

@@ -45,7 +45,7 @@ cluster_sites <- function(methytmle, window_size = 1000) {
 fdr_msa <- function(pvals, total_obs) {
   pvals_not_tested <- rep(1, total_obs - length(pvals))
   pvals_all <- c(pvals, pvals_not_tested)
-  fdr_adj <- p.adjust(pvals_all, method = "fdr")
+  fdr_adj <- stats::p.adjust(pvals_all, method = "fdr")
   fdr_out <- fdr_adj[seq_len(pvals)]
   return(fdr_out)
 }

### 2017-08-18 R/methyvim.R
fix horrid doc errors

@@ -146,11 +146,11 @@ methyvim <- function(data_grs,
   #=============================================================================
   # set up parallelization if so desired
   # ============================================================================
-  set_parallel(parallel = parallel,
-               future_param = future_param,
-               bppar_type = bppar_type)
-  #n_cores <- parallel::detectCores()
-  #doParallel::registerDoParallel(n_cores)
+  #set_parallel(parallel = parallel,
+  #             future_param = future_param,
+  #             bppar_type = bppar_type)
+  n_cores <- parallel::detectCores()
+  doParallel::registerDoParallel(n_cores)
 
   #=============================================================================
   # check if there is missing data in the phenotype-level matrix and drop if so
@@ -202,41 +202,41 @@ methyvim <- function(data_grs,
     ## TODO: THIS IS FOR TESTING ONLY
     methy_tmle_ind <- methy_tmle_ind[seq_len(sites_comp)]
 
-    methy_vim_out <- BiocParallel::bplapply(X = methy_tmle_ind,
-                                            FUN = methyvim_ate,
-                                            methytmle_screened = methy_tmle,
-                                            var_of_interest = var_of_interest,
-                                            type = type,
-                                            corr = corr_max,
-                                            obs_per_covar = obs_per_covar,
-                                            family = tmle_args$family,
-                                            g_lib = tmle_args$g_lib,
-                                            Q_lib = tmle_args$Q_lib,
-                                            return_ic = FALSE
-                                           )
-    methy_vim_out <- do.call(rbind.data.frame, methy_vim_out)
+    #methy_vim_out <- BiocParallel::bplapply(X = methy_tmle_ind,
+    #                                        FUN = methyvim_ate,
+    #                                        methytmle_screened = methy_tmle,
+    #                                        var_of_interest = var_of_interest,
+    #                                        type = type,
+    #                                        corr = corr_max,
+    #                                        obs_per_covar = obs_per_covar,
+    #                                        family = tmle_args$family,
+    #                                        g_lib = tmle_args$g_lib,
+    #                                        Q_lib = tmle_args$Q_lib,
+    #                                        return_ic = FALSE
+    #                                       )
+    #methy_vim_out <- do.call(rbind.data.frame, methy_vim_out)
 
-    #methy_vim_out <- foreach::foreach(i_site = methy_tmle_ind,
[…]

### 2017-08-18 R/tmle_ate.R

@@ -7,6 +7,7 @@
 #'
 #' @param target_site Numeric ...
 #' @param methytmle_screened An object of class \code{methytmle}...
+#' @param var_of_interest ...
 #' @param type Character ...
 #' @param corr Numeric ...
 #' @param obs_per_covar Numeric ...
@@ -18,6 +19,8 @@
 #' @importFrom minfi getBeta getM
 #' @importFrom tmle tmle
 #' @importFrom stats cor
+#'
+#' @export
 #'
 methyvim_ate <- function(target_site,
                          methytmle_screened,

### 2017-08-18 R/tmle_npvi.R

@@ -8,16 +8,17 @@
 #'
 #' @param methy_tmle An object of class \code{methytmle}...
 #'        ...
-#' @param catch_inputs_ate List...
+#' @param catch_inputs_npvi List...
 #'        ...
 #'
 #' @importFrom minfi getBeta getM
 #' @importFrom stats cor quantile qnorm
 #' @importFrom tmle.npvi tmle.npvi setConfLevel getPsi getSic getPValue
 #'
-
-methyvim_ate <- function(methy_tmle,
-                         catch_inputs_ate) {
+#' @export
+#'
+methyvim_npvi <- function(methy_tmle,
+                         catch_inputs_npvi) {
 
   # extract the variable of interest and ensure numeric
   var_of_interest <- colData(methy_tmle_screened)[, catch_inputs_npvi$var]

### 2017-08-18 R/methyvim.R
fix vignette parallelization error

@@ -149,8 +149,10 @@ methyvim <- function(data_grs,
   #set_parallel(parallel = parallel,
   #             future_param = future_param,
   #             bppar_type = bppar_type)
-  n_cores <- parallel::detectCores()
-  doParallel::registerDoParallel(n_cores)
+  if (parallel == TRUE) {
+    n_cores <- parallel::detectCores()
+    doParallel::registerDoParallel(n_cores)
+  }
 
   #=============================================================================
   # check if there is missing data in the phenotype-level matrix and drop if so

### 2017-08-18 README.Rmd
mostly adds reqs to travis+appveyor

@@
           month = {},
-          publisher={Faculty of 1000 Ltd},
+          publisher = {},
           volume = {},
@@
             J},
-          title = {methyvim: Differential methylation analysis with targeted
-            esitmates of nonparametric variable importance measures},
-          journal = {F1000Research}
+          title = {methyvim: Nonparametric differential methylation analysis
+            with targeted estimates of variable importance measures},
+          journal = {}
         }

### 2017-08-26 R/methyvim.R
PAM and function documentation

@@ -36,7 +36,7 @@
 #' @param corr_max Numeric indicating the maximum correlation that a neighboring
 #'        site can have with the target site.
 #' @param obs_per_covar Numeric indicating the number of observations needed for
-#'        for covariate included in (W/downstream analysis). This ensures the
+#'        for covariate included in W for downstream analysis. This ensures the
 #'        data is sufficient to control for the covariates.
 #' @param parallel Logical indicating whether parallelization ought to be used.
 #'        See the documentation of \code{set_parallel} for more information, as
@@ -90,7 +90,7 @@ methyvim <- function(data_grs,
                      filter = c("limma", "npvi", "adaptest"),
                      filter_cutoff = 0.05,
                      window_bp = 1e3,
-                     corr_max = 0.70,
+                     corr_max = 0.75,
                      obs_per_covar = 20,
                      parallel = TRUE,
                      future_param = NULL,

### 2017-08-26 R/tmle_ate.R

@@ -26,7 +26,7 @@ methyvim_ate <- function(target_site,
                          methytmle_screened,
                          var_of_interest,
                          type = c("Beta", "Mval"),
-                         corr = 0.80,
+                         corr = 0.70,
                          obs_per_covar = 20,
                          g_lib = c("SL.mean", "SL.glm"),
                          Q_lib = c("SL.mean", "SL.glm"),
@@ -91,10 +91,9 @@ methyvim_ate <- function(target_site,
 
     # use PAM to reduce W by selecting medoids
     if (!is.null(w_no_corr) & nrow(w_in) > w_max) {
-      message("PAM will be used to reduce W but is not yet implemented.")
-      # write utility function to perform PAM clustering and select medoids
-      # TODO: w <- cluster_w_pam(w)
-    }
+      message("Implementing PAM to reduce W")
+      w_in <- cluster::pam(x = t(w_in), k = w_max, diss = FALSE)
+    } 
 
     # strictly enforces the assumption of positivity by discretizing W
     if (!is.null(w_no_corr)) {

### 2017-08-26 sandbox/npvi_exp.Rmd

---
title: "On using the R package `tmle.npvi`"
author: "Nima Hejazi"
date: "`r Sys.Date()`"
output: html_document
---

```{r get-pkg, eval=FALSE}
library(devtools)
remove.packages("tmle.npvi")
install_github("achambaz/tmle.npvi")
```

Now, let's simply generate some data to be used with the `tmle.npvi` package.
The observed data structure will be of the form $O = (Y, X, W)$, with each of
the variables of interest being of the form: $Y \sim N(0, 1)$, $X \sim N(0, 1)$,
and $W = (W_1, W_2, W_3)$, with $W_i \sim N(0, 1)$. In order to ensure that the
exposure/treatment variable of interest ($X$) has sufficient mass at a null
value, we set $X \in [-1, 1] \mapsto 0$: that is, all observed values of $X$
within _one standard deviation_ of the mean are set to _zero_.

```{r sim}
suppressMessages(library(tmle.npvi))
set.seed(46781)
n_obs <- 1000
n_var <- 5

obs <- as.data.frame(replicate(n_var, rnorm(n_obs)))
colnames(obs) <- c("Y", "X", paste0("W", seq_len(n_var - 2)))
obs$X[obs$X > -1 & obs$X < 1] <- 0
head(obs)
```

Above, we examine the `data.frame` that represents the observed data structure
and confirm that it appears to meet the specifications required for invoking the
`tmle.npvi` procedure.

Now, we use the eponymous function of the `tmle.npvi` package to obtain a
targeted minimum loss-based estimate of the NPVI parameter. We then
appropriately set the confidence level and examine the output structure.

```{r npvi}
alpha <- 0.05
npvi <- tmle.npvi(obs, f = identity, flavor = "learning")
setConfLevel(npvi, 1 - alpha)
npvi
```

The object of class `npvi` returned by the estimation procedure provides a
wealth of information. We are interested in the point estimate of the estimator
of the parameter of interest $\psi$ as well as the uncertainty (as measured by
standard deviation) associated with this estimate.

__This leads to a few questions:__

1. While the point estimate is rather small ($-0.0621$) the standard error of
   the estimate is roughly an order of magnitude larger. Why is this?
2. While we note that the standard deviation of the estimator of $\psi$ is
   readily accessible via the `tmle.npvi::getPsiSd` function, this value is not
   used in computing the confidence interval (rather the standard error is) --
   why is this?
3. From this simulated data, we note that there is not a straightforward
   relationship between the standard deviation and standard error, contrary to
   our expectations. Why is this the case?

In particular, we would expect the confidence interval for the estimator of
$\psi$ to be computable as follows:

```{r npvi-stats}
psi <- getPsi(npvi)
sd <- getPsiSd(npvi)
CI = psi + c(-1,1) * sd * qnorm(1 - alpha/2) / sqrt(n_obs)
out <- c(CI[1], psi, CI[2])
names(out) <- c("CI_lower", "Est", "CI_upper")
out
```

Examining this CI, we note that it is considerably tighter than that reported by
the `print` method associated with the `npvi` object we computed.

### 2017-08-26 R/tmle_ate.R
function documentation

@@ -5,16 +5,32 @@
 #' variable of interest (which ought to be binarized) as a treatment/exposure
 #' (A), using the neighbors of a given CpG site as the adjustment set (W).
 #'
-#' @param target_site Numeric ...
-#' @param methytmle_screened An object of class \code{methytmle}...
-#' @param var_of_interest ...
-#' @param type Character ...
-#' @param corr Numeric ...
-#' @param obs_per_covar Numeric ...
-#' @param g_lib Character or vector of characters...
-#' @param Q_lib Character or vector of characters...
-#' @param family Character ...
-#' @param return_ic Logical ...
+#' @param target_site Numeric indicating the column containing the screened
+#'        CpG site indices that will be looped over in TMLE procedure.
+#' @param methytmle_screened An object of class \code{methytmle} with clustered
+#'        sites based on genomic windows containing the screened CpG site
+#'        indices column.
+#' @param var_of_interest Numeric indicating the column index of the binarized
+#'        variable of interest, treated as an exposure.
+#' @param type Character indicating the particular measure of DNA methylation to
+#'        be used as the observed data in the estimation procedure, either Beta
+#'        values or M-values. The data are accessed via \code{minfi::getBeta} or
+#'        \code{minfi::getM}.
+#' @param corr Numeric indicating the maximum correlation that a neighboring
+#'        site can have with the target site.
+#' @param obs_per_covar Numeric indicating the number of observations needed for
+#'        for covariate included in W for downstream analysis. This ensures the
+#'        data is sufficient to control for the covariates.
+#' @param g_lib Character or vector of characters indicating the algorithms to
+#'        be implemented in SuperLearner if \code{tmle_type} is set to "glm".
+#' @param Q_lib Character or vector of characters indicating the algorithms to
+#'        be implemented in SuperLearner if \code{tmle_type} is set to
+#'        "super_learning".
+#' @param family Character indicating the distribution to be implemented to
+#'        describe the error distribution for regressions, generally "gaussian"
+#'        for a continuous outcome and "binomial" for a binary outcome.
+#' @param return_ic Logical indicating whether an influence curve estimate
+#'        should be returned for each site that passed through the filter.
 #'
 #' @importFrom minfi getBeta getM
 #' @importFrom tmle tmle
@@ -26,7 +42,7 @@ methyvim_ate <- function(target_site,
                          methytmle_screened,
[…]

### 2017-08-27 README.Rmd
builds after rr, vignette, readme tweaks

@@
 The statistical methodology implemented computes targeted minimum loss-based
-estimates of well-studied variable importance measures:
+estimates of several well-characterized variable importance measures:
 
-1. The _average treatment effect_ (ATE) for discrete exposures/treatments,
-2. A _nonparametric variable importance measure_ (NPVI) for continuous exposures
-[@chambaz2012estimation].
+For discrete-valued treatments or exposures:
+* The _average treatment effect_ (ATE):
+  $$ \Psi_{ATE} = E[E[Y_j \mid A = 1, W_{-j}] - E[Y_j \mid A = 0, W_{-j}]],$$
+  where $j$ indexes a target CpG site of interest and $-j$ gives all CpG sites
+  in the same neighborhood as the target site except the target site itself
+  (i.e., the neighboring sites).
+
+* The _relative risk_ (RR), where
+  $$ \dots,$$
+  where...
+
+For continous-valued treatments or exposures:
+* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation].
 
@@
 [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi).
-2. Nonparametric VIMs are estimated for the specified parameter, currently using
-routines from the R packages
-[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) (to estimate the
-NPVI parameter) or [`tmle`](https://CRAN.R-project.org/package=tmle) (to
-estimate the ATE).
+2. Nonparametric VIMs are estimated for the specified parameter, currently
+adapting routines from the
+[`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) and
+[`tmle`](https://CRAN.R-project.org/package=tmle) R packages.
 3. Since pre-screening is performed prior to estimating VIMs, we make use of a
@@
 Note about shrinkage of influence curves, adapting @smyth2004linear.
+Specifically, this is meant to reference my working paper with Alan.
 

### 2017-08-27 R/methyvim.R

@@ -85,7 +85,7 @@
 methyvim <- function(data_grs,
                      sites_comp = 10,
                      var_int = 1,
-                     vim = c("ate"),
+                     vim = c("ate", "rr", "npvi"),
                      type = c("Beta", "Mval"),
                      filter = c("limma"),
                      filter_cutoff = 0.05,
@@ -242,8 +242,8 @@ methyvim <- function(data_grs,
     methy_vim_out <- as.data.frame(methy_vim_out)
 
     # TMLE procedure is now done, so let's just make the output object pretty...
-    colnames(methy_vim_out) <- c("lower_CI_ATE", "est_ATE", "upper_CI_ATE",
-                                 "Var", "pval", "n_neighbors_all",
+    colnames(methy_vim_out) <- c("lowerCI_ATE", "est_ATE", "upperCI_ATE",
+                                 "Var_ATE", "pval", "n_neighbors_all",
                                  "n_neighbors_w", "max_corr_w")
     rownames(methy_vim_out) <- cpg_screened_names
     methy_tmle@vim <- methy_vim_out
@@ -295,8 +295,8 @@ methyvim <- function(data_grs,
     methy_vim_out <- as.data.frame(methy_vim_out)
 
     # TMLE procedure is now done, so let's just make the output object pretty...
-    colnames(methy_vim_out) <- c("lower_CI_ATE", "est_ATE", "upper_CI_ATE",
-                                 "Var", "pval", "n_neighbors_all",
+    colnames(methy_vim_out) <- c("lowerCI_logRR", "est_logRR", "upperCI_logRR",
+                                 "Var_logRR", "pval", "n_neighbors_all",
                                  "n_neighbors_w", "max_corr_w")
     rownames(methy_vim_out) <- cpg_screened_names
     methy_tmle@vim <- methy_vim_out
@@ -307,8 +307,6 @@ methyvim <- function(data_grs,
   } else if (vim == "npvi") {
     stop("Support for TMLE-NPVI is planned but not yet implemented.")
     #methy_tmle@vim <- methyvim_npvi(methy_tmle = methy_tmle)
-  } else {
-    stop("The specified variable importance parameter is not available.")
   }
   return(methy_tmle)
   # NOTE: what else do we do before returning output...

### 2017-08-27 R/plots.R

@@ -1,7 +1,33 @@
-#' p-value histograms
+#' Common plots for methytmle objects
+#'
+#' Several types of plots providing...
+#'
+#'
+#' @param x object of class \code{biotmle} as produced by an appropriate call to
+#'        \code{biomarkertmle}
+#' @param type character describing whether to provide a plot of unadjusted or
+#'        adjusted p-values (adjustment performed via Benjamini-Hochberg)
+#' @param ... additional arguments passed \code{plot} as necessary
+#'
+#'
+#' @return object of class \code{ggplot} containing one of several types of
+#'         plots: a heatmap of the top 50 CpG sites, a volcano plot displaying
+#'         the parameter estimate against the log-transformed raw p-values, or
+#'         side-by-side histograms of the raw and corrected p-values.
+#'
+#' @export
+#'
+#' @method plot methytmle
+#'
 
-#' volcano plot
+plot.methytmle <- function(x, ..., type = c("heat", "volcano", "pvals")) {
+  message("Plot methods have not yet been implemented.")
+}
 
-#' heatmap method
+################################################################################
 
-#' annotation plot using GViz
+#' a heatmap for methytmle objects with more features goes here...
+
+################################################################################
+
+#' fancy annotation plot using GViz goes here...

### 2017-08-27 R/tmle_classic.R

@@ -166,10 +166,12 @@ methyvim_tmle <- function(target_site,
     var_rescaled <- est_raw[4] * ((b - a)^2)
     res <- c(est_rescaled, var_rescaled, est_raw[5], n_neighbors_total,
              n_neighbors_reduced, max_corr_w)
-  } else if (target_param == "rr" & family != "binomial") {
+  } else if (target_param == "rr" & family == "binomial") {
     est <- out$estimates$RR
-    est_raw <- c(est$CI[1], est$psi, est$CI[2], est$var.psi, est$pvalue)
-    res <- c(est_raw, n_neighbors_total, n_neighbors_reduced, max_corr_w)
+    est_ci_log <- log(est$CI)
+    est_out <- c(est_ci_log[1], est$log.psi, est_ci_log[2], est$var.log.psi,
+                 est$pvalue)
+    res <- c(est_out, n_neighbors_total, n_neighbors_reduced, max_corr_w)
   } else if (target_param == "rr" & family == "gaussian") {
     stop("The Relative Risk is not estimable with a gaussian error family.")
   }

### 2017-08-27 README.Rmd
fix vignette deps

@@
 For discrete-valued treatments or exposures:
+
 * The _average treatment effect_ (ATE):
@@
 For continous-valued treatments or exposures:
+
 * A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation].

### 2017-08-27 R/methyvim.R

@@ -100,7 +100,8 @@ methyvim <- function(data_grs,
                      tmle_type = c("glm", "sl"),
                      tmle_args = list(family = "binomial",
                                       g_lib = c("SL.mean", "SL.glm"),
-                                      Q_lib = c("SL.mean", "SL.glm", "SL.gam"),
+                                      Q_lib = c("SL.mean", "SL.glm", "SL.gam",
+                                                "SL.earth"),
                                       npvi_cutoff = 0.25,
                                       npvi_descr = NULL)
                     ) {

### 2017-08-28 R/methyvim.R
rm cluster placeholder script, export parallel function, adds arg checking

@@ -90,7 +90,7 @@ methyvim <- function(data_grs,
                      filter = c("limma"),
                      filter_cutoff = 0.05,
                      window_bp = 1e3,
-                     corr_max = 0.70,
+                     corr_max = 0.75,
                      obs_per_covar = 20,
                      parallel = TRUE,
                      future_param = NULL,
@@ -99,11 +99,8 @@ methyvim <- function(data_grs,
                      shrink_ic = FALSE,
                      tmle_type = c("glm", "sl"),
                      tmle_args = list(family = "binomial",
-                                      g_lib = c("SL.mean", "SL.glm"),
-                                      Q_lib = c("SL.mean", "SL.glm", "SL.gam",
-                                                "SL.earth"),
-                                      npvi_cutoff = 0.25,
-                                      npvi_descr = NULL)
+                                      g_lib = NULL, Q_lib = NULL,
+                                      npvi_cutoff = 0.25, npvi_descr = NULL)
                     ) {
 # if you say glm it will choose mean and glm
 # if you choose super learner it will choose more algorithms
@@ -173,8 +170,6 @@ methyvim <- function(data_grs,
     methy_tmle <- limma_screen(methytmle = methy_tmle,
                                var_int = var_int,
                                type = type)
-  } else {
-    stop("The designated filtering method has not yet been implemented.")
   }
 
   #=============================================================================
@@ -186,7 +181,6 @@ methyvim <- function(data_grs,
   # TMLE procedure for the Average Treatment Effect (ATE) parameter
   # ============================================================================
   if (vim == "ate") {
-
     # make sure that the outcome data is of class numeric
     var_of_interest <- as.numeric(SummarizedExperiment::colData(methy_tmle)[, var_int])
     if (class(var_of_interest) != "numeric") {
@@ -206,16 +200,17 @@ methyvim <- function(data_grs,
     methy_tmle_ind <- methy_tmle_ind[seq_len(sites_comp)]
 
     #methy_vim_out <- BiocParallel::bplapply(X = methy_tmle_ind,
-    #                                        FUN = methyvim_ate,
+    #                                        FUN = methyvim_tmle,
     #                                        methytmle_screened = methy_tmle,
     #                                        var_of_interest = var_of_interest,
     #                                        type = type,
     #                                        corr = corr_max,
     #                                        obs_per_covar = obs_per_covar,
+    #                                        target_param = "ate",
[…]

### 2017-08-28 R/tmle_classic.R

@@ -31,15 +31,24 @@ methyvim_tmle <- function(target_site,
                           corr,
                           obs_per_covar,
                           target_param = c("ate", "rr"),
-                          g_lib = c("SL.mean", "SL.glm"),
-                          Q_lib = c("SL.mean", "SL.glm"),
+                          g_lib = c("SL.mean", "SL.glm", "SL.glm.interaction"),
+                          Q_lib = c("SL.mean", "SL.glm", "SL.gam", "SL.earth"),
                           family = c("gaussian", "binomial"),
                           return_ic = FALSE
                          ) {
-  ### check arguments where possible
-  #type <- match.arg(type)
-  #family <- match.arg(family)
+  ### set treatment mechanism and outcome regression libraries
+  if(is.null(g_lib)) {
+    g_lib <- c("SL.mean", "SL.glm", "SL.glm.interaction")
+  }
+  if(is.null(Q_lib)) {
+    Q_lib <- c("SL.mean", "SL.glm", "SL.gam", "SL.earth")
+  }
 
+  ### check arguments where possible
+  type <- match.arg(type)
+  target_param <- match.arg(target_param)
+  family <- match.arg(family)
+  
   ### get neighboring site
   in_cluster <- which(methytmle_screened@clusters %in%
                       methytmle_screened@clusters[target_site])

### 2017-08-28 R/utils.R

@@ -77,6 +77,8 @@ fdr_msa <- function(pvals, total_obs) {
 #' @importFrom future plan multiprocess sequential
 #' @importFrom doFuture registerDoFuture
 #'
+#' @export
+#'
 set_parallel <- function(parallel = c(TRUE, FALSE),
                          future_param = NULL,
                          bppar_type = NULL) {

### 2017-08-28 R/check_inputs.R
rethink input checking, fix use of futures, alter appveyor warnings

@@ -4,9 +4,6 @@
 #'
 #' @param data ...
 #' @param var ...
-#' @param vim ...
-#' @param type ..
-#' @param filter ...
 #' @param filter_cutoff ...
 #' @param window ...
 #' @param corr ...
@@ -14,20 +11,14 @@
 #' @param parallel ...
 #' @param future ...
 #' @param bppar ...
-#' @param return_ic ...
-#' @param shrink_ic ..
 #' @param tmle_type ...
 #' @param tmle_args ...
 #'
 #' @return Options to be passed to various \code{tmle_*} functions.
 #'
-check_inputs <- function(data, var, vim, type,
-                         filter, filter_cutoff, window, corr, obs_per_covar,
-                         parallel, future, bppar,
-                         return_ic, shrink_ic,
-                         tmle_type, tmle_args) {
+check_inputs <- function(data, var, filter_cutoff, window, corr, obs_per_covar,
+                         parallel, future, bppar, tmle_type, tmle_args) {
 
-  message("This function is only partially implemented currently.")
   # NOTE: make sure to check combinations of the variable of interest and the
   # stated parameter of interest (i.e., VIM). In particular, we will use the
   # following heuristic to decide between NPVI and ATE: if the variable of
@@ -37,27 +28,6 @@ check_inputs <- function(data, var, vim, type,
   # heuristic, but use it to check that the user has specified the ATE vs. NPVI
   # parameters correctly.
 
-  if (vim == "npvi" & is.null(tmle_args$npvi_descr)) {
-    npvi_descr_defaults <- list(f = identity, iter = 10, cvControl = 2,
-                                nMax = 30,
-                                stoppingCriteria = list(mic = 0.001,
-                                                        div = 0.001,
-                                                        psi = 0.01)
-                               )
-    tmle_args$npvi_descr <- npvi_descr_defaults
-  }
-
-  if (tmle_type == "glm") {
-    if (vim == "ate") {
-      # set GLM libraries for "tmle" package
-      g_lib <- "SL.mean"
-      Q_lib <- "SL.glm"
-      tmle_args$g_lib <- g_lib
-      tmle_args$Q_lib <- Q_lib
-    } else if (vim == "npvi") {
-      message("finding a way to select GLM flavor for tmle.npvi")
-    }
-  }
-
   #  TODO: all other arguments in catch_inputs
+  message("Checking function inputs has not yet been implemented.")
 }

### 2017-08-28 R/methyvim.R

@@ -1,3 +1,5 @@
+utils::globalVariables(c("colData<-"))
+
 #' Differential Methylation Statistics with Variable Importance Measures
 #'
 #' Computes the Targeted Minimum Loss-Based Estimate of a specified statistical
@@ -79,6 +81,9 @@
 #' @importFrom parallel detectCores
 #' @importFrom doParallel registerDoParallel
 #' @importFrom foreach foreach "%dopar%"
+#' @importFrom BiocParallel register bpprogressbar DoparParam
+#' @importFrom future plan multiprocess sequential
+#' @importFrom doFuture registerDoFuture
 #'
 #' @export
 #'
@@ -117,23 +122,52 @@ methyvim <- function(data_grs,
   # ============================================================================
   # check that inputs satisfy expectations
   ## and clean up check_inputs (e.g., rm NPVI stuff if ATE is specified)
-  #check_inputs(data = data_grs,
-  #             var = var_int,
-  #             vim = vim,
-  #             type = type,
-  #             filter = filter,
-  #             filter_cutoff = filter_cutoff,
-  #             window = window_bp,
-  #             corr = corr_max,
-  #             obs_per_covar = obs_per_covar,
-  #             parallel = parallel,
-  #             future = future_param,
-  #             bppar = bppar_type,
-  #             return_ic = return_ic,
-  #             shrink_ic = shrink_ic,
-  #             tmle_type = tmle_type,
-  #             tmle_args = tmle_args
-  #            )
+  check_inputs(data = data_grs,
+               var = var_int,
+               filter_cutoff = filter_cutoff,
+               window = window_bp,
+               corr = corr_max,
+               obs_per_covar = obs_per_covar,
+               parallel = parallel,
+               future = future_param,
+               bppar = bppar_type,
+               tmle_type = tmle_type,
+               tmle_args = tmle_args
+              )
+
+  # ============================================================================
+  # set treatment mechanism and outcome regression libraries
+  # ============================================================================
+  if(is.null(tmle_args$g_lib)) {
+    tmle_args$g_lib <- c("SL.mean", "SL.glm", "SL.glm.interaction")
+  }
+  if(is.null(tmle_args$Q_lib)) {
+    tmle_args$Q_lib <- c("SL.mean", "SL.glm", "SL.gam", "SL.earth")
+  }
+
+  # ============================================================================
+  # modify arguments to TMLE functions based on type of type requested
[…]

### 2017-08-28 R/methytmle.R
records target parameter in object

@@ -13,6 +13,7 @@
        slots = list(call = "call",
                     screen_ind = "numeric",
                     clusters = "numeric",
+                    param = "character",
                     vim = "data.frame",
                     ic = "matrix"),
        contains = "GenomicRatioSet"

### 2017-08-28 R/methyvim.R

@@ -170,7 +170,7 @@ methyvim <- function(data_grs,
   }
 
   # ============================================================================
-  # invoke S4 class constructor for "methadapt" object
+  # invoke S4 class constructor for object
   # ============================================================================
   methy_tmle <- .methytmle(data_grs)
   methy_tmle@call <- call
@@ -212,7 +212,7 @@ methyvim <- function(data_grs,
   methy_tmle <- cluster_sites(methytmle = methy_tmle)
 
   #=============================================================================
-  # TMLEs for the Average Treatment Effect (ATE) and Riks Ratio (RR) parameters
+  # TMLEs for the Average Treatment Effect (ATE) and Risk Ratio (RR) parameters
   # ============================================================================
   if (vim %in% c("ate", "rr")) {
     # make sure that the outcome data is of class numeric
@@ -277,11 +277,13 @@ methyvim <- function(data_grs,
       colnames(methy_vim_out) <- c("lowerCI_ATE", "est_ATE", "upperCI_ATE",
                                    "Var_ATE", "pval", "n_neighbors_all",
                                    "n_neighbors_w", "max_corr_w")
+      methy_tmle@param <- "Average Treatment Effect"
     } else {
       colnames(methy_vim_out) <- c("lowerCI_logRR", "est_logRR",
                                    "upperCI_logRR", "Var_logRR", "pval",
                                    "n_neighbors_all", "n_neighbors_w",
                                    "max_corr_w")
+      methy_tmle@param <- "Risk Ratio"
     }
 
     rownames(methy_vim_out) <- cpg_screened_names

### 2017-08-28 R/methyvim.R
remove foreach

@@ -78,9 +78,6 @@ utils::globalVariables(c("colData<-"))
 #'
 #' @importFrom SummarizedExperiment colData
 #' @importFrom BiocParallel bplapply
-#' @importFrom parallel detectCores
-#' @importFrom doParallel registerDoParallel
-#' @importFrom foreach foreach "%dopar%"
 #' @importFrom BiocParallel register bpprogressbar DoparParam
 #' @importFrom future plan multiprocess sequential
 #' @importFrom doFuture registerDoFuture
@@ -181,10 +178,6 @@ methyvim <- function(data_grs,
   set_parallel(parallel = parallel,
                future_param = future_param,
                bppar_type = bppar_type)
-  #if (parallel == TRUE) {
-  #  n_cores <- parallel::detectCores()
-  #  doParallel::registerDoParallel(n_cores)
-  #}
 
   #=============================================================================
   # check if there is missing data in the phenotype-level matrix and drop if so
@@ -231,11 +224,8 @@ methyvim <- function(data_grs,
     # object of screened CpG site indices to loop over in TMLE procedure
     methy_tmle_ind <- seq_along(methy_tmle@screen_ind)
 
-    ## TODO: THIS IS FOR TESTING ONLY
-    methy_tmle_ind <- methy_tmle_ind[seq_len(sites_comp)]
-
-    methy_vim_out <- BiocParallel::bplapply(methy_tmle_ind,
-                                            methyvim_tmle,
+    methy_vim_out <- BiocParallel::bplapply(methy_tmle_ind[seq_len(sites_comp)],
+                                            FUN = methyvim_tmle,
                                             methytmle_screened = methy_tmle,
                                             var_of_interest = var_of_interest,
                                             type = type,
@@ -248,29 +238,6 @@ methyvim <- function(data_grs,
                                             return_ic = return_ic
                                            )
     methy_vim_out <- do.call(rbind.data.frame, methy_vim_out)
-
-    #methy_vim_out <- foreach::foreach(i_site = methy_tmle_ind,
-    #                                  .export = ls(envir = globalenv()),
-    #                                  .packages = c("tmle", "SuperLearner"),
-    #                                  .combine = rbind) %dopar% {
-    #
-    #  message(paste("Computing targeted estimate for site", i_site, "of",
-    #                length(methy_tmle_ind)))
-    #
-    #  out <- methyvim_tmle(target_site = i_site,
-    #                       methytmle_screened = methy_tmle,
-    #                       var_of_interest = var_of_interest,
[…]

### 2017-08-29 README.Rmd
pkg clean up

@@
 ---
-output: github_document
-bibliography: refs.bib
+output: rmarkdown::github_document
+bibliography: README-refs.bib
 ---
@@
 
-* The _average treatment effect_ (ATE):
-  $$ \Psi_{ATE} = E[E[Y_j \mid A = 1, W_{-j}] - E[Y_j \mid A = 0, W_{-j}]],$$
-  where $j$ indexes a target CpG site of interest and $-j$ gives all CpG sites
-  in the same neighborhood as the target site except the target site itself
-  (i.e., the neighboring sites).
+* The _average treatment effect_ (ATE): The effect of a binary exposure or
+  treatment on the observed methylation at a target CpG site is estimated,
+  controlling for the observed methylation at all other CpG sites in the same
+  neighborhood as the target site, based on an additive form. In particular, the
+  parameter estimate represents the __additive difference__ in methylation that
+  would have been observed at the target site had all observations received the
+  treatment versus the scenario in which none received the treatment.
 
-* The _relative risk_ (RR), where
-  $$ \dots,$$
-  where...
+* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
+  observed methylation at a target CpG site is estimated, controlling for the
+  observed methylation at all other CpG sites in the same neighborhood as the
+  target site, based on an geometric form. In particular, the parameter estimate
+  represents the __multiplicative difference__ in methylation that would have
+  been observed at the target site had all observations received the treatment
+  versus the scenario in which none received the treatment.
 
@@
 
-* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation].
+* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+  The effect of continous-valued exposure or treatment (the observed methylation
+  at a target CpG site) on an outcome of interest is estimated, controlling for
+  the observed methylation at all other CpG sites in the same neighborhood as
+  the target (treatment) site, based on a parameter that compares values of the
+  treatment against a reference value taken to be the null. In particular, the
+  implementation provided is designed to assess the effect of differential
+  methylation at the target CpG site on a (typically) phenotype-level outcome
+  of interest (e.g., survival), in effect providing an nonparametric evaluation
+  of the impact of methylation at the target site on said outcome.
+
+_In all cases, an estimator of the target parameter is constructed via targeted
+minimum loss-based estimation._
 
@@
 
-&copy; 2017 [Nima S. Hejazi](http://nimahejazi.org), [Alan E.
-Hubbard](http://sph.berkeley.edu/alan-hubbard), [Mark J. van der
-Laan](https://www.stat.berkeley.edu/~laan/)
+&copy; 2017 [Nima S. Hejazi](http://nimahejazi.org)
 
@@
 `LICENSE` for details.
-
----
-
-## References

### 2017-08-29 R/methyvim.R

@@ -76,7 +76,7 @@ utils::globalVariables(c("colData<-"))
 #'         regressions, as well as the original data rotated into influence
 #'         curve space may be returned, if so requested.
 #'
-#' @importFrom SummarizedExperiment colData
+#' @importFrom SummarizedExperiment colData RangedSummarizedExperiment
 #' @importFrom BiocParallel bplapply
 #' @importFrom BiocParallel register bpprogressbar DoparParam
 #' @importFrom future plan multiprocess sequential

### 2017-08-29 R/methytmle.R
add plot method, vignette fixes, minor bug fixes

@@ -13,6 +13,7 @@
        slots = list(call = "call",
                     screen_ind = "numeric",
                     clusters = "numeric",
+                    var_int = "numeric",
                     param = "character",
                     vim = "data.frame",
                     ic = "matrix"),

### 2017-08-29 R/methyvim.R

@@ -12,8 +12,7 @@ utils::globalVariables(c("colData<-"))
 #'        standard data structures associated with DNA Methylation experiments.
 #'        Consult the documentation for \code{minfi} to construct such objects.
 #' @param sites_comp Numeric indicating the number of sites for which a variable
-#'        importance measure is to be estimated post-screening. This arugment is
-#'        TEMPORARY, USED ONLY FOR TESTING PURPOSES currently.
+#'        importance measure is to be estimated post-screening.
 #' @param var_int Numeric indicating the column index of the variable of
 #'        interest, whether exposure or outcome. If argument \code{vim} is set
 #'        to the ATE, then the variable of interest is treated as an exposure;
@@ -76,7 +75,7 @@ utils::globalVariables(c("colData<-"))
 #'         regressions, as well as the original data rotated into influence
 #'         curve space may be returned, if so requested.
 #'
-#' @importFrom SummarizedExperiment colData RangedSummarizedExperiment
+#' @importFrom SummarizedExperiment colData
 #' @importFrom BiocParallel bplapply
 #' @importFrom BiocParallel register bpprogressbar DoparParam
 #' @importFrom future plan multiprocess sequential
@@ -104,15 +103,16 @@ methyvim <- function(data_grs,
                                       g_lib = NULL, Q_lib = NULL,
                                       npvi_cutoff = 0.25, npvi_descr = NULL)
                     ) {
-# if you say glm it will choose mean and glm
-# if you choose super learner it will choose more algorithms
   # ============================================================================
-  # catch input and return in output object for user convenience
+  # catch input for user convenience and check input types where possible
   # ============================================================================
+  # catch function call
   call <- match.call(expand.dots = TRUE)
-  filter <- match.arg(filter)
-  type <- match.arg(type)
+  # type checking (in same order the arguments appear)
   vim <- match.arg(vim)
+  type <- match.arg(type)
+  filter <- match.arg(filter)
+  tmle_type <- match.arg(tmle_type)
 
   # ============================================================================
   # set treatment mechanism and outcome regression libraries
@@ -153,6 +153,7 @@ methyvim <- function(data_grs,
   # ============================================================================
   methy_tmle <- .methytmle(data_grs)
   methy_tmle@call <- call
+  methy_tmle@var_int <- var_int
 
   #=============================================================================
   # set up parallelization if so desired

### 2017-08-29 R/plots.R

@@ -1,33 +1,169 @@
-#' Common plots for methytmle objects
+utils::globalVariables(c("..count..", "color", "log_pval", "param pval",
+                         "param", "pval"))
+
+#' Plot utility for methytmle objects
 #'
 #' Several types of plots providing...
 #'
 #'
-#' @param x object of class \code{biotmle} as produced by an appropriate call to
-#'        \code{biomarkertmle}
-#' @param type character describing whether to provide a plot of unadjusted or
-#'        adjusted p-values (adjustment performed via Benjamini-Hochberg)
-#' @param ... additional arguments passed \code{plot} as necessary
+#' @param x Object of class \code{methytmle} as produced by an appropriate call
+#'        to \code{methyvim}.
+#' @param ... additional arguments passed \code{superheat}. This argument is
+#'        used only if \code{type} below is set to the heatmap option. Please
+#'        consult the documentation of the \code{superheat} package for options
+#'        that may be appropriately set.
+#' @param type Character describing the type of plot to be generated: one among
+#'        the following collection: heatmap, volcano plot, or histograms of
+#'        raw and FDR-corrected p-values.
+#' @param num_sites Numeric indcating the number of CpG sites to be shown in the
+#'        plot. This argument is only used if \code{type} is set to the heatmap
+#'        option. If the number of sites analyzed is greater than this cutoff,
+#'        sites to be displayed are chosen by ranking sites based on the raw
+#'        (marginal) p-values.
+#' @param param_bound Numeric for a threshold indicating the magnitude of the
+#'        size of the effect considered to be interesting. This is used to
+#'        assign groupings and colors to individual CpG sites. This argument is
+#'        only used when \code{type} is set to the volcano plot option.
+#' @param pval_bound Numeric for a threshold indicating the magnitude of
+#'        p-values deemed to be interesting. This is used to assign groupings
+#'        and colors to individual CpG sites. This argument is only used when
+#'        \code{type} is set to the volcano plot option.
 #'
-#'
-#' @return object of class \code{ggplot} containing one of several types of
-#'         plots: a heatmap of the top 50 CpG sites, a volcano plot displaying
[…]

### 2017-08-29 R/utils.R

@@ -48,7 +48,7 @@ fdr_msa <- function(pvals, total_obs) {
   pvals_not_tested <- rep(1, total_obs - length(pvals))
   pvals_all <- c(pvals, pvals_not_tested)
   fdr_adj <- stats::p.adjust(pvals_all, method = "fdr")
-  fdr_out <- fdr_adj[seq_len(pvals)]
+  fdr_out <- fdr_adj[seq_along(pvals)]
   return(fdr_out)
 }
 

### 2017-09-15 paper/paper_BiocF1000.tex

--- 2017-05-19 paper_BiocF1000/paper_BiocF1000.tex
@@
 \newtheorem{example}{Example}
+\theoremstyle{definition}
+\newtheorem{exercise}{Exercise}
 \theoremstyle{remark}
 \newtheorem*{remark}{Remark}
+\newtheorem*{solution}{Solution}
 \begin{document}
@@
 
-\author[1]{Nima Hejazi}
+\author[1]{Nima S. Hejazi}
+\author[1]{Rachael Phillips}
 \author[2]{Alan E. Hubbard}
@@
 \section*{Keywords}
-Please list up to eight keywords to help readers interested in your
-article find it more easily.
+DNA methylation, differential methylation, epigenetics, causal
+inference, variable importance, targeted minimum loss-based estimation.
 
@@
 using square brackets and use an @ plus the key for the entry you are
-referencing {[}@Smith:2012qr{]}. You can combine multiple entries by
-separating them with a semi-colon {[}@Smith:2012qr; @Smith:2013jd{]}.
+referencing (Smith and Jones 2012). You can combine multiple entries by
+separating them with a semi-colon (Smith and Jones 2012; Jones and Smith
+2013).
 
@@
 
+DNA methylation is a fundamental epigenetic process known to play an
+important role in controlling gene expression. It remains perhaps the
+best studied biological mechanism amongst all known epigenetic processes
+of its kind. With relatively recent advances in computational biology
+and bioinformatics, probing DNA methylation has become rather easily
+possible -- in fact, modern assays allow for DNA methylation signatures
+at up to \(850,000\) CpG sites to be measured simultaneously. Most
+statistical techniques available for the analysis of data produced by
+such assays rely on (generalized) linear models. Here, we provide an
+alternative to such approaches. Specifically, we provide a range of
+\emph{variable importance measures} (\textbf{VIM}), parameters that
+arise in statistical causal inference, for which targeted minimum
+loss-based estimates may be readily computed based on the data made
+available by DNA methylation assays. \texttt{methyvim} is an R package
+that provides facilities for performing differential methylation
+analyses within exactly this scope.
+
+For a general discussion of the framework of targeted minimum loss-based
+estimation and the role this approach plays in statistical causal
+inference, the interested reader is invited to consult M. J. van der
+Laan and Rose (2011) and M. J. van der Laan and Rose (2017). For a more
+general introduction to (statistical) causal inference, Pearl (2009) and
+Hernan and Robins (2018, forthcoming) may be of interest.
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsection{Methodology}\label{methodology}
+
+The core functionality of this package is made available via the
+eponymous \texttt{methyvim} function, which implements a statistical
+algorithm designed to compute targeted estimates of VIMs, defined in
+such a way that the VIMs represent parameters of scientific interest in
+computational biology experiments; moreover, these VIMs are defined such
+that they may be estimated in a manner that is very nearly
+assumption-free, that is, within a fully nonparametric statistical
+model. \textbf{The statistical algorithm consists in several major
+steps:}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Pre-screening of genomic sites is used to isolate a subset of sites
+  for which there is cursory evidence of differential methylation. For
+  the sake of computational feasibility, targeted minimum loss-based
+  estimates of VIMs are computed only for this subset of sites.
+  Currently, the available screening approaches adapts core routines
+  from the
+  \href{http://bioconductor.org/packages/release/bioc/html/limma.html}{\texttt{limma}}
+  R package, thought future releases will support functionality from
+  other packages (e.g.,
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}).
+\item
+  Nonparametric estimates of VIMs (for the specified target parameter)
+  are currently computed by adapting routines from the
+  \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R
+  package. Future releases will support doubly-robust estimates of these
+  VIMs (via the
+  \href{https://cran.r-project.org/web/packages/drtmle/index.html}{\texttt{drtmle}}
+  package) and add parameters for continuous treatments/exposures (via
+  the
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}
+  package).
+\item
+  Since pre-screening is performed prior to estimating VIMs, we make use
+  of a multiple testing correction uniquely suited to such settings. Due
+  to the multiple testing nature of the estimation problem, a variant of
+  the Benjamini \& Hochberg procedure for controlling the False
+  Discovery Rate (FDR) is applied (Benjamini and Hochberg 1995).
+  Specifically, we apply the modified marginal Benjamini \& Hochberg
+  step-up False Discovery Rate controlling procedure for multi-stage
+  analyses (FDR-MSA), which is guaranteed to control the FDR as if all
+  sites were tested (Tuglus and van der Laan 2009).
+\end{enumerate}
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsection{Parameters of Interest}\label{parameters-of-interest}
+
+For discrete-valued treatments or exposures:
+
+\begin{itemize}
+\item
+  The \emph{average treatment effect} (ATE): The effect of a binary
+  exposure or treatment on the observed methylation at a target CpG site
+  is estimated, controlling for the observed methylation at all other
+  CpG sites in the same neighborhood as the target site, based on an
+  additive form. In particular, the parameter estimate represents the
+  \textbf{additive difference} in methylation that would have been
+  observed at the target site had all observations received the
+  treatment versus the scenario in which none received the treatment.
+\item
+  The \emph{relative risk} (RR): The effect of a binary exposure or
+  treatment on the observed methylation at a target CpG site is
+  estimated, controlling for the observed methylation at all other CpG
+  sites in the same neighborhood as the target site, based on an
+  geometric form. In particular, the parameter estimate represents the
+  \textbf{multiplicative difference} in methylation that would have been
+  observed at the target site had all observations received the
+  treatment versus the scenario in which none received the treatment.
+\end{itemize}
+
+Support for continuous-valued treatments or exposures is \emph{planned
+but not yet available}, though work is underway to incorporate into our
+methodology the following
+
+\begin{itemize}
+\item
+  A \emph{nonparametric variable importance measure} (NPVI) (Chambaz,
+  Neuvial, and van der Laan 2012): The effect of continous-valued
+  exposure or treatment (the observed methylation at a target CpG site)
+  on an outcome of interest is estimated, controlling for the observed
+  methylation at all other CpG sites in the same neighborhood as the
+  target (treatment) site, based on a parameter that compares values of
+  the treatment against a reference value taken to be the null. In
+  particular, the implementation provided is designed to assess the
+  effect of differential methylation at the target CpG site on a
+  (typically) phenotype-level outcome of interest (e.g., survival), in
+  effect providing an nonparametric evaluation of the impact of
+  methylation at the target site on said outcome.
+\end{itemize}
+
+\emph{As previously noted, in all cases, an estimator of the target
+parameter is constructed via targeted minimum loss-based estimation.}
+
+Having now discussed the foundational principles of the estimation
+procedure employed and the statistical algorithm implemented, it is best
+to proceed by examining \texttt{methyvim} by example.
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsection{Preliminaries: Setting up the
+Data}\label{preliminaries-setting-up-the-data}
+
+First, we'll load the \texttt{methyvim} package and the example data
+contained in the \texttt{methyvimData} package that accompanies it:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(methyvim)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## methyvim: Nonparametric Differential Methylation Analysis with Variable Importance Measures
+\end{verbatim}
+
+\begin{verbatim}
+## Version: 0.99.1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(methyvimData)}
+\end{Highlighting}
+\end{Shaded}
+
+Now, let's load the data set and seed the RNG:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{data}\NormalTok{(grsExample)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{479253}\NormalTok{)}
+\NormalTok{grsExample}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## class: GenomicRatioSet 
+## dim: 400 210 
+## metadata(0):
+## assays(2): Beta M
+## rownames(400): cg23578515 cg06747907 ... cg01715842 cg09895959
+## rowData names(0):
+## colnames(210): V2 V3 ... V397 V398
+## colData names(2): exp outcome
+## Annotation
+##   array: IlluminaHumanMethylationEPIC
+## Preprocessing
+##   Method: NA
+##   minfi version: NA
+##   Manifest version: NA
+\end{verbatim}
+
+The example data object is of class \texttt{GenomicRatioSet}, provided
+by the \texttt{minfi} package. The summary provided by the
+\texttt{print} method gives a wealth of information on the experiment
+that generated the data -- since we are working with a simulated data
+set, we need not concern ourselves with much of this information. While
+most slots of objects of this S4 class have descriptive names, the
+interested analyst might consider consulting the documentation of the
+\texttt{minfi} package for an in-depth description (Aryee et al. 2014).
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsection{Quantifying the Effect of an Exposure on DNA
+Methylation}\label{quantifying-the-effect-of-an-exposure-on-dna-methylation}
+
+\subsubsection{The Average Treatment Effect as Variable Importance
+Measure}\label{the-average-treatment-effect-as-variable-importance-measure}
+
+\ldots{}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{suppressMessages}\NormalTok{(}
+\NormalTok{  methyvim_out_ate <-}\StringTok{ }\KeywordTok{methyvim}\NormalTok{(}\DataTypeTok{data_grs =}\NormalTok{ grsExample, }\DataTypeTok{sites_comp =} \DecValTok{10}\NormalTok{,}
+                               \DataTypeTok{var_int =} \DecValTok{1}\NormalTok{, }\DataTypeTok{vim =} \StringTok{"ate"}\NormalTok{, }\DataTypeTok{type =} \StringTok{"Mval"}\NormalTok{,}
+                               \DataTypeTok{filter =} \StringTok{"limma"}\NormalTok{, }\DataTypeTok{filter_cutoff =} \FloatTok{0.05}\NormalTok{,}
+                               \DataTypeTok{parallel =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{tmle_type =} \StringTok{"sl"}
+\NormalTok{                              )}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in set_parallel(parallel = parallel, future_param = future_param, : Sequential evaluation is strongly discouraged. 
+##  Proceed with caution.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{methyvim_out_ate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## class: methytmle 
+## dim: 400 210 
+## metadata(0):
+## assays(2): Beta M
+## rownames(400): cg23578515 cg06747907 ... cg01715842 cg09895959
+## rowData names(0):
+## colnames(210): V2 V3 ... V397 V398
+## colData names(2): exp outcome
+## Annotation
+##   array: IlluminaHumanMethylationEPIC
+## Preprocessing
+##   Method: NA
+##   minfi version: NA
+##   Manifest version: NA
+\end{verbatim}
+
+As is clear from examining the object \texttt{methyvim\_out\_ate}, the
+output resembles exactly that returned when examining objects of class
+\texttt{GenomicRatioSet} from the \texttt{minfi} R package. In
+particular, the returned \texttt{methytmle} object is merely a modified
+form (in particular, a subclass) of the input \texttt{GenomicRatioSet}
+object -- thus, it contains all of the original slots, with all
+experimental data intact. Several extra pieces of information are
+contained within the output object as well\_.
+
+We can take a look at the results produced from the estimation procedure
+by examining the \texttt{"vim"} \texttt{slot} of the \texttt{methytmle}
+object:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{head}\NormalTok{(}\KeywordTok{slot}\NormalTok{(methyvim_out_ate, }\StringTok{"vim"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##            lowerCI_ATE     est_ATE upperCI_ATE    Var_ATE      pval
+## cg22913481  -0.3790600 -0.10205080   0.1749584 0.01997451 0.4702524
+## cg15131207  -0.3121142 -0.10416890   0.1037764 0.01125605 0.3261739
+## cg10613282  -0.4272500 -0.16315679   0.1009364 0.01815525 0.2259382
+## cg15857610  -0.3186500 -0.07896784   0.1607144 0.01495407 0.5184354
+## cg24775884  -0.4696981  0.03989038   0.5494788 0.06759693 0.8780607
+## cg22954484  -0.3289972  0.07614832   0.4812939 0.04272775 0.7125840
+##            n_neighbors_all n_neighbors_w max_corr_w
+## cg22913481               0             0         NA
+## cg15131207               0             0         NA
+## cg10613282               0             0         NA
+## cg15857610               5             4  0.7720705
+## cg24775884               5             2  0.8897566
+## cg22954484               5             2  0.8643936
+\end{verbatim}
+
+From the table displayed, we note that we have access to point estimates
+of the ATE (``est\_ATE'') as well as lower and upper confidence interval
+bounds for each estimate (``lowerCI\_ATE'' and ``upperCI\_ATE'',
+respectively). Additional statistical information we have access to
+include the variance (``Var\_ATE'') of the estimate as well as the
+p-value (``pval'') associated with each estimate (based on Wald-style
+testing procedures). Beyond these, key bioinformatical quantities (with
+respect to the algorithm outlined above) are also returned; these
+include the total number of neighbors of the target site, the number of
+neighboring sites controlled for when estimating the effect of exposure
+on DNA methylation, and, finally, the maximum correlation between the
+target site and any given site in its full set of neighbors.
+
+\ldots{}
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsubsection{The Risk Ratio as Variable Importance
+Measure}\label{the-risk-ratio-as-variable-importance-measure}
+
+\ldots{}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{methyvim_out_rr <-}\StringTok{ }\KeywordTok{methyvim}\NormalTok{(}\DataTypeTok{data_grs =}\NormalTok{ grsExample, }\DataTypeTok{sites_comp =} \DecValTok{10}\NormalTok{,}
+                            \DataTypeTok{var_int =} \DecValTok{1}\NormalTok{, }\DataTypeTok{vim =} \StringTok{"rr"}\NormalTok{, }\DataTypeTok{type =} \StringTok{"Mval"}\NormalTok{,}
+                            \DataTypeTok{filter =} \StringTok{"limma"}\NormalTok{, }\DataTypeTok{filter_cutoff =} \FloatTok{0.05}\NormalTok{,}
+                            \DataTypeTok{parallel =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{tmle_type =} \StringTok{"sl"}
+\NormalTok{                           )}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in set_parallel(parallel = parallel, future_param = future_param, : Sequential evaluation is strongly discouraged. 
+##  Proceed with caution.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{methyvim_out_rr}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## class: methytmle 
+## dim: 400 210 
+## metadata(0):
+## assays(2): Beta M
+## rownames(400): cg23578515 cg06747907 ... cg01715842 cg09895959
+## rowData names(0):
+## colnames(210): V2 V3 ... V397 V398
+## colData names(2): exp outcome
+## Annotation
+##   array: IlluminaHumanMethylationEPIC
+## Preprocessing
+##   Method: NA
+##   minfi version: NA
+##   Manifest version: NA
+\end{verbatim}
+
+\ldots{}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{head}\NormalTok{(}\KeywordTok{slot}\NormalTok{(methyvim_out_rr, }\StringTok{"vim"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##            lowerCI_logRR    est_logRR upperCI_logRR    Var_logRR      pval
+## cg22913481   -0.15978435 -0.042980967    0.07382242 0.0035513930 0.4707649
+## cg15131207   -0.10832796 -0.036151215    0.03602553 0.0013560711 0.3262445
+## cg10613282   -0.18349779 -0.069866731    0.04376433 0.0033611044 0.2281579
+## cg15857610   -0.07196023 -0.018317791    0.03532465 0.0007490399 0.5033043
+## cg24775884   -0.07914787  0.006261306    0.09167049 0.0018988776 0.8857479
+## cg22954484   -0.06728830  0.015365555    0.09801941 0.0017783369 0.7155826
+##            n_neighbors_all n_neighbors_w max_corr_w
+## cg22913481               0             0         NA
+## cg15131207               0             0         NA
+## cg10613282               0             0         NA
+## cg15857610               5             4  0.7720705
+## cg24775884               5             2  0.8897566
+## cg22954484               5             2  0.8643936
+\end{verbatim}
+
+\ldots{}
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsection{Visualization of Results}\label{visualization-of-results}
+
+\ldots{}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(methyvim_out_rr, }\DataTypeTok{type =} \StringTok{"volcano"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\ldots{}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(methyvim_out_rr, }\DataTypeTok{type =} \StringTok{"pvals"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\ldots{}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(methyvim_out_rr, }\DataTypeTok{type =} \StringTok{"heatmap"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\ldots{}
+
+\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}
+
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
+\hypertarget{refs}{}
+\hypertarget{ref-aryee2014minfi}{}
+Aryee, Martin J, Andrew E Jaffe, Hector Corrada-Bravo, Christine
+Ladd-Acosta, Andrew P Feinberg, Kasper D Hansen, and Rafael A Irizarry.
+2014. ``Minfi: A Flexible and Comprehensive Bioconductor Package for the
+Analysis of Infinium DNA Methylation Microarrays.''
+\emph{Bioinformatics} 30 (10). Oxford University Press (OUP): 1363--9.
+doi:\href{https://doi.org/10.1093/bioinformatics/btu049}{10.1093/bioinformatics/btu049}.
+
+\hypertarget{ref-benjamini1995controlling}{}
+Benjamini, Yoav, and Yosef Hochberg. 1995. ``Controlling the False
+Discovery Rate: A Practical and Powerful Approach to Multiple Testing.''
+\emph{Journal of the Royal Statistical Society. Series B
+(Methodological)}. JSTOR, 289--300.
+
+\hypertarget{ref-chambaz2012estimation}{}
+Chambaz, Antoine, Pierre Neuvial, and Mark J van der Laan. 2012.
+``Estimation of a Non-Parametric Variable Importance Measure of a
+Continuous Exposure.'' \emph{Electronic Journal of Statistics} 6. NIH
+Public Access: 1059.
+
+\hypertarget{ref-hernan2018causal}{}
+Hernan, Miguel A, and James M Robins. 2018, forthcoming. \emph{Causal
+Inference}. Chapman \& Hall/Crc Texts in Statistical Science. Taylor \&
+Francis.
+
+\hypertarget{ref-Smith:2013jd}{}
+Jones, A. B., and J. M. Smith. 2013. ``Article Title.'' \emph{Journal
+Title} 13 (52). Publisher: 123--456.
+
+\hypertarget{ref-pearl2009causality}{}
+Pearl, Judea. 2009. \emph{Causality: Models, Reasoning, and Inference}.
+Cambridge University Press.
+
+\hypertarget{ref-Smith:2012qr}{}
+Smith, J. M., and A. B. Jones. 2012. \emph{Book Title}. 7th ed.
+Publisher.
+
+\hypertarget{ref-tuglus2009modified}{}
+Tuglus, Catherine, and Mark J. van der Laan. 2009. ``Modified FDR
+Controlling Procedure for Multi-Stage Analyses.'' \emph{Statistical
+Applications in Genetics and Molecular Biology} 8 (1). Walter de
+Gruyter: 1--15.
+doi:\href{https://doi.org/10.2202/1544-6115.1397}{10.2202/1544-6115.1397}.
+
+\hypertarget{ref-vdl2011targeted}{}
+van der Laan, Mark J, and Sherri Rose. 2011. \emph{Targeted Learning:
+Causal Inference for Observational and Experimental Data}. Springer
+Science \& Business Media.
+
+\hypertarget{ref-vdl2017targeted}{}
+---------. 2017. \emph{Targeted Learning in Data Science: Causal
+Inference for Complex Longitudinal Studies}. Springer Science \&
+Business Media.
+
 \end{document}

### 2017-09-15 paper/paper_BiocF1000.Rmd

--- 2017-07-14 paper_F1000Bioc/paper_BiocF1000.Rmd
@@
 author:
-  - name: Nima Hejazi
+  - name: Nima S. Hejazi
+    affiliation: Address of author1
+  - name: Rachael Phillips
     affiliation: Address of author1
@@
     affiliation: Address of author2
-abstract: Abstracts should be up to 300 words and provide a succinct summary of the article.
-Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article.
-Citations should not be used in the abstract, and the use of abbreviations should be minimized.
-keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
-#bibliography: paper_BiocF1000.bib
+abstract: Abstracts should be up to 300 words and provide a succinct summary of
+  the article. Although the abstract should explain why the article might be
+  interesting, care should be taken not to inappropriately over-emphasise the
+  importance of the work described in the article. Citations should not be used
+  in the abstract, and the use of abbreviations should be minimized.
+keywords: DNA methylation, differential methylation, epigenetics, causal
+  inference, variable importance, targeted minimum loss-based estimation.
+bibliography: paper_BiocF1000.bib
 output: BiocWorkflowTools::f1000_article
@@
 
-The introduction provides context as to why the software tool was developed and what need it addresses.
-It is good scholarly practice to mention previously developed tools that address similar needs, and why the current tool is needed.
+The introduction provides context as to why the software tool was developed and
+what need it addresses. It is good scholarly practice to mention previously
+developed tools that address similar needs, and why the current tool is needed.
 
@@
 ## Implementation
-For software tool papers, this section should address how the tool works and any relevant technical details required for implementation of the tool by other developers.
+For software tool papers, this section should address how the tool works and any
+relevant technical details required for implementation of the tool by other
+developers.
 
 ## Operation
-This part of the methods should include the minimal system requirements needed to run the software and an overview of the workflow for the tool for users of the tool.
+This part of the methods should include the minimal system requirements needed
+to run the software and an overview of the workflow for the tool for users of
+the tool.
 
 # Results <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and should be written as a traditional results section.
+This section is only required if the paper includes novel data or analyses, and
+should be written as a traditional results section.
 
@@
 This section is required if the paper does not include novel data or analyses.
-Examples of input and output files should be provided with some explanatory context.
-Any novel or complex variable parameters should also be explained in sufficient detail to allow users to understand and use the tool's functionality.
+Examples of input and output files should be provided with some explanatory
+context.  Any novel or complex variable parameters should also be explained in
+sufficient detail to allow users to understand and use the tool's functionality.
 
 # Discussion <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and should be written in the same style as a traditional discussion section.
-Please include a brief discussion of allowances made (if any) for controlling bias or unwanted sources of variability, and the limitations of any novel datasets.
+This section is only required if the paper includes novel data or analyses, and
+should be written in the same style as a traditional discussion section.
+Please include a brief discussion of allowances made (if any) for controlling
+bias or unwanted sources of variability, and the limitations of any novel
+datasets.
 
 # Conclusions <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and should be written as a traditional conclusion.
+This section is only required if the paper includes novel data or analyses, and
+should be written as a traditional conclusion.
 
 # Summary <!-- Optional - only if NO new datasets are included -->
-This section is required if the paper does not include novel data or analyses.  It allows authors to briefly summarize the key points from the article.
+This section is required if the paper does not include novel data or analyses.
+It allows authors to briefly summarize the key points from the article.
 
 # Data availability <!-- Optional - only if novel data or analyses are included -->
-Please add details of where any datasets that are mentioned in the paper, and that have not have not previously been formally published, can be found.
-If previously published datasets are mentioned, these should be cited in the references, as per usual scholarly conventions.
+Please add details of where any datasets that are mentioned in the paper, and
+that have not have not previously been formally published, can be found. If
+previously published datasets are mentioned, these should be cited in the
+references, as per usual scholarly conventions.
 
 # Software availability
-This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.
+This section will be generated by the Editorial Office before publication.
+Authors are asked to provide some initial information to assist the Editorial
+Office, as detailed below.
 
@@
 In order to give appropriate credit to each author of an article, the individual
-contributions of each author to the manuscript should be detailed in this section. We
-recommend using author initials and then stating briefly how they contributed.
+contributions of each author to the manuscript should be detailed in this
+section. We recommend using author initials and then stating briefly how they
+contributed.
 
 # Competing interests
-All financial, personal, or professional competing interests for any of the authors that
-could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article.
-If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.
+All financial, personal, or professional competing interests for any of the
+authors that could be construed to unduly influence the content of the article
+must be disclosed and will be displayed alongside the article. If there are no
+relevant competing interests to declare, please add the following: 'No competing
+interests were disclosed'.
 
 # Grant information
-Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc. Please do not list funding that you have that is not relevant to this
-specific piece of research. For each funder, please state the funder’s name, the grant
-number where applicable, and the individual to whom the grant was assigned.
-If your work was not funded by any grants, please include the line: 'The author(s)
-declared that no grants were involved in supporting this work.'
+Please state who funded the work discussed in this article, whether it is your
+employer, a grant funder etc. Please do not list funding that you have that is
+not relevant to this specific piece of research. For each funder, please state
+the funder’s name, the grant number where applicable, and the individual to whom
+the grant was assigned. If your work was not funded by any grants, please
+include the line: 'The author(s) declared that no grants were involved in
+supporting this work.'
 
@@
 This section should acknowledge anyone who contributed to the research or the
-article but who does not qualify as an author based on the criteria provided earlier
-(e.g. someone or an organization that provided writing assistance). Please state how
-they contributed; authors should obtain permission to acknowledge from all those
-mentioned in the Acknowledgments section.
+article but who does not qualify as an author based on the criteria provided
+earlier (e.g. someone or an organization that provided writing assistance).
+Please state how they contributed; authors should obtain permission to
+acknowledge from all those mentioned in the Acknowledgments section.
 
@@
 # USING R MARKDOWN
-Some examples of commonly used markdown syntax are listed below, to help you get started.
+Some examples of commonly used markdown syntax are listed below, to help you get
+started.
 
 ## Cross-references
-For portability between different output formats, use the syntax introduced by *bookdown*, such as `(\#label)` for labels and `\@ref(label)` for cross-references.
-The following sections provide examples of referencing tables, figures, and equations.
+For portability between different output formats, use the syntax introduced by
+*bookdown*, such as `(\#label)` for labels and `\@ref(label)` for
+cross-references. The following sections provide examples of referencing tables,
+figures, and equations.
 
@@
 
-You can include refences in a standard Bibtex file.
-The name of this file is given in the header of the markdown document (in our case it is *sample.bib*).
-References to entries in the Bibtex file are made using square brackets and use an @ plus the key for the entry you are referencing [@Smith:2012qr].
-You can combine multiple entries by separating them with a semi-colon [@Smith:2012qr; @Smith:2013jd].
+You can include refences in a standard Bibtex file.  The name of this file is
+given in the header of the markdown document (in our case it is *sample.bib*).
+References to entries in the Bibtex file are made using square brackets and use
+an @ plus the key for the entry you are referencing [@Smith:2012qr].  You can
+combine multiple entries by separating them with a semi-colon
+[@Smith:2012qr; @Smith:2013jd].
 
@@
         - for sub-items
+
+DNA methylation is a fundamental epigenetic process known to play an important
+role in controlling gene expression. It remains perhaps the best studied
+biological mechanism amongst all known epigenetic processes of its kind. With
+relatively recent advances in computational biology and bioinformatics, probing
+DNA methylation has become rather easily possible -- in fact, modern assays
+allow for DNA methylation signatures at up to $850,000$ CpG sites to be measured
+simultaneously. Most statistical techniques available for the analysis of data
+produced by such assays rely on (generalized) linear models. Here, we provide an
+alternative to such approaches. Specifically, we provide a range of _variable
+importance measures_ (__VIM__), parameters that arise in statistical causal
+inference, for which targeted minimum loss-based estimates may be readily
+computed based on the data made available by DNA methylation assays. `methyvim`
+is an R package that provides facilities for performing differential methylation
+analyses within exactly this scope.
+
+For a general discussion of the framework of targeted minimum loss-based
+estimation and the role this approach plays in statistical causal inference, the
+interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
+For a more general introduction to (statistical) causal inference,
+@pearl2009causality and @hernan2018causal may be of interest.
+
+---
+
+## Methodology
+
+The core functionality of this package is made available via the eponymous
+`methyvim` function, which implements a statistical algorithm designed to
+compute targeted estimates of VIMs, defined in such a way that the VIMs
+represent parameters of scientific interest in computational biology
+experiments; moreover, these VIMs are defined such that they may be estimated in
+a manner that is very nearly assumption-free, that is, within a fully
+nonparametric statistical model. __The statistical algorithm consists in several
+major steps:__
+
+1. Pre-screening of genomic sites is used to isolate a subset of sites for
+  which there is cursory evidence of differential methylation. For the sake of
+  computational feasibility, targeted minimum loss-based estimates of VIMs are
+  computed only for this subset of sites. Currently, the available screening
+  approaches adapts core routines from the
+  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
+  package, thought future releases will support functionality from other
+  packages (e.g., [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)).
+
+2. Nonparametric estimates of VIMs (for the specified target parameter) are
+  currently computed by adapting routines from the
+  [`tmle`](https://CRAN.R-project.org/package=tmle) R package. Future releases
+  will support doubly-robust estimates of these VIMs (via the
+  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
+  and add parameters for continuous treatments/exposures (via the
+  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).
+
+3. Since pre-screening is performed prior to estimating VIMs, we make use of a
+  multiple testing correction uniquely suited to such settings. Due to the
+  multiple testing nature of the estimation problem, a variant of the Benjamini
+  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
+  [@benjamini1995controlling]. Specifically, we apply the modified marginal
+  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
+  multi-stage analyses (FDR-MSA), which is guaranteed to control the FDR as if
+  all sites were tested [@tuglus2009modified].
+
+---
+
+## Parameters of Interest
+
+For discrete-valued treatments or exposures:
+
+* The _average treatment effect_ (ATE): The effect of a binary exposure or
+  treatment on the observed methylation at a target CpG site is estimated,
+  controlling for the observed methylation at all other CpG sites in the same
+  neighborhood as the target site, based on an additive form. In particular, the
+  parameter estimate represents the __additive difference__ in methylation that
+  would have been observed at the target site had all observations received the
+  treatment versus the scenario in which none received the treatment.
+
+* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
+  observed methylation at a target CpG site is estimated, controlling for the
+  observed methylation at all other CpG sites in the same neighborhood as the
+  target site, based on an geometric form. In particular, the parameter estimate
+  represents the __multiplicative difference__ in methylation that would have
+  been observed at the target site had all observations received the treatment
+  versus the scenario in which none received the treatment.
+
+Support for continuous-valued treatments or exposures is _planned but not yet
+available_, though work is underway to incorporate into our methodology the
+following
+
+* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+  The effect of continous-valued exposure or treatment (the observed methylation
+  at a target CpG site) on an outcome of interest is estimated, controlling for
+  the observed methylation at all other CpG sites in the same neighborhood as
+  the target (treatment) site, based on a parameter that compares values of the
+  treatment against a reference value taken to be the null. In particular, the
+  implementation provided is designed to assess the effect of differential
+  methylation at the target CpG site on a (typically) phenotype-level outcome
+  of interest (e.g., survival), in effect providing an nonparametric evaluation
+  of the impact of methylation at the target site on said outcome.
+
+_As previously noted, in all cases, an estimator of the target parameter is
+constructed via targeted minimum loss-based estimation._
+
+Having now discussed the foundational principles of the estimation procedure
+employed and the statistical algorithm implemented, it is best to proceed by
+examining `methyvim` by example.
+
+---
+
+## Preliminaries: Setting up the Data
+
+```{r reqs, echo=FALSE}
+suppressMessages(library(tmle))
+suppressMessages(library(minfi))
+suppressMessages(library(SummarizedExperiment))
+```
+
+First, we'll load the `methyvim` package and the example data contained in the
+`methyvimData` package that accompanies it:
+
+```{r prelims}
+library(methyvim)
+library(methyvimData)
+```
+
+Now, let's load the data set and seed the RNG:
+
+```{r get-data}
+data(grsExample)
+set.seed(479253)
+grsExample
+```
+
+The example data object is of class `GenomicRatioSet`, provided by the `minfi`
+package. The summary provided by the `print` method gives a wealth of
+information on the experiment that generated the data -- since we are working
+with a simulated data set, we need not concern ourselves with much of this
+information. While most slots of objects of this S4 class have descriptive
+names, the interested analyst might consider consulting the documentation of the
+`minfi` package for an in-depth description [@aryee2014minfi].
+
+---
+
+## Quantifying the Effect of an Exposure on DNA Methylation
+
+### The Average Treatment Effect as Variable Importance Measure
+
+...
+
+```{r methyvim-ate}
+suppressMessages(
+  methyvim_out_ate <- methyvim(data_grs = grsExample, sites_comp = 10,
+                               var_int = 1, vim = "ate", type = "Mval",
+                               filter = "limma", filter_cutoff = 0.05,
+                               parallel = FALSE, tmle_type = "sl"
+                              )
+)
+methyvim_out_ate
+```
+
+As is clear from examining the object `methyvim_out_ate`, the output resembles
+exactly that returned when examining objects of class `GenomicRatioSet` from the
+`minfi` R package. In particular, the returned `methytmle` object is merely a
+modified form (in particular, a subclass) of the input `GenomicRatioSet` object
+-- thus, it contains all of the original slots, with all experimental data
+intact. Several extra pieces of information are contained within the output
+object as well_.
+
+We can take a look at the results produced from the estimation procedure by
+examining the `"vim"` `slot` of the `methytmle` object:
+
+```{r methyvim-ate-print}
+head(slot(methyvim_out_ate, "vim"))
+```
+
+From the table displayed, we note that we have access to point estimates of the
+ATE ("est_ATE") as well as lower and upper confidence interval bounds for each
+estimate ("lowerCI_ATE" and "upperCI_ATE", respectively). Additional statistical
+information we have access to include the variance ("Var_ATE") of the estimate
+as well as the p-value ("pval") associated with each estimate (based on
+Wald-style testing procedures). Beyond these, key bioinformatical quantities
+(with respect to the algorithm outlined above) are also returned; these include
+the total number of neighbors of the target site, the number of neighboring
+sites controlled for when estimating the effect of exposure on DNA methylation,
+and, finally, the maximum correlation between the target site and any given site
+in its full set of neighbors.
+
+...
+
+---
+
+### The Risk Ratio as Variable Importance Measure
+
+...
+
+```{r methyvim-rr}
+methyvim_out_rr <- methyvim(data_grs = grsExample, sites_comp = 10,
+                            var_int = 1, vim = "rr", type = "Mval",
+                            filter = "limma", filter_cutoff = 0.05,
+                            parallel = FALSE, tmle_type = "sl"
+                           )
+methyvim_out_rr
+```
+...
+
+```{r methyvim-rr-print}
+head(slot(methyvim_out_rr, "vim"))
+```
+...
+
+---
+
+## Visualization of Results
+
+...
+
+```{r methyvim-plot-volcano, eval=FALSE}
+plot(methyvim_out_rr, type = "volcano")
+```
+
+...
+
+```{r methyvim-plot-pvals, eval=FALSE}
+plot(methyvim_out_rr, type = "pvals")
+```
+
+...
+
+```{r methyvim-plot-heatmap, eval=FALSE}
+plot(methyvim_out_rr, type = "heatmap")
+```
+
+...
+
+---
+
+## References

### 2017-09-15 tests/testthat/test_methyvim.R
fix tests, examples, vignette

@@ -6,21 +6,22 @@ suppressMessages(library(minfi))
 suppressMessages(library(SummarizedExperiment))
 library(methyvimData)
 data(grsExample)
+var_int <- colData(grsExample)[, 1]
 
 
 # run TMLE procedure for the ATE parameter over M-values with Limma filtering
 methyvim_out_ate <- suppressWarnings(
-  methyvim(data_grs = grsExample, sites_comp = 3, var_int = 1, vim = "ate",
+  methyvim(data_grs = grsExample, var_int = var_int, vim = "ate",
            type = "Mval", filter = "limma", filter_cutoff = 0.05,
-           parallel = FALSE, tmle_type = "sl"
+           sites_comp = 3, parallel = FALSE, tmle_type = "sl"
           )
 )
 
 # run TMLE procedure for the RR parameter over M-values with Limma filtering
 methyvim_out_rr <- suppressWarnings(
-  methyvim(data_grs = grsExample, sites_comp = 3, var_int = 1, vim = "rr",
+  methyvim(data_grs = grsExample, var_int = var_int, vim = "rr",
            type = "Mval", filter = "limma", filter_cutoff = 0.05,
-           parallel = FALSE, tmle_type = "sl"
+           sites_comp = 3, parallel = FALSE, tmle_type = "sl"
           )
 )
 

### 2017-10-03 paper/paper_BiocF1000.Rmd
use minfi data
minor vignette additions/corrections

@@
   - name: Rachael Phillips
-    affiliation: Address of author1
+    affiliation: orchid.org/0000-0002-8474-591X
   - name: Alan E. Hubbard
@@
   in the abstract, and the use of abbreviations should be minimized.
+
+  I will do this section last
+
 keywords: DNA methylation, differential methylation, epigenetics, causal
-  inference, variable importance, targeted minimum loss-based estimation.
+  inference, variable importance, targeted minimum loss-based estimation
 bibliography: paper_BiocF1000.bib
@@
 
+--------------------------------------------------------------------------------
+DNA methylation is a fundamental epigenetic process known to play an important
+role in controlling gene expression. It remains perhaps the best studied
+biological mechanism amongst all known epigenetic processes of its kind. With
+relatively recent advances in computational biology and bioinformatics, probing
+DNA methylation has become rather easily possible -- in fact, modern assays
+allow for DNA methylation signatures at up to $850,000$ CpG sites to be measured
+simultaneously. Most statistical techniques available for the analysis of data
+produced by such assays rely on (generalized) linear models. Here, we provide an
+alternative to such approaches. Specifically, we provide a range of _variable
+importance measures_ (__VIM__), parameters that arise in statistical causal
+inference, for which targeted minimum loss-based estimates may be readily
+computed based on the data made available by DNA methylation assays. `methyvim`
+is an R package that provides facilities for performing differential methylation
+analyses within exactly this scope.
+
+For a general discussion of the framework of targeted minimum loss-based
+estimation and the role this approach plays in statistical causal inference, the
+interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
+For a more general introduction to (statistical) causal inference,
+@pearl2009causality and @hernan2018causal may be of interest.
+--------------------------------------------------------------------------------
+
+This methodology targets the individual importance of each CpG site by taking
+advantage of the prediction
+power of machine learning
+algorithms. The importance measurement is derived in the presence of the
+confounding variables of a individual CpG site and hence is stable, unwavering
+to the redundant information among correlated sites.
+
 # Methods
@@
 developers.
+--------------------------------------------------------------------------------
+The core functionality of this package is made available via the eponymous
+`methyvim` function, which implements a statistical algorithm designed to
+compute targeted estimates of VIMs, defined in such a way that the VIMs
+represent parameters of scientific interest in computational biology
+experiments; moreover, these VIMs are defined such that they may be estimated in
+a manner that is very nearly assumption-free, that is, within a fully
+nonparametric statistical model. __The statistical algorithm consists in several
+major steps:__
+
+1. Pre-screening of genomic sites is used to isolate a subset of sites for
+  which there is cursory evidence of differential methylation. For the sake of
+  computational feasibility, targeted minimum loss-based estimates of VIMs are
+  computed only for this subset of sites. Currently, the available screening
+  approaches adapts core routines from the
+  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
+  package, thought future releases will support functionality from other
+  packages (e.g., [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)).
+
+2. Nonparametric estimates of VIMs (for the specified target parameter) are
+  currently computed by adapting routines from the
+  [`tmle`](https://CRAN.R-project.org/package=tmle) R package. Future releases
+  will support doubly-robust estimates of these VIMs (via the
+  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
+  and add parameters for continuous treatments/exposures (via the
+  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).
+
+3. Since pre-screening is performed prior to estimating VIMs, we make use of a
+  multiple testing correction uniquely suited to such settings. Due to the
+  multiple testing nature of the estimation problem, a variant of the Benjamini
+  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
+  [@benjamini1995controlling]. Specifically, we apply the modified marginal
+  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
+  multi-stage analyses (FDR-MSA), which is guaranteed to control the FDR as if
+  all sites were tested [@tuglus2009modified].
+  ## Parameters of Interest
+
+  For discrete-valued treatments or exposures:
+
+  * The _average treatment effect_ (ATE): The effect of a binary exposure or
+    treatment on the observed methylation at a target CpG site is estimated,
+    controlling for the observed methylation at all other CpG sites in the same
+    neighborhood as the target site, based on an additive form. In particular, the
+    parameter estimate represents the __additive difference__ in methylation that
+    would have been observed at the target site had all observations received the
+    treatment versus the scenario in which none received the treatment.
+
+  * The _relative risk_ (RR): The effect of a binary exposure or treatment on the
+    observed methylation at a target CpG site is estimated, controlling for the
+    observed methylation at all other CpG sites in the same neighborhood as the
+    target site, based on a geometric form. In particular, the parameter estimate
+    represents the __multiplicative difference__ in methylation that would have
+    been observed at the target site had all observations received the treatment
+    versus the scenario in which none received the treatment.
+
+  Support for continuous-valued treatments or exposures is _planned but not yet
+  available_, though work is underway to incorporate into our methodology the
+  following
+
+  * A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+    The effect of continous-valued exposure or treatment (the observed methylation
+    at a target CpG site) on an outcome of interest is estimated, controlling for
+    the observed methylation at all other CpG sites in the same neighborhood as
+    the target (treatment) site, based on a parameter that compares values of the
+    treatment against a reference value taken to be the null. In particular, the
+    implementation provided is designed to assess the effect of differential
+    methylation at the target CpG site on a (typically) phenotype-level outcome
+    of interest (e.g., survival), in effect providing an nonparametric evaluation
+    of the impact of methylation at the target site on said outcome.
+  _As previously noted, in all cases, an estimator of the target parameter is
+  constructed via targeted minimum loss-based estimation._
+--------------------------------------------------------------------------------
+
+This method provides
+easily interpretable statistical inference based on asymptotic theories and is
+a fairly fast procedure with a proper design of the pre-screening algorithm and
+and suitable importance measure.
 
@@
 
+--------------------------------------------------------------------------------
+Having now discussed the foundational principles of the estimation procedure
+employed and the statistical algorithm implemented, it is best to proceed by
+examining `methyvim` by example.
+
+  # Preliminaries: Setting up the Data
+
+```{r reqs, echo=FALSE}
+suppressMessages(library(tmle))
+suppressMessages(library(minfi))
+suppressMessages(library(SummarizedExperiment))
+```
+
+First, we'll load the `methyvim` package and the example data contained in the
+`methyvimData` package that accompanies it:
+
+```{r prelims}
+library(methyvim)
+library(methyvimData)
+```
+
+Now, let's load the data set and seed the RNG:
+
+```{r get-data}
+data(grsExample)
+set.seed(479253)
+grsExample
+```
+
+The example data object is of class `GenomicRatioSet`, provided by the `minfi`
+package. The summary provided by the `print` method gives a wealth of
+information on the experiment that generated the data -- since we are working
+with a simulated data set, we need not concern ourselves with much of this
+information. While most slots of objects of this S4 class have descriptive
+names, the interested analyst might consider consulting the documentation of the
+`minfi` package for an in-depth description [@aryee2014minfi].
+---
+  ## Quantifying the Effect of an Exposure on DNA Methylation
+
+            ### The Average Treatment Effect as Variable Importance Measure
+
+...
+
+```{r methyvim-ate}
+suppressMessages(
+  methyvim_out_ate <- methyvim(data_grs = grsExample, sites_comp = 10,
+                               var_int = 1, vim = "ate", type = "Mval",
+                               filter = "limma", filter_cutoff = 0.05,
+                               parallel = FALSE, tmle_type = "sl"
+                              )
+)
+methyvim_out_ate
+```
+As is clear from examining the object `methyvim_out_ate`, the output resembles
+exactly that returned when examining objects of class `GenomicRatioSet` from the
+`minfi` R package. In particular, the returned `methytmle` object is merely a
+modified form (in particular, a subclass) of the input `GenomicRatioSet` object
+-- thus, it contains all of the original slots, with all experimental data
+intact. Several extra pieces of information are contained within the output
+object as well_.
+We can take a look at the results produced from the estimation procedure by
+examining the `"vim"` `slot` of the `methytmle` object:
+
+```{r methyvim-ate-print}
+head(slot(methyvim_out_ate, "vim"))
+```
+
+From the table displayed, we note that we have access to point estimates of the
+ATE ("est_ATE") as well as lower and upper confidence interval bounds for each
+estimate ("lowerCI_ATE" and "upperCI_ATE", respectively). Additional statistical
+information we have access to include the variance ("Var_ATE") of the estimate
+as well as the p-value ("pval") associated with each estimate (based on
+Wald-style testing procedures). Beyond these, key bioinformatical quantities
+(with respect to the algorithm outlined above) are also returned; these include
+the total number of neighbors of the target site, the number of neighboring
+sites controlled for when estimating the effect of exposure on DNA methylation,
+and, finally, the maximum correlation between the target site and any given site
+in its full set of neighbors.
+
+...
+
+          ### The Risk Ratio as Variable Importance Measure
+
+...
+
+```{r methyvim-rr}
+methyvim_out_rr <- methyvim(data_grs = grsExample, sites_comp = 10,
+                            var_int = 1, vim = "rr", type = "Mval",
+                            filter = "limma", filter_cutoff = 0.05,
+                            parallel = FALSE, tmle_type = "sl"
+                           )
+methyvim_out_rr
+```
+...
+
+```{r methyvim-rr-print}
+head(slot(methyvim_out_rr, "vim"))
+```
+...
+
+  ## Visualization of Results
+
+...
+
+```{r methyvim-plot-volcano, eval=FALSE}
+plot(methyvim_out_rr, type = "volcano")
+```
+
+...
+
+```{r methyvim-plot-pvals, eval=FALSE}
+plot(methyvim_out_rr, type = "pvals")
+```
+
+...
+
+```{r methyvim-plot-heatmap, eval=FALSE}
+plot(methyvim_out_rr, type = "heatmap")
+```
+
+...
+--------------------------------------------------------------------------------
+
 # Results <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and
-should be written as a traditional results section.
+implementing the minfi data analysis soon
 
@@
 
-DNA methylation is a fundamental epigenetic process known to play an important
-role in controlling gene expression. It remains perhaps the best studied
-biological mechanism amongst all known epigenetic processes of its kind. With
-relatively recent advances in computational biology and bioinformatics, probing
-DNA methylation has become rather easily possible -- in fact, modern assays
-allow for DNA methylation signatures at up to $850,000$ CpG sites to be measured
-simultaneously. Most statistical techniques available for the analysis of data
-produced by such assays rely on (generalized) linear models. Here, we provide an
-alternative to such approaches. Specifically, we provide a range of _variable
-importance measures_ (__VIM__), parameters that arise in statistical causal
-inference, for which targeted minimum loss-based estimates may be readily
-computed based on the data made available by DNA methylation assays. `methyvim`
-is an R package that provides facilities for performing differential methylation
-analyses within exactly this scope.
-
-For a general discussion of the framework of targeted minimum loss-based
-estimation and the role this approach plays in statistical causal inference, the
-interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
-For a more general introduction to (statistical) causal inference,
-@pearl2009causality and @hernan2018causal may be of interest.
-
----
-
-## Methodology
-
-The core functionality of this package is made available via the eponymous
-`methyvim` function, which implements a statistical algorithm designed to
-compute targeted estimates of VIMs, defined in such a way that the VIMs
-represent parameters of scientific interest in computational biology
-experiments; moreover, these VIMs are defined such that they may be estimated in
-a manner that is very nearly assumption-free, that is, within a fully
-nonparametric statistical model. __The statistical algorithm consists in several
-major steps:__
-
-1. Pre-screening of genomic sites is used to isolate a subset of sites for
-  which there is cursory evidence of differential methylation. For the sake of
-  computational feasibility, targeted minimum loss-based estimates of VIMs are
-  computed only for this subset of sites. Currently, the available screening
-  approaches adapts core routines from the
-  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
-  package, thought future releases will support functionality from other
-  packages (e.g., [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)).
-
-2. Nonparametric estimates of VIMs (for the specified target parameter) are
-  currently computed by adapting routines from the
-  [`tmle`](https://CRAN.R-project.org/package=tmle) R package. Future releases
-  will support doubly-robust estimates of these VIMs (via the
-  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
-  and add parameters for continuous treatments/exposures (via the
-  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).
-
-3. Since pre-screening is performed prior to estimating VIMs, we make use of a
-  multiple testing correction uniquely suited to such settings. Due to the
-  multiple testing nature of the estimation problem, a variant of the Benjamini
-  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
-  [@benjamini1995controlling]. Specifically, we apply the modified marginal
-  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
-  multi-stage analyses (FDR-MSA), which is guaranteed to control the FDR as if
-  all sites were tested [@tuglus2009modified].
-
----
-
-## Parameters of Interest
-
-For discrete-valued treatments or exposures:
-
-* The _average treatment effect_ (ATE): The effect of a binary exposure or
-  treatment on the observed methylation at a target CpG site is estimated,
-  controlling for the observed methylation at all other CpG sites in the same
-  neighborhood as the target site, based on an additive form. In particular, the
-  parameter estimate represents the __additive difference__ in methylation that
-  would have been observed at the target site had all observations received the
-  treatment versus the scenario in which none received the treatment.
-
-* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
-  observed methylation at a target CpG site is estimated, controlling for the
-  observed methylation at all other CpG sites in the same neighborhood as the
-  target site, based on an geometric form. In particular, the parameter estimate
-  represents the __multiplicative difference__ in methylation that would have
-  been observed at the target site had all observations received the treatment
-  versus the scenario in which none received the treatment.
-
-Support for continuous-valued treatments or exposures is _planned but not yet
-available_, though work is underway to incorporate into our methodology the
-following
-
-* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
-  The effect of continous-valued exposure or treatment (the observed methylation
-  at a target CpG site) on an outcome of interest is estimated, controlling for
-  the observed methylation at all other CpG sites in the same neighborhood as
-  the target (treatment) site, based on a parameter that compares values of the
-  treatment against a reference value taken to be the null. In particular, the
-  implementation provided is designed to assess the effect of differential
-  methylation at the target CpG site on a (typically) phenotype-level outcome
-  of interest (e.g., survival), in effect providing an nonparametric evaluation
-  of the impact of methylation at the target site on said outcome.
-
-_As previously noted, in all cases, an estimator of the target parameter is
-constructed via targeted minimum loss-based estimation._
-
-Having now discussed the foundational principles of the estimation procedure
-employed and the statistical algorithm implemented, it is best to proceed by
-examining `methyvim` by example.
-
----
-
-## Preliminaries: Setting up the Data
-
-```{r reqs, echo=FALSE}
-suppressMessages(library(tmle))
-suppressMessages(library(minfi))
-suppressMessages(library(SummarizedExperiment))
-```
-
-First, we'll load the `methyvim` package and the example data contained in the
-`methyvimData` package that accompanies it:
-
-```{r prelims}
-library(methyvim)
-library(methyvimData)
-```
-
-Now, let's load the data set and seed the RNG:
-
-```{r get-data}
-data(grsExample)
-set.seed(479253)
-grsExample
-```
-
-The example data object is of class `GenomicRatioSet`, provided by the `minfi`
-package. The summary provided by the `print` method gives a wealth of
-information on the experiment that generated the data -- since we are working
-with a simulated data set, we need not concern ourselves with much of this
-information. While most slots of objects of this S4 class have descriptive
-names, the interested analyst might consider consulting the documentation of the
-`minfi` package for an in-depth description [@aryee2014minfi].
-
----
-
-## Quantifying the Effect of an Exposure on DNA Methylation
-
-### The Average Treatment Effect as Variable Importance Measure
-
-...
-
-```{r methyvim-ate}
-suppressMessages(
-  methyvim_out_ate <- methyvim(data_grs = grsExample, sites_comp = 10,
-                               var_int = 1, vim = "ate", type = "Mval",
-                               filter = "limma", filter_cutoff = 0.05,
-                               parallel = FALSE, tmle_type = "sl"
-                              )
-)
-methyvim_out_ate
-```
-
-As is clear from examining the object `methyvim_out_ate`, the output resembles
-exactly that returned when examining objects of class `GenomicRatioSet` from the
-`minfi` R package. In particular, the returned `methytmle` object is merely a
-modified form (in particular, a subclass) of the input `GenomicRatioSet` object
--- thus, it contains all of the original slots, with all experimental data
-intact. Several extra pieces of information are contained within the output
-object as well_.
-
-We can take a look at the results produced from the estimation procedure by
-examining the `"vim"` `slot` of the `methytmle` object:
-
-```{r methyvim-ate-print}
-head(slot(methyvim_out_ate, "vim"))
-```
-
-From the table displayed, we note that we have access to point estimates of the
-ATE ("est_ATE") as well as lower and upper confidence interval bounds for each
-estimate ("lowerCI_ATE" and "upperCI_ATE", respectively). Additional statistical
-information we have access to include the variance ("Var_ATE") of the estimate
-as well as the p-value ("pval") associated with each estimate (based on
-Wald-style testing procedures). Beyond these, key bioinformatical quantities
-(with respect to the algorithm outlined above) are also returned; these include
-the total number of neighbors of the target site, the number of neighboring
-sites controlled for when estimating the effect of exposure on DNA methylation,
-and, finally, the maximum correlation between the target site and any given site
-in its full set of neighbors.
-
-...
-
----
-
-### The Risk Ratio as Variable Importance Measure
-
-...
-
-```{r methyvim-rr}
-methyvim_out_rr <- methyvim(data_grs = grsExample, sites_comp = 10,
-                            var_int = 1, vim = "rr", type = "Mval",
-                            filter = "limma", filter_cutoff = 0.05,
-                            parallel = FALSE, tmle_type = "sl"
-                           )
-methyvim_out_rr
-```
-...
-
-```{r methyvim-rr-print}
-head(slot(methyvim_out_rr, "vim"))
-```
-...
-
----
-
-## Visualization of Results
-
-...
-
-```{r methyvim-plot-volcano, eval=FALSE}
-plot(methyvim_out_rr, type = "volcano")
-```
-
-...
-
-```{r methyvim-plot-pvals, eval=FALSE}
-plot(methyvim_out_rr, type = "pvals")
-```
-
-...
-
-```{r methyvim-plot-heatmap, eval=FALSE}
-plot(methyvim_out_rr, type = "heatmap")
-```
-
-...
-
----
-
 ## References

### 2017-10-06 paper/paper_BiocF1000.Rmd
almost done with f1000, minor vignette corrections

--- 2017-07-14 paper_F1000Bioc/paper_BiocF1000.Rmd
@@
 ---
-title: "*F1000Research* Software Tool Article Template"
+title: '`methyvim`: an R-package for Variable Importance Measures for
+Differential Methylation Analysis'
 author:
-  - name: Nima Hejazi
+  - name: Nima S. Hejazi
     affiliation: Address of author1
+  - name: Rachael Phillips
+    affiliation: orchid.org/0000-0002-8474-591X
   - name: Alan E. Hubbard
@@
     affiliation: Address of author2
-abstract: Abstracts should be up to 300 words and provide a succinct summary of the article.
-Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article.
-Citations should not be used in the abstract, and the use of abbreviations should be minimized.
-keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
-#bibliography: paper_BiocF1000.bib
+abstract:
+  We present a general algorithm for the nonparametric estimation of
+  treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
+  leverages variable importance measures, a class of parameters that arise in
+  causal inference and are defined in such a manner that they provide targeted
+  estimates of the relative importance of individual CpG site with respect to
+  a binary exposure/treatment assigned at the level of subjects. Additionally,
+  this procedure is computationally efficient; incorporating a preliminary
+  screening step to isolate a subset of sites for which there is cursory
+  evidence of differential methylation and a multiple testing correction step
+  to control the False Discovery Rate as if all sites were tested. This
+  technique for analysis of differentially methylated positions provides an
+  avenue to move beyond linear models without the loss of interpretable
+  statistical inference.
+
+keywords: DNA methylation, differential methylation, epigenetics, causal
+  inference, variable importance, targeted minimum loss-based estimation
+bibliography: paper_BiocF1000.bib
 output: BiocWorkflowTools::f1000_article
@@
 
-The introduction provides context as to why the software tool was developed and what need it addresses.
-It is good scholarly practice to mention previously developed tools that address similar needs, and why the current tool is needed.
+DNA methylation is a fundamental epigenetic process known to play an important
+role in the regulation of gene expression. DNA CpG methylation involves the
+addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
+ring structure to form 5-methylcytosine. Numerous biological and medical studies
+have implicated DNA CpG methylation as playing a role in disease and
+development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
+have been developed to study the molecular mechanisms of this epigenetic
+process. Modern assays, like the Illumina _Infinium_ Methylation assay,
+allow for quantitative interrogation of DNA methylation of CpG sites scattered
+across the genome at single-nucleotide resolution; moreover, much effort
+has been invested, by the bioinformatics community, in the development of tools  
+for properly removing technological effects that may contaminate biological
+signatures measured by such assays. Despite these advances in both biological
+and bioninformatical techniques, most statistical methods available for
+differential analysis of data produced by such assays rely on over-simplified
+(often generalized linear) models. This is because typical statistical
+estimation approaches are not achievable for such high-dimensional data without restrictive modeling assumptions. When these assumptions are violated, estimates
+of the population-level effect of an exposure or treatment may have large bias.
+Data-adaptive estimation procedures that utilize machine learning can control
+for high-dimensional confounders when estimating treatment/exposure effects.
+However, interpretable statistical inference (i.e. confidence intervals,
+hypothesis tests) resulting from these data-adaptive estimates is challenging [@libbrecht2015machine]. In this paper, we present an alternative such
+statistical analysis approaches in the form of nonparametric estimation
+procedures that provide simple and well interpretable statistical inference.
+Inspired by concepts of counterfactual effects from the causal inference
+literature and machine learning algorithms, we provide a computationally
+efficient technique for obtaining targeted estimates of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
+the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
+and thus provide reliable inference. Counterfactuals are "counter"-fact, defined
+in terms of an unrealistic yet ideal experiment. In the context of DNA
+methylation and for a binary exposure/treatment, we define the counterfactual
+outcomes as the Beta-value/M-value a CpG site would have if all subjects were
+exposed/treated and the Beta-value/M-value a CpG site would have if all
+subjects were not exposed/treated. Of course, these counterfactual outcomes are
+impossible to observe but they are very helpful for us to asses the importance
+of a CpG site relative to a subjects treatment/exposure and they can be
+statistically estimated [@van2006statistical]. Our VIM definition uses the
+concepts of counterfactuals and the framework for estimation is built on
+targeted maximum likelihood estimation or targeted minimum loss-based estimation
+(TMLE) [@van2006targeted]. This methodology targets the individual importance
+of each CpG site by taking advantage of the prediction power of machine learning
+algorithms. The mathematical properties of this approach have been explored in [@tuglus2008targeted] and we focus on our contribution to it's application,
+`methyvim`, an R package that accommodates this technique for differential
+methylation analysis.
+
+For a general discussion of the framework of targeted minimum loss-based
+estimation and the role this approach plays in statistical causal inference, the
+interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
+For a more general introduction to (statistical) causal inference,
+@pearl2009causality and @hernan2018causal may be of interest.
 
 # Methods
+
 ## Implementation
-For software tool papers, this section should address how the tool works and any relevant technical details required for implementation of the tool by other developers.
+
+The core functionality of this package is made available via the eponymous
+`methyvim` function, which implements a statistical algorithm designed to
+compute targeted estimates of VIMs, defined in such a way that the VIMs
+represent parameters of scientific interest in computational biology
+experiments; moreover, these VIMs are defined such that they may be estimated in
+a manner that is very nearly assumption-free, that is, within a _fully
+nonparametric statistical model_. __The statistical algorithm consists in
+several major steps:__
+
+1. Pre-screening of genomic sites is used to isolate a subset of sites for
+  which there is cursory evidence of differential methylation. For the sake of
+  computational feasibility, targeted minimum loss-based estimates of VIMs are
+  computed only for this subset of sites. Currently, the available screening
+  approach adapts core routines from the
+  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
+  package. Future releases will support functionality from other
+  packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
+  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
+  style of the function for performing screening via `limma`, users may write
+  their own screening functions and are invited to contribute such functions to
+  the core software package by opening pull requests at the GitHub repository.
+
+2. Nonparametric estimates of VIMs, for the specified parameter of interest, are
+  computed at each of the CpG sites passing the screening step. The VIMs are
+  defined in such a way that the estimated effects is of an exposure/treatment
+  on the methylation status of a target CpG site, controlling for the observed
+  methylation status of the neighbors of that site. Currently, routines are
+  adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
+  Future releases will support doubly-robust estimates of these VIMs (via the
+  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
+  and add parameters for continuous treatments/exposures (via the
+  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).
+
+3. Since pre-screening is performed prior to estimating VIMs, we make use of a
+  multiple testing correction uniquely suited to such settings. Due to the
+  multiple testing nature of the estimation problem, a variant of the Benjamini
+  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
+  [@benjamini1995controlling]. Specifically, we apply the "modified marginal
+  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
+  multi-stage analyses" (FDR-MSA), which is guaranteed to control the FDR as if
+  all sites were tested (i.e., without screening) [@tuglus2009modified].
+
+#### Parameters of Interest
+
+For the CpG sites that passed the pre-screening step, the user chooses the
+parameter of interest that will be computed at each site. In all cases, an
+estimator of the parameter of interest is constructed via targeted minimum
+loss-based estimation.
+
+For discrete-valued treatments or exposures:
+
+* The _average treatment effect_ (ATE): The effect of a binary exposure or
+  treatment on the observed methylation at a target CpG site is estimated,
+  controlling for the observed methylation at all other CpG sites in the same
+  neighborhood as the target site, based on an additive form. Often denoted
+  $\psi_0 = \psi_0(1) - \psi_0(0)$, the parameter estimate represents the
+  __additive difference__ in methylation that would have been observed at the
+  target site had all observations received the treatment versus the
+  counterfactual under which none received the treatment.
+
+* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
+  observed methylation at a target CpG site is estimated, controlling for the
+  observed methylation at all other CpG sites in the same neighborhood as the
+  target site, based on a geometric form. Often denoted,
+  $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, the parameter estimate represents the
+  __multiplicative difference__ in methylation that would have
+  been observed at the target site had all observations received the treatment
+  versus the counterfactual under which none received the treatment.
+
+Support for continuous-valued treatments or exposures is _planned but not yet
+available_, though work is underway to incorporate into our methodology the
+following
+
+* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+  The effect of continuous-valued exposure or treatment (the observed
+  methylation at a target CpG site) on an outcome of interest is estimated,
+  controlling for the observed methylation at all other CpG sites in the same neighborhood as the target (treatment) site, based on a parameter that
+  compares values of the treatment against a reference value taken to be the
+  null. In particular, the implementation provided is designed to assess the
+  effect of differential methylation at the target CpG site on a (typically) phenotype-level outcome of interest (e.g., survival), in effect providing an nonparametric evaluation of the impact of methylation at the target site on
+  said outcome.
+
+We have adopted a class `methytmle` to help organize the functionality within
+this package. The `methytmle` class builds upon the `GenomicRatioSet` class
+provided by the `minfi` package so all of the slots of `GenomicRatioSet` are
+contained in a `methytmle` object. The new class introduced in the `methyvim`
+package includes several new slots:
+
+* `call` - the form of the original call to the `methyvim` function.
+* `screen_ind` - indices identifying CpG sites that pass the screening process.
+* `clusters` - non-unique IDs corresponding to the manner in wich sites are
+  treated as neighbors. These are assigned by genomic distance (bp) and respect
+  chromosome boundaries (produced via a call to `bumphunter::clusterMaker`).
+* `var_int` - the treatment/exposure status for each subject. Currently, these
+  must be binary, due to the definition of the supported targeted parameters.
+* `param` - the name of the target parameter from which the estimated VIMs are
+  defined.
+* `vim` - a table of statistical results obtained from estimating VIMs for each
+  of the CpG sites that pass the screening procedure.
+* `ic` - the measured array values for each of the CpG sites passing the
+  screening, transformed into influence curve space based on the chosen target
+  parameter.
+
+The interested analyst might consider consulting the documentation of the
+`minfi` package for an in-depth description of all of the other slots that
+appear in this class [@aryee2014minfi].
+
+We refer the reader to the package vignette, "`methyvim`: Variable Importance
+Measures for Differential Methylation" hosted with the package on Bioconductor,
+for a detailed description of the methods used in this package.
+(NEED CITATION FOR VIGNETTE)
 
 ## Operation
-This part of the methods should include the minimal system requirements needed to run the software and an overview of the workflow for the tool for users of the tool.
 
-# Results <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and should be written as a traditional results section.
+A standard computer with the latest version of R and Bioconductor 3.6 installed
+will handle applications of the `methyvim` package.
 
-# Use Cases <!-- Optional - only if NO new datasets are included -->
-This section is required if the paper does not include novel data or analyses. 
-Examples of input and output files should be provided with some explanatory context.
-Any novel or complex variable parameters should also be explained in sufficient detail to allow users to understand and use the tool's functionality.
+# Use Cases
 
-# Discussion <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and should be written in the same style as a traditional discussion section.
-Please include a brief discussion of allowances made (if any) for controlling bias or unwanted sources of variability, and the limitations of any novel datasets.
+To examine the practical applications and the full set of utilities of the
+`methyvim` package, we will use a publicly available example data set produced
+by the Illumina 450K array, from the `minfiData` R package.
 
-# Conclusions <!-- Optional - only if novel data or analyses are included -->
-This section is only required if the paper includes novel data or analyses, and should be written as a traditional conclusion.
+### Preliminaries: Setting up the Data
+
+We begin by loading the package and the data set. After loading the data, which
+comes in the form of a raw `MethylSet` object, we perform some further
+processing by mapping to the genome (with `mapToGenome`) and converting the
+values from the methylated and unmethylated channels to Beta-values
+(via `ratioConvert`). These two steps together produce an object of class `GenomicRatioSet`, provided by the `minfi` package.
+
+```{r setup-minfidata}
+suppressMessages(library(minfiData))
+data(MsetEx)
+mset <- mapToGenome(MsetEx)
+grs <- ratioConvert(mset)
+grs
+```
+We can create an object of class `methytmle` from any `GenomicRatioSet` object
+simply invoking the S4 class constructor:
+
+```{r make-methytmle}
+mtmle <- .methytmle(grs)
+```
+Additionally, a `GenomicRatioSet` can be created from a matrix with the function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.
+
+### Differential Methylation Analysis
+
+For this example analysis, we'll treat the condition of the patients as the
+exposure/treatment variable of interest. The `methyvim` function requires that
+this variable either be `numeric` or easily coercible to `numeric`. To
+facilitate this, we'll simply convert the covariate (currently a `character`):
+
+```{r, minfidata-maketx}
+var_int <- (as.numeric(as.factor(colData(grs)$status)) - 1)
+```
+__n.b.__, the re-coding process results in "normal" patients being assigned a
+value of 1 and cancer patients a 0.
+
+Now, we are ready to analyze the effects of cancer status on DNA methylation
+using this data set. We proceed as follows with a targeted minimum loss-based
+estimate of the Average Treatment Effect.
+
+```{r, minfidata-methyvim}
+suppressMessages(
+  methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
+                                  vim = "ate", type = "Beta", filter = "limma",
+                                  filter_cutoff = 0.20, obs_per_covar = 2,
+                                  parallel = FALSE, sites_comp = 125,
+                                  tmle_type = "glm"
+                                 )
+)
+```
+Note that we set the `obs_per_covar` argument to a relatively low value (2,
+where the recommended default is 20) for the purposes of this example as the
+sample size is only 10. We do this only to exemplify the estimation procedure
+and it is important to point out that such low values for `obs_per_covar` will
+compromise the quality of inference obtained because this setting directly
+affects the definition of the target parameter.
+
+Further, note that here we apply the `glm` flavor of the `tmle_type` argument,
+which produces faster results by fitting models for the propensity score and
+outcome regressions using a limited number of parametric models. By contrast,
+the `sl` (for "Super Learning") flavor fits these two regressions using highly
+nonparametric and data-adaptive procedures (i.e., via machine learning).
+Obtaining the estimates via GLMs results in each of the regression steps
+being less robust than if nonparametric regressions were used.
+
+We can view a table of results by examining the `vim` slot of the produced
+`methytmle` object.
+
+```{r vim-cancer-ate}
+head(slot(methyvim_cancer_ate, "vim"))
+```
+
+Finally, we may compute FDR-corrected p-values, by applying a modified procedure
+for controlling the False Discovery Rate for multi-stage analyses (FDR-MSA)
+[@tuglus2009modified]. We do this by simply applying the `fdr_msa` function.
+
+```{r fdr-msa}
+fdr_p <- fdr_msa(pvals = slot(methyvim_cancer_ate, "vim")$pval,
+                 total_obs = nrow(methyvim_cancer_ate))
+```
+
+Having explored the results of our analysis numerically, we now proceed to use
+the visualization tools provided with the `methyvim` R package to further
+enhance our understanding of the results.
+
+### Visualization of Results
+
+While making allowance for users to explore the full set of results produced by
+the estimation procedure (by way of exposing these directly to the user), the
+`methyvim` package also provides _three_ (3) visualization utilities that
+produce plots commonly used in examining the results of differential methylation
+analyses.
+
+A simple call to `plot` produces side-by-side histograms of the raw p-values
+computed as part of the estimation process and the corrected p-values obtained
+from using the FDR-MSA procedure.
+
+```{r methyvim-pvals-both}
+plot(methyvim_cancer_ate)
+```
+__Remark:__ The plots displayed above may also be generated separately by
+explicitly setting the argument "type" to `plot.methytmle`. For a plot of the
+raw p-values, specify `type = "raw_pvals"`, and for a plot of the FDR-corrected
+p-values, specify `type = "fdr_pvals"`.
+
+While histograms of the p-values may be generally useful in inspecting the
+results of the estimation procedure, a more common plot used in examining the
+results of differential methylation procedures is the volcano plot, which plots
+the parameter estimate along the x-axis and $-\text{log}_{10}(\text{p-value})$
+along the y-axis. We implement such a plot in the `methyvolc` function:
+
+```{r methyvim-volcano}
+methyvolc(methyvim_cancer_ate)
+```
+
+The purpose of such a plot is to ensure that very low (possibly statistically
+significant) p-values do not arise from cases of low variance. This appears to
+be the case in the plot above (notice that most parameter estimates are _near
+zero_, even in cases where the raw p-values are quite low).
+
+Yet another popular plot for visualizing effects in such settings is the
+heatmap, which plots estimates of the raw methylation effects (as measured by
+the assay) across subjects using a heat gradient. We implement this in the
+`methyheat` function:
+
+```{r methyvim-heatmap}
+methyheat(methyvim_cancer_ate)
+```
+
+Invoking `methyheat` in this manner produces a plot of the top sites ($25$, by
+default) based on the raw p-value, using the raw methylation measures in the
+plot. This uses the exceptional `superheat` R package [@barter2017superheat].
 
 # Summary <!-- Optional - only if NO new datasets are included -->
-This section is required if the paper does not include novel data or analyses.  It allows authors to briefly summarize the key points from the article.
 
-# Data availability <!-- Optional - only if novel data or analyses are included -->
-Please add details of where any datasets that are mentioned in the paper, and that have not have not previously been formally published, can be found.
-If previously published datasets are mentioned, these should be cited in the references, as per usual scholarly conventions.
+This section is required if the paper does not include novel data or analyses.
+It allows authors to briefly summarize the key points from the article.
 
 # Software availability
-This section will be generated by the Editorial Office before publication. Authors are asked to provide some initial information to assist the Editorial Office, as detailed below.
+This section will be generated by the Editorial Office before publication.
+Authors are asked to provide some initial information to assist the Editorial
+Office, as detailed below.
 
@@
 
+Software available from:
+
+Latest source code: https://github.com/nhejazi/methyvim
+
+Archived source code as at time of publication:
+
+Software license:
+
 # Author contributions
-In order to give appropriate credit to each author of an article, the individual
-contributions of each author to the manuscript should be detailed in this section. We
-recommend using author initials and then stating briefly how they contributed.
+
+NH designed and implemented the tool, applied the tool to the use case, and
+drafted the vignette. RP helped design the tool and drafted the article.
 
 # Competing interests
-All financial, personal, or professional competing interests for any of the authors that
-could be construed to unduly influence the content of the article must be disclosed and will be displayed alongside the article.
-If there are no relevant competing interests to declare, please add the following: 'No competing interests were disclosed'.
+All financial, personal, or professional competing interests for any of the
+authors that could be construed to unduly influence the content of the article
+must be disclosed and will be displayed alongside the article. If there are no
+relevant competing interests to declare, please add the following: 'No competing
+interests were disclosed'.
 
 # Grant information
-Please state who funded the work discussed in this article, whether it is your employer, a grant funder etc. Please do not list funding that you have that is not relevant to this
-specific piece of research. For each funder, please state the funder’s name, the grant
-number where applicable, and the individual to whom the grant was assigned.
-If your work was not funded by any grants, please include the line: 'The author(s)
-declared that no grants were involved in supporting this work.'
+Please state who funded the work discussed in this article, whether it is your
+employer, a grant funder etc. Please do not list funding that you have that is
+not relevant to this specific piece of research. For each funder, please state
+the funder’s name, the grant number where applicable, and the individual to whom
+the grant was assigned. If your work was not funded by any grants, please
+include the line: 'The author(s) declared that no grants were involved in
+supporting this work.'
 
@@
 
-# USING R MARKDOWN
-Some examples of commonly used markdown syntax are listed below, to help you get started.
-
-## Cross-references
-For portability between different output formats, use the syntax introduced by *bookdown*, such as `(\#label)` for labels and `\@ref(label)` for cross-references.
-The following sections provide examples of referencing tables, figures, and equations.
-
-## Citations
-
-You can include refences in a standard Bibtex file.
-The name of this file is given in the header of the markdown document (in our case it is *sample.bib*).
-References to entries in the Bibtex file are made using square brackets and use an @ plus the key for the entry you are referencing [@Smith:2012qr].
-You can combine multiple entries by separating them with a semi-colon [@Smith:2012qr; @Smith:2013jd].
-
-## Code chunks
-You can embed an R code chunk like this:
-```{r firstCodeChunk}
-x <- 1:10
-x
-```
-
-If you specify a figure caption to a code chunk using the chunk option `fig.cap`, the plot will be automatically labeled and numbered.
-The figure label is generated from the label of the code chunk by prefixing it with `fig:`, e.g., see Figure \@ref(fig:plot).
-
-```{r plot, fig.width=4, fig.height=4, fig.cap='A caption to our sample plot.'}
-plot(x)
-```
-
-## Tables
-
-Markdown syntax tends to lack some of the more sophisticated formatting features available in LaTeX, so you may need to edit the tables later to get the desired format.
-
-| First name  | Last Name | Grade |
-| ----------- | --------- | ----- |
-| John        | Doe       |   7.5 |
-| Richard     | Miles     |     2 |
-
-Table: Caption to table.
-
-Just like figures, tables with captions will also be numbered and can be referenced.
-Captions are entered as a paragraph beginning with the string "Table:" (or just ":"), which may appear either before or after the table.
-A label for the table should appear in the beginning of the caption in the form of `(\#tab:label)`, e.g., see Table \@ref(tab:table).
-
-: (\#tab:table) A table with text justification.
-
-| First name  | Last Name | Grade |
-| ----------- | :-------: | ----: |
-| John        | Doe       |   7.5 |
-| Richard     | Miles     |     2 |
-
-## Figures
-
-You can include static figures (i.e. no generated by code) using the `include_graphics()` function from the **knitr** package, in a standard code chunk.
-
-```{r frog-picture, out.width='50%', fig.align='center', fig.cap = 'Your figure legend goes here; it should be succinct, while still explaining all symbols and abbreviations.'}
-knitr::include_graphics('frog.jpg')
-```
-
-You can again use the `fig.cap` option to provide the figure caption, and reference the image based on the code chunk label.
-You can also use options such as `fig.align` and `fig.width` to adjust the position and size of the image within the final document, e.g.
-Figure \@ref(fig:frog-picture) is a frog.
-
-Alternatively, you can use the standard markdown syntax like so:
-
-![This is a smaller version of the same picture, inserted using the standard markdown syntax](frog.jpg){width=25%}
-
-Please give figures appropriate filenames, e.g.: figure1.pdf, figure2.png.
-
-Figure legends should briefly describe the key messages of the figure such that the figure can stand alone from the main text.
-However, all figures should also be discussed in the article text.
-Each legend should have a concise title of no more than 15 words.
-The legend itself should be succinct, while still explaining all symbols and abbreviations.
-Avoid lengthy descriptions of methods.
-
-For any figures reproduced from another publication (as long as appropriate permission has been obtained from the copyright holder —see under the heading 'Submission'), please include a line in the legend to state that: 'This figure has been reproduced with kind permission from [include original publication citation]'.
-
-## Mathematics
-
-You can use LaTeX syntax to typeset mathematical expressions.
-Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-$$S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i$$
-denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
-
-To number and refer to equations, put them in the equation environments and assign labels to them, as for Equation \@ref(eq:binom).
-
-\begin{equation}
-  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
-  (\#eq:binom)
-\end{equation}
-
-## Lists
-
-You can make ordered lists
-
-1. Like this,
-2. and like this.
-
-or bullet points
-
-* Like this,
-* and like this.
-    + Use indentation
-        - for sub-items
+# References

### 2017-10-07 paper/paper_BiocF1000.Rmd
adding the Summary and final touches

@@
 ---
-title: '`methyvim`: an R-package for Variable Importance Measures for
+title: '`methyvim`: an R package for Variable Importance Measures for
 Differential Methylation Analysis'
@@
 have been developed to study the molecular mechanisms of this epigenetic
-process. Modern assays, like the Illumina _Infinium_ Methylation assay,
+process. Modern assays, like the Illumina Infinium Methylation assay,
 allow for quantitative interrogation of DNA methylation of CpG sites scattered
@@
 literature and machine learning algorithms, we provide a computationally
-efficient technique for obtaining targeted estimates of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
+efficient technique for obtaining targeted estimates of nonparametric variable
+importance measures (VIM) for a pre-screened set of CpGs, controlling for
 the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
@@
 algorithms. The mathematical properties of this approach have been explored in [@tuglus2008targeted] and we focus on our contribution to it's application,
-`methyvim`, an R package that accommodates this technique for differential
+methyvim, an R package that accommodates this technique for differential
 methylation analysis.
@@
 The core functionality of this package is made available via the eponymous
-`methyvim` function, which implements a statistical algorithm designed to
+methyvim function, which implements a statistical algorithm designed to
 compute targeted estimates of VIMs, defined in such a way that the VIMs
@@
 experiments; moreover, these VIMs are defined such that they may be estimated in
-a manner that is very nearly assumption-free, that is, within a _fully
-nonparametric statistical model_. __The statistical algorithm consists in
-several major steps:__
+a manner that is very nearly assumption-free, that is, within a fully
+nonparametric statistical model. The statistical algorithm consists in
+several major steps:
 
@@
   approach adapts core routines from the
-  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
+  [limma](http://bioconductor.org/packages/release/bioc/html/limma.html) R
   package. Future releases will support functionality from other
-  packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
-  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
-  style of the function for performing screening via `limma`, users may write
+  packages (e.g., [randomForest](https://cran.r-project.org/web/packages/randomForest/index.html),
+  [tmle.npvi](https://CRAN.R-project.org/package=tmle.npvi)). Following the
+  style of the function for performing screening via limma, users may write
   their own screening functions and are invited to contribute such functions to
@@
   methylation status of the neighbors of that site. Currently, routines are
-  adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
+  adapted from the [tmle](https://CRAN.R-project.org/package=tmle) R package.
   Future releases will support doubly-robust estimates of these VIMs (via the
-  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
+  [drtmle](https://cran.r-project.org/web/packages/drtmle/index.html) package)
   and add parameters for continuous treatments/exposures (via the
-  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).
+  [tmle.npvi](https://CRAN.R-project.org/package=tmle.npvi) package).
 
@@
 
-* The _average treatment effect_ (ATE): The effect of a binary exposure or
+* The average treatment effect (ATE): The effect of a binary exposure or
   treatment on the observed methylation at a target CpG site is estimated,
@@
   $\psi_0 = \psi_0(1) - \psi_0(0)$, the parameter estimate represents the
-  __additive difference__ in methylation that would have been observed at the
+  additive difference in methylation that would have been observed at the
   target site had all observations received the treatment versus the
@@
 
-* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
+* The relative risk (RR): The effect of a binary exposure or treatment on the
   observed methylation at a target CpG site is estimated, controlling for the
@@
   $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, the parameter estimate represents the
-  __multiplicative difference__ in methylation that would have
+  multiplicative difference in methylation that would have
   been observed at the target site had all observations received the treatment
@@
 
-Support for continuous-valued treatments or exposures is _planned but not yet
-available_, though work is underway to incorporate into our methodology the
+Estimating the VIM corresponding to the parameters above, for discrete-valued
+treatments or exposures, requires two separate regression steps: one for the
+treatment mechanism (propensity score) and one for the outcome regression.
+Technical details on the nature of these regressions are discussed in
+@hernan2018causal, and details for estimating these regressions in the
+framework of targeted minimum loss-based estimation are discussed in
+@vdl2011targeted.
+
+Support for continuous-valued treatments or exposures is planned but not yet
+available, though work is underway to incorporate into our methodology the
 following
 
-* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]:
   The effect of continuous-valued exposure or treatment (the observed
@@
 
-We have adopted a class `methytmle` to help organize the functionality within
-this package. The `methytmle` class builds upon the `GenomicRatioSet` class
-provided by the `minfi` package so all of the slots of `GenomicRatioSet` are
-contained in a `methytmle` object. The new class introduced in the `methyvim`
+#### Class methytmle
+
+We have adopted a class methytmle to help organize the functionality within
+this package. The methytmle class builds upon the GenomicRatioSet class
+provided by the minfi package so all of the slots of GenomicRatioSet are
+contained in a methytmle object. The new class introduced in the methyvim
 package includes several new slots:
 
-* `call` - the form of the original call to the `methyvim` function.
-* `screen_ind` - indices identifying CpG sites that pass the screening process.
-* `clusters` - non-unique IDs corresponding to the manner in wich sites are
+* __call__ - the form of the original call to the methyvim function.
+* __screen_ind__ - indices identifying CpG sites that pass the screening process.
+* __clusters__ - non-unique IDs corresponding to the manner in wich sites are
   treated as neighbors. These are assigned by genomic distance (bp) and respect
-  chromosome boundaries (produced via a call to `bumphunter::clusterMaker`).
-* `var_int` - the treatment/exposure status for each subject. Currently, these
+  chromosome boundaries (produced via a call to bumphunter::clusterMaker).
+* __var_int__ - the treatment/exposure status for each subject. Currently, these
   must be binary, due to the definition of the supported targeted parameters.
-* `param` - the name of the target parameter from which the estimated VIMs are
+* __param__ - the name of the target parameter from which the estimated VIMs are
   defined.
-* `vim` - a table of statistical results obtained from estimating VIMs for each
-  of the CpG sites that pass the screening procedure.
-* `ic` - the measured array values for each of the CpG sites passing the
+* __vim__ - a table of statistical results obtained from estimating VIMs for
+  each of the CpG sites that pass the screening procedure.
+* __ic__ - the measured array values for each of the CpG sites passing the
   screening, transformed into influence curve space based on the chosen target
@@
 
-The interested analyst might consider consulting the documentation of the
-`minfi` package for an in-depth description of all of the other slots that
-appear in this class [@aryee2014minfi].
-
-We refer the reader to the package vignette, "`methyvim`: Variable Importance
+We refer the reader to the package vignette, "methyvim: Variable Importance
 Measures for Differential Methylation" hosted with the package on Bioconductor,
@@
 A standard computer with the latest version of R and Bioconductor 3.6 installed
-will handle applications of the `methyvim` package.
+will handle applications of the methyvim package.
 
@@
 
-### Preliminaries: Setting up the Data
+#### Preliminaries: Setting up the Data
 
@@
 
-### Differential Methylation Analysis
+#### Differential Methylation Analysis
 
@@
 
-### Visualization of Results
+#### Visualization of Results
 
@@
 significant) p-values do not arise from cases of low variance. This appears to
-be the case in the plot above (notice that most parameter estimates are _near
-zero_, even in cases where the raw p-values are quite low).
+be the case in the plot above (notice that most parameter estimates are near
+zero, even in cases where the raw p-values are quite low).
 
@@
 
-# Summary <!-- Optional - only if NO new datasets are included -->
+# Summary
 
-This section is required if the paper does not include novel data or analyses.
-It allows authors to briefly summarize the key points from the article.
+Here we introduce the R package methyvim, an algorithm for differential
+methylation analysis that moves beyond linear models. Additionally, this
+estimation procedure produces straightforward statistical inference. Inspired by
+concepts of counterfactual effects from the causal inference literature and
+machine learning algorithms, we provide a computationally efficient technique
+for obtaining targeted estimates of nonparametric variable importance measures
+for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
+all sites were tested, and the estimation framework is built on targeted minimum loss-based estimation. By taking advantage of the prediction power of machine
+learning algorithms, this methodology targets the individual importance of each
+CpG site, providing a fully nonparametric statistical model for the analysis of
+differential methylation.
 
 # Software availability
-This section will be generated by the Editorial Office before publication.
-Authors are asked to provide some initial information to assist the Editorial
-Office, as detailed below.
-
-1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
-2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
-3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
-4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
-5. Software license (AUTHOR TO PROVIDE; required)
 
@@
 drafted the vignette. RP helped design the tool and drafted the article.
+AH and ML serve as the advisors for the development of this tool.
 
 # Competing interests
-All financial, personal, or professional competing interests for any of the
-authors that could be construed to unduly influence the content of the article
-must be disclosed and will be displayed alongside the article. If there are no
-relevant competing interests to declare, please add the following: 'No competing
-interests were disclosed'.
+
+No competing interests were disclosed.
 
 # Grant information
-Please state who funded the work discussed in this article, whether it is your
-employer, a grant funder etc. Please do not list funding that you have that is
-not relevant to this specific piece of research. For each funder, please state
-the funder’s name, the grant number where applicable, and the individual to whom
-the grant was assigned. If your work was not funded by any grants, please
-include the line: 'The author(s) declared that no grants were involved in
-supporting this work.'
 
-# Acknowledgments
-This section should acknowledge anyone who contributed to the research or the
-article but who does not qualify as an author based on the criteria provided
-earlier (e.g. someone or an organization that provided writing assistance).
-Please state how they contributed; authors should obtain permission to
-acknowledge from all those mentioned in the Acknowledgments section.
-
-Please do not list grant funding in this section.
+The author(s) declared that no grants were involved in supporting this work.
 

### 2017-10-07 paper/paper_BiocF1000.Rmd
a few more minor things

--- 2017-10-06 paper/paper_BiocF1000.Rmd
@@
     affiliation: Address of author1
-  - name: Rachael Phillips
+  - name: Rachael V. Phillips
     affiliation: orchid.org/0000-0002-8474-591X
@@
     affiliation: Address of author2
-abstract:
-  We present a general algorithm for the nonparametric estimation of
+abstract: |
+  We present `methyvim`, a general algorithm for the nonparametric estimation of
   treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
@@
   statistical inference.
-
 keywords: DNA methylation, differential methylation, epigenetics, causal
@@
 literature and machine learning algorithms, we provide a computationally
-efficient technique for obtaining targeted estimates of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
+efficient technique for obtaining targeted estimates of nonparametric _variable
+importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
 the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
@@
 in terms of an unrealistic yet ideal experiment. In the context of DNA
-methylation and for a binary exposure/treatment, we define the counterfactual
+methylation and for a binary treatment, we define the counterfactual
 outcomes as the Beta-value/M-value a CpG site would have if all subjects were
-exposed/treated and the Beta-value/M-value a CpG site would have if all
-subjects were not exposed/treated. Of course, these counterfactual outcomes are
-impossible to observe but they are very helpful for us to asses the importance
-of a CpG site relative to a subjects treatment/exposure and they can be
-statistically estimated [@van2006statistical]. Our VIM definition uses the
-concepts of counterfactuals and the framework for estimation is built on
-targeted maximum likelihood estimation or targeted minimum loss-based estimation
-(TMLE) [@van2006targeted]. This methodology targets the individual importance
-of each CpG site by taking advantage of the prediction power of machine learning
-algorithms. The mathematical properties of this approach have been explored in [@tuglus2008targeted] and we focus on our contribution to it's application,
+treated and the Beta-value/M-value a CpG site would have if all subjects were
+not treated. Of course, these counterfactual outcomes are impossible to observe
+but they are very helpful for us to asses the importance of a CpG site relative
+to a subjects treatment/exposure and they can be statistically estimated [@van2006statistical]. Our VIM definition uses the concepts of counterfactuals
+and the framework for estimation is built on _targeted maximum likelihood
+estimation or targeted minimum loss-based estimation_ (__TMLE__)
+[@van2006targeted]. This methodology targets the individual importance of each
+CpG site by taking advantage of the prediction power of machine learning
+algorithms. The mathematical properties of this approach have been explored in
+[@tuglus2008targeted] and we focus on our contribution to it's application,
 `methyvim`, an R package that accommodates this technique for differential
@@
 nonparametric statistical model_. __The statistical algorithm consists in
-several major steps:__
+several major steps__:
 
@@
   [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
-  package. Future releases will support functionality from other
-  packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
+  package. Future releases will support functionality from other packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
@@
 
-2. Nonparametric estimates of VIMs, for the specified parameter of interest, are
+2. Nonparametric estimates of VIMs, for the specified target parameter, are
   computed at each of the CpG sites passing the screening step. The VIMs are
@@
 For the CpG sites that passed the pre-screening step, the user chooses the
-parameter of interest that will be computed at each site. In all cases, an
+parameter of interest that will be computed at each site. _In all cases, an
 estimator of the parameter of interest is constructed via targeted minimum
-loss-based estimation.
+loss-based estimation_.
 
@@
   $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, the parameter estimate represents the
-  __multiplicative difference__ in methylation that would have
+  multiplicative difference in methylation that would have
   been observed at the target site had all observations received the treatment
   versus the counterfactual under which none received the treatment.
+
+Estimating the VIM corresponding to the parameters above, for discrete-valued
+treatments or exposures, requires two separate regression steps: one for the
+treatment mechanism (propensity score) and one for the outcome regression.
+Technical details on the nature of these regressions are discussed in
+@hernan2018causal, and details for estimating these regressions in the
+framework of targeted minimum loss-based estimation are discussed in
+@vdl2011targeted.
 
@@
 
-* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]:
   The effect of continuous-valued exposure or treatment (the observed
@@
   said outcome.
+
+#### Class methytmle
 
@@
   defined.
-* `vim` - a table of statistical results obtained from estimating VIMs for each
-  of the CpG sites that pass the screening procedure.
+* `vim` - a table of statistical results obtained from estimating VIMs for
+  each of the CpG sites that pass the screening procedure.
 * `ic` - the measured array values for each of the CpG sites passing the
@@
   parameter.
-
-The interested analyst might consider consulting the documentation of the
-`minfi` package for an in-depth description of all of the other slots that
-appear in this class [@aryee2014minfi].
 
@@
 
-# Summary <!-- Optional - only if NO new datasets are included -->
+# Summary
 
-This section is required if the paper does not include novel data or analyses.
-It allows authors to briefly summarize the key points from the article.
+Here we introduce the R package `methyvim`, an algorithm for differential
+methylation analysis that moves beyond linear models. Additionally, this
+estimation procedure produces straightforward statistical inference. Inspired by
+concepts of counterfactual effects from the causal inference literature and
+machine learning algorithms, we provide a computationally efficient technique
+for obtaining targeted estimates of nonparametric variable importance measures
+for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
+all sites were tested, and the estimation framework is built on targeted minimum loss-based estimation. By taking advantage of the prediction power of machine
+learning algorithms, this methodology targets the individual importance of each
+CpG site, providing a fully nonparametric statistical model for the analysis of
+differential methylation.
 
 # Software availability
-This section will be generated by the Editorial Office before publication.
-Authors are asked to provide some initial information to assist the Editorial
-Office, as detailed below.
-
-1. URL link to where the software can be downloaded from or used by a non-coder (AUTHOR TO PROVIDE; optional)
-2. URL link to the author's version control system repository containing the source code (AUTHOR TO PROVIDE; required)
-3. Link to source code as at time of publication (*F1000Research* TO GENERATE)
-4. Link to archived source code as at time of publication (*F1000Research* TO GENERATE)
-5. Software license (AUTHOR TO PROVIDE; required)
 

### 2017-10-11 README.Rmd
add initial pkgdown materials
WIP => Active; misc other changes

@@
 [![Coverage Status](https://img.shields.io/codecov/c/github/nhejazi/methyvim/master.svg)](https://codecov.io/github/nhejazi/methyvim?branch=master)
-[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
+[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
 [![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
@@
 Note about shrinkage of influence curves, adapting @smyth2004linear.
-Specifically, this is meant to reference my working paper with Alan.
-
 More exposition here....

### 2017-10-11 R/zzz.R

@@ -1,7 +1,7 @@
 .onAttach <- function(...) {
   packageStartupMessage(paste("methyvim: Nonparametric Differential",
-                              "Methylation Analysis \n with Targeted Estimates",
-                              "of Variable Importance Measures"))
+                              "Methylation Analysis \n using Targeted",
+                              "Estimates of Variable Importance Measures"))
   packageStartupMessage("Version: ",
                         utils::packageDescription("methyvim")$Version)
 }

### 2017-10-11 README.Rmd
add zenodo; update site

@@
 [![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
+[![DOI](https://zenodo.org/badge/79256902.svg)](https://zenodo.org/badge/latestdoi/79256902)
 

### 2017-10-17 README.Rmd

@@
             J},
-          title = {methyvim: Nonparametric differential methylation analysis
-            with targeted estimates of variable importance measures},
+          title = {methyvim: Targeted and model-free analysis of differential
+            methylation in R},
           journal = {}

### 2017-10-18 paper/paper_BiocF1000.Rmd
minor formatting
f1000 paper edits, minor cleanup

@@
   - name: Nima S. Hejazi
-    affiliation: Address of author1
+    affiliation: 1
+    orcid: 0000-0002-7127-2789
   - name: Rachael V. Phillips
-    affiliation: orchid.org/0000-0002-8474-591X
+    affiliation: 1
+    orcid: 0000-0002-8474-591X
   - name: Alan E. Hubbard
-    affiliation: Address of author2
+    affiliation: 1
+    orcid: 0000-0002-3769-0127
   - name: Mark J. van der Laan
-    affiliation: Address of author2
+    affiliation: 1, 2
+    orcid: TBA
+affiliations:
+  - name: Division of Biostatistics, University of California, Berkeley
+    index: 1
+  - name: Department of Statistics, University of California, Berkeley
+    index: 2
 abstract: |
   We present `methyvim`, a general algorithm for the nonparametric estimation of
-  treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
+  treatment/exposure effects on DNA methylation at CpG sites, complete with
+  straightforward statistical inference for such estimates. This approach
   leverages variable importance measures, a class of parameters that arise in
@@
 (often generalized linear) models. This is because typical statistical
-estimation approaches are not achievable for such high-dimensional data without restrictive modeling assumptions. When these assumptions are violated, estimates
+estimation approaches are not achievable for such high-dimensional data without
+restrictive modeling assumptions. When these assumptions are violated, estimates
 of the population-level effect of an exposure or treatment may have large bias.
@@
 However, interpretable statistical inference (i.e. confidence intervals,
-hypothesis tests) resulting from these data-adaptive estimates is challenging [@libbrecht2015machine]. In this paper, we present an alternative such
+hypothesis tests) resulting from these data-adaptive estimates is challenging
+[@libbrecht2015machine]. In this paper, we present an alternative such
 statistical analysis approaches in the form of nonparametric estimation
@@
 importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
-the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
-and thus provide reliable inference. Counterfactuals are "counter"-fact, defined
-in terms of an unrealistic yet ideal experiment. In the context of DNA
-methylation and for a binary treatment, we define the counterfactual
-outcomes as the Beta-value/M-value a CpG site would have if all subjects were
-treated and the Beta-value/M-value a CpG site would have if all subjects were
-not treated. Of course, these counterfactual outcomes are impossible to observe
-but they are very helpful for us to asses the importance of a CpG site relative
-to a subjects treatment/exposure and they can be statistically estimated [@van2006statistical]. Our VIM definition uses the concepts of counterfactuals
-and the framework for estimation is built on _targeted maximum likelihood
-estimation or targeted minimum loss-based estimation_ (__TMLE__)
-[@van2006targeted]. This methodology targets the individual importance of each
-CpG site by taking advantage of the prediction power of machine learning
-algorithms. The mathematical properties of this approach have been explored in
-[@tuglus2008targeted] and we focus on our contribution to it's application,
-`methyvim`, an R package that accommodates this technique for differential
-methylation analysis.
+the False Discovery Rate (FDR) as if all sites were tested. Under
+counterfactual assumptions, these estimators have asymptotic normal sampling
+distributions and thus provide reliable inference. Counterfactuals are
+"counter"-fact, defined in terms of an unrealistic yet ideal experiment. In the
+context of DNA methylation and for a binary treatment, we define the
+counterfactual outcomes as the Beta-value/M-value a CpG site would have if all
+subjects were treated and the Beta-value/M-value a CpG site would have if all
+subjects were not treated. Of course, these counterfactual outcomes are
+impossible to observe but they are very helpful for us to asses the importance
+of a CpG site relative to a subjects treatment/exposure and they can be
+statistically estimated [@vdl2006statistical]. Our VIM definition uses the
+concepts of counterfactuals and the framework for estimation is built on
+_targeted maximum likelihood estimation_ or _targeted minimum loss-based
+estimation_ (__TMLE__) [@vdl2006targeted]. This methodology targets the
+individual importance of each CpG site by taking advantage of the prediction
+power of machine learning algorithms. The mathematical properties of this
+approach have been explored in [@tuglus2008targeted] and we focus on our
+contribution to it's application, `methyvim`, an R package that make this
+technique for differential methylation analysis easily accessible.
 
@@
    approach adapts core routines from the
-  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
-  package. Future releases will support functionality from other packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
+   [`limma`](http://bioconductor.org/packages/limma) R
+   package. Future releases will support functionality from other packages
+   (e.g., [`randomForest`](https://CRAN.R-project.org/package=randomForest),
    [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
@@
    Future releases will support doubly-robust estimates of these VIMs (via the
-  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
+   [`drtmle`](https://cran.r-project.org/package=drtmle) package)
    and add parameters for continuous treatments/exposures (via the
@@
    multiple testing nature of the estimation problem, a variant of the Benjamini
-  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
-  [@benjamini1995controlling]. Specifically, we apply the "modified marginal
-  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
-  multi-stage analyses" (FDR-MSA), which is guaranteed to control the FDR as if
-  all sites were tested (i.e., without screening) [@tuglus2009modified].
+   & Hochberg procedure for controlling the False Discovery Rate (FDR) is
+   applied [@benjamini1995controlling]. Specifically, we apply the "modified
+   marginal Benjamini & Hochberg step-up False Discovery Rate controlling
+   procedure for multi-stage analyses" (FDR-MSA), which is guaranteed to
+   control the FDR as if all sites were tested (i.e., without screening)
+   [@tuglus2009modified].
 
@@
   methylation at a target CpG site) on an outcome of interest is estimated,
-  controlling for the observed methylation at all other CpG sites in the same neighborhood as the target (treatment) site, based on a parameter that
+  controlling for the observed methylation at all other CpG sites in the same
+  neighborhood as the target (treatment) site, based on a parameter that
   compares values of the treatment against a reference value taken to be the
   null. In particular, the implementation provided is designed to assess the
-  effect of differential methylation at the target CpG site on a (typically) phenotype-level outcome of interest (e.g., survival), in effect providing an nonparametric evaluation of the impact of methylation at the target site on
+  effect of differential methylation at the target CpG site on a (typically)
+  phenotype-level outcome of interest (e.g., survival), in effect providing an
+  nonparametric evaluation of the impact of methylation at the target site on
   said outcome.
 
-#### Class methytmle
+#### Class `methytmle`
 
@@
 values from the methylated and unmethylated channels to Beta-values
-(via `ratioConvert`). These two steps together produce an object of class `GenomicRatioSet`, provided by the `minfi` package.
+(via `ratioConvert`). These two steps together produce an object of class
+`GenomicRatioSet`, provided by the `minfi` package.
 
@@
 ```
-Additionally, a `GenomicRatioSet` can be created from a matrix with the function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.
+Additionally, a `GenomicRatioSet` can be created from a matrix with the
+function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.
 
@@
 for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
-all sites were tested, and the estimation framework is built on targeted minimum loss-based estimation. By taking advantage of the prediction power of machine
-learning algorithms, this methodology targets the individual importance of each
-CpG site, providing a fully nonparametric statistical model for the analysis of
-differential methylation.
+all sites were tested, and the estimation framework is built on targeted
+minimum loss-based estimation. By taking advantage of the prediction power of
+machine learning algorithms, this methodology targets the individual importance
+of each CpG site, providing a fully nonparametric statistical model for the
+analysis of differential methylation.
 
@@
 
-Software available from:
+Latest source code (development version): https://github.com/nhejazi/methyvim
 
-Latest source code: https://github.com/nhejazi/methyvim
+Bioconductor (stable release): https://bioconductor.org/packages/methyvim
 
@@
 
-Software license:
+https://github.com/nhejazi/methyvim
+
+Software license: MIT, copyright Nima S. Hejazi
 
@@
 
-NH designed and implemented the tool, applied the tool to the use case, and
-drafted the vignette. RP helped design the tool and drafted the article.
-AH and ML serve as the advisors for the development of this tool.
+NH designed and implemented the software package, applied the tool to the use
+cases, and co-drafted the present article. RP helped in designing the software
+and co-drafted the present article. AH and ML serve as the advisors for the
+development of this software and general statistical algorithm.
 
@@
 
-The author(s) declared that no grants were involved in supporting this work.
+NH was supported in part by the National Library Of Medicine of the National
+Institutes of Health under Award Number T32LM012417, by [SUPERFUND GRANT], and
+by [NINA HOLLAND GRANT]. RP was supported by [SUPERFUND GRANT]. The content of
+this work is solely the responsibility of the authors and does not necessarily
+represent the official views of the various funding sources and agencies.
 

### 2017-10-18 R/screen_limma.R

@@ -36,14 +36,8 @@ limma_screen <- function(methytmle, var_int, type, cutoff = 0.05) {
   design <- as.matrix(cbind(rep(1, times = length(var_int)), var_int))
 
   # create expression object for modeling
-  # M-values have the property of having far more centrality than Beta-values.
-  # Parametric techniques work best with an underlying Gaussian distribution to
-  # the data so M values might be more appropriate to use for analysis.
   if (type == "Beta") {
-    Betas <- minfi::getBeta(methytmle)
-    # shifting the betas from 0 and 1 prevents Inf/Nan Mvals
-    newBetas <- Harman::shiftBetas(methytmle_exprs, shiftBy=1e-4)
-    methytmle_exprs <- wateRmelon::Beta2M(newBetas)
+    methytmle_exprs <- minfi::getBeta(methytmle)
   } else if (type == "Mval") {
     methytmle_exprs <- minfi::getM(methytmle)
   }

### 2017-10-18 R/zzz.R

@@ -1,8 +1,6 @@
 .onAttach <- function(...) {
-  packageStartupMessage(paste("methyvim: Nonparametric Differential",
-                              "Methylation Analysis \n using Targeted",
-                              "Estimates of Variable Importance Measures"))
+  packageStartupMessage(paste("methyvim: Targeted Learning for Differential",
+                              "Methylation Analysis"))
   packageStartupMessage("Version: ",
                         utils::packageDescription("methyvim")$Version)
 }
-

### 2017-10-18 R/screen_limma.R
F1000 paper updates (#30)
* use minfi data

* add paper refs

* started to add a bit to the introduction &  moved the text from the repository to the appropriate sections in the article. I will use the text from the repo as a rough guide to draft something more cohesive

* revisions/updates to F1000 manuscript (#28)

* minor vignette additions/corrections

* almost done with f1000, minor vignette corrections

* adding the Summary and final touches

* a few more minor things

* fixing merge conflict

* minor formatting

* suggestion: convert betas to mvals in the limma screenas the mvals have f

@@ -54,4 +54,3 @@ limma_screen <- function(methytmle, var_int, type, cutoff = 0.05) {
   methytmle@screen_ind <- as.numeric(indices_pass)
   return(methytmle)
 }
-

### 2017-10-18 R/screen_limma.R
add beta-shifting; borked initially

@@ -2,7 +2,9 @@
 #'
 #' Reduces the \code{methytmle} object by way of a \code{limma} model with
 #' empirical Bayes shrinkage to include only the CpG sites below the preset
-#' p-value cutoff. INTERNAL USE ONLY.
+#' p-value cutoff. Since M-values display far more centrality in distribution
+#' than do Beta-values, this parametric screening technique converts from Beta-
+#' to M-values before fitting a model. INTERNAL USE ONLY.
 #'
 #' @param methytmle An object of class \code{methytmle}.
 #' @param var_int A \code{numeric} vector containing subject-level measurements
@@ -27,6 +29,8 @@
 #'
 #' @importFrom limma lmFit eBayes topTable
 #' @importFrom minfi getBeta getM
+#' @importFrom Harman shiftBetas
+#' @importFrom lumi beta2m
 #
 limma_screen <- function(methytmle, var_int, type, cutoff = 0.05) {
 
@@ -37,7 +41,10 @@ limma_screen <- function(methytmle, var_int, type, cutoff = 0.05) {
 
   # create expression object for modeling
   if (type == "Beta") {
-    methytmle_exprs <- minfi::getBeta(methytmle)
+    betas <- minfi::getBeta(methytmle)
+    # shifting the betas away from 0 and 1 prevents Inf/Nan M-values
+    betas_shifted <- Harman::shiftBetas(betas, shiftBy = 1e-3)
+    methytmle_exprs <- lumi::beta2m(betas_shifted)
   } else if (type == "Mval") {
     methytmle_exprs <- minfi::getM(methytmle)
   }

### 2017-10-18 paper/paper_BiocF1000.Rmd
update+rebuild f1000 paper

@@
 ---
-title: '`methyvim`: an R package for Variable Importance Measures for
-Differential Methylation Analysis'
+title: '`methyvim`: Targeted and Model-free Analysis of Differential Methylation
+in R'
 author:
   - name: Nima S. Hejazi
-    affiliation: 1
+    affiliation: Division of Biostatistics, University of California, Berkeley
     orcid: 0000-0002-7127-2789
   - name: Rachael V. Phillips
-    affiliation: 1
+    affiliation: Division of Biostatistics, University of California, Berkeley
     orcid: 0000-0002-8474-591X
   - name: Alan E. Hubbard
-    affiliation: 1
+    affiliation: Division of Biostatistics, University of California, Berkeley
     orcid: 0000-0002-3769-0127
   - name: Mark J. van der Laan
-    affiliation: 1, 2
+    affiliation: Division of Biostatistics, University of California, Berkeley
     orcid: TBA
-affiliations:
-  - name: Division of Biostatistics, University of California, Berkeley
-    index: 1
-  - name: Department of Statistics, University of California, Berkeley
-    index: 2
 abstract: |
@@
 keywords: DNA methylation, differential methylation, epigenetics, causal
-  inference, variable importance, targeted minimum loss-based estimation
+  inference, statistical variable importance, targeted minimum loss-based
+  estimation, machine learning
 bibliography: paper_BiocF1000.bib
@@
 hypothesis tests) resulting from these data-adaptive estimates is challenging
-[@libbrecht2015machine]. In this paper, we present an alternative such
-statistical analysis approaches in the form of nonparametric estimation
-procedures that provide simple and well interpretable statistical inference.
-Inspired by concepts of counterfactual effects from the causal inference
-literature and machine learning algorithms, we provide a computationally
-efficient technique for obtaining targeted estimates of nonparametric _variable
-importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
-the False Discovery Rate (FDR) as if all sites were tested. Under
-counterfactual assumptions, these estimators have asymptotic normal sampling
-distributions and thus provide reliable inference. Counterfactuals are
-"counter"-fact, defined in terms of an unrealistic yet ideal experiment. In the
-context of DNA methylation and for a binary treatment, we define the
-counterfactual outcomes as the Beta-value/M-value a CpG site would have if all
-subjects were treated and the Beta-value/M-value a CpG site would have if all
-subjects were not treated. Of course, these counterfactual outcomes are
-impossible to observe but they are very helpful for us to asses the importance
-of a CpG site relative to a subjects treatment/exposure and they can be
-statistically estimated [@vdl2006statistical]. Our VIM definition uses the
-concepts of counterfactuals and the framework for estimation is built on
-_targeted maximum likelihood estimation_ or _targeted minimum loss-based
-estimation_ (__TMLE__) [@vdl2006targeted]. This methodology targets the
-individual importance of each CpG site by taking advantage of the prediction
-power of machine learning algorithms. The mathematical properties of this
-approach have been explored in [@tuglus2008targeted] and we focus on our
-contribution to it's application, `methyvim`, an R package that make this
-technique for differential methylation analysis easily accessible.
+[@libbrecht2015machine].
+
+In this paper, we present an alternative such statistical analysis approaches in
+the form of nonparametric estimation procedures that provide simple and well
+interpretable statistical inference. Inspired by concepts of counterfactual
+effects from the causal inference literature and machine learning algorithms, we
+provide a computationally efficient technique for obtaining targeted estimates
+of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set
+of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
+tested. Under counterfactual assumptions, these estimators have asymptotic
+normal sampling distributions and thus provide reliable inference.
+Counterfactuals are "counter"-fact, defined in terms of an unrealistic yet ideal
+experiment. In the context of DNA methylation and for a binary treatment, we
+define the counterfactual outcomes as the observed methylation value (whether
+Beta- or M-values) a CpG site would have if all subjects were treated and the
+methylation value a CpG site would have if all subjects were not treated. Of
+course, these counterfactual outcomes are impossible to observe but they are
+very helpful for us to asses the importance of a CpG site relative to a subjects
+treatment/exposure and they can be statistically estimated
+[@vdl2006statistical]. Our VIM definition uses the concepts of counterfactuals
+and the framework for estimation is built on _targeted maximum likelihood
+estimation_ or _targeted minimum loss-based estimation_ (__TMLE__)
+[@vdl2006targeted]. This methodology targets the individual importance of each
+CpG site by taking advantage of the prediction power of machine learning
+algorithms. The mathematical properties of this approach have been explored in
+[@tuglus2008targeted] and we focus on our contribution to it's application,
+`methyvim`, an R package that make this technique for differential methylation
+analysis easily accessible.
 
@@
 
-We refer the reader to the package vignette, "`methyvim`: Variable Importance
-Measures for Differential Methylation" hosted with the package on Bioconductor,
-for a detailed description of the methods used in this package.
-(NEED CITATION FOR VIGNETTE)
+We refer the reader to the package vignette, "`methyvim`: Targeted Learning for
+Differential Methylation Analysis," included in any distribution of the software
+package, for further details.
 
@@
 ```{r make-methytmle}
+library(methyvim)
 mtmle <- .methytmle(grs)
@@
 
-```{r methyvim-pvals-both}
-plot(methyvim_cancer_ate)
+```{r methyvim-pvals-raw}
+plot(methyvim_cancer_ate, type = "raw_pvals")
 ```
-__Remark:__ The plots displayed above may also be generated separately by
-explicitly setting the argument "type" to `plot.methytmle`. For a plot of the
-raw p-values, specify `type = "raw_pvals"`, and for a plot of the FDR-corrected
-p-values, specify `type = "fdr_pvals"`.
+
+```{r methyvim-pvals-fdr}
+plot(methyvim_cancer_ate, type = "fdr_pvals")
+```
+
+__Remark:__ The plots displayed above may also be generated as side-by-side
+histograms in a single plot object. This is the default for the `plot` method
+and may easily be invoked by specifying no additional arguments to the `plot`
+function, unlike in the above.
 
@@
 
-Archived source code as at time of publication:
+Archived source code as at time of
+publication: https://github.com/nhejazi/methyvim
 
-https://github.com/nhejazi/methyvim
-
-Software license: MIT, copyright Nima S. Hejazi
+Software license: The MIT License, copyright Nima S. Hejazi
 

### 2017-10-19 README.Rmd
minor name updates

@@
 
-> Targeted learning of variable importance measures for the analysis of
-> differential methylation experiments
+> Targeted Data-Adaptive Estimation and Inference for Differential Methylation
+> Analysis
 
@@
             J},
-          title = {methyvim: Targeted and model-free analysis of differential
-            methylation in R},
+          title = {methyvim: Targeted and model-free differential methylation
+            analysis in R},
           journal = {}

### 2017-10-19 paper/paper_BiocF1000.Rmd
minor name updates

@@
 ---
-title: '`methyvim`: Targeted and Model-free Analysis of Differential Methylation
-in R'
+title: '`methyvim`: Targeted and Model-free Differential Methylation Analysis in
+  R'
 author:
@@
 
-We refer the reader to the package vignette, "`methyvim`: Targeted Learning for
-Differential Methylation Analysis," included in any distribution of the software
-package, for further details.
+We refer the reader to the package vignette, "`methyvim`: Targeted Data-Adaptive
+Estimation and Inference for Differential Methylation Analysis," included in any
+distribution of the software package, for further details.
 

### 2017-11-09 README-withrefs.Rmd
minor fixes

--- 2017-10-19 README.Rmd
@@
 ---
-output: rmarkdown::github_document
+output:
+  rmarkdown::github_document
 bibliography: README-refs.bib
@@
 
-<!--
 For standard use, install from [Bioconductor](https://bioconductor.org):
+
 ```{r bioc-installation, eval = FALSE}
@@
 ```
--->
 
-Install the most recent _stable release_ from GitHub via
+To contribute, install the bleeding-edge _development version_ from GitHub via
 [`devtools`](https://www.rstudio.com/products/rpackages/devtools/):
@@
 
-<!--
-To contribute, install the _development version_ from GitHub via
-[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):
+Current and prior [Bioconductor](https://bioconductor.org) releases are
+available under branches with numbers prefixed by "RELEASE_". For example, to
+install the version of this package available via Bioconductor 3.6, use
 
 ```{r gh-develop-installation, eval = FALSE}
-devtools::install_github("nhejazi/methyvim", ref = "develop")
+devtools::install_github("nhejazi/methyvim", ref = "RELEASE_3_6")
 ```
--->
 

### 2017-11-09 R/methyvim.R

@@ -239,15 +239,14 @@ methyvim <- function(data_grs,
 
     # TMLE procedure is now done, so let's just make the output object pretty...
     if (vim == "ate") {
-      colnames(methy_vim_out) <- c("lowerCI_ATE", "est_ATE", "upperCI_ATE",
-                                   "Var_ATE", "pval", "n_neighbors_all",
-                                   "n_neighbors_w", "max_corr_w")
+      colnames(methy_vim_out) <- c("lowCI_ATE", "est_ATE", "upCI_ATE",
+                                   "var_ATE", "pval", "n_W", "n_W_contr",
+                                   "max_cor_n_W")
       methy_tmle@param <- "Average Treatment Effect"
     } else {
-      colnames(methy_vim_out) <- c("lowerCI_logRR", "est_logRR",
-                                   "upperCI_logRR", "Var_logRR", "pval",
-                                   "n_neighbors_all", "n_neighbors_w",
-                                   "max_corr_w")
+      colnames(methy_vim_out) <- c("lowCI_logRR", "est_logRR", "upCI_logRR",
+                                   "var_logRR", "pval", "n_W", "n_W_contr",
+                                   "max_cor_n_W")
       methy_tmle@param <- "Risk Ratio"
     }
 

### 2017-11-29 README.Rmd
rework readme; add funding

@@
 
+## Funding
+
+The development of this software was supported in part through a grant from the
+National Library of Medicine of the NIH (T32 LM012417).
+
+---
+
 ## License

### 2017-12-21 README-norefs.Rmd

--- 2017-11-29 README.Rmd
@@
 ---
-output:
-  rmarkdown::github_document
-bibliography: README-refs.bib
+output: rmarkdown::github_document
 ---
@@
 
-* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
+* A _nonparametric variable importance measure_ (NPVI):
   The effect of continous-valued exposure or treatment (the observed methylation
@@
 multiple testing nature of the estimation problem, a variant of the Benjamini &
-Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
-[@benjamini1995controlling]. Specifically, we apply the modified marginal
-Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
-multi-stage analyses (FDR-MSA) [@tuglus2009modified].
-
-For a general discussion of the framework of targeted minimum loss-based
-estimation, the many applications of this methodology, and the role the
-framework plays in statistical causal inference, the recommended references are
-@vdl2011targeted and @vdl2017targeted. @hernan2018causal and @pearl2009causality
-may be of interest to those desiring a more general introduction to statistical
-causal inference.
-
-<!--
-Note about shrinkage of influence curves, adapting @smyth2004linear.
-More exposition here....
--->
+Hochberg procedure for controlling the False Discovery Rate (FDR) is applied.
+Specifically, we apply the modified marginal Benjamini & Hochberg step-up False
+Discovery Rate controlling procedure for multi-stage analyses (FDR-MSA).
 

### 2017-12-21 paper/paper_BiocF1000.tex

--- 2017-05-19 paper_BiocF1000/paper_BiocF1000.tex
@@
 
-\title{\emph{F1000Research} Software Tool Article Template}
+\title{\texttt{methyvim}: Targeted and Model-free Analysis of Differential
+Methylation in R}
 
-\author[1]{Nima Hejazi}
-\author[2]{Alan E. Hubbard}
-\author[2]{Mark J. van der Laan}
-\affil[1]{Address of author1}
-\affil[2]{Address of author2}
+\author[1]{Nima S. Hejazi}
+\author[1]{Rachael V. Phillips}
+\author[1]{Alan E. Hubbard}
+\author[1]{Mark J. van der Laan}
+\affil[1]{Division of Biostatistics, University of California, Berkeley}
 
@@
 \begin{abstract}
-Abstracts should be up to 300 words and provide a succinct summary of
-the article. Although the abstract should explain why the article might
-be interesting, care should be taken not to inappropriately
-over-emphasise the importance of the work described in the article.
-Citations should not be used in the abstract, and the use of
-abbreviations should be minimized.
+We present \texttt{methyvim}, a general algorithm for the nonparametric
+estimation of treatment/exposure effects on DNA methylation at CpG
+sites, complete with straightforward statistical inference for such
+estimates. This approach leverages variable importance measures, a class
+of parameters that arise in causal inference and are defined in such a
+manner that they provide targeted estimates of the relative importance
+of individual CpG site with respect to a binary exposure/treatment
+assigned at the level of subjects. Additionally, this procedure is
+computationally efficient; incorporating a preliminary screening step to
+isolate a subset of sites for which there is cursory evidence of
+differential methylation and a multiple testing correction step to
+control the False Discovery Rate as if all sites were tested. This
+technique for analysis of differentially methylated positions provides
+an avenue to move beyond linear models without the loss of interpretable
+statistical inference.
 \end{abstract}
@@
 \section*{Keywords}
-Please list up to eight keywords to help readers interested in your
-article find it more easily.
+DNA methylation, differential methylation, epigenetics, causal
+inference, statistical variable importance, targeted minimum loss-based
+estimation, machine learning
 
@@
 
-The introduction provides context as to why the software tool was
-developed and what need it addresses. It is good scholarly practice to
-mention previously developed tools that address similar needs, and why
-the current tool is needed.
+DNA methylation is a fundamental epigenetic process known to play an
+important role in the regulation of gene expression. DNA CpG methylation
+involves the addition of a methyl group (\(\text{CH}_3\)) to the fifth
+carbon of the cytosine ring structure to form 5-methylcytosine. Numerous
+biological and medical studies have implicated DNA CpG methylation as
+playing a role in disease and development (Robertson 2005). Perhaps
+unsurprisingly then, biotechnologies have been developed to study the
+molecular mechanisms of this epigenetic process. Modern assays, like the
+Illumina \emph{Infinium} Methylation assay, allow for quantitative
+interrogation of DNA methylation of CpG sites scattered across the
+genome at single-nucleotide resolution; moreover, much effort has been
+invested, by the bioinformatics community, in the development of tools
+for properly removing technological effects that may contaminate
+biological signatures measured by such assays. Despite these advances in
+both biological and bioninformatical techniques, most statistical
+methods available for differential analysis of data produced by such
+assays rely on over-simplified (often generalized linear) models. This
+is because typical statistical estimation approaches are not achievable
+for such high-dimensional data without restrictive modeling assumptions.
+When these assumptions are violated, estimates of the population-level
+effect of an exposure or treatment may have large bias. Data-adaptive
+estimation procedures that utilize machine learning can control for
+high-dimensional confounders when estimating treatment/exposure effects.
+However, interpretable statistical inference (i.e.~confidence intervals,
+hypothesis tests) resulting from these data-adaptive estimates is
+challenging (Libbrecht and Noble 2015).
+
+In this paper, we present an alternative such statistical analysis
+approaches in the form of nonparametric estimation procedures that
+provide simple and well interpretable statistical inference. Inspired by
+concepts of counterfactual effects from the causal inference literature
+and machine learning algorithms, we provide a computationally efficient
+technique for obtaining targeted estimates of nonparametric
+\emph{variable importance measures} (\textbf{VIM}) for a pre-screened
+set of CpGs, controlling for the False Discovery Rate (FDR) as if all
+sites were tested. Under counterfactual assumptions, these estimators
+have asymptotic normal sampling distributions and thus provide reliable
+inference. Counterfactuals are ``counter''-fact, defined in terms of an
+unrealistic yet ideal experiment. In the context of DNA methylation and
+for a binary treatment, we define the counterfactual outcomes as the
+observed methylation value (whether Beta- or M-values) a CpG site would
+have if all subjects were treated and the methylation value a CpG site
+would have if all subjects were not treated. Of course, these
+counterfactual outcomes are impossible to observe but they are very
+helpful for us to asses the importance of a CpG site relative to a
+subjects treatment/exposure and they can be statistically estimated (M.
+J. van der Laan 2006). Our VIM definition uses the concepts of
+counterfactuals and the framework for estimation is built on
+\emph{targeted maximum likelihood estimation} or \emph{targeted minimum
+loss-based estimation} (\textbf{TMLE}) (M. J. van der Laan and Rubin
+2006). This methodology targets the individual importance of each CpG
+site by taking advantage of the prediction power of machine learning
+algorithms. The mathematical properties of this approach have been
+explored in (Tuglus and van der Laan 2008) and we focus on our
+contribution to it's application, \texttt{methyvim}, an R package that
+make this technique for differential methylation analysis easily
+accessible.
+
+For a general discussion of the framework of targeted minimum loss-based
+estimation and the role this approach plays in statistical causal
+inference, the interested reader is invited to consult M. J. van der
+Laan and Rose (2011) and M. J. van der Laan and Rose (2017). For a more
+general introduction to (statistical) causal inference, Pearl (2009) and
+Hernan and Robins (2018, forthcoming) may be of interest.
 
@@
 
-For software tool papers, this section should address how the tool works
-and any relevant technical details required for implementation of the
-tool by other developers.
-
-\subsection{Operation}\label{operation}
-
-This part of the methods should include the minimal system requirements
-needed to run the software and an overview of the workflow for the tool
-for users of the tool.
-
-\section{Results }\label{results}
-
-This section is only required if the paper includes novel data or
-analyses, and should be written as a traditional results section.
-
-\section{Use Cases }\label{use-cases}
-
-This section is required if the paper does not include novel data or
-analyses. Examples of input and output files should be provided with
-some explanatory context. Any novel or complex variable parameters
-should also be explained in sufficient detail to allow users to
-understand and use the tool's functionality.
-
-\section{Discussion }\label{discussion}
-
-This section is only required if the paper includes novel data or
-analyses, and should be written in the same style as a traditional
-discussion section. Please include a brief discussion of allowances made
-(if any) for controlling bias or unwanted sources of variability, and
-the limitations of any novel datasets.
-
-\section{Conclusions }\label{conclusions}
-
-This section is only required if the paper includes novel data or
-analyses, and should be written as a traditional conclusion.
-
-\section{Summary }\label{summary}
-
-This section is required if the paper does not include novel data or
-analyses. It allows authors to briefly summarize the key points from the
-article.
-
-\section{Data availability }\label{data-availability}
-
-Please add details of where any datasets that are mentioned in the
-paper, and that have not have not previously been formally published,
-can be found. If previously published datasets are mentioned, these
-should be cited in the references, as per usual scholarly conventions.
-
-\section{Software availability}\label{software-availability}
-
-This section will be generated by the Editorial Office before
-publication. Authors are asked to provide some initial information to
-assist the Editorial Office, as detailed below.
+The core functionality of this package is made available via the
+eponymous \texttt{methyvim} function, which implements a statistical
+algorithm designed to compute targeted estimates of VIMs, defined in
+such a way that the VIMs represent parameters of scientific interest in
+computational biology experiments; moreover, these VIMs are defined such
+that they may be estimated in a manner that is very nearly
+assumption-free, that is, within a \emph{fully nonparametric statistical
+model}. \textbf{The statistical algorithm consists in several major
+steps}:
 
@@
 \item
-  URL link to where the software can be downloaded from or used by a
-  non-coder (AUTHOR TO PROVIDE; optional)
+  Pre-screening of genomic sites is used to isolate a subset of sites
+  for which there is cursory evidence of differential methylation. For
+  the sake of computational feasibility, targeted minimum loss-based
+  estimates of VIMs are computed only for this subset of sites.
+  Currently, the available screening approach adapts core routines from
+  the \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R
+  package. Future releases will support functionality from other
+  packages (e.g.,
+  \href{https://CRAN.R-project.org/package=randomForest}{\texttt{randomForest}},
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}).
+  Following the style of the function for performing screening via
+  \texttt{limma}, users may write their own screening functions and are
+  invited to contribute such functions to the core software package by
+  opening pull requests at the GitHub repository.
 \item
-  URL link to the author's version control system repository containing
-  the source code (AUTHOR TO PROVIDE; required)
+  Nonparametric estimates of VIMs, for the specified target parameter,
+  are computed at each of the CpG sites passing the screening step. The
+  VIMs are defined in such a way that the estimated effects is of an
+  exposure/treatment on the methylation status of a target CpG site,
+  controlling for the observed methylation status of the neighbors of
+  that site. Currently, routines are adapted from the
+  \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R
+  package. Future releases will support doubly-robust estimates of these
+  VIMs (via the
+  \href{https://cran.r-project.org/package=drtmle}{\texttt{drtmle}}
+  package) and add parameters for continuous treatments/exposures (via
+  the
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}
+  package).
 \item
-  Link to source code as at time of publication (\emph{F1000Research} TO
-  GENERATE)
-\item
-  Link to archived source code as at time of publication
-  (\emph{F1000Research} TO GENERATE)
-\item
-  Software license (AUTHOR TO PROVIDE; required)
+  Since pre-screening is performed prior to estimating VIMs, we make use
+  of a multiple testing correction uniquely suited to such settings. Due
+  to the multiple testing nature of the estimation problem, a variant of
+  the Benjamini \& Hochberg procedure for controlling the False
+  Discovery Rate (FDR) is applied (Benjamini and Hochberg 1995).
+  Specifically, we apply the ``modified marginal Benjamini \& Hochberg
+  step-up False Discovery Rate controlling procedure for multi-stage
+  analyses'' (FDR-MSA), which is guaranteed to control the FDR as if all
+  sites were tested (i.e., without screening) (Tuglus and van der Laan
+  2009).
 \end{enumerate}
 
-\section{Author contributions}\label{author-contributions}
+\paragraph{Parameters of Interest}\label{parameters-of-interest}
 
-In order to give appropriate credit to each author of an article, the
-individual contributions of each author to the manuscript should be
-detailed in this section. We recommend using author initials and then
-stating briefly how they contributed.
+For the CpG sites that passed the pre-screening step, the user chooses
+the parameter of interest that will be computed at each site. \emph{In
+all cases, an estimator of the parameter of interest is constructed via
+targeted minimum loss-based estimation}.
 
-\section{Competing interests}\label{competing-interests}
+For discrete-valued treatments or exposures:
 
-All financial, personal, or professional competing interests for any of
-the authors that could be construed to unduly influence the content of
-the article must be disclosed and will be displayed alongside the
-article. If there are no relevant competing interests to declare, please
-add the following: `No competing interests were disclosed'.
+\begin{itemize}
+\item
+  The average treatment effect (ATE): The effect of a binary exposure or
+  treatment on the observed methylation at a target CpG site is
+  estimated, controlling for the observed methylation at all other CpG
+  sites in the same neighborhood as the target site, based on an
+  additive form. Often denoted \(\psi_0 = \psi_0(1) - \psi_0(0)\), the
+  parameter estimate represents the additive difference in methylation
+  that would have been observed at the target site had all observations
+  received the treatment versus the counterfactual under which none
+  received the treatment.
+\item
+  The relative risk (RR): The effect of a binary exposure or treatment
+  on the observed methylation at a target CpG site is estimated,
+  controlling for the observed methylation at all other CpG sites in the
+  same neighborhood as the target site, based on a geometric form. Often
+  denoted, \(\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}\), the parameter
+  estimate represents the multiplicative difference in methylation that
+  would have been observed at the target site had all observations
+  received the treatment versus the counterfactual under which none
+  received the treatment.
+\end{itemize}
 
-\section{Grant information}\label{grant-information}
+Estimating the VIM corresponding to the parameters above, for
+discrete-valued treatments or exposures, requires two separate
+regression steps: one for the treatment mechanism (propensity score) and
+one for the outcome regression. Technical details on the nature of these
+regressions are discussed in Hernan and Robins (2018, forthcoming), and
+details for estimating these regressions in the framework of targeted
+minimum loss-based estimation are discussed in M. J. van der Laan and
+Rose (2011).
 
-Please state who funded the work discussed in this article, whether it
-is your employer, a grant funder etc. Please do not list funding that
-you have that is not relevant to this specific piece of research. For
-each funder, please state the funder's name, the grant number where
-applicable, and the individual to whom the grant was assigned. If your
-work was not funded by any grants, please include the line: `The
-author(s) declared that no grants were involved in supporting this
-work.'
+Support for continuous-valued treatments or exposures is \emph{planned
+but not yet available}, though work is underway to incorporate into our
+methodology the following
 
-\section{Acknowledgments}\label{acknowledgments}
+\begin{itemize}
+\item
+  A nonparametric variable importance measure (NPVI) (Chambaz, Neuvial,
+  and van der Laan 2012): The effect of continuous-valued exposure or
+  treatment (the observed methylation at a target CpG site) on an
+  outcome of interest is estimated, controlling for the observed
+  methylation at all other CpG sites in the same neighborhood as the
+  target (treatment) site, based on a parameter that compares values of
+  the treatment against a reference value taken to be the null. In
+  particular, the implementation provided is designed to assess the
+  effect of differential methylation at the target CpG site on a
+  (typically) phenotype-level outcome of interest (e.g., survival), in
+  effect providing an nonparametric evaluation of the impact of
+  methylation at the target site on said outcome.
+\end{itemize}
 
-This section should acknowledge anyone who contributed to the research
-or the article but who does not qualify as an author based on the
-criteria provided earlier (e.g.~someone or an organization that provided
-writing assistance). Please state how they contributed; authors should
-obtain permission to acknowledge from all those mentioned in the
-Acknowledgments section.
+\paragraph{\texorpdfstring{Class
+\texttt{methytmle}}{Class methytmle}}\label{class-methytmle}
 
-Please do not list grant funding in this section.
+We have adopted a class \texttt{methytmle} to help organize the
+functionality within this package. The \texttt{methytmle} class builds
+upon the \texttt{GenomicRatioSet} class provided by the \texttt{minfi}
+package so all of the slots of \texttt{GenomicRatioSet} are contained in
+a \texttt{methytmle} object. The new class introduced in the
+\texttt{methyvim} package includes several new slots:
 
-\section{USING R MARKDOWN}\label{using-r-markdown}
+\begin{itemize}
+\item
+  \texttt{call} - the form of the original call to the \texttt{methyvim}
+  function.
+\item
+  \texttt{screen\_ind} - indices identifying CpG sites that pass the
+  screening process.
+\item
+  \texttt{clusters} - non-unique IDs corresponding to the manner in wich
+  sites are treated as neighbors. These are assigned by genomic distance
+  (bp) and respect chromosome boundaries (produced via a call to
+  \texttt{bumphunter::clusterMaker}).
+\item
+  \texttt{var\_int} - the treatment/exposure status for each subject.
+  Currently, these must be binary, due to the definition of the
+  supported targeted parameters.
+\item
+  \texttt{param} - the name of the target parameter from which the
+  estimated VIMs are defined.
+\item
+  \texttt{vim} - a table of statistical results obtained from estimating
+  VIMs for each of the CpG sites that pass the screening procedure.
+\item
+  \texttt{ic} - the measured array values for each of the CpG sites
+  passing the screening, transformed into influence curve space based on
+  the chosen target parameter.
+\end{itemize}
 
-Some examples of commonly used markdown syntax are listed below, to help
-you get started.
+We refer the reader to the package vignette, ``\texttt{methyvim}:
+Targeted Learning for Differential Methylation Analysis,'' included in
+any distribution of the software package, for further details.
 
-\subsection{Cross-references}\label{cross-references}
+\subsection{Operation}\label{operation}
 
-For portability between different output formats, use the syntax
-introduced by \emph{bookdown}, such as
-\texttt{(\textbackslash{}\#label)} for labels and
-\texttt{\textbackslash{}@ref(label)} for cross-references. The following
-sections provide examples of referencing tables, figures, and equations.
+A standard computer with the latest version of R and Bioconductor 3.6
+installed will handle applications of the \texttt{methyvim} package.
 
-\subsection{Citations}\label{citations}
+\section{Use Cases}\label{use-cases}
 
-You can include refences in a standard Bibtex file. The name of this
-file is given in the header of the markdown document (in our case it is
-\emph{sample.bib}). References to entries in the Bibtex file are made
-using square brackets and use an @ plus the key for the entry you are
-referencing {[}@Smith:2012qr{]}. You can combine multiple entries by
-separating them with a semi-colon {[}@Smith:2012qr; @Smith:2013jd{]}.
+To examine the practical applications and the full set of utilities of
+the \texttt{methyvim} package, we will use a publicly available example
+data set produced by the Illumina 450K array, from the
+\texttt{minfiData} R package.
 
-\subsection{Code chunks}\label{code-chunks}
+\paragraph{Preliminaries: Setting up the
+Data}\label{preliminaries-setting-up-the-data}
 
-You can embed an R code chunk like this:
+We begin by loading the package and the data set. After loading the
+data, which comes in the form of a raw \texttt{MethylSet} object, we
+perform some further processing by mapping to the genome (with
+\texttt{mapToGenome}) and converting the values from the methylated and
+unmethylated channels to Beta-values (via \texttt{ratioConvert}). These
+two steps together produce an object of class \texttt{GenomicRatioSet},
+provided by the \texttt{minfi} package.
 
@@
 \begin{Highlighting}[]
-\NormalTok{x <-}\StringTok{ }\DecValTok{1}\OperatorTok{:}\DecValTok{10}
-\NormalTok{x}
+\KeywordTok{suppressMessages}\NormalTok{(}\KeywordTok{library}\NormalTok{(minfiData))}
+\KeywordTok{data}\NormalTok{(MsetEx)}
+\NormalTok{mset <-}\StringTok{ }\KeywordTok{mapToGenome}\NormalTok{(MsetEx)}
+\NormalTok{grs <-}\StringTok{ }\KeywordTok{ratioConvert}\NormalTok{(mset)}
+\NormalTok{grs}
 \end{Highlighting}
@@
 \begin{verbatim}
-##  [1]  1  2  3  4  5  6  7  8  9 10
+## class: GenomicRatioSet 
+## dim: 485512 6 
+## metadata(0):
+## assays(2): Beta CN
+## rownames(485512): cg13869341 cg14008030 ... cg08265308 cg14273923
+## rowData names(0):
+## colnames(6): 5723646052_R02C02 5723646052_R04C01 ...
+##   5723646053_R05C02 5723646053_R06C02
+## colData names(13): Sample_Name Sample_Well ... Basename filenames
+## Annotation
+##   array: IlluminaHumanMethylation450k
+##   annotation: ilmn12.hg19
+## Preprocessing
+##   Method: Raw (no normalization or bg correction)
+##   minfi version: 1.21.2
+##   Manifest version: 0.4.0
 \end{verbatim}
 
-If you specify a figure caption to a code chunk using the chunk option
-\texttt{fig.cap}, the plot will be automatically labeled and numbered.
-The figure label is generated from the label of the code chunk by
-prefixing it with \texttt{fig:}, e.g., see Figure \ref{fig:plot}.
+We can create an object of class \texttt{methytmle} from any
+\texttt{GenomicRatioSet} object simply invoking the S4 class
+constructor:
 
@@
 \begin{Highlighting}[]
-\KeywordTok{plot}\NormalTok{(x)}
+\KeywordTok{library}\NormalTok{(methyvim)}
 \end{Highlighting}
@@
 
-\begin{figure}
-\centering
-\includegraphics{paper_BiocF1000_files/figure-latex/plot-1.pdf}
-\caption{\label{fig:plot}A caption to our sample plot.}
-\end{figure}
+\begin{verbatim}
+## methyvim: Targeted Learning for Differential Methylation Analysis
+\end{verbatim}
 
-\subsection{Tables}\label{tables}
-
-Markdown syntax tends to lack some of the more sophisticated formatting
-features available in LaTeX, so you may need to edit the tables later to
-get the desired format.
-
-\begin{table}[htbp]
-\caption{Caption to table.}
-\centering
-\begin{tabledata}{@{}lll@{}}
-\header First name & Last Name & Grade\\
-\row John & Doe & 7.5\\
-\row Richard & Miles & 2\\
-\end{tabledata}
-\end{table}
-
-Just like figures, tables with captions will also be numbered and can be
-referenced. Captions are entered as a paragraph beginning with the
-string ``Table:'' (or just ``:''), which may appear either before or
-after the table. A label for the table should appear in the beginning of
-the caption in the form of \texttt{(\textbackslash{}\#tab:label)}, e.g.,
-see Table \ref{tab:table}.
-
-\begin{table}[htbp]
-\caption{\label{tab:table} A table with text justification.}
-\centering
-\begin{tabledata}{@{}lcr@{}}
-\header First name & Last Name & Grade\\
-\row John & Doe & 7.5\\
-\row Richard & Miles & 2\\
-\end{tabledata}
-\end{table}
-
-\subsection{Figures}\label{figures}
-
-You can include static figures (i.e.~no generated by code) using the
-\texttt{include\_graphics()} function from the \textbf{knitr} package,
-in a standard code chunk.
+\begin{verbatim}
+## Version: 0.99.8
+\end{verbatim}
 
@@
 \begin{Highlighting}[]
-\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include_graphics}\NormalTok{(}\StringTok{'frog.jpg'}\NormalTok{)}
+\NormalTok{mtmle <-}\StringTok{ }\KeywordTok{.methytmle}\NormalTok{(grs)}
 \end{Highlighting}
@@
 
-\begin{figure}
+Additionally, a \texttt{GenomicRatioSet} can be created from a matrix
+with the function \texttt{makeGenomicRatioSetFromMatrix} provided by the
+\texttt{minfi} package.
 
-{\centering \includegraphics[width=0.5\linewidth]{frog}
+\paragraph{Differential Methylation
+Analysis}\label{differential-methylation-analysis}
 
-}
+For this example analysis, we'll treat the condition of the patients as
+the exposure/treatment variable of interest. The \texttt{methyvim}
+function requires that this variable either be \texttt{numeric} or
+easily coercible to \texttt{numeric}. To facilitate this, we'll simply
+convert the covariate (currently a \texttt{character}):
 
-\caption{Your figure legend goes here; it should be succinct, while still explaining all symbols and abbreviations.}\label{fig:frog-picture}
-\end{figure}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{var_int <-}\StringTok{ }\NormalTok{(}\KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{as.factor}\NormalTok{(}\KeywordTok{colData}\NormalTok{(grs)}\OperatorTok{$}\NormalTok{status)) }\OperatorTok{-}\StringTok{ }\DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-You can again use the \texttt{fig.cap} option to provide the figure
-caption, and reference the image based on the code chunk label. You can
-also use options such as \texttt{fig.align} and \texttt{fig.width} to
-adjust the position and size of the image within the final document,
-e.g.~Figure \ref{fig:frog-picture} is a frog.
+\textbf{n.b.}, the re-coding process results in ``normal'' patients
+being assigned a value of 1 and cancer patients a 0.
 
-Alternatively, you can use the standard markdown syntax like so:
+Now, we are ready to analyze the effects of cancer status on DNA
+methylation using this data set. We proceed as follows with a targeted
+minimum loss-based estimate of the Average Treatment Effect.
 
-\begin{figure}
-\centering
-\includegraphics[width=0.25000\textwidth]{frog.jpg}
-\caption{This is a smaller version of the same picture, inserted using
-the standard markdown syntax}
-\end{figure}
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{suppressMessages}\NormalTok{(}
+\NormalTok{  methyvim_cancer_ate <-}\StringTok{ }\KeywordTok{methyvim}\NormalTok{(}\DataTypeTok{data_grs =}\NormalTok{ grs, }\DataTypeTok{var_int =}\NormalTok{ var_int,}
+                                  \DataTypeTok{vim =} \StringTok{"ate"}\NormalTok{, }\DataTypeTok{type =} \StringTok{"Beta"}\NormalTok{, }\DataTypeTok{filter =} \StringTok{"limma"}\NormalTok{,}
+                                  \DataTypeTok{filter_cutoff =} \FloatTok{0.20}\NormalTok{, }\DataTypeTok{obs_per_covar =} \DecValTok{2}\NormalTok{,}
+                                  \DataTypeTok{parallel =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{sites_comp =} \DecValTok{125}\NormalTok{,}
+                                  \DataTypeTok{tmle_type =} \StringTok{"glm"}
+\NormalTok{                                 )}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-Please give figures appropriate filenames, e.g.: figure1.pdf,
-figure2.png.
+\begin{verbatim}
+## Warning in set_parallel(parallel = parallel, future_param = future_param, : Sequential evaluation is strongly discouraged. 
+##  Proceed with caution.
+\end{verbatim}
 
-Figure legends should briefly describe the key messages of the figure
-such that the figure can stand alone from the main text. However, all
-figures should also be discussed in the article text. Each legend should
-have a concise title of no more than 15 words. The legend itself should
-be succinct, while still explaining all symbols and abbreviations. Avoid
-lengthy descriptions of methods.
+Note that we set the \texttt{obs\_per\_covar} argument to a relatively
+low value (2, where the recommended default is 20) for the purposes of
+this example as the sample size is only 10. We do this only to exemplify
+the estimation procedure and it is important to point out that such low
+values for \texttt{obs\_per\_covar} will compromise the quality of
+inference obtained because this setting directly affects the definition
+of the target parameter.
 
-For any figures reproduced from another publication (as long as
-appropriate permission has been obtained from the copyright holder
----see under the heading `Submission'), please include a line in the
-legend to state that: `This figure has been reproduced with kind
-permission from {[}include original publication citation{]}'.
+Further, note that here we apply the \texttt{glm} flavor of the
+\texttt{tmle\_type} argument, which produces faster results by fitting
+models for the propensity score and outcome regressions using a limited
+number of parametric models. By contrast, the \texttt{sl} (for ``Super
+Learning'') flavor fits these two regressions using highly nonparametric
+and data-adaptive procedures (i.e., via machine learning). Obtaining the
+estimates via GLMs results in each of the regression steps being less
+robust than if nonparametric regressions were used.
 
-\subsection{Mathematics}\label{mathematics}
+We can view a table of results by examining the \texttt{vim} slot of the
+produced \texttt{methytmle} object.
 
-You can use LaTeX syntax to typeset mathematical expressions. Let
-\(X_1, X_2, \ldots, X_n\) be a sequence of independent and identically
-distributed random variables with \(\text{E}[X_i] = \mu\) and
-\(\text{Var}[X_i] = \sigma^2 < \infty\), and let
-\[S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i\] denote their mean. Then as \(n\)
-approaches infinity, the random variables \(\sqrt{n}(S_n - \mu)\)
-converge in distribution to a normal \(\mathcal{N}(0, \sigma^2)\).
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{head}\NormalTok{(}\KeywordTok{slot}\NormalTok{(methyvim_cancer_ate, }\StringTok{"vim"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
-To number and refer to equations, put them in the equation environments
-and assign labels to them, as for Equation \eqref{eq:binom}.
+\begin{verbatim}
+##            lowerCI_ATE      est_ATE   upperCI_ATE      Var_ATE
+## cg14008030 -0.11956597 -0.031415962  0.0567340489 2.022705e-03
+## cg20253340 -0.08850637 -0.058866142 -0.0292259165 2.286919e-04
+## cg21870274 -0.09499057 -0.029118982  0.0367526091 1.129495e-03
+## cg17308840 -0.04626018 -0.007152452  0.0319552773 3.981191e-04
+## cg00645010 -0.02677328 -0.013465554 -0.0001578256 4.609945e-05
+## cg27534567  0.06745648  0.115711854  0.1639672225 6.061486e-04
+##                    pval n_neighbors_all n_neighbors_w max_corr_w
+## cg14008030 4.848468e-01               0             0         NA
+## cg20253340 9.917426e-05               0             0         NA
+## cg21870274 3.862537e-01               2             1  0.9443580
+## cg17308840 7.199943e-01               2             1  0.9443580
+## cg00645010 4.734007e-02               2             2  0.5236810
+## cg27534567 2.602936e-06               1             0  0.9362968
+\end{verbatim}
 
-\begin{equation}
-  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
-  \label{eq:binom}
-\end{equation}
+Finally, we may compute FDR-corrected p-values, by applying a modified
+procedure for controlling the False Discovery Rate for multi-stage
+analyses (FDR-MSA) (Tuglus and van der Laan 2009). We do this by simply
+applying the \texttt{fdr\_msa} function.
 
-\subsection{Lists}\label{lists}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fdr_p <-}\StringTok{ }\KeywordTok{fdr_msa}\NormalTok{(}\DataTypeTok{pvals =} \KeywordTok{slot}\NormalTok{(methyvim_cancer_ate, }\StringTok{"vim"}\NormalTok{)}\OperatorTok{$}\NormalTok{pval,}
+                 \DataTypeTok{total_obs =} \KeywordTok{nrow}\NormalTok{(methyvim_cancer_ate))}
+\end{Highlighting}
+\end{Shaded}
 
-You can make ordered lists
+Having explored the results of our analysis numerically, we now proceed
+to use the visualization tools provided with the \texttt{methyvim} R
+package to further enhance our understanding of the results.
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Like this,
-\item
-  and like this.
-\end{enumerate}
+\paragraph{Visualization of Results}\label{visualization-of-results}
 
-or bullet points
+While making allowance for users to explore the full set of results
+produced by the estimation procedure (by way of exposing these directly
+to the user), the \texttt{methyvim} package also provides \emph{three}
+(3) visualization utilities that produce plots commonly used in
+examining the results of differential methylation analyses.
 
-\begin{itemize}
-\item
-  Like this,
-\item
-  and like this.
+A simple call to \texttt{plot} produces side-by-side histograms of the
+raw p-values computed as part of the estimation process and the
+corrected p-values obtained from using the FDR-MSA procedure.
 
-  \begin{itemize}
-  \item
-    Use indentation
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(methyvim_cancer_ate, }\DataTypeTok{type =} \StringTok{"raw_pvals"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-    \begin{itemize}
-    \item
-      for sub-items
-    \end{itemize}
-  \end{itemize}
-\end{itemize}
+\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-raw-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(methyvim_cancer_ate, }\DataTypeTok{type =} \StringTok{"fdr_pvals"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-fdr-1.pdf}
+
+\textbf{Remark:} The plots displayed above may also be generated as
+side-by-side histograms in a single plot object. This is the default for
+the \texttt{plot} method and may easily be invoked by specifying no
+additional arguments to the \texttt{plot} function, unlike in the above.
+
+While histograms of the p-values may be generally useful in inspecting
+the results of the estimation procedure, a more common plot used in
+examining the results of differential methylation procedures is the
+volcano plot, which plots the parameter estimate along the x-axis and
+\(-\text{log}_{10}(\text{p-value})\) along the y-axis. We implement such
+a plot in the \texttt{methyvolc} function:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{methyvolc}\NormalTok{(methyvim_cancer_ate)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-volcano-1.pdf}
+
+The purpose of such a plot is to ensure that very low (possibly
+statistically significant) p-values do not arise from cases of low
+variance. This appears to be the case in the plot above (notice that
+most parameter estimates are near zero, even in cases where the raw
+p-values are quite low).
+
+Yet another popular plot for visualizing effects in such settings is the
+heatmap, which plots estimates of the raw methylation effects (as
+measured by the assay) across subjects using a heat gradient. We
+implement this in the \texttt{methyheat} function:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{methyheat}\NormalTok{(methyvim_cancer_ate)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1.pdf}
+
+Invoking \texttt{methyheat} in this manner produces a plot of the top
+sites (\(25\), by default) based on the raw p-value, using the raw
+methylation measures in the plot. This uses the exceptional
+\texttt{superheat} R package ({\textbf{???}}).
+
+\section{Summary}\label{summary}
+
+Here we introduce the R package \texttt{methyvim}, an algorithm for
+differential methylation analysis that moves beyond linear models.
+Additionally, this estimation procedure produces straightforward
+statistical inference. Inspired by concepts of counterfactual effects
+from the causal inference literature and machine learning algorithms, we
+provide a computationally efficient technique for obtaining targeted
+estimates of nonparametric variable importance measures for a
+pre-screened set of CpGs, controlling for the False Discovery Rate as if
+all sites were tested, and the estimation framework is built on targeted
+minimum loss-based estimation. By taking advantage of the prediction
+power of machine learning algorithms, this methodology targets the
+individual importance of each CpG site, providing a fully nonparametric
+statistical model for the analysis of differential methylation.
+
+\section{Software availability}\label{software-availability}
+
+Latest source code (development version):
+\url{https://github.com/nhejazi/methyvim}
+
+Bioconductor (stable release):
+\url{https://bioconductor.org/packages/methyvim}
+
+Archived source code as at time of publication:
+\url{https://github.com/nhejazi/methyvim}
+
+Software license: The MIT License, copyright Nima S. Hejazi
+
+\section{Author contributions}\label{author-contributions}
+
+NH designed and implemented the software package, applied the tool to
+the use cases, and co-drafted the present article. RP helped in
+designing the software and co-drafted the present article. AH and ML
+serve as the advisors for the development of this software and general
+statistical algorithm.
+
+\section{Competing interests}\label{competing-interests}
+
+No competing interests were disclosed.
+
+\section{Grant information}\label{grant-information}
+
+NH was supported in part by the National Library Of Medicine of the
+National Institutes of Health under Award Number T32LM012417, by
+{[}SUPERFUND GRANT{]}, and by {[}NINA HOLLAND GRANT{]}. RP was supported
+by {[}SUPERFUND GRANT{]}. The content of this work is solely the
+responsibility of the authors and does not necessarily represent the
+official views of the various funding sources and agencies.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\hypertarget{ref-benjamini1995controlling}{}
+Benjamini, Yoav, and Yosef Hochberg. 1995. ``Controlling the False
+Discovery Rate: A Practical and Powerful Approach to Multiple Testing.''
+\emph{Journal of the Royal Statistical Society. Series B
+(Methodological)}. JSTOR, 289--300.
+
+\hypertarget{ref-chambaz2012estimation}{}
+Chambaz, Antoine, Pierre Neuvial, and Mark J van der Laan. 2012.
+``Estimation of a Non-Parametric Variable Importance Measure of a
+Continuous Exposure.'' \emph{Electronic Journal of Statistics} 6. NIH
+Public Access: 1059.
+
+\hypertarget{ref-hernan2018causal}{}
+Hernan, Miguel A, and James M Robins. 2018, forthcoming. \emph{Causal
+Inference}. Chapman \& Hall/Crc Texts in Statistical Science. Taylor \&
+Francis.
+
+\hypertarget{ref-libbrecht2015machine}{}
+Libbrecht, Maxwell W, and William Stafford Noble. 2015. ``Machine
+Learning in Genetics and Genomics.'' \emph{Nature Reviews. Genetics} 16
+(6). NIH Public Access: 321.
+
+\hypertarget{ref-pearl2009causality}{}
+Pearl, Judea. 2009. \emph{Causality: Models, Reasoning, and Inference}.
+Cambridge University Press.
+
+\hypertarget{ref-robertson2005dna}{}
+Robertson, Keith D. 2005. ``DNA Methylation and Human Disease.''
+\emph{Nature Reviews. Genetics} 6 (8). Nature Publishing Group: 597.
+
+\hypertarget{ref-tuglus2008targeted}{}
+Tuglus, Catherine, and Mark J van der Laan. 2008. ``Targeted Methods for
+Biomarker Discovery, the Search for a Standard.'' bepress.
+
+\hypertarget{ref-tuglus2009modified}{}
+Tuglus, Catherine, and Mark J. van der Laan. 2009. ``Modified FDR
+Controlling Procedure for Multi-Stage Analyses.'' \emph{Statistical
+Applications in Genetics and Molecular Biology} 8 (1). Walter de
+Gruyter: 1--15.
+doi:\href{https://doi.org/10.2202/1544-6115.1397}{10.2202/1544-6115.1397}.
+
+\hypertarget{ref-vdl2006statistical}{}
+van der Laan, Mark J. 2006. ``Statistical Inference for Variable
+Importance.'' \emph{The International Journal of Biostatistics} 2 (1).
+
+\hypertarget{ref-vdl2011targeted}{}
+van der Laan, Mark J, and Sherri Rose. 2011. \emph{Targeted Learning:
+Causal Inference for Observational and Experimental Data}. Springer
+Science \& Business Media.
+
+\hypertarget{ref-vdl2017targeted}{}
+---------. 2017. \emph{Targeted Learning in Data Science: Causal
+Inference for Complex Longitudinal Studies}. Springer Science \&
+Business Media.
+
+\hypertarget{ref-vdl2006targeted}{}
+van der Laan, Mark J, and Daniel Rubin. 2006. ``Targeted Maximum
+Likelihood Learning.'' \emph{The International Journal of Biostatistics}
+2 (1).
 

### 2018-01-30 README.Rmd
minor admin updates
add funding notices

@@
   rmarkdown::github_document
-bibliography: README-refs.bib
+bibliography: "`r here::here('README-refs.bib')`"
 ---
@@
 
-The development of this software was supported in part through a grant from the
-National Library of Medicine of the NIH (T32 LM012417).
+The development of this software was supported in part through grants from the
+National Institutes of Health: [T32 LM012417](https://projectreporter.nih.gov/project_info_description.cfm?aid=9248418&icde=37849831&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC&pball=), [R01 ES021369](https://projectreporter.nih.gov/project_info_description.cfm?aid=9210551&icde=37849782&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC&pball=), and [P42 ES004705-29](https://projectreporter.nih.gov/project_info_details.cfm?aid=9260357&map=y).
 

### 2018-01-30 README.Rmd
fix funding notices

--- 2017-11-29 README.Rmd
@@
 
-<!--
-Note about shrinkage of influence curves, adapting @smyth2004linear.
-More exposition here....
--->
-
 ---
@@
 
-<!--
 ## Example
 
-This is a basic example which shows you how to solve a common problem:
+For details on how to best use `methyvim` to perform data-adaptive estimation
+and obtain nonparametric inference in the context of differential methylation
+analyses, please consult the most recent [package vignette](http://bioconductor.org/packages/release/bioc/vignettes/methyvim/inst/doc/using_methyvim.html)
+available through the [Bioconductor
+project](https://bioconductor.org/packages/methvim).
 
-```{r example}
-## basic example code
-```
--->
+---
 
@@
 
-        @article{hejazi2017methyvim,
+        @article{hejazi2018methyvim,
           doi = {},
           url = {},
-          year  = {2017},
+          year  = {2018},
           month = {},
@@
           volume = {},
-          author = {Hejazi, Nima S and Hubbard, Alan E and {van der Laan}, Mark
-            J},
+          author = {Hejazi, Nima S and Phillips, Rachael V and Hubbard, Alan E
+            and {van der Laan}, Mark J},
           title = {methyvim: Targeted and model-free differential methylation
@@
 
-The development of this software was supported in part through a grant from the
-National Library of Medicine of the NIH (T32 LM012417).
+The development of this software was supported in part through grants from the
+National Institutes of Health: [T32 LM012417-02](https://projectreporter.nih.gov/project_info_description.cfm?aid=9248418&icde=37849831&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC&pball=), [R01 ES021369-05](https://projectreporter.nih.gov/project_info_description.cfm?aid=9210551&icde=37849782&ddparam=&ddvalue=&ddsub=&cr=1&csb=default&cs=ASC&pball=), and [P42 ES004705-29](https://projectreporter.nih.gov/project_info_details.cfm?aid=9260357&map=y).
 
@@
 
-&copy; 2017 [Nima S. Hejazi](http://nimahejazi.org)
+&copy; 2017-2018 [Nima S. Hejazi](https://nimahejazi.org)
 

### 2018-04-16 R/fdr_msa.R
update pkgdown site; re-style; ggsci for colors

@@ -1,0 +1,32 @@
+#' FDR-MSA correction
+#'
+#' Modified FDR Controlling Procedure for Multi-Stage Analyses (MJ van der Laan
+#' and C Tuglus, 2009, <doi:10.2202/1544-6115.1397>)
+#'
+#' @param pvals Numeric vector containing the p-values that result from any
+#'  chosen statistical hypothesis testing procedure.
+#' @param total_obs Numeric indicating the total number of observations that
+#'  would have been available for testing prior to the selection procedure
+#'  employed in the multi-stage analysis performed.
+#'
+#' @return A \code{numeric} vector of corrected p-values, controlling the False
+#'  Discovery Rate, using the method of Tuglus and van der Laan.
+#'
+#' @importFrom stats p.adjust
+#'
+#' @export
+#'
+#' @examples
+#' g <- 1e4
+#' n <- 1e2
+#' p <- abs(rnorm(n, mean = 1e-8, sd = 1e-2))
+#' # treating the vector p as one of p-values, FDR-MSA may be applied
+#' fdr_p <- fdr_msa(pvals = p, total_obs = g)
+#
+fdr_msa <- function(pvals, total_obs) {
+  pvals_not_tested <- rep(1, total_obs - length(pvals))
+  pvals_all <- c(pvals, pvals_not_tested)
+  fdr_adj <- stats::p.adjust(pvals_all, method = "fdr")
+  fdr_out <- fdr_adj[seq_along(pvals)]
+  return(fdr_out)
+}

### 2018-04-16 R/force_positivity.R

@@ -1,0 +1,55 @@
+#' Enforce the Assumption of Positivity
+#'
+#' Discretize continuous variables in the adjustment set (W) of a TMLE procedure
+#' in order to avoid practical violations of the assumption of positivity.
+#' Discretizes the numeric columns of an input matrix such that the newly
+#' created levels of each variable individually contain at least a specified
+#' mass when considering each level against levels of the treatment variable.
+#' INTERNAL USE ONLY.
+#'
+#' @param A Numeric giving the levels of the (discretized) treatment variable.
+#' @param W Data.Frame or Matrix containing the covariates in the adjustment set
+#'  to be discretized against the levels of the treatment variable.
+#' @param pos_min Numeric indicating the minimum mass (as a proportion) of the
+#'  observations to be included in any cell of the table composed of the levels
+#'  of the treatment against levels of an adjustment covariate.
+#' @param q_init Numeric indicating the initial number of levels to discretize a
+#'  given adjustment variable into. This defaults to quantiles.
+#'
+#' @return A numeric vector with the adjustment variables re-coded into discrete
+#'  levels respecting the minimum mass requested in each table comparing levels
+#'  of the treatment against levels of an adjustment covariate.
+#'
+#' @keywords internal
+#'
+#' @importFrom gtools quantcut
+#
+force_positivity <- function(A, W, pos_min = 0.1, q_init = 10) {
+  stopifnot(length(A) == nrow(W))
+
+  if (class(W) != "data.frame") W <- as.data.frame(W) # cover use of "ncol"
+  out_w <- NULL # concatenate W columnwise as we discretize each covar below
+
+  for (obs_w in seq_len(ncol(W))) {
+    in_w <- as.numeric(W[, obs_w])
+    discr_w <- as.numeric(as.factor(gtools::quantcut(x = in_w, q = q_init)))
+    check <- sum((table(A, discr_w) / length(A)) < pos_min)
+    next_guess_q <- q_init
+    while (check > 0) {
+      next_guess_q <- (next_guess_q - 1)
+      discr_w <- as.numeric(as.factor(gtools::quantcut(
+        x = in_w,
+        q = next_guess_q
+      )))
+      check <- sum((table(A, discr_w) / length(A)) < pos_min)
+    }
+    out_w <- cbind(out_w, discr_w)
+  }
+  out <- as.data.frame(out_w)
+  colnames(out) <- colnames(W)
[…]

### 2018-04-16 R/methyvim.R

@@ -78,12 +78,12 @@ utils::globalVariables(c("colData<-"))
 #'  Current choices include \code{tmle}, \code{drtmle}, and \code{tmle.npvi}.
 #'
 #' @return An object of class \code{methytmle}, with all unique slots filled in,
-#'         in particular, including indices of CpG sites that pass screening,
-#'         cluster of neighboring CpG sites, and a matrix of the results of the
-#'         estimation procedure performed for the given variable importance
-#'         measure. Optionally, estimates of the propensity score and outcome
-#'         regressions, as well as the original data rotated into influence
-#'         curve space may be returned, if so requested.
+#'  in particular, including indices of CpG sites that pass screening, cluster
+#'  of neighboring CpG sites, and a matrix of the results of the estimation
+#'  procedure performed for the given variable importance measure. Optionally,
+#'  estimates of the propensity score and outcome regressions, as well as the
+#'  original data rotated into influence curve space may be returned, if so
+#'  requested.
 #'
 #' @importFrom SummarizedExperiment colData
 #' @importFrom BiocParallel bplapply
@@ -122,11 +122,14 @@ methyvim <- function(data_grs,
                      return_ic = FALSE,
                      shrink_ic = FALSE,
                      tmle_type = c("glm", "sl"),
-                     tmle_args = list(family = "binomial",
-                                      g_lib = NULL, Q_lib = NULL,
-                                      npvi_cutoff = 0.25, npvi_descr = NULL),
-                     tmle_backend = c("tmle", "drtmle", "tmle.npvi")
-                    ) {
+                     tmle_args = list(
+                       family = "binomial",
+                       g_lib = c("SL.mean", "SL.glm", "SL.bayesglm", "SL.gam"),
+                       Q_lib = c("SL.mean", "SL.glm", "SL.gam", "SL.earth"),
+                       npvi_cutoff = 0.25,
+                       npvi_descr = NULL
+                     ),
+                     tmle_backend = c("tmle", "drtmle", "tmle.npvi")) {
   # ============================================================================
   # catch input for user convenience and check input types where possible
   # ============================================================================
@@ -139,20 +142,11 @@ methyvim <- function(data_grs,
   type <- match.arg(type)
   filter <- match.arg(filter)
   tmle_type <- match.arg(tmle_type)
+  tmle_backend <- match.arg(tmle_backend)
 
   # check that variable of interest is the correct length
   if (length(var_int) != ncol(data_grs)) {
     stop("Variable of interest is not the same size as number of observations.")
-  }
-
[…]

### 2018-04-16 R/plots.R

@@ -1,30 +1,29 @@
-utils::globalVariables(c("..count..", "color", "log_pval", "param pval",
-                         "param", "pval"))
+utils::globalVariables(c(
+  "..count..", "color", "log_pval", "param pval",
+  "param", "pval"
+))
 
 #' Plot p-values of methytmle objects
 #'
 #' @param x Object of class \code{methytmle} as produced by an appropriate call
-#'        to \code{methyvim}.
+#'  to \code{methyvim}.
 #' @param ... Additional arguments passed \code{plot} as necessary.
 #' @param type The type of plot to build: one of side-by-side histograms (type
-#'        "both") comparing raw p-values to FDR-adjusted p-values (using the
-#'        FDR-MSA correction) or either of these two histogram separately. Set
-#'        this argument to "raw_pvals" for a histogram of the raw p-values, and
-#'        to "fdr_pvals" for a histogram of the FDR-corrected p-values.
+#'  "both") comparing raw p-values to FDR-adjusted p-values (using the FDR-MSA
+#'  correction) or either of these two histogram separately. Set this argument
+#'  to "raw_pvals" for a histogram of the raw p-values, and to "fdr_pvals" for
+#'  a histogram of the FDR-corrected p-values.
 #'
 #' @return Object of class \code{ggplot} containing a histogram or side-by-side
-#'         histograms of the raw (marginal) and corrected p-values, with the
-#'         latter computed automatically using the method of Tuglus and van der
-#'         Laan.
+#'  histograms of the raw (marginal) and corrected p-values, with the latter
+#'  computed automatically using the method of Tuglus and van der Laan.
 #'
 #' @importFrom magrittr "%>%"
 #' @importFrom dplyr select slice arrange transmute
-#' @importFrom ggplot2 ggplot aes geom_point geom_histogram xlab ylab ggtitle
-#'             scale_colour_manual scale_fill_gradientn guides guide_legend
-#'             theme_minimal
-#' @importFrom wesanderson wes_palette
+#' @importFrom ggplot2 ggplot aes geom_histogram xlab ylab ggtitle guides
+#'  guide_legend theme_bw
+#' @importFrom ggsci scale_fill_gsea
 #' @importFrom gridExtra grid.arrange
-#' @importFrom superheat superheat
 #'
 #' @export
 #'
[…]

### 2018-04-16 R/screen_limma.R

@@ -29,7 +29,6 @@
 #' @importFrom minfi getBeta getM
 #
 limma_screen <- function(methytmle, var_int, type, cutoff = 0.05) {
-
   stopifnot(class(methytmle) == "methytmle")
 
   # setup design matrix

### 2018-04-16 R/tmle_classic.R

@@ -58,8 +58,7 @@ methyvim_tmle <- function(target_site,
                           g_lib = c("SL.mean", "SL.glm", "SL.glm.interaction"),
                           Q_lib = c("SL.mean", "SL.glm", "SL.gam", "SL.earth"),
                           family = c("gaussian", "binomial"),
-                          return_ic = FALSE
-                         ) {
+                          return_ic = FALSE) {
   ### check arguments where possible
   type <- match.arg(type)
   target_param <- match.arg(target_param)
@@ -67,7 +66,7 @@ methyvim_tmle <- function(target_site,
 
   ### get neighboring site
   in_cluster <- which(methytmle_screened@clusters %in%
-                      methytmle_screened@clusters[target_site])
+    methytmle_screened@clusters[target_site])
 
   ### remove target site from the set of neighbors
   only_neighbors <- in_cluster[in_cluster != target_site]
@@ -147,14 +146,15 @@ methyvim_tmle <- function(target_site,
     max_corr_w <- max(cor(y, t(w)))
 
     # compute the ATE
-    out <- tmle::tmle(Y = as.numeric(y_star),
-                      A = as.numeric(var_of_interest),
-                      W = as.data.frame(w_pos),
-                      Q.SL.library = Q_lib,
-                      g.SL.library = g_lib,
-                      family = family,
-                      verbose = FALSE
-                     )
+    out <- tmle::tmle(
+      Y = as.numeric(y_star),
+      A = as.numeric(var_of_interest),
+      W = as.data.frame(w_pos),
+      Q.SL.library = Q_lib,
+      g.SL.library = g_lib,
+      family = family,
+      verbose = FALSE
+    )
   } else {
     n_neighbors_total <- 0
     n_neighbors_reduced <- 0
@@ -164,22 +164,23 @@ methyvim_tmle <- function(target_site,
     w_int <- as.data.frame(rep(1, length(var_of_interest)))
 
     # compute the ATE
-    out <- tmle::tmle(Y = as.numeric(y_star),
-                      A = as.numeric(var_of_interest),
-                      W = as.data.frame(w_int),
-                      Q.SL.library = Q_lib,
-                      g.SL.library = g_lib,
-                      family = family,
[…]

### 2018-04-16 R/zzz.R

@@ -1,6 +1,8 @@
 .onAttach <- function(...) {
-  packageStartupMessage(paste0("methyvim v",
-                               utils::packageDescription("methyvim")$Version,
-                               ": Targeted Variable Importance for ",
-                               "Differential Methylation Analysis"))
+  packageStartupMessage(paste0(
+    "methyvim v",
+    utils::packageDescription("methyvim")$Version,
+    ": Targeted Variable Importance for ",
+    "Differential Methylation Analysis"
+  ))
 }

### 2018-04-16 tests/testthat/test_methyvim.R

@@ -11,18 +11,20 @@ var_int <- colData(grsExample)[, 1]
 
 # run TMLE procedure for the ATE parameter over M-values with Limma filtering
 methyvim_out_ate <- suppressWarnings(
-  methyvim(data_grs = grsExample, var_int = var_int, vim = "ate",
-           type = "Mval", filter = "limma", filter_cutoff = 0.05,
-           sites_comp = 3, parallel = FALSE, tmle_type = "sl"
-          )
+  methyvim(
+    data_grs = grsExample, var_int = var_int, vim = "ate",
+    type = "Mval", filter = "limma", filter_cutoff = 0.05,
+    sites_comp = 3, parallel = FALSE, tmle_type = "sl"
+  )
 )
 
 # run TMLE procedure for the RR parameter over M-values with Limma filtering
 methyvim_out_rr <- suppressWarnings(
-  methyvim(data_grs = grsExample, var_int = var_int, vim = "rr",
-           type = "Mval", filter = "limma", filter_cutoff = 0.05,
-           sites_comp = 3, parallel = FALSE, tmle_type = "sl"
-          )
+  methyvim(
+    data_grs = grsExample, var_int = var_int, vim = "rr",
+    type = "Mval", filter = "limma", filter_cutoff = 0.05,
+    sites_comp = 3, parallel = FALSE, tmle_type = "sl"
+  )
 )
 
 
@@ -65,7 +67,8 @@ test_that("Cluster IDs are of appropriate length and unique length", {
 })
 
 test_that("Variable importance results within tolerance for the RR parameter", {
-  expect_lt(sum(range(methyvim_out_rr@vim$pval) - c(0.2281579, 0.4707649)),
-            0.01)
+  expect_lt(
+    sum(range(methyvim_out_rr@vim$pval) - c(0.2281579, 0.4707649)),
+    0.01
+  )
 })
-

### 2018-04-16 tests/testthat/test_set_parallel.R

@@ -56,4 +56,3 @@ test_that("sets up transparent future plan when so asked", {
   set_parallel(parallel = TRUE, future_param = "transparent")
   expect_true("transparent" %in% class(plan()))
 })
-

### 2018-04-16 README.Rmd
Minor fixes (#32)
* update readme and site

* update pkgdown site; re-style; ggsci for colors

* update pkgdown; tweak CI params

@@
 
-__Author:__ [Nima Hejazi](http://nimahejazi.org)
+__Author:__ [Nima Hejazi](https://nimahejazi.org)
 
@@
 
-For standard use, install from [Bioconductor](https://bioconductor.org):
+For standard use, install from
+[Bioconductor](https://bioconductor.org/packages/methyvim):
 
@@
 
-For details on how to best use `methyvim` to perform data-adaptive estimation
-and obtain nonparametric inference in the context of differential methylation
-analyses, please consult the most recent [package vignette](http://bioconductor.org/packages/release/bioc/vignettes/methyvim/inst/doc/using_methyvim.html)
+For details on how to best use the `methyvim` R package, please consult the most
+recent [package
+vignette](http://bioconductor.org/packages/release/bioc/vignettes/methyvim/inst/doc/using_methyvim.html)
 available through the [Bioconductor
-project](https://bioconductor.org/packages/methvim).
+Project](https://bioconductor.org/packages/methvim).
 
@@
 
-It is our hope that `methyvim` will grow to be widely adopted as a tool for the
-nonparametric assessment of variable importance in studies of differential
-methylation. To that end, contributions are very welcome, though we ask that
-interested contributors consult our [`contribution
-guidelines`](https://github.com/nhejazi/methyvim/blob/master/CONTRIBUTING.md)
+Contributions are very welcome. Interested contributors should consult our
+[contribution
+guidelines](https://github.com/nhejazi/methyvim/blob/master/CONTRIBUTING.md)
 prior to submitting a pull request.

### 2018-04-16 R/methytmle.R

@@ -16,14 +16,15 @@
 #' methy_tmle <- .methytmle(grsExample)
 #
 .methytmle <- methods::setClass(
-       Class = "methytmle",
-       slots = list(call = "call",
-                    screen_ind = "numeric",
-                    clusters = "numeric",
-                    var_int = "numeric",
-                    param = "character",
-                    vim = "data.frame",
-                    ic = "matrix"),
-       contains = "GenomicRatioSet"
+  Class = "methytmle",
+  slots = list(
+    call = "call",
+    screen_ind = "numeric",
+    clusters = "numeric",
+    var_int = "numeric",
+    param = "character",
+    vim = "data.frame",
+    ic = "matrix"
+  ),
+  contains = "GenomicRatioSet"
 )
-

### 2018-04-16 R/utils.R

@@ -1,77 +1,3 @@
-#' CpG Neighborhoods from Genomic Distance
-#'
-#' Clustering of CpG sites to define CpG neighborhoods based on distance (bp).
-#' INTERNAL USE ONLY.
-#'
-#' @param methytmle Object of class \code{methytmle} produced from an object of
-#'        class \code{GenomicRatioSet}, but containing extra slots.
-#' @param window_size Numeric giving the number of base pairs used to define
-#'        neighborhoods along the genome. CpG sites within this distance (bp) of
-#'        one another are denoted as neighbors. Chromosome boundaries and other
-#'        biological constraints are respected (see the documentation available
-#'        for \code{bumphunter::clusterMaker} for details).
-#'
-#' @return An object of class \code{methytmle} with the "clusters" slot filled
-#'        in. The "clusters" slot contains a \code{numeric} vector as long as
-#'        the number of CpG sites. Each entry in the vector is a neighborhood
-#'        assignment used in the estimation procedure.
-#'
-#' @keywords internal
-#'
-#' @importFrom bumphunter clusterMaker
-#' @importFrom SummarizedExperiment rowRanges
-#' @importFrom GenomeInfoDb seqnames
-#' @importFrom BiocGenerics start
-#' @importFrom IRanges ranges
-#
-cluster_sites <- function(methytmle, window_size = 1000) {
-  gr <- SummarizedExperiment::rowRanges(methytmle)
-  pos <- BiocGenerics::start(IRanges::ranges(gr))
-  clusters <- bumphunter::clusterMaker(chr = GenomeInfoDb::seqnames(gr),
-                                       pos = pos,
-                                       assumeSorted = FALSE,
-                                       maxGap = window_size)
-  methytmle@clusters <- as.numeric(clusters)
-  return(methytmle)
-}
-
-################################################################################
-
-#' FDR-MSA correction
-#'
-#' Modified FDR Controlling Procedure for Multi-Stage Analyses (MJ van der Laan
-#' and C Tuglus, 2009, <doi:10.2202/1544-6115.1397>)
-#'
-#' @param pvals Numeric vector containing the p-values that result from any
-#'        chosen statistical hypothesis testing procedure.
-#' @param total_obs Numeric indicating the total number of observations that
-#'        would have been available for testing prior to the selection procedure
[…]

### 2018-04-16 tests/testthat.R

@@ -3,4 +3,3 @@ library(methyvim)
 
 set.seed(43719)
 test_check("methyvim")
-

### 2018-04-16 tests/testthat/test_screen_limma.R

@@ -5,10 +5,14 @@ library(methyvimData)
 data(grsExample)
 var_int <- colData(grsExample)[, 1]
 methytmle <- .methytmle(grsExample)
-screened <- methyvim:::limma_screen(methytmle, var_int = var_int,
-                                    type = "Mval")
-screened_betas <- methyvim:::limma_screen(methytmle, var_int = var_int,
-                                          type = "Beta")
+screened <- methyvim:::limma_screen(methytmle,
+  var_int = var_int,
+  type = "Mval"
+)
+screened_betas <- methyvim:::limma_screen(methytmle,
+  var_int = var_int,
+  type = "Beta"
+)
 
 
 # check the screen_ind slot
@@ -41,4 +45,3 @@ test_that("classes of screened object and of unscreened object match", {
 test_that("class types of screened object and of unscreened object match", {
   expect_equivalent(typeof(screened), typeof(methytmle))
 })
-

### 2018-04-16 tests/testthat/test_tmle_classic.R

@@ -11,92 +11,111 @@ clustered <- methyvim:::cluster_sites(screened)
 
 
 # consistentcy of output for ATE and RR over target site with M-values
-methyvim_mvals_ate <- methyvim:::methyvim_tmle(target_site = 10,
-                                               methytmle_screened = clustered,
-                                               var_of_interest = var_int,
-                                               type = "Mval",
-                                               corr = 0.75,
-                                               obs_per_covar = 20,
-                                               target_param = "ate",
-                                               family = "binomial",
-                                               return_ic = FALSE
-                                              )
+methyvim_mvals_ate <- methyvim:::methyvim_tmle(
+  target_site = 10,
+  methytmle_screened = clustered,
+  var_of_interest = var_int,
+  type = "Mval",
+  corr = 0.75,
+  obs_per_covar = 20,
+  target_param = "ate",
+  family = "binomial",
+  return_ic = FALSE
+)
 
 test_that("ATE procedure with M-values is consistent for target site", {
-  expect_equal(methyvim_mvals_ate,
-               c(-0.38903184, -0.17995147, 0.02912890, 0.01137927, 0.09161596,
-                 0.000000000, 0.000000000, NA))
+  expect_equal(
+    methyvim_mvals_ate,
+    c(
+      -0.38903184, -0.17995147, 0.02912890, 0.01137927, 0.09161596,
+      0.000000000, 0.000000000, NA
+    )
+  )
 })
 
-methyvim_mvals_rr <- methyvim:::methyvim_tmle(target_site = 10,
-                                              methytmle_screened = clustered,
-                                              var_of_interest = var_int,
-                                              type = "Mval",
-                                              corr = 0.75,
-                                              obs_per_covar = 20,
-                                              target_param = "rr",
-                                              family = "binomial",
-                                              return_ic = FALSE
-                                             )
+methyvim_mvals_rr <- methyvim:::methyvim_tmle(
+  target_site = 10,
+  methytmle_screened = clustered,
+  var_of_interest = var_int,
+  type = "Mval",
+  corr = 0.75,
+  obs_per_covar = 20,
+  target_param = "rr",
+  family = "binomial",
+  return_ic = FALSE
+)
 
 test_that("RR procedure with M-values is consistent for target site", {
-  expect_equal(methyvim_mvals_rr,
[…]

### 2018-04-16 paper/paper_BiocF1000.Rmd
F1000 (#34)
* use minfi data

* add paper refs

* started to add a bit to the introduction &  moved the text from the repository to the appropriate sections in the article. I will use the text from the repo as a rough guide to draft something more cohesive

* revisions/updates to F1000 manuscript (#28)

* minor vignette additions/corrections

* almost done with f1000, minor vignette corrections

* adding the Summary and final touches

* a few more minor things

* fixing merge conflict

* minor formatting

* suggestion: convert betas to mvals in the limma screenas the mvals have far more centra

@@
   - name: Nima S. Hejazi
-    affiliation: Division of Biostatistics, University of California, Berkeley
+    affiliation:
+      - Group in Biostatistics, University of California, Berkeley
+      - Center for Computational Biology, University of California, Berkeley
     orcid: 0000-0002-7127-2789
+    email: nhejazi@berkeley.edu
   - name: Rachael V. Phillips
-    affiliation: Division of Biostatistics, University of California, Berkeley
+    affiliation:
+      - Group in Biostatistics, University of California, Berkeley
     orcid: 0000-0002-8474-591X
   - name: Alan E. Hubbard
-    affiliation: Division of Biostatistics, University of California, Berkeley
+    affiliation:
+      - Group in Biostatistics, University of California, Berkeley
     orcid: 0000-0002-3769-0127
   - name: Mark J. van der Laan
-    affiliation: Division of Biostatistics, University of California, Berkeley
-    orcid: TBA
+    affiliation:
+      - Group in Biostatistics, University of California, Berkeley
+      - Department of Statistics, University of California, Berkeley
 abstract: |
@@
 NH designed and implemented the software package, applied the tool to the use
-cases, and co-drafted the present article. RP helped in designing the software
-and co-drafted the present article. AH and ML serve as the advisors for the
-development of this software and general statistical algorithm.
+cases presented, and co-drafted the present manuscript. RP helped in designing
+the software and co-drafted the present article. AH and ML served as advisors
+for the development of this software and general statistical algorithm.
 
@@
 
-NH was supported in part by the National Library Of Medicine of the National
-Institutes of Health under Award Number T32LM012417, by [SUPERFUND GRANT], and
-by [NINA HOLLAND GRANT]. RP was supported by [SUPERFUND GRANT]. The content of
-this work is solely the responsibility of the authors and does not necessarily
-represent the official views of the various funding sources and agencies.
+NH was supported in part by the National Library of Medicine of the National
+Institutes of Health under Award Number T32-LM012417, by P42-ES004705, and
+by R01-ES021369. RP was supported by P42-ES004705. The content of this work is
+solely the responsibility of the authors and does not necessarily represent the
+official views of the various funding sources and agencies.
 

### 2018-04-16 paper/paper_BiocF1000.tex
F1000 (#34)
* use minfi data

* add paper refs

* started to add a bit to the introduction &  moved the text from the repository to the appropriate sections in the article. I will use the text from the repo as a rough guide to draft something more cohesive

* revisions/updates to F1000 manuscript (#28)

* minor vignette additions/corrections

* almost done with f1000, minor vignette corrections

* adding the Summary and final touches

* a few more minor things

* fixing merge conflict

* minor formatting

* suggestion: convert betas to mvals in the limma screenas the mvals have far more centra

@@
 \usepackage[numbers]{natbib}
+\bibliographystyle{unsrtnat}
 
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
 \newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
 \newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
 \newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
 
@@
 \setcounter{secnumdepth}{0}
+
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
 
@@
 
-\title{\texttt{methyvim}: Targeted and Model-free Analysis of Differential
-Methylation in R}
+\title{\texttt{methyvim}: Targeted and Model-free Differential Methylation Analysis in R}
 
-\author[1]{Nima S. Hejazi}
+\author[1,2]{Nima S. Hejazi\thanks{\ttfamily nhejazi@berkeley.edu}}
 \author[1]{Rachael V. Phillips}
 \author[1]{Alan E. Hubbard}
-\author[1]{Mark J. van der Laan}
-\affil[1]{Division of Biostatistics, University of California, Berkeley}
+\author[1,3]{Mark J. van der Laan}
+\affil[1]{Group in Biostatistics, University of California, Berkeley}
+\affil[2]{Center for Computational Biology, University of California, Berkeley}
+\affil[3]{Department of Statistics, University of California, Berkeley}
 
@@
 \begin{abstract}
-We present \texttt{methyvim}, a general algorithm for the nonparametric
-estimation of treatment/exposure effects on DNA methylation at CpG
-sites, complete with straightforward statistical inference for such
-estimates. This approach leverages variable importance measures, a class
-of parameters that arise in causal inference and are defined in such a
-manner that they provide targeted estimates of the relative importance
-of individual CpG site with respect to a binary exposure/treatment
-assigned at the level of subjects. Additionally, this procedure is
-computationally efficient; incorporating a preliminary screening step to
-isolate a subset of sites for which there is cursory evidence of
-differential methylation and a multiple testing correction step to
-control the False Discovery Rate as if all sites were tested. This
-technique for analysis of differentially methylated positions provides
-an avenue to move beyond linear models without the loss of interpretable
+We present \texttt{methyvim}, a general algorithm for the nonparametric estimation of
+treatment/exposure effects on DNA methylation at CpG sites, complete with
+straightforward statistical inference for such estimates. This approach
+leverages variable importance measures, a class of parameters that arise in
+causal inference and are defined in such a manner that they provide targeted
+estimates of the relative importance of individual CpG site with respect to
+a binary exposure/treatment assigned at the level of subjects. Additionally,
+this procedure is computationally efficient; incorporating a preliminary
+screening step to isolate a subset of sites for which there is cursory
+evidence of differential methylation and a multiple testing correction step
+to control the False Discovery Rate as if all sites were tested. This
+technique for analysis of differentially methylated positions provides an
+avenue to move beyond linear models without the loss of interpretable
 statistical inference.
@@
 \section*{Keywords}
-DNA methylation, differential methylation, epigenetics, causal
-inference, statistical variable importance, targeted minimum loss-based
-estimation, machine learning
+DNA methylation, differential methylation, epigenetics, causal inference, statistical variable importance, targeted minimum loss-based estimation, machine learning
 
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-DNA methylation is a fundamental epigenetic process known to play an
-important role in the regulation of gene expression. DNA CpG methylation
-involves the addition of a methyl group (\(\text{CH}_3\)) to the fifth
-carbon of the cytosine ring structure to form 5-methylcytosine. Numerous
-biological and medical studies have implicated DNA CpG methylation as
-playing a role in disease and development (Robertson 2005). Perhaps
-unsurprisingly then, biotechnologies have been developed to study the
-molecular mechanisms of this epigenetic process. Modern assays, like the
-Illumina \emph{Infinium} Methylation assay, allow for quantitative
-interrogation of DNA methylation of CpG sites scattered across the
-genome at single-nucleotide resolution; moreover, much effort has been
-invested, by the bioinformatics community, in the development of tools
-for properly removing technological effects that may contaminate
-biological signatures measured by such assays. Despite these advances in
-both biological and bioninformatical techniques, most statistical
-methods available for differential analysis of data produced by such
-assays rely on over-simplified (often generalized linear) models. This
-is because typical statistical estimation approaches are not achievable
-for such high-dimensional data without restrictive modeling assumptions.
-When these assumptions are violated, estimates of the population-level
-effect of an exposure or treatment may have large bias. Data-adaptive
-estimation procedures that utilize machine learning can control for
-high-dimensional confounders when estimating treatment/exposure effects.
+DNA methylation is a fundamental epigenetic process known to play an important
+role in the regulation of gene expression. DNA CpG methylation involves the
+addition of a methyl group (\(\text{CH}_3\)) to the fifth carbon of the cytosine
+ring structure to form 5-methylcytosine. Numerous biological and medical studies
+have implicated DNA CpG methylation as playing a role in disease and
+development \citep{robertson2005dna}. Perhaps unsurprisingly then, biotechnologies
+have been developed to study the molecular mechanisms of this epigenetic
+process. Modern assays, like the Illumina \emph{Infinium} Methylation assay,
+allow for quantitative interrogation of DNA methylation of CpG sites scattered
+across the genome at single-nucleotide resolution; moreover, much effort
+has been invested, by the bioinformatics community, in the development of tools
+for properly removing technological effects that may contaminate biological
+signatures measured by such assays. Despite these advances in both biological
+and bioninformatical techniques, most statistical methods available for
+differential analysis of data produced by such assays rely on over-simplified
+(often generalized linear) models. This is because typical statistical
+estimation approaches are not achievable for such high-dimensional data without
+restrictive modeling assumptions. When these assumptions are violated, estimates
+of the population-level effect of an exposure or treatment may have large bias.
+Data-adaptive estimation procedures that utilize machine learning can control
+for high-dimensional confounders when estimating treatment/exposure effects.
 However, interpretable statistical inference (i.e.~confidence intervals,
-hypothesis tests) resulting from these data-adaptive estimates is
-challenging (Libbrecht and Noble 2015).
+hypothesis tests) resulting from these data-adaptive estimates is challenging
+\citep{libbrecht2015machine}.
 
-In this paper, we present an alternative such statistical analysis
-approaches in the form of nonparametric estimation procedures that
-provide simple and well interpretable statistical inference. Inspired by
-concepts of counterfactual effects from the causal inference literature
-and machine learning algorithms, we provide a computationally efficient
-technique for obtaining targeted estimates of nonparametric
-\emph{variable importance measures} (\textbf{VIM}) for a pre-screened
-set of CpGs, controlling for the False Discovery Rate (FDR) as if all
-sites were tested. Under counterfactual assumptions, these estimators
-have asymptotic normal sampling distributions and thus provide reliable
-inference. Counterfactuals are ``counter''-fact, defined in terms of an
-unrealistic yet ideal experiment. In the context of DNA methylation and
-for a binary treatment, we define the counterfactual outcomes as the
-observed methylation value (whether Beta- or M-values) a CpG site would
-have if all subjects were treated and the methylation value a CpG site
-would have if all subjects were not treated. Of course, these
-counterfactual outcomes are impossible to observe but they are very
-helpful for us to asses the importance of a CpG site relative to a
-subjects treatment/exposure and they can be statistically estimated (M.
-J. van der Laan 2006). Our VIM definition uses the concepts of
-counterfactuals and the framework for estimation is built on
-\emph{targeted maximum likelihood estimation} or \emph{targeted minimum
-loss-based estimation} (\textbf{TMLE}) (M. J. van der Laan and Rubin
-2006). This methodology targets the individual importance of each CpG
-site by taking advantage of the prediction power of machine learning
-algorithms. The mathematical properties of this approach have been
-explored in (Tuglus and van der Laan 2008) and we focus on our
-contribution to it's application, \texttt{methyvim}, an R package that
-make this technique for differential methylation analysis easily
-accessible.
+In this paper, we present an alternative such statistical analysis approaches in
+the form of nonparametric estimation procedures that provide simple and well
+interpretable statistical inference. Inspired by concepts of counterfactual
+effects from the causal inference literature and machine learning algorithms, we
+provide a computationally efficient technique for obtaining targeted estimates
+of nonparametric \emph{variable importance measures} (\textbf{VIM}) for a pre-screened set
+of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
+tested. Under counterfactual assumptions, these estimators have asymptotic
+normal sampling distributions and thus provide reliable inference.
+Counterfactuals are ``counter''-fact, defined in terms of an unrealistic yet ideal
+experiment. In the context of DNA methylation and for a binary treatment, we
+define the counterfactual outcomes as the observed methylation value (whether
+Beta- or M-values) a CpG site would have if all subjects were treated and the
+methylation value a CpG site would have if all subjects were not treated. Of
+course, these counterfactual outcomes are impossible to observe but they are
+very helpful for us to asses the importance of a CpG site relative to a subjects
+treatment/exposure and they can be statistically estimated
+\citep{vdl2006statistical}. Our VIM definition uses the concepts of counterfactuals
+and the framework for estimation is built on \emph{targeted maximum likelihood
+estimation} or \emph{targeted minimum loss-based estimation} (\textbf{TMLE})
+\citep{vdl2006targeted}. This methodology targets the individual importance of each
+CpG site by taking advantage of the prediction power of machine learning
+algorithms. The mathematical properties of this approach have been explored in
+\citep{tuglus2008targeted} and we focus on our contribution to it's application,
+\texttt{methyvim}, an R package that make this technique for differential methylation
+analysis easily accessible.
 
 For a general discussion of the framework of targeted minimum loss-based
-estimation and the role this approach plays in statistical causal
-inference, the interested reader is invited to consult M. J. van der
-Laan and Rose (2011) and M. J. van der Laan and Rose (2017). For a more
-general introduction to (statistical) causal inference, Pearl (2009) and
-Hernan and Robins (2018, forthcoming) may be of interest.
+estimation and the role this approach plays in statistical causal inference, the
+interested reader is invited to consult \citet{vdl2011targeted} and \citet{vdl2017targeted}.
+For a more general introduction to (statistical) causal inference,
+\citet{pearl2009causality} and \citet{hernan2018causal} may be of interest.
 
-\section{Methods}\label{methods}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-\subsection{Implementation}\label{implementation}
+\hypertarget{implementation}{%
+\subsection{Implementation}\label{implementation}}
 
-The core functionality of this package is made available via the
-eponymous \texttt{methyvim} function, which implements a statistical
-algorithm designed to compute targeted estimates of VIMs, defined in
-such a way that the VIMs represent parameters of scientific interest in
-computational biology experiments; moreover, these VIMs are defined such
-that they may be estimated in a manner that is very nearly
-assumption-free, that is, within a \emph{fully nonparametric statistical
-model}. \textbf{The statistical algorithm consists in several major
-steps}:
+The core functionality of this package is made available via the eponymous
+\texttt{methyvim} function, which implements a statistical algorithm designed to
+compute targeted estimates of VIMs, defined in such a way that the VIMs
+represent parameters of scientific interest in computational biology
+experiments; moreover, these VIMs are defined such that they may be estimated in
+a manner that is very nearly assumption-free, that is, within a \emph{fully
+nonparametric statistical model}. \textbf{The statistical algorithm consists in
+several major steps}:
 
@@
 \item
-  Pre-screening of genomic sites is used to isolate a subset of sites
-  for which there is cursory evidence of differential methylation. For
-  the sake of computational feasibility, targeted minimum loss-based
-  estimates of VIMs are computed only for this subset of sites.
-  Currently, the available screening approach adapts core routines from
-  the \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R
-  package. Future releases will support functionality from other
-  packages (e.g.,
-  \href{https://CRAN.R-project.org/package=randomForest}{\texttt{randomForest}},
-  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}).
-  Following the style of the function for performing screening via
-  \texttt{limma}, users may write their own screening functions and are
-  invited to contribute such functions to the core software package by
-  opening pull requests at the GitHub repository.
+  Pre-screening of genomic sites is used to isolate a subset of sites for
+  which there is cursory evidence of differential methylation. For the sake of
+  computational feasibility, targeted minimum loss-based estimates of VIMs are
+  computed only for this subset of sites. Currently, the available screening
+  approach adapts core routines from the
+  \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R
+  package. Future releases will support functionality from other packages
+  (e.g., \href{https://CRAN.R-project.org/package=randomForest}{\texttt{randomForest}},
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}). Following the
+  style of the function for performing screening via \texttt{limma}, users may write
+  their own screening functions and are invited to contribute such functions to
+  the core software package by opening pull requests at the GitHub repository.
 \item
-  Nonparametric estimates of VIMs, for the specified target parameter,
-  are computed at each of the CpG sites passing the screening step. The
-  VIMs are defined in such a way that the estimated effects is of an
-  exposure/treatment on the methylation status of a target CpG site,
-  controlling for the observed methylation status of the neighbors of
-  that site. Currently, routines are adapted from the
-  \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R
-  package. Future releases will support doubly-robust estimates of these
-  VIMs (via the
-  \href{https://cran.r-project.org/package=drtmle}{\texttt{drtmle}}
-  package) and add parameters for continuous treatments/exposures (via
-  the
-  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}
-  package).
+  Nonparametric estimates of VIMs, for the specified target parameter, are
+  computed at each of the CpG sites passing the screening step. The VIMs are
+  defined in such a way that the estimated effects is of an exposure/treatment
+  on the methylation status of a target CpG site, controlling for the observed
+  methylation status of the neighbors of that site. Currently, routines are
+  adapted from the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package.
+  Future releases will support doubly-robust estimates of these VIMs (via the
+  \href{https://cran.r-project.org/package=drtmle}{\texttt{drtmle}} package)
+  and add parameters for continuous treatments/exposures (via the
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}} package).
 \item
-  Since pre-screening is performed prior to estimating VIMs, we make use
-  of a multiple testing correction uniquely suited to such settings. Due
-  to the multiple testing nature of the estimation problem, a variant of
-  the Benjamini \& Hochberg procedure for controlling the False
-  Discovery Rate (FDR) is applied (Benjamini and Hochberg 1995).
-  Specifically, we apply the ``modified marginal Benjamini \& Hochberg
-  step-up False Discovery Rate controlling procedure for multi-stage
-  analyses'' (FDR-MSA), which is guaranteed to control the FDR as if all
-  sites were tested (i.e., without screening) (Tuglus and van der Laan
-  2009).
+  Since pre-screening is performed prior to estimating VIMs, we make use of a
+  multiple testing correction uniquely suited to such settings. Due to the
+  multiple testing nature of the estimation problem, a variant of the Benjamini
+  \& Hochberg procedure for controlling the False Discovery Rate (FDR) is
+  applied \citep{benjamini1995controlling}. Specifically, we apply the ``modified
+  marginal Benjamini \& Hochberg step-up False Discovery Rate controlling
+  procedure for multi-stage analyses'' (FDR-MSA), which is guaranteed to
+  control the FDR as if all sites were tested (i.e., without screening)
+  \citep{tuglus2009modified}.
 \end{enumerate}
 
-\paragraph{Parameters of Interest}\label{parameters-of-interest}
+\hypertarget{parameters-of-interest}{%
+\paragraph{Parameters of Interest}\label{parameters-of-interest}}
 
-For the CpG sites that passed the pre-screening step, the user chooses
-the parameter of interest that will be computed at each site. \emph{In
-all cases, an estimator of the parameter of interest is constructed via
-targeted minimum loss-based estimation}.
+For the CpG sites that passed the pre-screening step, the user chooses the
+parameter of interest that will be computed at each site. \emph{In all cases, an
+estimator of the parameter of interest is constructed via targeted minimum
+loss-based estimation}.
 
@@
   The average treatment effect (ATE): The effect of a binary exposure or
-  treatment on the observed methylation at a target CpG site is
-  estimated, controlling for the observed methylation at all other CpG
-  sites in the same neighborhood as the target site, based on an
-  additive form. Often denoted \(\psi_0 = \psi_0(1) - \psi_0(0)\), the
-  parameter estimate represents the additive difference in methylation
-  that would have been observed at the target site had all observations
-  received the treatment versus the counterfactual under which none
-  received the treatment.
+  treatment on the observed methylation at a target CpG site is estimated,
+  controlling for the observed methylation at all other CpG sites in the same
+  neighborhood as the target site, based on an additive form. Often denoted
+  \(\psi_0 = \psi_0(1) - \psi_0(0)\), the parameter estimate represents the
+  additive difference in methylation that would have been observed at the
+  target site had all observations received the treatment versus the
+  counterfactual under which none received the treatment.
 \item
-  The relative risk (RR): The effect of a binary exposure or treatment
-  on the observed methylation at a target CpG site is estimated,
-  controlling for the observed methylation at all other CpG sites in the
-  same neighborhood as the target site, based on a geometric form. Often
-  denoted, \(\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}\), the parameter
-  estimate represents the multiplicative difference in methylation that
-  would have been observed at the target site had all observations
-  received the treatment versus the counterfactual under which none
-  received the treatment.
+  The relative risk (RR): The effect of a binary exposure or treatment on the
+  observed methylation at a target CpG site is estimated, controlling for the
+  observed methylation at all other CpG sites in the same neighborhood as the
+  target site, based on a geometric form. Often denoted,
+  \(\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}\), the parameter estimate represents the
+  multiplicative difference in methylation that would have
+  been observed at the target site had all observations received the treatment
+  versus the counterfactual under which none received the treatment.
 \end{itemize}
 
-Estimating the VIM corresponding to the parameters above, for
-discrete-valued treatments or exposures, requires two separate
-regression steps: one for the treatment mechanism (propensity score) and
-one for the outcome regression. Technical details on the nature of these
-regressions are discussed in Hernan and Robins (2018, forthcoming), and
-details for estimating these regressions in the framework of targeted
-minimum loss-based estimation are discussed in M. J. van der Laan and
-Rose (2011).
+Estimating the VIM corresponding to the parameters above, for discrete-valued
+treatments or exposures, requires two separate regression steps: one for the
+treatment mechanism (propensity score) and one for the outcome regression.
+Technical details on the nature of these regressions are discussed in
+\citet{hernan2018causal}, and details for estimating these regressions in the
+framework of targeted minimum loss-based estimation are discussed in
+\citet{vdl2011targeted}.
 
-Support for continuous-valued treatments or exposures is \emph{planned
-but not yet available}, though work is underway to incorporate into our
-methodology the following
+Support for continuous-valued treatments or exposures is \emph{planned but not yet
+available}, though work is underway to incorporate into our methodology the
+following
 
@@
 \item
-  A nonparametric variable importance measure (NPVI) (Chambaz, Neuvial,
-  and van der Laan 2012): The effect of continuous-valued exposure or
-  treatment (the observed methylation at a target CpG site) on an
-  outcome of interest is estimated, controlling for the observed
-  methylation at all other CpG sites in the same neighborhood as the
-  target (treatment) site, based on a parameter that compares values of
-  the treatment against a reference value taken to be the null. In
-  particular, the implementation provided is designed to assess the
-  effect of differential methylation at the target CpG site on a
-  (typically) phenotype-level outcome of interest (e.g., survival), in
-  effect providing an nonparametric evaluation of the impact of
-  methylation at the target site on said outcome.
+  A nonparametric variable importance measure (NPVI) \citep{chambaz2012estimation}:
+  The effect of continuous-valued exposure or treatment (the observed
+  methylation at a target CpG site) on an outcome of interest is estimated,
+  controlling for the observed methylation at all other CpG sites in the same
+  neighborhood as the target (treatment) site, based on a parameter that
+  compares values of the treatment against a reference value taken to be the
+  null. In particular, the implementation provided is designed to assess the
+  effect of differential methylation at the target CpG site on a (typically)
+  phenotype-level outcome of interest (e.g., survival), in effect providing an
+  nonparametric evaluation of the impact of methylation at the target site on
+  said outcome.
 \end{itemize}
 
-\paragraph{\texorpdfstring{Class
-\texttt{methytmle}}{Class methytmle}}\label{class-methytmle}
+\hypertarget{class-methytmle}{%
+\paragraph{\texorpdfstring{Class \texttt{methytmle}}{Class methytmle}}\label{class-methytmle}}
 
-We have adopted a class \texttt{methytmle} to help organize the
-functionality within this package. The \texttt{methytmle} class builds
-upon the \texttt{GenomicRatioSet} class provided by the \texttt{minfi}
-package so all of the slots of \texttt{GenomicRatioSet} are contained in
-a \texttt{methytmle} object. The new class introduced in the
-\texttt{methyvim} package includes several new slots:
+We have adopted a class \texttt{methytmle} to help organize the functionality within
+this package. The \texttt{methytmle} class builds upon the \texttt{GenomicRatioSet} class
+provided by the \texttt{minfi} package so all of the slots of \texttt{GenomicRatioSet} are
+contained in a \texttt{methytmle} object. The new class introduced in the \texttt{methyvim}
+package includes several new slots:
 
@@
 \item
-  \texttt{call} - the form of the original call to the \texttt{methyvim}
-  function.
+  \texttt{call} - the form of the original call to the \texttt{methyvim} function.
 \item
-  \texttt{screen\_ind} - indices identifying CpG sites that pass the
-  screening process.
+  \texttt{screen\_ind} - indices identifying CpG sites that pass the screening process.
 \item
-  \texttt{clusters} - non-unique IDs corresponding to the manner in wich
-  sites are treated as neighbors. These are assigned by genomic distance
-  (bp) and respect chromosome boundaries (produced via a call to
-  \texttt{bumphunter::clusterMaker}).
+  \texttt{clusters} - non-unique IDs corresponding to the manner in wich sites are
+  treated as neighbors. These are assigned by genomic distance (bp) and respect
+  chromosome boundaries (produced via a call to \texttt{bumphunter::clusterMaker}).
 \item
-  \texttt{var\_int} - the treatment/exposure status for each subject.
-  Currently, these must be binary, due to the definition of the
-  supported targeted parameters.
+  \texttt{var\_int} - the treatment/exposure status for each subject. Currently, these
+  must be binary, due to the definition of the supported targeted parameters.
 \item
-  \texttt{param} - the name of the target parameter from which the
-  estimated VIMs are defined.
+  \texttt{param} - the name of the target parameter from which the estimated VIMs are
+  defined.
 \item
-  \texttt{vim} - a table of statistical results obtained from estimating
-  VIMs for each of the CpG sites that pass the screening procedure.
+  \texttt{vim} - a table of statistical results obtained from estimating VIMs for
+  each of the CpG sites that pass the screening procedure.
 \item
-  \texttt{ic} - the measured array values for each of the CpG sites
-  passing the screening, transformed into influence curve space based on
-  the chosen target parameter.
+  \texttt{ic} - the measured array values for each of the CpG sites passing the
+  screening, transformed into influence curve space based on the chosen target
+  parameter.
 \end{itemize}
 
-We refer the reader to the package vignette, ``\texttt{methyvim}:
-Targeted Learning for Differential Methylation Analysis,'' included in
-any distribution of the software package, for further details.
+We refer the reader to the package vignette, ``\texttt{methyvim}: Targeted Data-Adaptive
+Estimation and Inference for Differential Methylation Analysis,'' included in any
+distribution of the software package, for further details.
 
-\subsection{Operation}\label{operation}
+\hypertarget{operation}{%
+\subsection{Operation}\label{operation}}
 
-A standard computer with the latest version of R and Bioconductor 3.6
-installed will handle applications of the \texttt{methyvim} package.
+A standard computer with the latest version of R and Bioconductor 3.6 installed
+will handle applications of the \texttt{methyvim} package.
 
-\section{Use Cases}\label{use-cases}
+\hypertarget{use-cases}{%
+\section{Use Cases}\label{use-cases}}
 
-To examine the practical applications and the full set of utilities of
-the \texttt{methyvim} package, we will use a publicly available example
-data set produced by the Illumina 450K array, from the
-\texttt{minfiData} R package.
+To examine the practical applications and the full set of utilities of the
+\texttt{methyvim} package, we will use a publicly available example data set produced
+by the Illumina 450K array, from the \texttt{minfiData} R package.
 
-\paragraph{Preliminaries: Setting up the
-Data}\label{preliminaries-setting-up-the-data}
+\hypertarget{preliminaries-setting-up-the-data}{%
+\paragraph{Preliminaries: Setting up the Data}\label{preliminaries-setting-up-the-data}}
 
-We begin by loading the package and the data set. After loading the
-data, which comes in the form of a raw \texttt{MethylSet} object, we
-perform some further processing by mapping to the genome (with
-\texttt{mapToGenome}) and converting the values from the methylated and
-unmethylated channels to Beta-values (via \texttt{ratioConvert}). These
-two steps together produce an object of class \texttt{GenomicRatioSet},
-provided by the \texttt{minfi} package.
+We begin by loading the package and the data set. After loading the data, which
+comes in the form of a raw \texttt{MethylSet} object, we perform some further
+processing by mapping to the genome (with \texttt{mapToGenome}) and converting the
+values from the methylated and unmethylated channels to Beta-values
+(via \texttt{ratioConvert}). These two steps together produce an object of class
+\texttt{GenomicRatioSet}, provided by the \texttt{minfi} package.
 
@@
 
-We can create an object of class \texttt{methytmle} from any
-\texttt{GenomicRatioSet} object simply invoking the S4 class
-constructor:
+We can create an object of class \texttt{methytmle} from any \texttt{GenomicRatioSet} object
+simply invoking the S4 class constructor:
 
@@
 \begin{verbatim}
-## methyvim: Targeted Learning for Differential Methylation Analysis
-\end{verbatim}
-
-\begin{verbatim}
-## Version: 0.99.8
+## methyvim v1.1.0: Targeted Variable Importance for Differential Methylation Analysis
 \end{verbatim}
@@
 
-Additionally, a \texttt{GenomicRatioSet} can be created from a matrix
-with the function \texttt{makeGenomicRatioSetFromMatrix} provided by the
-\texttt{minfi} package.
+Additionally, a \texttt{GenomicRatioSet} can be created from a matrix with the
+function \texttt{makeGenomicRatioSetFromMatrix} provided by the \texttt{minfi} package.
 
-\paragraph{Differential Methylation
-Analysis}\label{differential-methylation-analysis}
+\hypertarget{differential-methylation-analysis}{%
+\paragraph{Differential Methylation Analysis}\label{differential-methylation-analysis}}
 
-For this example analysis, we'll treat the condition of the patients as
-the exposure/treatment variable of interest. The \texttt{methyvim}
-function requires that this variable either be \texttt{numeric} or
-easily coercible to \texttt{numeric}. To facilitate this, we'll simply
-convert the covariate (currently a \texttt{character}):
+For this example analysis, we'll treat the condition of the patients as the
+exposure/treatment variable of interest. The \texttt{methyvim} function requires that
+this variable either be \texttt{numeric} or easily coercible to \texttt{numeric}. To
+facilitate this, we'll simply convert the covariate (currently a \texttt{character}):
 
@@
 
-\textbf{n.b.}, the re-coding process results in ``normal'' patients
-being assigned a value of 1 and cancer patients a 0.
+\textbf{n.b.}, the re-coding process results in ``normal'' patients being assigned a
+value of 1 and cancer patients a 0.
 
-Now, we are ready to analyze the effects of cancer status on DNA
-methylation using this data set. We proceed as follows with a targeted
-minimum loss-based estimate of the Average Treatment Effect.
+Now, we are ready to analyze the effects of cancer status on DNA methylation
+using this data set. We proceed as follows with a targeted minimum loss-based
+estimate of the Average Treatment Effect.
 
@@
 
-Note that we set the \texttt{obs\_per\_covar} argument to a relatively
-low value (2, where the recommended default is 20) for the purposes of
-this example as the sample size is only 10. We do this only to exemplify
-the estimation procedure and it is important to point out that such low
-values for \texttt{obs\_per\_covar} will compromise the quality of
-inference obtained because this setting directly affects the definition
-of the target parameter.
+Note that we set the \texttt{obs\_per\_covar} argument to a relatively low value (2,
+where the recommended default is 20) for the purposes of this example as the
+sample size is only 10. We do this only to exemplify the estimation procedure
+and it is important to point out that such low values for \texttt{obs\_per\_covar} will
+compromise the quality of inference obtained because this setting directly
+affects the definition of the target parameter.
 
-Further, note that here we apply the \texttt{glm} flavor of the
-\texttt{tmle\_type} argument, which produces faster results by fitting
-models for the propensity score and outcome regressions using a limited
-number of parametric models. By contrast, the \texttt{sl} (for ``Super
-Learning'') flavor fits these two regressions using highly nonparametric
-and data-adaptive procedures (i.e., via machine learning). Obtaining the
-estimates via GLMs results in each of the regression steps being less
-robust than if nonparametric regressions were used.
+Further, note that here we apply the \texttt{glm} flavor of the \texttt{tmle\_type} argument,
+which produces faster results by fitting models for the propensity score and
+outcome regressions using a limited number of parametric models. By contrast,
+the \texttt{sl} (for ``Super Learning'') flavor fits these two regressions using highly
+nonparametric and data-adaptive procedures (i.e., via machine learning).
+Obtaining the estimates via GLMs results in each of the regression steps
+being less robust than if nonparametric regressions were used.
 
-We can view a table of results by examining the \texttt{vim} slot of the
-produced \texttt{methytmle} object.
+We can view a table of results by examining the \texttt{vim} slot of the produced
+\texttt{methytmle} object.
 
@@
 \begin{verbatim}
-##            lowerCI_ATE      est_ATE   upperCI_ATE      Var_ATE
-## cg14008030 -0.11956597 -0.031415962  0.0567340489 2.022705e-03
-## cg20253340 -0.08850637 -0.058866142 -0.0292259165 2.286919e-04
-## cg21870274 -0.09499057 -0.029118982  0.0367526091 1.129495e-03
-## cg17308840 -0.04626018 -0.007152452  0.0319552773 3.981191e-04
-## cg00645010 -0.02677328 -0.013465554 -0.0001578256 4.609945e-05
-## cg27534567  0.06745648  0.115711854  0.1639672225 6.061486e-04
-##                    pval n_neighbors_all n_neighbors_w max_corr_w
-## cg14008030 4.848468e-01               0             0         NA
-## cg20253340 9.917426e-05               0             0         NA
-## cg21870274 3.862537e-01               2             1  0.9443580
-## cg17308840 7.199943e-01               2             1  0.9443580
-## cg00645010 4.734007e-02               2             2  0.5236810
-## cg27534567 2.602936e-06               1             0  0.9362968
+##                 lwr_ci      est_ate       upr_ci      var_ate         pval
+## cg14008030 -0.11956597 -0.031415962  0.056734049 2.022705e-03 4.848468e-01
+## cg20253340 -0.08850637 -0.058866142 -0.029225917 2.286919e-04 9.917426e-05
+## cg21870274 -0.09499057 -0.029118982  0.036752609 1.129495e-03 3.862537e-01
+## cg17308840 -0.04626018 -0.007152452  0.031955277 3.981191e-04 7.199943e-01
+## cg00645010 -0.02697515 -0.013809263 -0.000643373 4.512199e-05 3.980386e-02
+## cg27534567  0.06745648  0.115711854  0.163967223 6.061486e-04 2.602936e-06
+##            n_neighbors n_neighbors_control max_cor_neighbors
+## cg14008030           0                   0                NA
+## cg20253340           0                   0                NA
+## cg21870274           2                   1         0.9443580
+## cg17308840           2                   1         0.9443580
+## cg00645010           2                   2         0.5236810
+## cg27534567           1                   0         0.9362968
 \end{verbatim}
 
-Finally, we may compute FDR-corrected p-values, by applying a modified
-procedure for controlling the False Discovery Rate for multi-stage
-analyses (FDR-MSA) (Tuglus and van der Laan 2009). We do this by simply
-applying the \texttt{fdr\_msa} function.
+Finally, we may compute FDR-corrected p-values, by applying a modified procedure
+for controlling the False Discovery Rate for multi-stage analyses (FDR-MSA)
+\citep{tuglus2009modified}. We do this by simply applying the \texttt{fdr\_msa} function.
 
@@
 
-Having explored the results of our analysis numerically, we now proceed
-to use the visualization tools provided with the \texttt{methyvim} R
-package to further enhance our understanding of the results.
+Having explored the results of our analysis numerically, we now proceed to use
+the visualization tools provided with the \texttt{methyvim} R package to further
+enhance our understanding of the results.
 
-\paragraph{Visualization of Results}\label{visualization-of-results}
+\hypertarget{visualization-of-results}{%
+\paragraph{Visualization of Results}\label{visualization-of-results}}
 
-While making allowance for users to explore the full set of results
-produced by the estimation procedure (by way of exposing these directly
-to the user), the \texttt{methyvim} package also provides \emph{three}
-(3) visualization utilities that produce plots commonly used in
-examining the results of differential methylation analyses.
+While making allowance for users to explore the full set of results produced by
+the estimation procedure (by way of exposing these directly to the user), the
+\texttt{methyvim} package also provides \emph{three} (3) visualization utilities that
+produce plots commonly used in examining the results of differential methylation
+analyses.
 
-A simple call to \texttt{plot} produces side-by-side histograms of the
-raw p-values computed as part of the estimation process and the
-corrected p-values obtained from using the FDR-MSA procedure.
+A simple call to \texttt{plot} produces side-by-side histograms of the raw p-values
+computed as part of the estimation process and the corrected p-values obtained
+from using the FDR-MSA procedure.
 
@@
 
-\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-raw-1.pdf}
+\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-raw-1} \end{center}
 
@@
 
-\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-fdr-1.pdf}
+\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-pvals-fdr-1} \end{center}
 
-\textbf{Remark:} The plots displayed above may also be generated as
-side-by-side histograms in a single plot object. This is the default for
-the \texttt{plot} method and may easily be invoked by specifying no
-additional arguments to the \texttt{plot} function, unlike in the above.
+\textbf{Remark:} The plots displayed above may also be generated as side-by-side
+histograms in a single plot object. This is the default for the \texttt{plot} method
+and may easily be invoked by specifying no additional arguments to the \texttt{plot}
+function, unlike in the above.
 
-While histograms of the p-values may be generally useful in inspecting
-the results of the estimation procedure, a more common plot used in
-examining the results of differential methylation procedures is the
-volcano plot, which plots the parameter estimate along the x-axis and
-\(-\text{log}_{10}(\text{p-value})\) along the y-axis. We implement such
-a plot in the \texttt{methyvolc} function:
+While histograms of the p-values may be generally useful in inspecting the
+results of the estimation procedure, a more common plot used in examining the
+results of differential methylation procedures is the volcano plot, which plots
+the parameter estimate along the x-axis and \(-\text{log}_{10}(\text{p-value})\)
+along the y-axis. We implement such a plot in the \texttt{methyvolc} function:
 
@@
 
-\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-volcano-1.pdf}
+\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-volcano-1} \end{center}
 
-The purpose of such a plot is to ensure that very low (possibly
-statistically significant) p-values do not arise from cases of low
-variance. This appears to be the case in the plot above (notice that
-most parameter estimates are near zero, even in cases where the raw
-p-values are quite low).
+The purpose of such a plot is to ensure that very low (possibly statistically
+significant) p-values do not arise from cases of low variance. This appears to
+be the case in the plot above (notice that most parameter estimates are near
+zero, even in cases where the raw p-values are quite low).
 
 Yet another popular plot for visualizing effects in such settings is the
-heatmap, which plots estimates of the raw methylation effects (as
-measured by the assay) across subjects using a heat gradient. We
-implement this in the \texttt{methyheat} function:
+heatmap, which plots estimates of the raw methylation effects (as measured by
+the assay) across subjects using a heat gradient. We implement this in the
+\texttt{methyheat} function:
 
@@
 
-\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1.pdf}
+\begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1} \end{center}
 
-Invoking \texttt{methyheat} in this manner produces a plot of the top
-sites (\(25\), by default) based on the raw p-value, using the raw
-methylation measures in the plot. This uses the exceptional
-\texttt{superheat} R package ({\textbf{???}}).
+Invoking \texttt{methyheat} in this manner produces a plot of the top sites (\(25\), by
+default) based on the raw p-value, using the raw methylation measures in the
+plot. This uses the exceptional \texttt{superheat} R package \citep{barter2017superheat}.
 
-\section{Summary}\label{summary}
+\hypertarget{summary}{%
+\section{Summary}\label{summary}}
 
-Here we introduce the R package \texttt{methyvim}, an algorithm for
-differential methylation analysis that moves beyond linear models.
-Additionally, this estimation procedure produces straightforward
-statistical inference. Inspired by concepts of counterfactual effects
-from the causal inference literature and machine learning algorithms, we
-provide a computationally efficient technique for obtaining targeted
-estimates of nonparametric variable importance measures for a
-pre-screened set of CpGs, controlling for the False Discovery Rate as if
+Here we introduce the R package \texttt{methyvim}, an algorithm for differential
+methylation analysis that moves beyond linear models. Additionally, this
+estimation procedure produces straightforward statistical inference. Inspired by
+concepts of counterfactual effects from the causal inference literature and
+machine learning algorithms, we provide a computationally efficient technique
+for obtaining targeted estimates of nonparametric variable importance measures
+for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
 all sites were tested, and the estimation framework is built on targeted
-minimum loss-based estimation. By taking advantage of the prediction
-power of machine learning algorithms, this methodology targets the
-individual importance of each CpG site, providing a fully nonparametric
-statistical model for the analysis of differential methylation.
+minimum loss-based estimation. By taking advantage of the prediction power of
+machine learning algorithms, this methodology targets the individual importance
+of each CpG site, providing a fully nonparametric statistical model for the
+analysis of differential methylation.
 
-\section{Software availability}\label{software-availability}
+\hypertarget{software-availability}{%
+\section{Software availability}\label{software-availability}}
 
-Latest source code (development version):
-\url{https://github.com/nhejazi/methyvim}
+Latest source code (development version): \url{https://github.com/nhejazi/methyvim}
 
-Bioconductor (stable release):
-\url{https://bioconductor.org/packages/methyvim}
+Bioconductor (stable release): \url{https://bioconductor.org/packages/methyvim}
 
-Archived source code as at time of publication:
-\url{https://github.com/nhejazi/methyvim}
+Archived source code as at time of
+publication: \url{https://github.com/nhejazi/methyvim}
 
@@
 
-\section{Author contributions}\label{author-contributions}
+\hypertarget{author-contributions}{%
+\section{Author contributions}\label{author-contributions}}
 
-NH designed and implemented the software package, applied the tool to
-the use cases, and co-drafted the present article. RP helped in
-designing the software and co-drafted the present article. AH and ML
-serve as the advisors for the development of this software and general
-statistical algorithm.
+NH designed and implemented the software package, applied the tool to the use
+cases presented, and co-drafted the present manuscript. RP helped in designing
+the software and co-drafted the present article. AH and ML served as advisors
+for the development of this software and general statistical algorithm.
 
-\section{Competing interests}\label{competing-interests}
+\hypertarget{competing-interests}{%
+\section{Competing interests}\label{competing-interests}}
 
@@
 
-\section{Grant information}\label{grant-information}
+\hypertarget{grant-information}{%
+\section{Grant information}\label{grant-information}}
 
-NH was supported in part by the National Library Of Medicine of the
-National Institutes of Health under Award Number T32LM012417, by
-{[}SUPERFUND GRANT{]}, and by {[}NINA HOLLAND GRANT{]}. RP was supported
-by {[}SUPERFUND GRANT{]}. The content of this work is solely the
-responsibility of the authors and does not necessarily represent the
+NH was supported in part by the National Library of Medicine of the National
+Institutes of Health under Award Number T32-LM012417, by P42-ES004705, and
+by R01-ES021369. RP was supported by P42-ES004705. The content of this work is
+solely the responsibility of the authors and does not necessarily represent the
 official views of the various funding sources and agencies.
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\hypertarget{ref-benjamini1995controlling}{}
-Benjamini, Yoav, and Yosef Hochberg. 1995. ``Controlling the False
-Discovery Rate: A Practical and Powerful Approach to Multiple Testing.''
-\emph{Journal of the Royal Statistical Society. Series B
-(Methodological)}. JSTOR, 289--300.
-
-\hypertarget{ref-chambaz2012estimation}{}
-Chambaz, Antoine, Pierre Neuvial, and Mark J van der Laan. 2012.
-``Estimation of a Non-Parametric Variable Importance Measure of a
-Continuous Exposure.'' \emph{Electronic Journal of Statistics} 6. NIH
-Public Access: 1059.
-
-\hypertarget{ref-hernan2018causal}{}
-Hernan, Miguel A, and James M Robins. 2018, forthcoming. \emph{Causal
-Inference}. Chapman \& Hall/Crc Texts in Statistical Science. Taylor \&
-Francis.
-
-\hypertarget{ref-libbrecht2015machine}{}
-Libbrecht, Maxwell W, and William Stafford Noble. 2015. ``Machine
-Learning in Genetics and Genomics.'' \emph{Nature Reviews. Genetics} 16
-(6). NIH Public Access: 321.
-
-\hypertarget{ref-pearl2009causality}{}
-Pearl, Judea. 2009. \emph{Causality: Models, Reasoning, and Inference}.
-Cambridge University Press.
-
-\hypertarget{ref-robertson2005dna}{}
-Robertson, Keith D. 2005. ``DNA Methylation and Human Disease.''
-\emph{Nature Reviews. Genetics} 6 (8). Nature Publishing Group: 597.
-
-\hypertarget{ref-tuglus2008targeted}{}
-Tuglus, Catherine, and Mark J van der Laan. 2008. ``Targeted Methods for
-Biomarker Discovery, the Search for a Standard.'' bepress.
-
-\hypertarget{ref-tuglus2009modified}{}
-Tuglus, Catherine, and Mark J. van der Laan. 2009. ``Modified FDR
-Controlling Procedure for Multi-Stage Analyses.'' \emph{Statistical
-Applications in Genetics and Molecular Biology} 8 (1). Walter de
-Gruyter: 1--15.
-doi:\href{https://doi.org/10.2202/1544-6115.1397}{10.2202/1544-6115.1397}.
-
-\hypertarget{ref-vdl2006statistical}{}
-van der Laan, Mark J. 2006. ``Statistical Inference for Variable
-Importance.'' \emph{The International Journal of Biostatistics} 2 (1).
-
-\hypertarget{ref-vdl2011targeted}{}
-van der Laan, Mark J, and Sherri Rose. 2011. \emph{Targeted Learning:
-Causal Inference for Observational and Experimental Data}. Springer
-Science \& Business Media.
-
-\hypertarget{ref-vdl2017targeted}{}
----------. 2017. \emph{Targeted Learning in Data Science: Causal
-Inference for Complex Longitudinal Studies}. Springer Science \&
-Business Media.
-
-\hypertarget{ref-vdl2006targeted}{}
-van der Laan, Mark J, and Daniel Rubin. 2006. ``Targeted Maximum
-Likelihood Learning.'' \emph{The International Journal of Biostatistics}
-2 (1).
+\renewcommand\refname{References}
+{\small\bibliography{paper\_BiocF1000.bib}}
 

### 2018-04-20 README.Rmd
update readme and site

@@
 [![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+[![Bioc Time](http://bioconductor.org/shields/years-in-bioc/methyvim.svg)](https://bioconductor.org/packages/release/bioc/html/methyvim.html)
+[![Bioc Downloads](http://bioconductor.org/shields/downloads/methyvim.svg)](https://bioconductor.org/packages/release/bioc/html/methyvim.html)
 [![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
@@
 available through the [Bioconductor
-Project](https://bioconductor.org/packages/methvim).
+project](https://bioconductor.org/packages/methvim).
 

### 2018-06-15 R/methytmle-class.R
fixing pkgdown

@@ -4,6 +4,9 @@
 #'
 #' @importFrom methods setClass
 #' @importClassesFrom minfi GenomicRatioSet
+#' @importClassesFrom SummarizedExperiment SummarizedExperiment
+#'  RangedSummarizedExperiment
+#' @import S4Vectors
 #'
 #' @export .methytmle
 #' @exportClass methytmle

### 2018-06-19 README.Rmd
fixing pkgdown
initial show method implementation; update bibs and resources

@@
 framework plays in statistical causal inference, the recommended references are
-@vdl2011targeted and @vdl2017targeted. @hernan2018causal and @pearl2009causality
+@vdl2011targeted and @vdl2018targeted. @hernan2018causal and @pearl2009causality
 may be of interest to those desiring a more general introduction to statistical

### 2018-06-19 R/methytmle.R

@@ -28,3 +28,18 @@
   ),
   contains = "GenomicRatioSet"
 )
+
+################################################################################
+
+#' @importFrom methods setMethod
+#
+methods::setMethod("show", "methytmle", function(object) {
+  print(noquote(paste("class:", class(object)[1])))
+  print(noquote(paste("data dimension:", paste(as.character(dim(object))[1],
+                      as.character(dim(object))[2]))))
+  print(noquote(paste("annotation:", object@annotation)))
+  print(noquote(paste("target parameter:", object@param)))
+  print(noquote("results:"))
+  print(object@vim)
+})
+

### 2018-06-19 R/utils.R

@@ -82,3 +82,4 @@ wrap_in_try <- function(fun, ...) {
     }, silent = TRUE)
   return(wrapped)
 }
+

### 2018-06-20 README.Rmd
fix class imports; update readme

--- 2018-04-20 README.Rmd
@@
   rmarkdown::github_document
-bibliography: "`r here::here('README-refs.bib')`"
+bibliography: "README-refs.bib"
 ---
@@
 For standard use, install from
-[Bioconductor](https://bioconductor.org/packages/methyvim):
+[Bioconductor](https://bioconductor.org/packages/methyvim) using
+[`BiocManager`](https://CRAN.R-project.org/package=BiocManager):
 
 ```{r bioc-installation, eval = FALSE}
-source("https://bioconductor.org/biocLite.R")
-biocLite("methyvim")
+if (!("BiocManager" %in% installed.packages())) {
+  install.packages("BiocManager")
+}
+BiocManager::install("methyvim")
 ```

### 2018-06-20 R/methytmle-class.R

@@ -6,7 +6,7 @@
 #' @importClassesFrom minfi GenomicRatioSet
 #' @importClassesFrom SummarizedExperiment SummarizedExperiment
 #'  RangedSummarizedExperiment
-#' @import S4Vectors
+#' @importClassesFrom S4Vectors Vector Annotated
 #'
 #' @export .methytmle
 #' @exportClass methytmle

### 2018-06-21 paper/paper_BiocF1000.Rmd
update f1000; misc other edits

@@
       - Department of Statistics, University of California, Berkeley
+    orcid: 0000-0002-XXXX-XXXX 
 abstract: |
-  We present `methyvim`, a general algorithm for the nonparametric estimation of
-  treatment/exposure effects on DNA methylation at CpG sites, complete with
-  straightforward statistical inference for such estimates. This approach
-  leverages variable importance measures, a class of parameters that arise in
-  causal inference and are defined in such a manner that they provide targeted
-  estimates of the relative importance of individual CpG site with respect to
-  a binary exposure/treatment assigned at the level of subjects. Additionally,
-  this procedure is computationally efficient; incorporating a preliminary
-  screening step to isolate a subset of sites for which there is cursory
-  evidence of differential methylation and a multiple testing correction step
-  to control the False Discovery Rate as if all sites were tested. This
-  technique for analysis of differentially methylated positions provides an
-  avenue to move beyond linear models without the loss of interpretable
-  statistical inference.
+  We present `methyvim`, an R package implementing a general algorithm for the
+  nonparametric estimation of treatment effects on DNA methylation at CpG sites
+  throughout the genome, complete with straightforward statistical inference for
+  such estimates. The approach leverages variable importance measures derived
+  from statistical parameters arising causal inference, defined in such a manner
+  that they may be used to obtain targeted estimates of the relative importance
+  of individual CpG site with respect to a binary treatment assigned at the
+  phenotype level. The procedure implemented is computationally efficient,
+  incorporating a preliminary screening step to isolate a subset of sites for
+  which there is cursory evidence of differential methylation as well as a
+  unique multiple testing correction to control the False Discovery Rate with
+  the same rigor as if all sites were tested. This technique for analysis of
+  differentially methylated positions provides an avenue to incorporate flexible
+  state-of-the-art machine learning algorithms into the estimation of
+  differential methylation effects without the loss of interpretable statistical
+  inference.
 keywords: DNA methylation, differential methylation, epigenetics, causal
@@
 DNA methylation is a fundamental epigenetic process known to play an important
-role in the regulation of gene expression. DNA CpG methylation involves the
-addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
-ring structure to form 5-methylcytosine. Numerous biological and medical studies
-have implicated DNA CpG methylation as playing a role in disease and
+role in the regulation of gene expression. DNA methylation occurs at CpG sites
+and involves the addition of a methyl group ($\text{CH}_3$) to the fifth carbon
+of the cytosine ring structure to form 5-methylcytosine. Numerous biological and
+medical studies have implicated DNA methylation as playing a role in disease and
 development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
-have been developed to study the molecular mechanisms of this epigenetic
-process. Modern assays, like the Illumina _Infinium_ Methylation assay,
-allow for quantitative interrogation of DNA methylation of CpG sites scattered
-across the genome at single-nucleotide resolution; moreover, much effort
-has been invested, by the bioinformatics community, in the development of tools
-for properly removing technological effects that may contaminate biological
-signatures measured by such assays. Despite these advances in both biological
-and bioninformatical techniques, most statistical methods available for
-differential analysis of data produced by such assays rely on over-simplified
-(often generalized linear) models. This is because typical statistical
-estimation approaches are not achievable for such high-dimensional data without
-restrictive modeling assumptions. When these assumptions are violated, estimates
-of the population-level effect of an exposure or treatment may have large bias.
-Data-adaptive estimation procedures that utilize machine learning can control
-for high-dimensional confounders when estimating treatment/exposure effects.
-However, interpretable statistical inference (i.e. confidence intervals,
-hypothesis tests) resulting from these data-adaptive estimates is challenging
-[@libbrecht2015machine].
+have been developed to rigorously probe the molecular mechanisms of this
+epigenetic process. Modern assays, like the Illumina _Infinium_ Methylation chip
+technology, allow for quantitative interrogation of DNA methylation at a large
+set of CpG sites scattered across the genome at single-nucleotide resolution;
+moreover, much effort has been invested, by the bioinformatics community, in the
+development of tools for properly removing technological effects that may
+contaminate biological signatures measured by such assays. Despite these
+advances in both biological and bioninformatical techniques, most statistical
+methods available for differential analysis of data produced by such assays rely
+on over-simplified models that fail to incorporate many of the recent advances
+in machine learning. Generally, typical statistical estimation approaches do not
+readily extend to such high-dimensional data without restrictive modeling
+assumptions and the use of inferentially costly hypothesis testing corrections.
+When these standard assumptions are violated, estimates of the population-level
+effect of an exposure or treatment may suffer from large bias. Data-adaptive
+estimation procedures that utilize machine learning can control for confounding
+even in high-dimensional settings; however, interpretable statistical inference
+(i.e., confidence intervals and hypothesis tests) resulting from these
+data-adaptive estimates is challenging to obtain [@libbrecht2015machine].
 
-In this paper, we present an alternative such statistical analysis approaches in
-the form of nonparametric estimation procedures that provide simple and well
-interpretable statistical inference. Inspired by concepts of counterfactual
-effects from the causal inference literature and machine learning algorithms, we
+In this paper, we briefly present an alternative to such statistical analysis
+approaches in the form of a nonparametric estimation procedure that provides
+simple and readily interpretable statistical inference, discussing at length a
+recent implementation of the methodology in the `methyvim` R package. Inspired
+by recent advances in statistical causal inference and machine learning, we
 provide a computationally efficient technique for obtaining targeted estimates
-of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set
-of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
-tested. Under counterfactual assumptions, these estimators have asymptotic
-normal sampling distributions and thus provide reliable inference.
-Counterfactuals are "counter"-fact, defined in terms of an unrealistic yet ideal
-experiment. In the context of DNA methylation and for a binary treatment, we
-define the counterfactual outcomes as the observed methylation value (whether
-Beta- or M-values) a CpG site would have if all subjects were treated and the
-methylation value a CpG site would have if all subjects were not treated. Of
-course, these counterfactual outcomes are impossible to observe but they are
-very helpful for us to asses the importance of a CpG site relative to a subjects
-treatment/exposure and they can be statistically estimated
-[@vdl2006statistical]. Our VIM definition uses the concepts of counterfactuals
-and the framework for estimation is built on _targeted maximum likelihood
-estimation_ or _targeted minimum loss-based estimation_ (__TMLE__)
-[@vdl2006targeted]. This methodology targets the individual importance of each
-CpG site by taking advantage of the prediction power of machine learning
-algorithms. The mathematical properties of this approach have been explored in
-[@tuglus2008targeted] and we focus on our contribution to it's application,
-`methyvim`, an R package that make this technique for differential methylation
-analysis easily accessible.
+of nonparametric _variable importance measures_ (__VIMs__), estimated at a set
+of pre-screened CpG sites, controlling for the False Discovery Rate (FDR) as if
+all sites were tested. Under standard assumptions (e.g., identifiability, strong
+ignorability) [@pearl2009causality], targeted minimum loss-based estimators of
+regular asymptotically linear estimators have sampling distributions that are
+asymptotically normal, allowing for reliable point estimation and the
+construction of Wald-style confidence intervals [@vdl2011targeted,
+@vdl2018targeted]. As counterfactuals are contrary-to-fact -- that is, defined
+in terms of missing data -- in the context of DNA methylation, we define the
+counterfactual outcomes under a binary treatment as the observed methylation
+(whether Beta- or M-) value a CpG site would have if all subjects were given the
+treatment and the methylation value a CpG site would have if treatment were
+withheld from all subjects. Although these counterfactual outcomes are, of
+course, impossible to observe, they have statistical analogs that may be
+reliably estimated from observed data [@pearl2009causality]. We describe an
+algorithm that incorporates the estimation of VIMs using _targeted minimum
+loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
+description and analysis of the statistical methodology to work-in-progress
+[@hejazi2018methylation]. This methodology assesses the individual importance of
+a given CpG site by utilizing state-of-the-art machine learning algorithms in
+the estimation of a targeted VIM, initially proposed and explored in
+[@tuglus2008targeted]. In the presnet work, we focus on a software package,
+`methyvim`, that implements a variant of this methodology, specifically tailored
+to differential methylation analysis.
 
@@
 # Methods
+
+## Data Structure
+
+We consider an observed data structure $O = (W_k, A, Y_j)$, where $Y_j, j = 1,
+\ldots J$ represents a given CpG site of interest, $A \in \{0, 1\}$ represents a
+binary phenotype-level treatment, and $W_k, k = 1, \ldots K(j)$ is a matrix of
+the observed values of CpG sites in the same neighborhood as $j$. We consider
+the case of observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
+take $O \sim \mathcal{P} \in \mathcal{M}$, which is to say that $O$ is governed
+by an unknown probability distribution $\mathcal{P}$, taken only to reside in a
+nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
+data-generating process. We consider having access to measurements on a large
+number $G$ of CpG sites (e.g., 850,000, as measured by the Illumina
+MethylationEPIC BeadChip arrays), and as a matter of practicality, let $J \leq
+G$, so that $J$ indexes only those CpG sites that pass an arbitrary
+pre-screening procedure; note that the equality is only obtained in the case
+that all sites pass the pre-screening procedure. Further, we let $K(j)$ index
+the set of all CpG sites so that $k$, the realization of $K(j)$ for a given CpG
+site $j$, is a set of the CpG sites that neighbor the CpG site $j$, where the
+definition of a neighborhood is left as user-specified; thus, $W_k$ is a matrix
+of observed methylation values for the set of CpG sites $k$ that neighbor a
+particular target site $j$. We consider estimating a target parameter ($\psi_j$)
+as a variable importance measure of the effect of changing a treatment $A$ on
+the methylation of a CpG site $j$, accounting for the observed methylation at
+the set of sites $k$ that neighbor the site $j$. After a complete run of the
+procedure, we have access to $j$ VIM estimates $\psi_j$, each corresponding to
+the degree to which a given CpG site $j$ was responsive to a change in the
+binary treatment $A$, after controlling for any confounding exerted by its
+neighboring CpG sites. When machine learning estimators are employed in
+estimating $\psi_j$, this produces a nonparametric variable importance measure
+of differential methylation, allowing for the identification of differentially
+methylated positions (DMPs).
 
@@
    and add parameters for continuous treatments/exposures (via the
-   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).
+   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) and
+   [`txshift`](https://github.com/nhejazi/txshift) R packages).
 
@@
    multiple testing nature of the estimation problem, a variant of the Benjamini
-   & Hochberg procedure for controlling the False Discovery Rate (FDR) is
-   applied [@benjamini1995controlling]. Specifically, we apply the "modified
-   marginal Benjamini & Hochberg step-up False Discovery Rate controlling
-   procedure for multi-stage analyses" (FDR-MSA), which is guaranteed to
-   control the FDR as if all sites were tested (i.e., without screening)
-   [@tuglus2009modified].
+   and Hochberg procedure for controlling the False Discovery Rate (FDR) is
+   applied [@benjamini1995controlling]. Specifically, we apply a modified
+   marginal Benjamini and Hochberg step-up False Discovery Rate controlling
+   procedure for multi-stage analyses (FDR-MSA), which has established
+   theoretical guarantees to control the FDR at the same rate as if all sites
+   were tested (i.e., without screening) [@tuglus2009modified].
 
@@
 
-For the CpG sites that passed the pre-screening step, the user chooses the
-parameter of interest that will be computed at each site. _In all cases, an
-estimator of the parameter of interest is constructed via targeted minimum
+For the CpG sites that pass the pre-screening step, a user-specified target
+parameter of interest is estimated independently at each site. _In all cases,
+an estimator of the parameter of interest is constructed via targeted minimum
 loss-based estimation_.
@@
 Support for continuous-valued treatments or exposures is _planned but not yet
-available_, though work is underway to incorporate into our methodology the
-following
+available_. Future releases will allow users to assess continuous-valued
+treatments by relying on parameters estimable through implementations available
+in the following software packages:
 
-* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]:
-  The effect of continuous-valued exposure or treatment (the observed
-  methylation at a target CpG site) on an outcome of interest is estimated,
-  controlling for the observed methylation at all other CpG sites in the same
-  neighborhood as the target (treatment) site, based on a parameter that
-  compares values of the treatment against a reference value taken to be the
-  null. In particular, the implementation provided is designed to assess the
-  effect of differential methylation at the target CpG site on a (typically)
-  phenotype-level outcome of interest (e.g., survival), in effect providing an
-  nonparametric evaluation of the impact of methylation at the target site on
-  said outcome.
+* `tmle.npvi` - A nonparametric variable importance measure (NPVI)
+  [@chambaz2012estimation]. The effect of a continuous-valued exposure or
+  treatment (the observed methylation at a target CpG site) on an outcome of
+  interest is estimated, controlling for the observed methylation value at all
+  other CpG sites in the same neighborhood as the target site. This uses a
+  parameter that compares values of the treatment against a user-specified
+  reference value taken to be the null. In particular, the implementation to be
+  provided is designed to assess the effect of differential methylation at the
+  target CpG site on an outcome of interest (e.g., survival), providing a
+  nonparametric evaluation of the impact of methylation at the target site.
+
+* `txshift` - Variable importance using a causal parameter defined as the
+  counterfactual outcome under a posited shift of a continuous-value treatment
+  of interest using stochastic intervention policies [@munoz2012population,
+  diaz2018stochastic, hejazi2018txshift]. The value of an outcome of interest
+  under an unobserved value of the (continuous-valued) treatment, specified
+  through a user-provided _additive shift_ of the observed treatment, may be
+  data-adaptively estimated. This allows for the effect of changes/shifts in the
+  observed methylation at a CpG site on an outcome of interest to be evalauted,
+  providing a nonparametric evalution of the relative importance of a particular
+  target CpG site with respect to another variable measured in the same study.
 
@@
 We can create an object of class `methytmle` from any `GenomicRatioSet` object
-simply invoking the S4 class constructor:
+simply invoking the S4 class constructor `.methytmle`:
 
@@
 library(methyvim)
-mtmle <- .methytmle(grs)
+grs_mtmle <- .methytmle(grs)
+grs_mtmle
 ```
@@
 
-```{r, minfidata-methyvim}
-suppressMessages(
+```{r, minfidata-methyvim, warning=FALSE}
 methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
@@
                                 filter_cutoff = 0.20, obs_per_covar = 2,
-                                  parallel = FALSE, sites_comp = 125,
+                                parallel = FALSE, sites_comp = 100,
                                 tmle_type = "glm"
-                                 )
                                )
@@
 We can view a table of results by examining the `vim` slot of the produced
-`methytmle` object.
+object, displayed using the accessor method of the same name:
 
 ```{r vim-cancer-ate}
-head(slot(methyvim_cancer_ate, "vim"))
+vim(methyvim_cancer_ate)
 ```
+
+Note that simply printing the object will also display an abbreviated version of
+these same results, alongside a bit of additional information and all of the
+standard information normally printed for the `GenomicRatioSet` superclass.
 
@@
 
-Here we introduce the R package `methyvim`, an algorithm for differential
-methylation analysis that moves beyond linear models. Additionally, this
-estimation procedure produces straightforward statistical inference. Inspired by
-concepts of counterfactual effects from the causal inference literature and
-machine learning algorithms, we provide a computationally efficient technique
-for obtaining targeted estimates of nonparametric variable importance measures
-for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
-all sites were tested, and the estimation framework is built on targeted
-minimum loss-based estimation. By taking advantage of the prediction power of
-machine learning algorithms, this methodology targets the individual importance
-of each CpG site, providing a fully nonparametric statistical model for the
-analysis of differential methylation.
+Here we introduce the R package `methyvim`, an implementation of a general
+algorithm for differential methylation analysis that allows for recent advances
+in causal inference and machine learning to be leveraged in computational
+biology settings. The estimation procedure produces straightforward statistical
+inference and takes great care to ensure computationally efficiency of the
+technique for obtaining targeted estimates of nonparametric variable importance
+measures. The software package includes techniques for pre-screening a set of
+CpG sites, controlling for the False Discovery Rate as if all sites were tested,
+and for visualzing the results of the analyses in a variety of ways. The anatomy
+of the software package is dissected and the design described in detail. The
+`methyvim` R package is available via the Bioconductor project.
 
@@
 
-Archived source code as at time of
-publication: https://github.com/nhejazi/methyvim
+Archived source code as at time of publication:
+https://github.com/nhejazi/methyvim/releases/tag/v1.3.1
 

### 2018-06-21 R/methytmle-class.R

@@ -47,6 +47,8 @@ methods::setMethod("show", "methytmle", function(object) {
 ################################################################################
 
 #' @importFrom methods setGeneric setMethod
+#'
+#' @export param
 #
 methods::setGeneric("param", function(object) standardGeneric("param"))
 methods::setMethod("param", "methytmle", function(object) {
@@ -56,6 +58,8 @@ methods::setMethod("param", "methytmle", function(object) {
 ################################################################################
 
 #' @importFrom methods setGeneric setMethod
+#'
+#' @export vim
 #
 methods::setGeneric("vim", function(object) standardGeneric("vim"))
 methods::setMethod("vim", "methytmle", function(object) {

### 2018-06-21 paper/paper_BiocF1000.Rmd
un-export methods; fix F1000 paper bib; new revisions to F1000 manuscript

@@
       - Department of Statistics, University of California, Berkeley
-    orcid: 0000-0002-XXXX-XXXX 
+    orcid: 0000-0003-1432-5511
 abstract: |
@@
   estimation, machine learning
-bibliography: paper_BiocF1000.bib
+bibliography: references.bib
 output: BiocWorkflowTools::f1000_article
 ---
+
+```{r setup, echo=FALSE}
+options(scipen=999)
+```
 
@@
 development of tools for properly removing technological effects that may
-contaminate biological signatures measured by such assays. Despite these
+contaminate biological signatures measured by such assays
+[@fortin2014functional, dedeurwaerder2013comprehensive]. Despite these
 advances in both biological and bioninformatical techniques, most statistical
@@
 provide a computationally efficient technique for obtaining targeted estimates
-of nonparametric _variable importance measures_ (__VIMs__), estimated at a set
-of pre-screened CpG sites, controlling for the False Discovery Rate (FDR) as if
-all sites were tested. Under standard assumptions (e.g., identifiability, strong
-ignorability) [@pearl2009causality], targeted minimum loss-based estimators of
-regular asymptotically linear estimators have sampling distributions that are
-asymptotically normal, allowing for reliable point estimation and the
-construction of Wald-style confidence intervals [@vdl2011targeted,
-@vdl2018targeted]. As counterfactuals are contrary-to-fact -- that is, defined
-in terms of missing data -- in the context of DNA methylation, we define the
-counterfactual outcomes under a binary treatment as the observed methylation
-(whether Beta- or M-) value a CpG site would have if all subjects were given the
-treatment and the methylation value a CpG site would have if treatment were
-withheld from all subjects. Although these counterfactual outcomes are, of
-course, impossible to observe, they have statistical analogs that may be
-reliably estimated from observed data [@pearl2009causality]. We describe an
-algorithm that incorporates the estimation of VIMs using _targeted minimum
-loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
-description and analysis of the statistical methodology to work-in-progress
-[@hejazi2018methylation]. This methodology assesses the individual importance of
-a given CpG site by utilizing state-of-the-art machine learning algorithms in
-the estimation of a targeted VIM, initially proposed and explored in
-[@tuglus2008targeted]. In the presnet work, we focus on a software package,
+of nonparametric _variable importance measures_ (VIMs) [@vdl2006statistical],
+estimated at a set of pre-screened CpG sites, controlling for the False
+Discovery Rate (FDR) as if all sites were tested. Under standard assumptions
+(e.g., identifiability, strong ignorability) [@pearl2009causality], targeted
+minimum loss-based estimators of regular asymptotically linear estimators have
+sampling distributions that are asymptotically normal, allowing for reliable
+point estimation and the construction of Wald-style confidence intervals
+[@vdl2011targeted, @vdl2018targeted]. As counterfactuals are contrary-to-fact --
+that is, defined in terms of missing data -- in the context of DNA methylation,
+we define the counterfactual outcomes under a binary treatment as the observed
+methylation (whether Beta- or M-) value a CpG site would have if all subjects
+were given the treatment and the methylation value a CpG site would have if
+treatment were withheld from all subjects. Although these counterfactual
+outcomes are, of course, impossible to observe, they have statistical analogs
+that may be reliably estimated from observed data [@pearl2009causality]. We
+describe an algorithm that incorporates the estimation of VIMs using _targeted
+minimum loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
+description and analysis of the statistical methodology to work outside the
+scope of the present manuscript. This methodology assesses the individual
+importance of a given CpG site by utilizing state-of-the-art machine learning
+algorithms in the estimation of a targeted VIM, initially proposed and explored
+in [@tuglus2008targeted]. In the presnet work, we focus on a software package,
 `methyvim`, that implements a variant of this methodology, specifically tailored
@@
 estimation and the role this approach plays in statistical causal inference, the
-interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
+interested reader is invited to consult @vdl2011targeted and @vdl2018targeted.
 For a more general introduction to (statistical) causal inference,
@@
   of interest using stochastic intervention policies [@munoz2012population,
-  diaz2018stochastic, hejazi2018txshift]. The value of an outcome of interest
+  @diaz2018stochastic, @hejazi2018txshift]. The value of an outcome of interest
   under an unobserved value of the (continuous-valued) treatment, specified
@@
 
-```{r, minfidata-methyvim, warning=FALSE}
+```{r, minfidata-methyvim, warning=FALSE, cache=TRUE}
 methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
@@
                                 filter_cutoff = 0.20, obs_per_covar = 2,
-                                parallel = FALSE, sites_comp = 100,
+                                parallel = FALSE, sites_comp = 250,
                                 tmle_type = "glm"
@@
 ```
-Note that we set the `obs_per_covar` argument to a relatively low value (2,
-where the recommended default is 20) for the purposes of this example as the
-sample size is only 10. We do this only to exemplify the estimation procedure
-and it is important to point out that such low values for `obs_per_covar` will
-compromise the quality of inference obtained because this setting directly
-affects the definition of the target parameter.
+Note that we set the `obs_per_covar` argument to a relatively low value (just 2,
+even though the recommended value, and default, is 20) for the purposes of this
+example as the sample size is only 10. We do this only to exemplify the
+estimation procedure and it is important to point out that such low values for
+`obs_per_covar` will compromise the quality of inference obtained because this
+setting directly affects the definition of the target parameter.
 
@@
 We can view a table of results by examining the `vim` slot of the produced
-object, displayed using the accessor method of the same name:
+object, most easily displayed by simply printing the resultant object:
 
 ```{r vim-cancer-ate}
-vim(methyvim_cancer_ate)
+methyvim_cancer_ate
 ```
-
-Note that simply printing the object will also display an abbreviated version of
-these same results, alongside a bit of additional information and all of the
-standard information normally printed for the `GenomicRatioSet` superclass.
 
@@
 ```{r fdr-msa}
-fdr_p <- fdr_msa(pvals = slot(methyvim_cancer_ate, "vim")$pval,
+fdr_p <- fdr_msa(pvals = methyvim_cancer_ate@vim$pval,
                  total_obs = nrow(methyvim_cancer_ate))
@@
 Archived source code as at time of publication:
-https://github.com/nhejazi/methyvim/releases/tag/v1.3.1
+https://github.com/nhejazi/methyvim/releases/tag/f1000
+
+Documentation (development version): https://code.nimahejazi.org/methyvim
 
@@
 cases presented, and co-drafted the present manuscript. RP helped in designing
-the software and co-drafted the present article. AH and ML served as advisors
-for the development of this software and general statistical algorithm.
+the software and co-drafted the present manuscript. AH and ML served as advisors
+for the development of this software and the general statistical algorithm it
+implements.
 

### 2018-06-21 paper/paper_BiocF1000.tex
un-export methods; fix F1000 paper bib; new revisions to F1000 manuscript

@@
 \begin{abstract}
-We present \texttt{methyvim}, a general algorithm for the nonparametric estimation of
-treatment/exposure effects on DNA methylation at CpG sites, complete with
-straightforward statistical inference for such estimates. This approach
-leverages variable importance measures, a class of parameters that arise in
-causal inference and are defined in such a manner that they provide targeted
-estimates of the relative importance of individual CpG site with respect to
-a binary exposure/treatment assigned at the level of subjects. Additionally,
-this procedure is computationally efficient; incorporating a preliminary
-screening step to isolate a subset of sites for which there is cursory
-evidence of differential methylation and a multiple testing correction step
-to control the False Discovery Rate as if all sites were tested. This
-technique for analysis of differentially methylated positions provides an
-avenue to move beyond linear models without the loss of interpretable
-statistical inference.
+We present \texttt{methyvim}, an R package implementing a general algorithm for the
+nonparametric estimation of treatment effects on DNA methylation at CpG sites
+throughout the genome, complete with straightforward statistical inference for
+such estimates. The approach leverages variable importance measures derived
+from statistical parameters arising causal inference, defined in such a manner
+that they may be used to obtain targeted estimates of the relative importance
+of individual CpG site with respect to a binary treatment assigned at the
+phenotype level. The procedure implemented is computationally efficient,
+incorporating a preliminary screening step to isolate a subset of sites for
+which there is cursory evidence of differential methylation as well as a
+unique multiple testing correction to control the False Discovery Rate with
+the same rigor as if all sites were tested. This technique for analysis of
+differentially methylated positions provides an avenue to incorporate flexible
+state-of-the-art machine learning algorithms into the estimation of
+differential methylation effects without the loss of interpretable statistical
+inference.
 \end{abstract}
@@
 DNA methylation is a fundamental epigenetic process known to play an important
-role in the regulation of gene expression. DNA CpG methylation involves the
-addition of a methyl group (\(\text{CH}_3\)) to the fifth carbon of the cytosine
-ring structure to form 5-methylcytosine. Numerous biological and medical studies
-have implicated DNA CpG methylation as playing a role in disease and
+role in the regulation of gene expression. DNA methylation occurs at CpG sites
+and involves the addition of a methyl group (\(\text{CH}_3\)) to the fifth carbon
+of the cytosine ring structure to form 5-methylcytosine. Numerous biological and
+medical studies have implicated DNA methylation as playing a role in disease and
 development \citep{robertson2005dna}. Perhaps unsurprisingly then, biotechnologies
-have been developed to study the molecular mechanisms of this epigenetic
-process. Modern assays, like the Illumina \emph{Infinium} Methylation assay,
-allow for quantitative interrogation of DNA methylation of CpG sites scattered
-across the genome at single-nucleotide resolution; moreover, much effort
-has been invested, by the bioinformatics community, in the development of tools
-for properly removing technological effects that may contaminate biological
-signatures measured by such assays. Despite these advances in both biological
-and bioninformatical techniques, most statistical methods available for
-differential analysis of data produced by such assays rely on over-simplified
-(often generalized linear) models. This is because typical statistical
-estimation approaches are not achievable for such high-dimensional data without
-restrictive modeling assumptions. When these assumptions are violated, estimates
-of the population-level effect of an exposure or treatment may have large bias.
-Data-adaptive estimation procedures that utilize machine learning can control
-for high-dimensional confounders when estimating treatment/exposure effects.
-However, interpretable statistical inference (i.e.~confidence intervals,
-hypothesis tests) resulting from these data-adaptive estimates is challenging
-\citep{libbrecht2015machine}.
+have been developed to rigorously probe the molecular mechanisms of this
+epigenetic process. Modern assays, like the Illumina \emph{Infinium} Methylation chip
+technology, allow for quantitative interrogation of DNA methylation at a large
+set of CpG sites scattered across the genome at single-nucleotide resolution;
+moreover, much effort has been invested, by the bioinformatics community, in the
+development of tools for properly removing technological effects that may
+contaminate biological signatures measured by such assays
+\citep[dedeurwaerder2013comprehensive]{fortin2014functional}. Despite these
+advances in both biological and bioninformatical techniques, most statistical
+methods available for differential analysis of data produced by such assays rely
+on over-simplified models that fail to incorporate many of the recent advances
+in machine learning. Generally, typical statistical estimation approaches do not
+readily extend to such high-dimensional data without restrictive modeling
+assumptions and the use of inferentially costly hypothesis testing corrections.
+When these standard assumptions are violated, estimates of the population-level
+effect of an exposure or treatment may suffer from large bias. Data-adaptive
+estimation procedures that utilize machine learning can control for confounding
+even in high-dimensional settings; however, interpretable statistical inference
+(i.e., confidence intervals and hypothesis tests) resulting from these
+data-adaptive estimates is challenging to obtain \citep{libbrecht2015machine}.
 
-In this paper, we present an alternative such statistical analysis approaches in
-the form of nonparametric estimation procedures that provide simple and well
-interpretable statistical inference. Inspired by concepts of counterfactual
-effects from the causal inference literature and machine learning algorithms, we
+In this paper, we briefly present an alternative to such statistical analysis
+approaches in the form of a nonparametric estimation procedure that provides
+simple and readily interpretable statistical inference, discussing at length a
+recent implementation of the methodology in the \texttt{methyvim} R package. Inspired
+by recent advances in statistical causal inference and machine learning, we
 provide a computationally efficient technique for obtaining targeted estimates
-of nonparametric \emph{variable importance measures} (\textbf{VIM}) for a pre-screened set
-of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
-tested. Under counterfactual assumptions, these estimators have asymptotic
-normal sampling distributions and thus provide reliable inference.
-Counterfactuals are ``counter''-fact, defined in terms of an unrealistic yet ideal
-experiment. In the context of DNA methylation and for a binary treatment, we
-define the counterfactual outcomes as the observed methylation value (whether
-Beta- or M-values) a CpG site would have if all subjects were treated and the
-methylation value a CpG site would have if all subjects were not treated. Of
-course, these counterfactual outcomes are impossible to observe but they are
-very helpful for us to asses the importance of a CpG site relative to a subjects
-treatment/exposure and they can be statistically estimated
-\citep{vdl2006statistical}. Our VIM definition uses the concepts of counterfactuals
-and the framework for estimation is built on \emph{targeted maximum likelihood
-estimation} or \emph{targeted minimum loss-based estimation} (\textbf{TMLE})
-\citep{vdl2006targeted}. This methodology targets the individual importance of each
-CpG site by taking advantage of the prediction power of machine learning
-algorithms. The mathematical properties of this approach have been explored in
-\citep{tuglus2008targeted} and we focus on our contribution to it's application,
-\texttt{methyvim}, an R package that make this technique for differential methylation
-analysis easily accessible.
+of nonparametric \emph{variable importance measures} (VIMs) \citep{vdl2006statistical},
+estimated at a set of pre-screened CpG sites, controlling for the False
+Discovery Rate (FDR) as if all sites were tested. Under standard assumptions
+(e.g., identifiability, strong ignorability) \citep{pearl2009causality}, targeted
+minimum loss-based estimators of regular asymptotically linear estimators have
+sampling distributions that are asymptotically normal, allowing for reliable
+point estimation and the construction of Wald-style confidence intervals
+\citep[\citet{vdl2018targeted}]{vdl2011targeted}. As counterfactuals are contrary-to-fact --
+that is, defined in terms of missing data -- in the context of DNA methylation,
+we define the counterfactual outcomes under a binary treatment as the observed
+methylation (whether Beta- or M-) value a CpG site would have if all subjects
+were given the treatment and the methylation value a CpG site would have if
+treatment were withheld from all subjects. Although these counterfactual
+outcomes are, of course, impossible to observe, they have statistical analogs
+that may be reliably estimated from observed data \citep{pearl2009causality}. We
+describe an algorithm that incorporates the estimation of VIMs using \emph{targeted
+minimum loss-based estimation} (\textbf{TMLE}) \citep{vdl2006targeted}, deferring detailed
+description and analysis of the statistical methodology to work outside the
+scope of the present manuscript. This methodology assesses the individual
+importance of a given CpG site by utilizing state-of-the-art machine learning
+algorithms in the estimation of a targeted VIM, initially proposed and explored
+in \citep{tuglus2008targeted}. In the presnet work, we focus on a software package,
+\texttt{methyvim}, that implements a variant of this methodology, specifically tailored
+to differential methylation analysis.
 
@@
 estimation and the role this approach plays in statistical causal inference, the
-interested reader is invited to consult \citet{vdl2011targeted} and \citet{vdl2017targeted}.
+interested reader is invited to consult \citet{vdl2011targeted} and \citet{vdl2018targeted}.
 For a more general introduction to (statistical) causal inference,
@@
 \section{Methods}\label{methods}}
+
+\hypertarget{data-structure}{%
+\subsection{Data Structure}\label{data-structure}}
+
+We consider an observed data structure \(O = (W_k, A, Y_j)\), where \(Y_j, j = 1, \ldots J\) represents a given CpG site of interest, \(A \in \{0, 1\}\) represents a
+binary phenotype-level treatment, and \(W_k, k = 1, \ldots K(j)\) is a matrix of
+the observed values of CpG sites in the same neighborhood as \(j\). We consider
+the case of observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where we
+take \(O \sim \mathcal{P} \in \mathcal{M}\), which is to say that \(O\) is governed
+by an unknown probability distribution \(\mathcal{P}\), taken only to reside in a
+nonparametric statistical model \(\mathcal{M}\) that places no restrictions on the
+data-generating process. We consider having access to measurements on a large
+number \(G\) of CpG sites (e.g., 850,000, as measured by the Illumina
+MethylationEPIC BeadChip arrays), and as a matter of practicality, let \(J \leq G\), so that \(J\) indexes only those CpG sites that pass an arbitrary
+pre-screening procedure; note that the equality is only obtained in the case
+that all sites pass the pre-screening procedure. Further, we let \(K(j)\) index
+the set of all CpG sites so that \(k\), the realization of \(K(j)\) for a given CpG
+site \(j\), is a set of the CpG sites that neighbor the CpG site \(j\), where the
+definition of a neighborhood is left as user-specified; thus, \(W_k\) is a matrix
+of observed methylation values for the set of CpG sites \(k\) that neighbor a
+particular target site \(j\). We consider estimating a target parameter (\(\psi_j\))
+as a variable importance measure of the effect of changing a treatment \(A\) on
+the methylation of a CpG site \(j\), accounting for the observed methylation at
+the set of sites \(k\) that neighbor the site \(j\). After a complete run of the
+procedure, we have access to \(j\) VIM estimates \(\psi_j\), each corresponding to
+the degree to which a given CpG site \(j\) was responsive to a change in the
+binary treatment \(A\), after controlling for any confounding exerted by its
+neighboring CpG sites. When machine learning estimators are employed in
+estimating \(\psi_j\), this produces a nonparametric variable importance measure
+of differential methylation, allowing for the identification of differentially
+methylated positions (DMPs).
 
@@
   and add parameters for continuous treatments/exposures (via the
-  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}} package).
+  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}} and
+  \href{https://github.com/nhejazi/txshift}{\texttt{txshift}} R packages).
 \item
@@
   multiple testing nature of the estimation problem, a variant of the Benjamini
-  \& Hochberg procedure for controlling the False Discovery Rate (FDR) is
-  applied \citep{benjamini1995controlling}. Specifically, we apply the ``modified
-  marginal Benjamini \& Hochberg step-up False Discovery Rate controlling
-  procedure for multi-stage analyses'' (FDR-MSA), which is guaranteed to
-  control the FDR as if all sites were tested (i.e., without screening)
-  \citep{tuglus2009modified}.
+  and Hochberg procedure for controlling the False Discovery Rate (FDR) is
+  applied \citep{benjamini1995controlling}. Specifically, we apply a modified
+  marginal Benjamini and Hochberg step-up False Discovery Rate controlling
+  procedure for multi-stage analyses (FDR-MSA), which has established
+  theoretical guarantees to control the FDR at the same rate as if all sites
+  were tested (i.e., without screening) \citep{tuglus2009modified}.
 \end{enumerate}
@@
 
-For the CpG sites that passed the pre-screening step, the user chooses the
-parameter of interest that will be computed at each site. \emph{In all cases, an
-estimator of the parameter of interest is constructed via targeted minimum
+For the CpG sites that pass the pre-screening step, a user-specified target
+parameter of interest is estimated independently at each site. \emph{In all cases,
+an estimator of the parameter of interest is constructed via targeted minimum
 loss-based estimation}.
@@
 Support for continuous-valued treatments or exposures is \emph{planned but not yet
-available}, though work is underway to incorporate into our methodology the
-following
+available}. Future releases will allow users to assess continuous-valued
+treatments by relying on parameters estimable through implementations available
+in the following software packages:
 
@@
 \item
-  A nonparametric variable importance measure (NPVI) \citep{chambaz2012estimation}:
-  The effect of continuous-valued exposure or treatment (the observed
-  methylation at a target CpG site) on an outcome of interest is estimated,
-  controlling for the observed methylation at all other CpG sites in the same
-  neighborhood as the target (treatment) site, based on a parameter that
-  compares values of the treatment against a reference value taken to be the
-  null. In particular, the implementation provided is designed to assess the
-  effect of differential methylation at the target CpG site on a (typically)
-  phenotype-level outcome of interest (e.g., survival), in effect providing an
-  nonparametric evaluation of the impact of methylation at the target site on
-  said outcome.
+  \texttt{tmle.npvi} - A nonparametric variable importance measure (NPVI)
+  \citep{chambaz2012estimation}. The effect of a continuous-valued exposure or
+  treatment (the observed methylation at a target CpG site) on an outcome of
+  interest is estimated, controlling for the observed methylation value at all
+  other CpG sites in the same neighborhood as the target site. This uses a
+  parameter that compares values of the treatment against a user-specified
+  reference value taken to be the null. In particular, the implementation to be
+  provided is designed to assess the effect of differential methylation at the
+  target CpG site on an outcome of interest (e.g., survival), providing a
+  nonparametric evaluation of the impact of methylation at the target site.
+\item
+  \texttt{txshift} - Variable importance using a causal parameter defined as the
+  counterfactual outcome under a posited shift of a continuous-value treatment
+  of interest using stochastic intervention policies \citep[
+  \citet{diaz2018stochastic}, \citet{hejazi2018txshift}]{munoz2012population}. The value of an outcome of interest
+  under an unobserved value of the (continuous-valued) treatment, specified
+  through a user-provided \emph{additive shift} of the observed treatment, may be
+  data-adaptively estimated. This allows for the effect of changes/shifts in the
+  observed methylation at a CpG site on an outcome of interest to be evalauted,
+  providing a nonparametric evalution of the relative importance of a particular
+  target CpG site with respect to another variable measured in the same study.
 \end{itemize}
@@
 We can create an object of class \texttt{methytmle} from any \texttt{GenomicRatioSet} object
-simply invoking the S4 class constructor:
+simply invoking the S4 class constructor \texttt{.methytmle}:
 
@@
 \begin{verbatim}
-## methyvim v1.1.0: Targeted Variable Importance for Differential Methylation Analysis
+## methyvim v1.3.1: Targeted Variable Importance for Differential Methylation Analysis
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\NormalTok{mtmle <-}\StringTok{ }\KeywordTok{.methytmle}\NormalTok{(grs)}
+\NormalTok{grs_mtmle <-}\StringTok{ }\KeywordTok{.methytmle}\NormalTok{(grs)}
+\NormalTok{grs_mtmle}
 \end{Highlighting}
 \end{Shaded}
+
+\begin{verbatim}
+## class: methytmle 
+## dim: 485512 6 
+## metadata(0):
+## assays(2): Beta CN
+## rownames(485512): cg13869341 cg14008030 ... cg08265308 cg14273923
+## rowData names(0):
+## colnames(6): 5723646052_R02C02 5723646052_R04C01 ...
+##   5723646053_R05C02 5723646053_R06C02
+## colData names(13): Sample_Name Sample_Well ... Basename filenames
+## Annotation
+##   array: IlluminaHumanMethylation450k
+##   annotation: ilmn12.hg19
+## Preprocessing
+##   Method: Raw (no normalization or bg correction)
+##   minfi version: 1.21.2
+##   Manifest version: 0.4.0
+## Target Parameter: 
+## Results: 
+## Object of class "data.frame"
+## data frame with 0 columns and 0 rows
+\end{verbatim}
 
@@
 \begin{Highlighting}[]
-\KeywordTok{suppressMessages}\NormalTok{(}
-\NormalTok{  methyvim_cancer_ate <-}\StringTok{ }\KeywordTok{methyvim}\NormalTok{(}\DataTypeTok{data_grs =}\NormalTok{ grs, }\DataTypeTok{var_int =}\NormalTok{ var_int,}
+\NormalTok{methyvim_cancer_ate <-}\StringTok{ }\KeywordTok{methyvim}\NormalTok{(}\DataTypeTok{data_grs =}\NormalTok{ grs, }\DataTypeTok{var_int =}\NormalTok{ var_int,}
                                 \DataTypeTok{vim =} \StringTok{"ate"}\NormalTok{, }\DataTypeTok{type =} \StringTok{"Beta"}\NormalTok{, }\DataTypeTok{filter =} \StringTok{"limma"}\NormalTok{,}
                                 \DataTypeTok{filter_cutoff =} \FloatTok{0.20}\NormalTok{, }\DataTypeTok{obs_per_covar =} \DecValTok{2}\NormalTok{,}
-                                  \DataTypeTok{parallel =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{sites_comp =} \DecValTok{125}\NormalTok{,}
+                                \DataTypeTok{parallel =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{sites_comp =} \DecValTok{250}\NormalTok{,}
                                 \DataTypeTok{tmle_type =} \StringTok{"glm"}
 \NormalTok{                               )}
-\NormalTok{)}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-## Warning in set_parallel(parallel = parallel, future_param = future_param, : Sequential evaluation is strongly discouraged. 
-##  Proceed with caution.
-\end{verbatim}
-
-Note that we set the \texttt{obs\_per\_covar} argument to a relatively low value (2,
-where the recommended default is 20) for the purposes of this example as the
-sample size is only 10. We do this only to exemplify the estimation procedure
-and it is important to point out that such low values for \texttt{obs\_per\_covar} will
-compromise the quality of inference obtained because this setting directly
-affects the definition of the target parameter.
+Note that we set the \texttt{obs\_per\_covar} argument to a relatively low value (just 2,
+even though the recommended value, and default, is 20) for the purposes of this
+example as the sample size is only 10. We do this only to exemplify the
+estimation procedure and it is important to point out that such low values for
+\texttt{obs\_per\_covar} will compromise the quality of inference obtained because this
+setting directly affects the definition of the target parameter.
 
@@
 We can view a table of results by examining the \texttt{vim} slot of the produced
-\texttt{methytmle} object.
+object, most easily displayed by simply printing the resultant object:
 
@@
 \begin{Highlighting}[]
-\KeywordTok{head}\NormalTok{(}\KeywordTok{slot}\NormalTok{(methyvim_cancer_ate, }\StringTok{"vim"}\NormalTok{))}
+\NormalTok{methyvim_cancer_ate}
 \end{Highlighting}
@@
 \begin{verbatim}
-##                 lwr_ci      est_ate       upr_ci      var_ate         pval
-## cg14008030 -0.11956597 -0.031415962  0.056734049 2.022705e-03 4.848468e-01
-## cg20253340 -0.08850637 -0.058866142 -0.029225917 2.286919e-04 9.917426e-05
-## cg21870274 -0.09499057 -0.029118982  0.036752609 1.129495e-03 3.862537e-01
-## cg17308840 -0.04626018 -0.007152452  0.031955277 3.981191e-04 7.199943e-01
-## cg00645010 -0.02697515 -0.013809263 -0.000643373 4.512199e-05 3.980386e-02
-## cg27534567  0.06745648  0.115711854  0.163967223 6.061486e-04 2.602936e-06
-##            n_neighbors n_neighbors_control max_cor_neighbors
-## cg14008030           0                   0                NA
-## cg20253340           0                   0                NA
-## cg21870274           2                   1         0.9443580
-## cg17308840           2                   1         0.9443580
-## cg00645010           2                   2         0.5236810
-## cg27534567           1                   0         0.9362968
+## class: methytmle 
+## dim: 485512 6 
+## metadata(0):
+## assays(2): Beta CN
+## rownames(485512): cg13869341 cg14008030 ... cg08265308 cg14273923
+## rowData names(0):
+## colnames(6): 5723646052_R02C02 5723646052_R04C01 ...
+##   5723646053_R05C02 5723646053_R06C02
+## colData names(13): Sample_Name Sample_Well ... Basename filenames
+## Annotation
+##   array: IlluminaHumanMethylation450k
+##   annotation: ilmn12.hg19
+## Preprocessing
+##   Method: Raw (no normalization or bg correction)
+##   minfi version: 1.21.2
+##   Manifest version: 0.4.0
+## Target Parameter: Average Treatment Effect
+## Results: 
+##                    lwr_ci       est_ate        upr_ci        var_ate
+## cg14008030 -0.11956597277 -0.0314159619  0.0567340489 0.002022705230
+## cg20253340 -0.08850636712 -0.0588661418 -0.0292259165 0.000228691940
+## cg21870274 -0.09499057246 -0.0291189817  0.0367526091 0.001129494604
+## cg17308840 -0.04626018088 -0.0071524518  0.0319552773 0.000398119136
+## cg00645010 -0.02677328296 -0.0134655543 -0.0001578256 0.000046099449
+## cg27534567  0.06745648461  0.1157118536  0.1639672225 0.000606148645
+## cg08258224  0.13650451332  0.3050884951  0.4736724770 0.007398104677
+## cg20275697 -0.30210263597 -0.1231299608  0.0558427144 0.008337988977
+## cg24373735 -0.04283533168  0.0076666975  0.0581687266 0.000663904349
+## cg12445832 -0.06082410546  0.0150395574  0.0909032203 0.001498150599
+## cg01097950 -0.04392159158  0.0658323796  0.1755863507 0.003135655504
+## cg01782097 -0.01082071508  0.0010232901  0.0128672954 0.000036516155
+##                                                      pval n_neighbors
+## cg14008030 0.48484680819856229572195616128738038241863251           0
+## cg20253340 0.00009917426050587487018107835101687896894873           0
+## cg21870274 0.38625371823607712595816110479063354432582855           2
+## cg17308840 0.71999433317147576438799205789109691977500916           2
+## cg00645010 0.04734006904146480232409288646522327326238155           2
+## cg27534567 0.00000260293555116907031197386393484016764432           1
+## cg08258224 0.00038959136906232085580809032521187873499002           1
+## cg20275697 0.17751545277429223168574878854997223243117332           0
+## cg24373735 0.76604893436462262457098404411226511001586914           0
+## cg12445832 0.69760217179702999068524604808771982789039612           0
+## cg01097950 0.23973765213124559325663653908122796565294266           0
+## cg01782097 0.86553022244577304533663664187770336866378784           1
+##            n_neighbors_control max_cor_neighbors
+## cg14008030                   0                NA
+## cg20253340                   0                NA
+## cg21870274                   1        0.94435796
+## cg17308840                   1        0.94435796
+## cg00645010                   2        0.52368097
+## cg27534567                   0        0.93629683
+## cg08258224                   0        0.93629683
+## cg20275697                   0                NA
+## cg24373735                   0                NA
+## cg12445832                   0                NA
+## cg01097950                   0                NA
+## cg01782097                   1       -0.39410834
+##  [ reached getOption("max.print") -- omitted 238 rows ]
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\NormalTok{fdr_p <-}\StringTok{ }\KeywordTok{fdr_msa}\NormalTok{(}\DataTypeTok{pvals =} \KeywordTok{slot}\NormalTok{(methyvim_cancer_ate, }\StringTok{"vim"}\NormalTok{)}\OperatorTok{$}\NormalTok{pval,}
+\NormalTok{fdr_p <-}\StringTok{ }\KeywordTok{fdr_msa}\NormalTok{(}\DataTypeTok{pvals =}\NormalTok{ methyvim_cancer_ate}\OperatorTok{@}\NormalTok{vim}\OperatorTok{$}\NormalTok{pval,}
                  \DataTypeTok{total_obs =} \KeywordTok{nrow}\NormalTok{(methyvim_cancer_ate))}
@@
 
-Here we introduce the R package \texttt{methyvim}, an algorithm for differential
-methylation analysis that moves beyond linear models. Additionally, this
-estimation procedure produces straightforward statistical inference. Inspired by
-concepts of counterfactual effects from the causal inference literature and
-machine learning algorithms, we provide a computationally efficient technique
-for obtaining targeted estimates of nonparametric variable importance measures
-for a pre-screened set of CpGs, controlling for the False Discovery Rate as if
-all sites were tested, and the estimation framework is built on targeted
-minimum loss-based estimation. By taking advantage of the prediction power of
-machine learning algorithms, this methodology targets the individual importance
-of each CpG site, providing a fully nonparametric statistical model for the
-analysis of differential methylation.
+Here we introduce the R package \texttt{methyvim}, an implementation of a general
+algorithm for differential methylation analysis that allows for recent advances
+in causal inference and machine learning to be leveraged in computational
+biology settings. The estimation procedure produces straightforward statistical
+inference and takes great care to ensure computationally efficiency of the
+technique for obtaining targeted estimates of nonparametric variable importance
+measures. The software package includes techniques for pre-screening a set of
+CpG sites, controlling for the False Discovery Rate as if all sites were tested,
+and for visualzing the results of the analyses in a variety of ways. The anatomy
+of the software package is dissected and the design described in detail. The
+\texttt{methyvim} R package is available via the Bioconductor project.
 
@@
 
-Archived source code as at time of
-publication: \url{https://github.com/nhejazi/methyvim}
+Archived source code as at time of publication:
+\url{https://github.com/nhejazi/methyvim/releases/tag/f1000}
+
+Documentation (development version): \url{https://code.nimahejazi.org/methyvim}
 
@@
 \renewcommand\refname{References}
-{\small\bibliography{paper\_BiocF1000.bib}}
+{\small\bibliography{references.bib}}
 

### 2018-06-21 R/methytmle-class.R

@@ -47,8 +47,6 @@ methods::setMethod("show", "methytmle", function(object) {
 ################################################################################
 
 #' @importFrom methods setGeneric setMethod
-#'
-#' @export param
 #
 methods::setGeneric("param", function(object) standardGeneric("param"))
 methods::setMethod("param", "methytmle", function(object) {
@@ -58,8 +56,6 @@ methods::setMethod("param", "methytmle", function(object) {
 ################################################################################
 
 #' @importFrom methods setGeneric setMethod
-#'
-#' @export vim
 #
 methods::setGeneric("vim", function(object) standardGeneric("vim"))
 methods::setMethod("vim", "methytmle", function(object) {

### 2018-06-22 paper/paper_BiocF1000.Rmd
integrate RVP revisions/comments

@@
   that they may be used to obtain targeted estimates of the relative importance
-  of individual CpG site with respect to a binary treatment assigned at the
+  of individual CpG sites with respect to a binary treatment assigned at the
   phenotype level. The procedure implemented is computationally efficient,
@@
 keywords: DNA methylation, differential methylation, epigenetics, causal
-  inference, statistical variable importance, targeted minimum loss-based
-  estimation, machine learning
+  inference, statistical variable importance, machine learning, targeted minimum
+  loss-based estimation
 bibliography: references.bib
@@
 DNA methylation is a fundamental epigenetic process known to play an important
-role in the regulation of gene expression. DNA methylation occurs at CpG sites
-and involves the addition of a methyl group ($\text{CH}_3$) to the fifth carbon
-of the cytosine ring structure to form 5-methylcytosine. Numerous biological and
-medical studies have implicated DNA methylation as playing a role in disease and
-development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
-have been developed to rigorously probe the molecular mechanisms of this
-epigenetic process. Modern assays, like the Illumina _Infinium_ Methylation chip
-technology, allow for quantitative interrogation of DNA methylation at a large
-set of CpG sites scattered across the genome at single-nucleotide resolution;
-moreover, much effort has been invested, by the bioinformatics community, in the
+role in the regulation of gene expression. DNA methylation mostly commonly
+occurs at CpG sites and involves the addition of a methyl group ($\text{CH}_3$)
+to the fifth carbon of the cytosine ring structure to form 5-methylcytosine.
+Numerous biological and medical studies have implicated DNA methylation as
+playing a role in disease and development [@robertson2005dna]. Perhaps
+unsurprisingly then, biotechnologies have been developed to rigorously probe the
+molecular mechanisms of this epigenetic process. Modern assays, like the
+Illumina _Infinium_ HumanMethylation BeadChip assay, allow for quantitative
+interrogation of DNA methylation, at single-nucleotide resolution, across a
+comprehensive set of CpG sites scattered across the genome; moreover, the
+computational biology community has invested significant effort in the
 development of tools for properly removing technological effects that may
 contaminate biological signatures measured by such assays
-[@fortin2014functional, dedeurwaerder2013comprehensive]. Despite these
+[@fortin2014functional, @dedeurwaerder2013comprehensive]. Despite these
 advances in both biological and bioninformatical techniques, most statistical
 methods available for differential analysis of data produced by such assays rely
-on over-simplified models that fail to incorporate many of the recent advances
-in machine learning. Generally, typical statistical estimation approaches do not
-readily extend to such high-dimensional data without restrictive modeling
-assumptions and the use of inferentially costly hypothesis testing corrections.
-When these standard assumptions are violated, estimates of the population-level
-effect of an exposure or treatment may suffer from large bias. Data-adaptive
-estimation procedures that utilize machine learning can control for confounding
-even in high-dimensional settings; however, interpretable statistical inference
-(i.e., confidence intervals and hypothesis tests) resulting from these
-data-adaptive estimates is challenging to obtain [@libbrecht2015machine].
+on over-simplified models that do not readily extend to such high-dimensional
+data structures without restrictive modeling assumptions and the use of
+inferentially costly hypothesis testing corrections. When these standard
+assumptions are violated, estimates of the population-level effect of an
+exposure or treatment may suffer from large bias. What's more, reliance on
+restrictive and misspecified statistical models naturally leads to biased effect
+estimates that are not only misleading in assessing effect sizes but also result
+in false discoveries as these biased estimates are subject to testing and
+inferential procedures. Such predictably unreliable methods serve only to
+produce findings that are later invalidated by replication studies and add still
+further complexity to discovering biological targets for potential therapeutics.
+Data-adaptive estimation procedures that utilize machine learning provide a way
+to overcome many of the problems common in classical methods, controlling for
+potential confounding even in high-dimensional settings; however, interpretable
+statistical inference (i.e., confidence intervals and hypothesis tests) from
+such data-adaptive estimates is challenging to obtain [@libbrecht2015machine].
 
@@
 we define the counterfactual outcomes under a binary treatment as the observed
-methylation (whether Beta- or M-) value a CpG site would have if all subjects
-were given the treatment and the methylation value a CpG site would have if
+methylation (whether Beta- or M-) values a CpG site would have if all subjects
+were given the treatment and the methylation values a CpG site would have if
 treatment were withheld from all subjects. Although these counterfactual
@@
 algorithms in the estimation of a targeted VIM, initially proposed and explored
-in [@tuglus2008targeted]. In the presnet work, we focus on a software package,
+in [@tuglus2008targeted]. In the present work, we focus on a software package,
 `methyvim`, that implements a variant of this methodology, specifically tailored
@@
 binary phenotype-level treatment, and $W_k, k = 1, \ldots K(j)$ is a matrix of
-the observed values of CpG sites in the same neighborhood as $j$. We consider
-the case of observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
-take $O \sim \mathcal{P} \in \mathcal{M}$, which is to say that $O$ is governed
-by an unknown probability distribution $\mathcal{P}$, taken only to reside in a
-nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
-data-generating process. We consider having access to measurements on a large
-number $G$ of CpG sites (e.g., 850,000, as measured by the Illumina
-MethylationEPIC BeadChip arrays), and as a matter of practicality, let $J \leq
-G$, so that $J$ indexes only those CpG sites that pass an arbitrary
-pre-screening procedure; note that the equality is only obtained in the case
-that all sites pass the pre-screening procedure. Further, we let $K(j)$ index
-the set of all CpG sites so that $k$, the realization of $K(j)$ for a given CpG
-site $j$, is a set of the CpG sites that neighbor the CpG site $j$, where the
-definition of a neighborhood is left as user-specified; thus, $W_k$ is a matrix
-of observed methylation values for the set of CpG sites $k$ that neighbor a
-particular target site $j$. We consider estimating a target parameter ($\psi_j$)
-as a variable importance measure of the effect of changing a treatment $A$ on
-the methylation of a CpG site $j$, accounting for the observed methylation at
-the set of sites $k$ that neighbor the site $j$. After a complete run of the
-procedure, we have access to $j$ VIM estimates $\psi_j$, each corresponding to
-the degree to which a given CpG site $j$ was responsive to a change in the
-binary treatment $A$, after controlling for any confounding exerted by its
-neighboring CpG sites. When machine learning estimators are employed in
+the observed values of CpG sites in the same neighborhood as $j$ as well as any
+potential phenotype-level confounders (e.g., sex, age). We consider having
+access to measurements on a large number $G$ of CpG sites (e.g., $850,000$, as
+measured by the Illumina MethylationEPIC BeadChip arrays), and as a matter of
+practicality, let $J \leq G$, so that $J$ indexes only those CpG sites that pass
+an arbitrary pre-screening procedure; note that equality is only obtained in the
+case that all sites pass the pre-screening procedure. Further, we let $K(j)$
+index the set of all CpG sites so that $k$, the realization of $K(j)$ for a
+given CpG site $j$, is a set of the CpG sites that neighbor the CpG site $j$,
+where the definition of a neighborhood is left as user-specified; thus, $W_k$ is
+a matrix of observed methylation values for the set of CpG sites $k$ that
+neighbor a particular target site $j$ as well as any potential phenotype-level
+confounders, where the latter does not vary between realizations of $K(j)$. As
+a variable importance measure, we see to estimate a target parameter ($\psi_j$)
+that quantifies the effect of changing a treatment $A$ on the methylation of a
+CpG site $j$, accounting for any potential confounding from phenotype-level
+covariates and the observed methylation at the set of sites $k$ that neighbor
+the site $j$. After a complete run of the procedure, we have access to $j$ VIM
+estimates $\psi_j$, each corresponding to the degree to which a given CpG site
+$j$ was responsive to a change in the binary treatment $A$, after controlling
+for potential confounding. When machine learning estimators are employed in
 estimating $\psi_j$, this produces a nonparametric variable importance measure
 of differential methylation, allowing for the identification of differentially
-methylated positions (DMPs).
+methylated positions (DMPs). We note that $\psi(j)$, for any given $j$, is a
+function of the _true_ data-generating distribution; so, to formalize, we
+consider observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
+have $O \sim P \in \mathcal{M}$, which is to say that the random variable $O$ is
+governed by an unknown probability distribution $P$, assumed only to reside in a
+nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
+data-generating process.
 
@@
 
-* `tmle.npvi` - A nonparametric variable importance measure (NPVI)
-  [@chambaz2012estimation]. The effect of a continuous-valued exposure or
+* A nonparametric variable importance measure (NPVI) [@chambaz2012estimation]
+  (R package `tmle.npvi`). The effect of a continuous-valued exposure or
   treatment (the observed methylation at a target CpG site) on an outcome of
@@
   parameter that compares values of the treatment against a user-specified
-  reference value taken to be the null. In particular, the implementation to be
-  provided is designed to assess the effect of differential methylation at the
-  target CpG site on an outcome of interest (e.g., survival), providing a
+  reference value taken to be the null value. In particular, the implementation
+  to be provided is designed to assess the effect of differential methylation at
+  the target CpG site on an outcome of interest (e.g., survival), providing a
   nonparametric evaluation of the impact of methylation at the target site.
 
-* `txshift` - Variable importance using a causal parameter defined as the
-  counterfactual outcome under a posited shift of a continuous-value treatment
-  of interest using stochastic intervention policies [@munoz2012population,
-  @diaz2018stochastic, @hejazi2018txshift]. The value of an outcome of interest
-  under an unobserved value of the (continuous-valued) treatment, specified
-  through a user-provided _additive shift_ of the observed treatment, may be
-  data-adaptively estimated. This allows for the effect of changes/shifts in the
-  observed methylation at a CpG site on an outcome of interest to be evalauted,
-  providing a nonparametric evalution of the relative importance of a particular
-  target CpG site with respect to another variable measured in the same study.
+* The causal effect of shifting the value of an observed intervention, defined
+  as the counterfactual outcome under a posited shift of a continuous-value
+  treatment of interest using stochastic intervention policies
+  [@munoz2012population, @diaz2018stochastic, @hejazi2018txshift] (R package
+  `txshift`). The value of an outcome of interest under an unobserved value of
+  the (continuous-valued) treatment, specified through a user-provided
+  _additive shift_ of the observed treatment, may be data-adaptively estimated.
+  This allows for the effect of changes/shifts in the observed methylation at a
+  CpG site on an outcome of interest to be evaluated, providing a nonparametric
+  evalution of the relative importance of a particular target CpG site with
+  respect to another variable measured in the same study.
 
@@
 ```{r methyvim-heatmap}
-methyheat(methyvim_cancer_ate)
+methyheat(methyvim_cancer_ate, smooth.heat = TRUE, left.label = "none")
 ```
 
-Invoking `methyheat` in this manner produces a plot of the top sites ($25$, by
-default) based on the raw p-value, using the raw methylation measures in the
-plot. This uses the exceptional `superheat` R package [@barter2017superheat].
+__Remark__: Invoking `methyheat` in this manner produces a plot of the top sites
+($25$, by default) based on the raw p-value, using the raw methylation measures
+in the plot. This uses the exceptional `superheat` R package
+[@barter2017superheat], to which we can easily pass additional parameters. In
+particulat, we hide the CpG site labels that would appear by default on the left
+of the heatmap (by setting `left.label = "none"`) to emphasize that this is only
+an example and _not_ a scientific discovery.
 

### 2018-06-22 paper/paper_BiocF1000.tex
integrate RVP revisions/comments

@@
 that they may be used to obtain targeted estimates of the relative importance
-of individual CpG site with respect to a binary treatment assigned at the
+of individual CpG sites with respect to a binary treatment assigned at the
 phenotype level. The procedure implemented is computationally efficient,
@@
 \section*{Keywords}
-DNA methylation, differential methylation, epigenetics, causal inference, statistical variable importance, targeted minimum loss-based estimation, machine learning
+DNA methylation, differential methylation, epigenetics, causal inference, statistical variable importance, machine learning, targeted minimum loss-based estimation
 
@@
 DNA methylation is a fundamental epigenetic process known to play an important
-role in the regulation of gene expression. DNA methylation occurs at CpG sites
-and involves the addition of a methyl group (\(\text{CH}_3\)) to the fifth carbon
-of the cytosine ring structure to form 5-methylcytosine. Numerous biological and
-medical studies have implicated DNA methylation as playing a role in disease and
-development \citep{robertson2005dna}. Perhaps unsurprisingly then, biotechnologies
-have been developed to rigorously probe the molecular mechanisms of this
-epigenetic process. Modern assays, like the Illumina \emph{Infinium} Methylation chip
-technology, allow for quantitative interrogation of DNA methylation at a large
-set of CpG sites scattered across the genome at single-nucleotide resolution;
-moreover, much effort has been invested, by the bioinformatics community, in the
+role in the regulation of gene expression. DNA methylation mostly commonly
+occurs at CpG sites and involves the addition of a methyl group (\(\text{CH}_3\))
+to the fifth carbon of the cytosine ring structure to form 5-methylcytosine.
+Numerous biological and medical studies have implicated DNA methylation as
+playing a role in disease and development \citep{robertson2005dna}. Perhaps
+unsurprisingly then, biotechnologies have been developed to rigorously probe the
+molecular mechanisms of this epigenetic process. Modern assays, like the
+Illumina \emph{Infinium} HumanMethylation BeadChip assay, allow for quantitative
+interrogation of DNA methylation, at single-nucleotide resolution, across a
+comprehensive set of CpG sites scattered across the genome; moreover, the
+computational biology community has invested significant effort in the
 development of tools for properly removing technological effects that may
 contaminate biological signatures measured by such assays
-\citep[dedeurwaerder2013comprehensive]{fortin2014functional}. Despite these
+\citep[\citet{dedeurwaerder2013comprehensive}]{fortin2014functional}. Despite these
 advances in both biological and bioninformatical techniques, most statistical
 methods available for differential analysis of data produced by such assays rely
-on over-simplified models that fail to incorporate many of the recent advances
-in machine learning. Generally, typical statistical estimation approaches do not
-readily extend to such high-dimensional data without restrictive modeling
-assumptions and the use of inferentially costly hypothesis testing corrections.
-When these standard assumptions are violated, estimates of the population-level
-effect of an exposure or treatment may suffer from large bias. Data-adaptive
-estimation procedures that utilize machine learning can control for confounding
-even in high-dimensional settings; however, interpretable statistical inference
-(i.e., confidence intervals and hypothesis tests) resulting from these
-data-adaptive estimates is challenging to obtain \citep{libbrecht2015machine}.
+on over-simplified models that do not readily extend to such high-dimensional
+data structures without restrictive modeling assumptions and the use of
+inferentially costly hypothesis testing corrections. When these standard
+assumptions are violated, estimates of the population-level effect of an
+exposure or treatment may suffer from large bias. What's more, reliance on
+restrictive and misspecified statistical models naturally leads to biased effect
+estimates that are not only misleading in assessing effect sizes but also result
+in false discoveries as these biased estimates are subject to testing and
+inferential procedures. Such predictably unreliable methods serve only to
+produce findings that are later invalidated by replication studies and add still
+further complexity to discovering biological targets for potential therapeutics.
+Data-adaptive estimation procedures that utilize machine learning provide a way
+to overcome many of the problems common in classical methods, controlling for
+potential confounding even in high-dimensional settings; however, interpretable
+statistical inference (i.e., confidence intervals and hypothesis tests) from
+such data-adaptive estimates is challenging to obtain \citep{libbrecht2015machine}.
 
@@
 algorithms in the estimation of a targeted VIM, initially proposed and explored
-in \citep{tuglus2008targeted}. In the presnet work, we focus on a software package,
+in \citep{tuglus2008targeted}. In the present work, we focus on a software package,
 \texttt{methyvim}, that implements a variant of this methodology, specifically tailored
@@
 binary phenotype-level treatment, and \(W_k, k = 1, \ldots K(j)\) is a matrix of
-the observed values of CpG sites in the same neighborhood as \(j\). We consider
-the case of observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where we
-take \(O \sim \mathcal{P} \in \mathcal{M}\), which is to say that \(O\) is governed
-by an unknown probability distribution \(\mathcal{P}\), taken only to reside in a
-nonparametric statistical model \(\mathcal{M}\) that places no restrictions on the
-data-generating process. We consider having access to measurements on a large
-number \(G\) of CpG sites (e.g., 850,000, as measured by the Illumina
-MethylationEPIC BeadChip arrays), and as a matter of practicality, let \(J \leq G\), so that \(J\) indexes only those CpG sites that pass an arbitrary
-pre-screening procedure; note that the equality is only obtained in the case
-that all sites pass the pre-screening procedure. Further, we let \(K(j)\) index
-the set of all CpG sites so that \(k\), the realization of \(K(j)\) for a given CpG
-site \(j\), is a set of the CpG sites that neighbor the CpG site \(j\), where the
-definition of a neighborhood is left as user-specified; thus, \(W_k\) is a matrix
-of observed methylation values for the set of CpG sites \(k\) that neighbor a
-particular target site \(j\). We consider estimating a target parameter (\(\psi_j\))
-as a variable importance measure of the effect of changing a treatment \(A\) on
-the methylation of a CpG site \(j\), accounting for the observed methylation at
-the set of sites \(k\) that neighbor the site \(j\). After a complete run of the
-procedure, we have access to \(j\) VIM estimates \(\psi_j\), each corresponding to
-the degree to which a given CpG site \(j\) was responsive to a change in the
-binary treatment \(A\), after controlling for any confounding exerted by its
-neighboring CpG sites. When machine learning estimators are employed in
+the observed values of CpG sites in the same neighborhood as \(j\) as well as any
+potential phenotype-level confounders (e.g., sex, age). We consider having
+access to measurements on a large number \(G\) of CpG sites (e.g., \(850,000\), as
+measured by the Illumina MethylationEPIC BeadChip arrays), and as a matter of
+practicality, let \(J \leq G\), so that \(J\) indexes only those CpG sites that pass
+an arbitrary pre-screening procedure; note that equality is only obtained in the
+case that all sites pass the pre-screening procedure. Further, we let \(K(j)\)
+index the set of all CpG sites so that \(k\), the realization of \(K(j)\) for a
+given CpG site \(j\), is a set of the CpG sites that neighbor the CpG site \(j\),
+where the definition of a neighborhood is left as user-specified; thus, \(W_k\) is
+a matrix of observed methylation values for the set of CpG sites \(k\) that
+neighbor a particular target site \(j\) as well as any potential phenotype-level
+confounders, where the latter does not vary between realizations of \(K(j)\). As
+a variable importance measure, we see to estimate a target parameter (\(\psi_j\))
+that quantifies the effect of changing a treatment \(A\) on the methylation of a
+CpG site \(j\), accounting for any potential confounding from phenotype-level
+covariates and the observed methylation at the set of sites \(k\) that neighbor
+the site \(j\). After a complete run of the procedure, we have access to \(j\) VIM
+estimates \(\psi_j\), each corresponding to the degree to which a given CpG site
+\(j\) was responsive to a change in the binary treatment \(A\), after controlling
+for potential confounding. When machine learning estimators are employed in
 estimating \(\psi_j\), this produces a nonparametric variable importance measure
 of differential methylation, allowing for the identification of differentially
-methylated positions (DMPs).
+methylated positions (DMPs). We note that \(\psi(j)\), for any given \(j\), is a
+function of the \emph{true} data-generating distribution; so, to formalize, we
+consider observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where we
+have \(O \sim P \in \mathcal{M}\), which is to say that the random variable \(O\) is
+governed by an unknown probability distribution \(P\), assumed only to reside in a
+nonparametric statistical model \(\mathcal{M}\) that places no restrictions on the
+data-generating process.
 
@@
 \item
-  \texttt{tmle.npvi} - A nonparametric variable importance measure (NPVI)
-  \citep{chambaz2012estimation}. The effect of a continuous-valued exposure or
+  A nonparametric variable importance measure (NPVI) \citep{chambaz2012estimation}
+  (R package \texttt{tmle.npvi}). The effect of a continuous-valued exposure or
   treatment (the observed methylation at a target CpG site) on an outcome of
@@
   parameter that compares values of the treatment against a user-specified
-  reference value taken to be the null. In particular, the implementation to be
-  provided is designed to assess the effect of differential methylation at the
-  target CpG site on an outcome of interest (e.g., survival), providing a
+  reference value taken to be the null value. In particular, the implementation
+  to be provided is designed to assess the effect of differential methylation at
+  the target CpG site on an outcome of interest (e.g., survival), providing a
   nonparametric evaluation of the impact of methylation at the target site.
 \item
-  \texttt{txshift} - Variable importance using a causal parameter defined as the
-  counterfactual outcome under a posited shift of a continuous-value treatment
-  of interest using stochastic intervention policies \citep[
-  \citet{diaz2018stochastic}, \citet{hejazi2018txshift}]{munoz2012population}. The value of an outcome of interest
-  under an unobserved value of the (continuous-valued) treatment, specified
-  through a user-provided \emph{additive shift} of the observed treatment, may be
-  data-adaptively estimated. This allows for the effect of changes/shifts in the
-  observed methylation at a CpG site on an outcome of interest to be evalauted,
-  providing a nonparametric evalution of the relative importance of a particular
-  target CpG site with respect to another variable measured in the same study.
+  The causal effect of shifting the value of an observed intervention, defined
+  as the counterfactual outcome under a posited shift of a continuous-value
+  treatment of interest using stochastic intervention policies
+  \citep[\citet{diaz2018stochastic}, \citet{hejazi2018txshift}]{munoz2012population} (R package
+  \texttt{txshift}). The value of an outcome of interest under an unobserved value of
+  the (continuous-valued) treatment, specified through a user-provided
+  \emph{additive shift} of the observed treatment, may be data-adaptively estimated.
+  This allows for the effect of changes/shifts in the observed methylation at a
+  CpG site on an outcome of interest to be evaluated, providing a nonparametric
+  evalution of the relative importance of a particular target CpG site with
+  respect to another variable measured in the same study.
 \end{itemize}
@@
 \begin{Highlighting}[]
-\KeywordTok{methyheat}\NormalTok{(methyvim_cancer_ate)}
+\KeywordTok{methyheat}\NormalTok{(methyvim_cancer_ate, }\DataTypeTok{smooth.heat =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{left.label =} \StringTok{"none"}\NormalTok{)}
 \end{Highlighting}
@@
 
+\begin{verbatim}
+## [1] 0.1 0.3 0.6 0.8 1.0
+\end{verbatim}
+
 \begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1} \end{center}
 
-Invoking \texttt{methyheat} in this manner produces a plot of the top sites (\(25\), by
-default) based on the raw p-value, using the raw methylation measures in the
-plot. This uses the exceptional \texttt{superheat} R package \citep{barter2017superheat}.
+\textbf{Remark}: Invoking \texttt{methyheat} in this manner produces a plot of the top sites
+(\(25\), by default) based on the raw p-value, using the raw methylation measures
+in the plot. This uses the exceptional \texttt{superheat} R package
+\citep{barter2017superheat}, to which we can easily pass additional parameters. In
+particulat, we hide the CpG site labels that would appear by default on the left
+of the heatmap (by setting \texttt{left.label\ =\ "none"}) to emphasize that this is only
+an example and \emph{not} a scientific discovery.
 

### 2018-06-29 inst/doc/using_methyvim.Rmd
rework paper; update methods

---
title: "Targeted Data-Adaptive Estimation and Inference for Differential
  Methylation Analysis"
author: "Nima Hejazi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
bibliography: vignette-refs.bib
abstract: |
  We present a general algorithm for the nonparametric estimation of effects of
  DNA methylation at CpG sites scattered across the genome, complete with honest
  statistical inference for such estimates. This approach leverages variable
  importance measures, a class of parameters that arise in the study of causal
  inference. The parameters we present are defined in such a manner that they
  provide targeted estimates of the relative importance of CpG sites in the
  case of binary exposures/treatments assigned at the level of subjects. Such
  parameters come equipped with rich scientific interpretations, providing an
  avenue to move beyond linear models, applying modern developments in machine
  learning to estimating quantities of scientific interest in computational
  biology.
vignette: >
  %\VignetteIndexEntry{Targeted Data-Adaptive Estimation and Inference for Differential Methylation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of
tools for properly removing technological effects that may contaminate
biological signatures measured by such assays. Despite these advances in both
biological and bioninformatical techniques, most statistical methods available
for the analysis of data produced by such assays rely on over-simplified (often
generalized linear) models.

Here, we present an alternative to such statistical analysis approaches, in the
form of nonparametric estimation procedures inspired by machine learning and
causal inference. Specifically, we provide a technique for obtaining estimates
of nonparametric _variable importance measures_ (__VIM__), parameters with rich
scientific interpretations under the standard (untestable) assumptions used in
statistical causal inference, defining a limited set of VIMs specifically with
respect to the type of data commonly produced by DNA methylation assays.
For VIMs defined in such a manner, targeted minimum loss-based estimates may be
readily computed based on the data made available by DNA methylation assays.
Our contribution, `methyvim` is an R package that provides facilities for
performing differential methylation analyses within exactly this scope.

As the substantive contribution of our work is an estimation procedure, we focus
on data produced by 450k and 850k (EPIC) arrays made by Illumina and assume that
data has been subjected to proper quality control and normalizaton procedures,
as outlined by others in the computational biology community
[@fortin2014functional, @dedeurwaerder2013comprehensive]. For a general
discussion of the framework of targeted minimum loss-based estimation and the
role this approach plays in statistical causal inference, the interested reader
is invited to consult @vdl2011targeted and @vdl2018targeted. For a more general
introduction to statistical causal inference, @pearl2009causality serves well.

---

## Methodology

The core functionality of this package is made available via the eponymous
`methyvim` function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a _fully
nonparametric statistical model_. __The statistical algorithm consists in
several major steps:__

1. Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approach adapts core routines from the
  [`limma`](http://bioconductor.org/packages/limma) R package. Future releases
  will support functionality from other packages (e.g.,
  [`randomForest`](https://CRAN.R-project.org/package=randomForest),
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
  style of the function for performing screening via `limma`, users may write
  their own screening functions and are invited to contribute such functions to
  the core software package by opening pull requests at the GitHub repository.

2. Nonparametric estimates of VIMs, for the specified target parameter, are
   computed at each of the CpG sites passing the screening step. The VIMs are
   defined in such a way that the estimated effects is of an exposure/treatment
   on the methylation status of a target CpG site, controlling for the observed
   methylation status of the neighbors of that site. Currently, routines are
   adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
   Future releases will support doubly-robust estimates of these VIMs (via the
   [`drtmle`](https://CRAN.R-project.org/package=drtmle) package) and add
   parameters for continuous treatments/exposures (via the
   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

3. Since pre-screening is performed prior to estimating VIMs, we make use of a
   multiple testing correction uniquely suited to such settings. Due to the
   multiple testing nature of the estimation problem, a variant of the Benjamini
   & Hochberg procedure for controlling the False Discovery Rate (FDR) is
   applied [@benjamini1995controlling]. Specifically, we apply the "modified
   marginal Benjamini & Hochberg step-up False Discovery Rate controlling
   procedure for multi-stage analyses" (FDR-MSA), which is guaranteed to
   control the FDR as if all sites were tested (i.e., without screening)
   [@tuglus2009modified].

---

## Parameters of Interest

For discrete-valued treatments or exposures:

* The _average treatment effect_ (ATE): The effect of a binary exposure or
  treatment on the observed methylation at a target CpG site is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target site, based on an additive form. In particular, the
  parameter estimate represents the __additive difference__ in methylation that
  would have been observed at the target site had all observations received the
  treatment versus the counterfactual under which none received the treatment.

* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
  observed methylation at a target CpG site is estimated, controlling for the
  observed methylation at all other CpG sites in the same neighborhood as the
  target site, based on a geometric form. In particular, the parameter estimate
  represents the __multiplicative difference__ in methylation that would have
  been observed at the target site had all observations received the treatment
  versus the counterfactual under which none received the treatment.

Support for continuous-valued treatments or exposures is _planned but not yet
available_, though work is underway to incorporate into our methodology the
following

* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
  The effect of continuous-valued exposure or treatment (the observed
  methylation at a target CpG site) on an outcome of interest is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target (treatment) site, based on a parameter that
  compares values of the treatment against a reference value taken to be the
  null. In particular, the implementation provided is designed to assess the
  effect of differential methylation at the target CpG site on a (typically)
  phenotype-level outcome of interest (e.g., survival), in effect providing an
  nonparametric evaluation of the impact of methylation at the target site on
  said outcome.

_As previously noted, in all cases, an estimator of the target parameter is
constructed via targeted minimum loss-based estimation._

Having now discussed the foundational principles of the estimation procedure
employed and the statistical algorithm implemented, it is best to proceed by
examining `methyvim` by example.

---

## Preliminaries: Setting up the Data

```{r reqs, echo=FALSE, message=FALSE}
suppressMessages(library(tmle))
suppressMessages(library(minfi))
suppressMessages(library(SummarizedExperiment))
```

First, we'll load the `methyvim` package and the example data contained in the
`methyvimData` package that accompanies it:

```{r prelims}
library(methyvim)
library(methyvimData)
```

Now, let's load the data set and seed the RNG:

```{r get-data}
set.seed(479253)
data(grsExample)
grsExample
var_int <- as.numeric(colData(grsExample)[, 1])
table(var_int)
```

The example data object is of class `GenomicRatioSet`, provided by the `minfi`
package. The summary provided by the `print` method gives a wealth of
information on the experiment that generated the data -- since we are working
with a simulated data set, we need not concern ourselves with much of this
information.

We can create an object of class `methytmle` from any `GenomicRatioSet` object
simply invoking the S4 class constructor:

```{r make-methytmle}
mtmle <- .methytmle(grsExample)
```

Since the `methytmle` class builds upon the `GenomicRatioSet` class, it contains
all of the slots of `GenomicRatioSet`. The new class introduced in the
`methyvim` package includes several new slots:

* `call` - the form of the original call to the `methyvim` function.
* `screen_ind` - indices identifying CpG sites that pass the screening process.
* `clusters` - non-unique IDs corresponding to the manner in wich sites are
  treated as neighbors. These are assigned by genomic distance (bp) and respect
  chromosome boundaries (produced via a call to `bumphunter::clusterMaker`).
* `var_int` - the treatment/exposure status for each subject. Currently, these
  must be binary, due to the definition of the supported targeted parameters.
* `param` - the name of the target parameter from which the estimated VIMs are
  defined.
* `vim` - a table of statistical results obtained from estimating VIMs for each
  of the CpG sites that pass the screening procedure.
* `ic` - the measured array values for each of the CpG sites passing the
  screening, transformed into influence curve space based on the chosen target
  parameter.

The interested analyst might consider consulting the documentation of the
`minfi` package for an in-depth description of all of the other slots that
appear in this class [@aryee2014minfi]. Having examined the core structure of
the package, it is time now to discuss the analytic capabilities implemented.

---

## Differential Methylation Based on a Binary Treatment or Exposure

### The Average Treatment Effect as Variable Importance Measure

The average treatment effect (ATE) is a canonical parameter that arises in
statistical causal inference, often denoted $\psi_0 = \psi_0(1) - \psi_0(0)$,
representing the difference in an outcome between the counterfactuals under
which all subjects received the treatment/exposure and under which none received
such treatment/exposure. Under additional (untestable) assumptions, this
parameter has a richer interpretation as a mean counterfactual outcome, wherein
the counterfactuals used in this definition define causal effects. When causal
assumptions remain unfulfilled or untested, this parameter may still be
estimated in the form of a nonparametric VIM.

Estimating such the VIM corresponding to such a parameter requires two separate
regression steps: one for the treatment mechanism (propensity score) and one for
the outcome regression. Technical details on the nature of these regressions are
discussed in @hernan2018causal, and details for estimating these regressions in
the framework of targeted minimum loss-based estimation are discussed in
@vdl2011targeted.

#### Super Learning for nonparametric parameter estimation

Nonparametric and data-adaptive regressions (i.e., machine learning) may be used
in the two regression steps outlined above. This is implemented using the super
learner algorithm, which produces optimal combinations of such regression
functions (a variant of stacked regressions) using cross-validation
[@vdl2007super, @breiman1996stacked, @wolpert1992stacked].

`methyvim` makes performing such estimation for CpG sites using a given VIM
essentially trivial:

```{r methyvim-ate-sl}
suppressMessages(
  methyvim_ate_sl <- methyvim(data_grs = grsExample, sites_comp = 25,
                              var_int = var_int, vim = "ate", type = "Mval",
                              filter = "limma", filter_cutoff = 0.10,
                              parallel = FALSE, tmle_type = "sl"
                             )
)
```

As is clear from examining the object `methyvim_ate_sl`, the output resembles
exactly that returned when examining objects of class `GenomicRatioSet` from the
`minfi` R package. In particular, the returned `methytmle` object is merely a
modified form (in particular, a subclass) of the input `GenomicRatioSet` object
-- thus, it contains all of the original slots, with all of the experimental
data intact. Several extra pieces of information are contained within the output
object as well_.

We can take a look at the results produced from the estimation procedure (stored
in the `vim` `slot` of the `methytmle` object) simply by invoking the custom S4
accessor function `vim` (note that the `show` method of the resultant object
also displays this same information, amongst other things):

```{r methyvim-ate-sl-print}
vim(methyvim_ate_sl)
```

From the table displayed, we note that we have access to point estimates of the
ATE ("est_ate") as well as lower and upper confidence interval bounds for each
estimate ("lwr_ci" and "upr_ci", respectively). Additional statistical
information we have access to include the variance ("var_ate") of the estimate
as well as the p-value ("pval") associated with each point estimate (based on
Wald-style testing procedures). Beyond these, key bioinformatical quantities,
with respect to the algorithm outlined above, are also returned; these include
the total number of neighbors of the target site ("n_neighbors"), the number of
neighboring sites controlled for when estimating the effect of exposure on DNA
methylation ("n_neighbors_control"), and, the maximum correlation between the
target site and any given site in its full set of neighbors
("max_cor_neighbors").

#### Generalized linear models for parameter estimation

In cases where nonparametric regressions may not be preferred (e.g., where time
constraints are of concern), generalized linear models (GLMs) may be used to fit
the two regression steps required for estimating VIMs for the ATE.

`methyvim` makes performing such estimation for CpG sites using a given VIM
essentially trivial:

```{r methyvim-ate-glm}
suppressMessages(
  methyvim_ate_glm <- methyvim(data_grs = grsExample, sites_comp = 25,
                               var_int = var_int, vim = "ate", type = "Mval",
                               filter = "limma", filter_cutoff = 0.10,
                               parallel = FALSE, tmle_type = "glm"
                              )
)
```

Just as before, we can take a look at the results produced from the estimation
procedure (stored in the `vim` `slot` of the `methytmle` object) simply by
invoking the custom S4 accessor function `vim` (note that the `show` method of
the resultant object also displays this same information, amongst other things):

```{r methyvim-ate-glm-print}
vim(methyvim_ate_glm)
```

_Remark:_ Here, the estimates are obtained via GLMs, making each of the
regression steps less robust than if nonparametric regressions were used. It is
expected that these estimates differ from those obtained previously.

---

### The Risk Ratio as Variable Importance Measure

The risk ratio (RR) is another popular parameter that arises in statistical
causal inference, denoted $\psi_0 = \frac{\psi_0(1)}{\psi_0(0)}$, representing
the multiplicative contrast of an outcome between the counterfactuals under
which all subjects received the treatment/exposure and under which none received
such treatment/exposure. Under additional (untestable) assumptions, this
parameter has a richer interpretation as a mean counterfactual outcome, wherein
the counterfactuals used in this definition define causal effects. When causal
assumptions remain unfulfilled or untested, this parameter may still be
estimated in the form of a nonparametric VIM.

Just as before (in the case of the ATE), there are two regression steps required
for estimating VIMs based on this parameter. We do so in a manner analogous to
that described previously.

#### Super Learning for nonparametric parameter estimation

Nonparametric and data-adaptive regressions (i.e., machine learning) may be used
in the two regression steps required for estimating a VIM based on the RR. This
is implemented using the Super Learner algorithm.

`methyvim` makes performing such estimation for CpG sites using a given VIM
essentially trivial:

```{r methyvim-rr-sl}
methyvim_rr_sl <- methyvim(data_grs = grsExample, sites_comp = 25,
                            var_int = var_int, vim = "rr", type = "Mval",
                            filter = "limma", filter_cutoff = 0.10,
                            parallel = FALSE, tmle_type = "sl"
                           )
```

Just as before, we can take a look at the results produced from the estimation
procedure (stored in the `vim` `slot` of the `methytmle` object) simply by
invoking the custom S4 accessor function `vim` (note that the `show` method of
the resultant object also displays this same information, amongst other things):

```{r methyvim-rr-sl-print}
vim(methyvim_rr_sl)
```

#### Generalized linear models for parameter estimation

In cases where nonparametric regressions may not be preferred (e.g., where time
constraints are of concern), generalized linear models (GLMs) may be used to fit
the two regression steps required for estimating a VIMs for the ATE.

`methyvim` makes performing such estimation for CpG sites using a given VIM
essentially trivial:

```{r methyvim-rr-glm}
methyvim_rr_glm <- methyvim(data_grs = grsExample, sites_comp = 25,
                            var_int = var_int, vim = "rr", type = "Mval",
                            filter = "limma", filter_cutoff = 0.10,
                            parallel = FALSE, tmle_type = "glm"
                           )
```

Just as before, we can take a look at the results produced from the estimation
procedure (stored in the `vim` `slot` of the `methytmle` object) simply by
invoking the custom S4 accessor function `vim` (note that the `show` method of
the resultant object also displays this same information, amongst other things):

```{r methyvim-rr-glm-print}
vim(methyvim_rr_glm)
```

_Remark:_ Here, the estimates are obtained via GLMs, making each of the
regression steps less robust than if nonparametric regressions were used. It is
expected that these estimates differ from those obtained previously.

---

## Data Analysis with `methyvim`

In order to explore practical applications of the `methyvim` package, as well as
the full set of utilities it provides, our toy example (of just $10$ CpG sites)
is unfortunately insufficient. To proceed, we will use a publicly available
example data set produced by the Illumina 450K array, from the `minfiData` R
package. Now, let's load the package and data set, and take a look

```{r setup-minfidata}
suppressMessages(library(minfiData))
data(MsetEx)
mset <- mapToGenome(MsetEx)
grs <- ratioConvert(mset)
grs
```

After loading the data, which comes in the form of a raw `MethylSet` object, we
perform some further processing by mapping to the genome (with `mapToGenome`)
and converting the values from the methylated and unmethylated channels to
Beta-values (via `ratioConvert`). These two steps together produce an object of
class `GenomicRatioSet`, like what we had worked with previously.

For this example analysis, we'll treat the condition of the patients as the
exposure/treatment variable of interest. The `methyvim` function requires that
this variable either be `numeric` or easily coercible to `numeric`. To
facilitate this, we'll simply convert the covariate (currently a `character`):

```{r, minfidata-maketx}
var_int <- (as.numeric(as.factor(colData(grs)$status)) - 1)
table(var_int)
```
__n.b.__, the re-coding process results in "normal" patients being assigned a
value of 1 and cancer patients a 0.

Now, we are ready to analyze the effects of cancer status on DNA methylation
using this data set. To do this with a targeted minimum loss-based estimate of
the Average Treatment Effect, we may proceed as follows:

```{r, minfidata-methyvim}
suppressMessages(
  methyvim_cancer_ate <- methyvim(data_grs = grs, var_int = var_int,
                                  vim = "ate", type = "Beta", filter = "limma",
                                  filter_cutoff = 0.20, obs_per_covar = 2,
                                  parallel = FALSE, sites_comp = 125,
                                  tmle_type = "glm"
                                 )
)
```
Note that we set the `obs_per_covar` argument to a relatively low value (2,
where the recommended default is 20) for the purposes of this example as the
sample size is only 10. We do this only to exemplify the estimation procedure
and it is important to point out that such low values for `obs_per_covar` will
compromise the quality of inference obtained because this setting directly
affects the definition of the target parameter.

Further, note that here we apply the `glm` flavor of the `tmle_type` argument,
which produces faster results by fitting models for the propensity score and
outcome regressions using a limited number of parametric models. By contrast,
the `sl` (for "Super Learning") flavor fits these two regressions using highly
nonparametric and data-adaptive procedures (i.e., via machine learning).

Let's examine the table of results by examining the `methytmle` object that was
produced by our call to the `methyvim` wrapper function:

```{r vim-cancer-ate}
methyvim_cancer_ate
```

Finally, we may compute FDR-corrected p-values, by applying a modified procedure
for controlling the False Discovery Rate for multi-stage analyses (FDR-MSA)
[@tuglus2009modified]. We do this by simply applying the `fdr_msa` function:

```{r fdr-msa}
fdr_p <- fdr_msa(pvals = vim(methyvim_cancer_ate)$pval,
                 total_obs = nrow(methyvim_cancer_ate))
```

Having explored the results of our analysis numerically, we now proceed to use
the visualization tools provided with the `methyvim` R package to further
enhance our understanding of the results.

### Visualization of Results

While making allowance for users to explore the full set of results produced by
the estimation procedure (by way of exposing these directly to the user), the
`methyvim` package also provides _three_ (3) visualization utilities that
produce plots commonly used in examining the results of differential methylation
analyses.

A simple call to `plot` produces side-by-side histograms of the raw p-values
computed as part of the estimation process and the corrected p-values obtained
from using the FDR-MSA procedure.

```{r methyvim-pvals-both}
plot(methyvim_cancer_ate)
```

__Remark:__ The plots displayed above may also be generated separately by
explicitly setting the argument "type" to `plot.methytmle`. For a plot of the
raw p-values, specify `type = "raw_pvals"`, and for a plot of the FDR-corrected
p-values, specify `type = "fdr_pvals"`.

While histograms of the p-values may be generally useful in inspecting the
results of the estimation procedure, a more common plot used in examining the
results of differential methylation procedures is the volcano plot, which plots
the parameter estimate along the x-axis and $-\text{log}_{10}(\text{p-value})$
along the y-axis. We implement such a plot in the `methyvolc` function:

```{r methyvim-volcano}
methyvolc(methyvim_cancer_ate)
```

The purpose of such a plot is to ensure that very low (possibly statistically
significant) p-values do not arise from cases of low variance. This appears to
be the case in the plot above (notice that most parameter estimates are _near
zero_, even in cases where the raw p-values are quite low).

Yet another popular plot for visualizing effects in such settings is the
heatmap, which plots estimates of the raw methylation effects (as measured by
the assay) across subjects using a heat gradient. We implement this in the
`methyheat` function:

```{r methyvim-heatmap}
methyheat(methyvim_cancer_ate)
```

Invoking `methyheat` in this manner produces a plot of the top sites ($25$, by
default) based on the raw p-value, using the raw methylation measures in the
plot. This uses the exceptional `superheat` R package [@barter2017superheat].

---

## Session Information

```{r session-info, echo=FALSE}
sessionInfo()
```

---

## References

### 2018-06-29 paper/paper_BiocF1000.Rmd
rework paper; update methods

@@
 ---
-title: '`methyvim`: Targeted and Model-free Differential Methylation Analysis in
-  R'
+title: '`methyvim`: Targeted, Robust, and Model-free Analysis of Differential
+  Methylation in R'
 author:
@@
   such estimates. The approach leverages variable importance measures derived
-  from statistical parameters arising causal inference, defined in such a manner
-  that they may be used to obtain targeted estimates of the relative importance
-  of individual CpG sites with respect to a binary treatment assigned at the
-  phenotype level. The procedure implemented is computationally efficient,
-  incorporating a preliminary screening step to isolate a subset of sites for
-  which there is cursory evidence of differential methylation as well as a
-  unique multiple testing correction to control the False Discovery Rate with
-  the same rigor as if all sites were tested. This technique for analysis of
-  differentially methylated positions provides an avenue to incorporate flexible
-  state-of-the-art machine learning algorithms into the estimation of
+  from statistical parameters arising in causal inference, defined in such a
+  manner that they may be used to obtain targeted estimates of the relative
+  importance of individual CpG sites with respect to a binary treatment assigned
+  at the phenotype level. The procedure implemented is computationally
+  efficient, incorporating a preliminary screening step to isolate a subset of
+  sites for which there is cursory evidence of differential methylation as well
+  as a unique multiple testing correction to control the False Discovery Rate
+  with the same rigor as if all sites were tested. This technique for analysis
+  of differentially methylated positions provides an avenue to incorporate
+  flexible state-of-the-art machine learning algorithms into the estimation of
   differential methylation effects without the loss of interpretable statistical
@@
 ---
-
-```{r setup, echo=FALSE}
-options(scipen=999)
-```
 
@@
 point estimation and the construction of Wald-style confidence intervals
-[@vdl2011targeted, @vdl2018targeted]. As counterfactuals are contrary-to-fact --
-that is, defined in terms of missing data -- in the context of DNA methylation,
+[@vdl2011targeted, @vdl2018targeted]. In the context of DNA methylation studies,
 we define the counterfactual outcomes under a binary treatment as the observed
 methylation (whether Beta- or M-) values a CpG site would have if all subjects
-were given the treatment and the methylation values a CpG site would have if
-treatment were withheld from all subjects. Although these counterfactual
-outcomes are, of course, impossible to observe, they have statistical analogs
-that may be reliably estimated from observed data [@pearl2009causality]. We
-describe an algorithm that incorporates the estimation of VIMs using _targeted
-minimum loss-based estimation_ (__TMLE__) [@vdl2006targeted], deferring detailed
-description and analysis of the statistical methodology to work outside the
-scope of the present manuscript. This methodology assesses the individual
-importance of a given CpG site by utilizing state-of-the-art machine learning
-algorithms in the estimation of a targeted VIM, initially proposed and explored
-in [@tuglus2008targeted]. In the present work, we focus on a software package,
-`methyvim`, that implements a variant of this methodology, specifically tailored
-to differential methylation analysis.
+were administered the treatment and the methylation values a CpG site would have
+if treatment were withheld from all subjects. Although these counterfactual
+outcomes are, of course, impossible to observe, they do have statistical analogs
+that may be reliably estimated (i.e., identified) from observed data under a
+small number of untestable assumptions [@pearl2009causality]. We describe an
+algorithm that incorporates, in its final step, the use _targeted minimum
+loss-based estimators_ (__TMLE__) [@vdl2006targeted] of a given VIM of interest,
+though we defer rigorous and detailed descriptions of this aspect of the
+statistical methodology to work outside the scope of the present manuscript
+[@vdl2006targeted, @vdl2011targeted, @vdl2018targeted]. The proposed methodology
+assesses the individual importance of a given CpG site, as a proposed measure of
+differential methylation, by utilizing state-of-the-art machine learning
+algorithms in deriving targeted estimates and robust inference of a VIM, as
+considered more broadly for biomarkers in @bembom2009biomarker and
+@tuglus2011targeted. In the present work, we focus on the `methyvim` software
+package, available through the Bioconductor project [@gentleman2004bioconductor,
+@huber2015orchestrating] for the R language and environment for statistical
+computing [@R], which implements a particular realization of this methodology
+specifically tailored for the analysis and identification of differentially
+methylated positions (DMPs).
 
-For a general discussion of the framework of targeted minimum loss-based
-estimation and the role this approach plays in statistical causal inference, the
-interested reader is invited to consult @vdl2011targeted and @vdl2018targeted.
-For a more general introduction to (statistical) causal inference,
-@pearl2009causality and @hernan2018causal may be of interest.
+For an extended discussion of the general framework of targeted minimum
+loss-based estimation and detailed accounts of how this approach may be brought
+to bear in developing answers to complex scientific problems through tatistical
+and causal inference, the interested reader is invited to consult
+@vdl2011targeted and @vdl2018targeted. For a more general introduction to causal
+inference, @pearl2009causality and @hernan2018causal may be of interest.
 
@@
 
-We consider an observed data structure $O = (W_k, A, Y_j)$, where $Y_j, j = 1,
-\ldots J$ represents a given CpG site of interest, $A \in \{0, 1\}$ represents a
-binary phenotype-level treatment, and $W_k, k = 1, \ldots K(j)$ is a matrix of
-the observed values of CpG sites in the same neighborhood as $j$ as well as any
-potential phenotype-level confounders (e.g., sex, age). We consider having
-access to measurements on a large number $G$ of CpG sites (e.g., $850,000$, as
-measured by the Illumina MethylationEPIC BeadChip arrays), and as a matter of
-practicality, let $J \leq G$, so that $J$ indexes only those CpG sites that pass
-an arbitrary pre-screening procedure; note that equality is only obtained in the
-case that all sites pass the pre-screening procedure. Further, we let $K(j)$
-index the set of all CpG sites so that $k$, the realization of $K(j)$ for a
-given CpG site $j$, is a set of the CpG sites that neighbor the CpG site $j$,
-where the definition of a neighborhood is left as user-specified; thus, $W_k$ is
-a matrix of observed methylation values for the set of CpG sites $k$ that
-neighbor a particular target site $j$ as well as any potential phenotype-level
-confounders, where the latter does not vary between realizations of $K(j)$. As
-a variable importance measure, we see to estimate a target parameter ($\psi_j$)
-that quantifies the effect of changing a treatment $A$ on the methylation of a
-CpG site $j$, accounting for any potential confounding from phenotype-level
-covariates and the observed methylation at the set of sites $k$ that neighbor
-the site $j$. After a complete run of the procedure, we have access to $j$ VIM
-estimates $\psi_j$, each corresponding to the degree to which a given CpG site
-$j$ was responsive to a change in the binary treatment $A$, after controlling
-for potential confounding. When machine learning estimators are employed in
-estimating $\psi_j$, this produces a nonparametric variable importance measure
-of differential methylation, allowing for the identification of differentially
-methylated positions (DMPs). We note that $\psi(j)$, for any given $j$, is a
-function of the _true_ data-generating distribution; so, to formalize, we
-consider observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where we
-have $O \sim P \in \mathcal{M}$, which is to say that the random variable $O$ is
-governed by an unknown probability distribution $P$, assumed only to reside in a
-nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
+We consider an observed data structure, on a single experimental subject (e.g.,
+a patient), $O = (W, A, (Y(j) : j))$, where $(Y(j) : j = 1, \ldots J)$ is the
+set of CpG sites measured by the assay in question, $A \in \{0, 1\}$ represents
+a binary phenotype-level treatment, and $W$ is a vector of the phenotype-level
+baseline covariates that are potential confounders (e.g., age, sex). We consider
+having access to measurements on a large number $J$ of CpG sites (e.g.,
+$850,000$, as measured by the Illumina _MethylationEPIC_ BeadChip arrays).
+Further, let $K: j \to S_j$ be a procedure that assigns to a given CpG site $j$
+a set of neighbors $S_j$ -- that is, $S_j$ is a collection of indices of the
+neighbors of $j$ just as $j$ indexes the full set of CpG sites; moreover, since
+$Y(j)$ is the measured methylation value for a given CpG site $j$, $Y(S_j)$ is
+the measured methylation values at the neighbors $S_j$ of $j$. Note that the
+definition of a neighborhood $\{j, S_j\}$ is left vague so as to facilitate the
+use of user-specified strategies in implementation. We consider the case of
+observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P \in
+\mathcal{M}$, which is to say that the random variable $O$ is governed by an
+unknown probability distribution $P$, assumed only to reside in a nonparametric
+statistical model $\mathcal{M}$ that places no restrictions on the
 data-generating process.
+
+## Variable Importance Measure
+
+With the data structure above in hand, we let the estimand of interest be a
+$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
+through the _true_ (and unknown) probability distribution $P$. As a motivating
+example, consider the case where $\Psi_j(P)$ is
+\begin{equation}\label{vim_param}
+  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
+    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
+\end{equation}
+where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
+values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
+parameter of interest is a variant of the _average treatment effect_, which has
+been the subject of much attention in statistical causal inference **CITATION
+NEEDED**. As a measure of variable importance, we seek to estimate this target
+parameter ($\Psi_j(P)$), which quantifies the effect of changing a treatment $A$
+on the methylation $Y(j)$ of a CpG site $j$, accounting for any potential
+confounding from phenotype-level covariates $W$ and observed methylation
+$Y(S_j)$ at the neighboring sites $S_j$ of $j$. We propose a procedure that
+estimates this target parameter $\Psi_j(P)$ across all CpG sites of interest
+$j: 1, \ldots, J$. When data-adaptive regression procedures (i.e., machine
+learning) are employed in estimating $\Psi_j(P)$, this produces a nonparametric
+variable importance measure of differential methylation, allowing for the
+identification of differentially methylated positions (DMPs) while avoiding many
+assumptions common in the use of standard parametric regression procedures. We
+propose estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which
+allows for data-adaptive regression procedures to be employed in a
+straightforward manner.
+
+## Pre-Screening Procedure
+
+As a matter of practicality, we consider only estimating the target VIM
+$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
+span the full set of assayed CpG sites. In order to determine this subset of CpG
+sites, we propose the use of a pre-screening procedure, which need be nothing
+more than a method for differential methylation analysis that is computationally
+less demanding than the method proposed here. Formally, let the pre-screening
+procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
+input a CpG site $j$ and returns a binary decision rule of whether to include
+CpG site $j$ in a subset to be evaluated further or not. Details on the
+implementation of pre-screening approaches is available in the sequel.
+
+## Reducing Neighbors via Clustering
+
+Due to the data-adaptive nature of the regression procedures employed in
+evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
+estimation, it is possible the a given CpG site $j$ may have _too many_
+neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
+the inclusion of too many neighbors when controlling for potential confounders
+may lead to instability in the estimates produced. In such cases, we propose and
+implement the use of a clustering technique (e.g., Partitioning Around Medoids)
+to select a _representative_ subset of neighbors. Formally, there is likely no
+best choice of a specific clustering algorithm, so we leave this aspect of the
+proposed algorithm as flexible for the user [@kleinberg2003impossibility].
+
+## Targeted Minimum Loss-Based Estimation and Inference
+
+Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
+minimum loss-based estimation (TMLE) to construct and evaluate an estimator
+$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
+target VIM is the average treatment effect, we make use of a TML estimator of
+this parameter, which has been implemented for a general case in the
+[`tmle`](https://CRAN.R-project.org/package=tmle) R package [@gruber2011tmle].
+Generally speaking, a TML estimator is constructed from a few simple components:
+(1) an estimator of the propensity score [@rosenbaum1983central] (often denoted
+by $g(A, W)$), (2) an estimator of the outcome regression (often denoted $Q(A,
+W)$), and (3) a targeting step applied to the initial estimator constructed from
+the aforementioned components [@gruber2009targeted]. In the sequel, we describe
+a few of the key mathematical properties of TML estimators relevant for
+statistical inference, deferring extended discussion of the technical details to
+@vdl2011targeted and @vdl2018targeted; moreover, as a matter of practicality, we
+make use of the [`tmle`](https://CRAN.R-project.org/package=tmle)  R package of
+@gruber2011tmle as a backend for computing estimators of relevant target
+parameters in our software package, so the interested user is encouraged to
+discuss the documentation of that software package for further details.
+
+As noted previously, data-adaptive regression techniques may be used in the
+construction of TML estimators. This is done by generating a stacked regression
+model from a given set of machine learning algorithms via the Super Learner
+algorithm [@wolpert1992stacked, breiman1996stacked, vdl2007super], which ensures
+that the resulting ensemble learner satisifies important optimality properties
+with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
+@vdl2003unified]. Importantly, TML estimators are well suited for inference,
+having an asymptotically normal limiting distribution
+[@tsiatis2007semiparametric, @vdl2006targeted] $\sqrt{n}(\psi_{n, j} -
+\Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2)$, where $\sigma_{\text{EIF}}^2$ is
+the variance of the efficient influence function of the target parameter
+evaluated at the observed data. This allows for confidence intervals to be
+constructed and and hypothesis tests (i.e., $H_{0, j}: \Psi_j(P) = 0$) to be
+performed in a straightforward manner.
+
+## Correction for Multiple Testing
+
+Given that we seek to estimate $\Psi_j(P)$ for a possibly large number of CpG
+sites $(j: 1, \ldots J)$, the need to perform corrections for multiple testing
+is clear. In order to curb the potential for false discoveries, we recommend the
+use of the Benjamini and Hochberg procedure for controlling the False Discovery
+Rate (FDR) [@benjamini1995controlling]; however, as the proposed procedure
+involves a pre-screening step, naive application of the Benjamini and Hochberg
+procedure (BH) is invalid -- instead, we rely on a modification of the procedure
+to control the FDR, with established theoretical guarantees when pre-screening
+is employed [@tuglus2009modified]. In brief, the modified marginal Benjamini and
+Hochberg procedure to control the FDR under pre-screening works by applying the
+standard BH procedure to a padded vector of p-values -- that is, letting $J$ be
+the number of CpG sites tested and $K$ the number of CpG sites filtered out (so
+that $J + K = P$, where $P$ is the original dimension of the genomic assay), the
+modified marginal BH procedure is the application of the original BH procedure
+to a vector of p-values, composed of the $J$ p-values from performing $J$
+hypothesis tests (of the form $H_{0, j}: \Psi_j(P) = 0$) and $K$ additional
+p-values automatically set to $1$ (for the hypothesis tests not performed on
+account of pre-screening). This procedure is guaranteed to control the FDR at
+the same desired rate when pre-screening is performed whereas naive application
+of the BH procedure fails to do so.
 
@@
 a manner that is very nearly assumption-free, that is, within a _fully
-nonparametric statistical model_. __The statistical algorithm consists in
-several major steps__:
+nonparametric statistical model_. __The statistical algorithm consists in the
+several major steps elaborated upon above and summarized below__:
 
-1. Pre-screening of genomic sites is used to isolate a subset of sites for
-   which there is cursory evidence of differential methylation. For the sake of
-   computational feasibility, targeted minimum loss-based estimates of VIMs are
-   computed only for this subset of sites. Currently, the available screening
-   approach adapts core routines from the
-   [`limma`](http://bioconductor.org/packages/limma) R
-   package. Future releases will support functionality from other packages
-   (e.g., [`randomForest`](https://CRAN.R-project.org/package=randomForest),
-   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
+1. _Pre-screening_ of genomic sites is used to isolate a subset of sites for
+   which there is cursory evidence of differential methylation. Currently, the
+   available screening approach adapts core routines from the
+   [`limma`](http://bioconductor.org/packages/limma) R package. Following the
    style of the function for performing screening via `limma`, users may write
    their own screening functions and are invited to contribute such functions to
-   the core software package by opening pull requests at the GitHub repository.
+   the core software package by opening pull requests at the GitHub repository:
+   \url{https://github.com/nhejazi/methyvim}.
 
@@
    computed at each of the CpG sites passing the screening step. The VIMs are
-   defined in such a way that the estimated effects is of an exposure/treatment
-   on the methylation status of a target CpG site, controlling for the observed
+   defined in such a way that the estimated effects is of an binary treatment on
+   the methylation status of a target CpG site, controlling for the observed
    methylation status of the neighbors of that site. Currently, routines are
    adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
-   Future releases will support doubly-robust estimates of these VIMs (via the
-   [`drtmle`](https://cran.r-project.org/package=drtmle) package)
-   and add parameters for continuous treatments/exposures (via the
-   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) and
-   [`txshift`](https://github.com/nhejazi/txshift) R packages).
 
-3. Since pre-screening is performed prior to estimating VIMs, we make use of a
-   multiple testing correction uniquely suited to such settings. Due to the
-   multiple testing nature of the estimation problem, a variant of the Benjamini
-   and Hochberg procedure for controlling the False Discovery Rate (FDR) is
-   applied [@benjamini1995controlling]. Specifically, we apply a modified
-   marginal Benjamini and Hochberg step-up False Discovery Rate controlling
-   procedure for multi-stage analyses (FDR-MSA), which has established
-   theoretical guarantees to control the FDR at the same rate as if all sites
-   were tested (i.e., without screening) [@tuglus2009modified].
+3. Since pre-screening is performed prior to estimating VIMs, we apply the
+   modified marginal Benjamini and Hochberg step-up False Discovery Rate
+   controlling procedure for multi-stage analyses (FDR-MSA), which is
+   well-suited for avoiding false positive discoveries when testing is only
+   performed on a subset of potential targets.
 
@@
 
-Support for continuous-valued treatments or exposures is _planned but not yet
-available_. Future releases will allow users to assess continuous-valued
-treatments by relying on parameters estimable through implementations available
-in the following software packages:
+#### Future Work
+
+Support for continuous-valued treatments or exposures is _planned and ongoing_
+but not yet available in stable releases of the `methyvim` software package.
+Future releases will allow users to assess continuous-valued treatments by
+relying on parameters estimable through implementations available in the
+following software packages:
 
@@
 
-We refer the reader to the package vignette, "`methyvim`: Targeted Data-Adaptive
-Estimation and Inference for Differential Methylation Analysis," included in any
-distribution of the software package, for further details.
+The `show` method of the `methytmle` class summarizes a selection of the above
+information for the user while masking some of the wealth of information given
+when calling the same method for `GenomicRatioSet`. All information contained in
+`GenomicRatioSet` objects is preserved in `methytmle` objects, so as to easen
+interoperability with other differential methylation software for experienced
+users. We refer the reader to the package vignette, "`methyvim`: Targeted
+Data-Adaptive Estimation and Inference for Differential Methylation Analysis,"
+included in any distribution of the software package, for further details.
 
@@
 
-```{r setup-minfidata}
-suppressMessages(library(minfiData))
+```{r setup-minfidata, message=FALSE}
+suppressMessages(
+  # numerous messages displayed at time of loading
+  library(minfiData)
+)
 data(MsetEx)
@@
 ```{r fdr-msa}
-fdr_p <- fdr_msa(pvals = methyvim_cancer_ate@vim$pval,
+fdr_p <- fdr_msa(pvals = vim(methyvim_cancer_ate)$pval,
                  total_obs = nrow(methyvim_cancer_ate))
@@
 
-```{r methyvim-pvals-raw}
+```{r methyvim-pvals-raw, fig.height=4, fig.width=6}
 plot(methyvim_cancer_ate, type = "raw_pvals")
@@
 
-```{r methyvim-pvals-fdr}
+```{r methyvim-pvals-fdr, fig.height=4, fig.width=6}
 plot(methyvim_cancer_ate, type = "fdr_pvals")
@@
 
-```{r methyvim-volcano}
+```{r methyvim-volcano, fig.height=4, fig.width=6}
 methyvolc(methyvim_cancer_ate)
@@
 
-```{r methyvim-heatmap}
+```{r methyvim-heatmap, fig.height=4, fig.width=6}
 methyheat(methyvim_cancer_ate, smooth.heat = TRUE, left.label = "none")
@@
 
-No competing interests were disclosed.
+No competing interests were disclosed at the time of publication.
 

### 2018-06-29 paper/paper_BiocF1000.tex
rework paper; update methods

@@
 
-\title{\texttt{methyvim}: Targeted and Model-free Differential Methylation Analysis in R}
+\title{\texttt{methyvim}: Targeted, Robust, and Model-free Analysis of Differential Methylation in R}
 
@@
 such estimates. The approach leverages variable importance measures derived
-from statistical parameters arising causal inference, defined in such a manner
-that they may be used to obtain targeted estimates of the relative importance
-of individual CpG sites with respect to a binary treatment assigned at the
-phenotype level. The procedure implemented is computationally efficient,
-incorporating a preliminary screening step to isolate a subset of sites for
-which there is cursory evidence of differential methylation as well as a
-unique multiple testing correction to control the False Discovery Rate with
-the same rigor as if all sites were tested. This technique for analysis of
-differentially methylated positions provides an avenue to incorporate flexible
-state-of-the-art machine learning algorithms into the estimation of
+from statistical parameters arising in causal inference, defined in such a
+manner that they may be used to obtain targeted estimates of the relative
+importance of individual CpG sites with respect to a binary treatment assigned
+at the phenotype level. The procedure implemented is computationally
+efficient, incorporating a preliminary screening step to isolate a subset of
+sites for which there is cursory evidence of differential methylation as well
+as a unique multiple testing correction to control the False Discovery Rate
+with the same rigor as if all sites were tested. This technique for analysis
+of differentially methylated positions provides an avenue to incorporate
+flexible state-of-the-art machine learning algorithms into the estimation of
 differential methylation effects without the loss of interpretable statistical
@@
 point estimation and the construction of Wald-style confidence intervals
-\citep[\citet{vdl2018targeted}]{vdl2011targeted}. As counterfactuals are contrary-to-fact --
-that is, defined in terms of missing data -- in the context of DNA methylation,
+\citep[\citet{vdl2018targeted}]{vdl2011targeted}. In the context of DNA methylation studies,
 we define the counterfactual outcomes under a binary treatment as the observed
 methylation (whether Beta- or M-) values a CpG site would have if all subjects
-were given the treatment and the methylation values a CpG site would have if
-treatment were withheld from all subjects. Although these counterfactual
-outcomes are, of course, impossible to observe, they have statistical analogs
-that may be reliably estimated from observed data \citep{pearl2009causality}. We
-describe an algorithm that incorporates the estimation of VIMs using \emph{targeted
-minimum loss-based estimation} (\textbf{TMLE}) \citep{vdl2006targeted}, deferring detailed
-description and analysis of the statistical methodology to work outside the
-scope of the present manuscript. This methodology assesses the individual
-importance of a given CpG site by utilizing state-of-the-art machine learning
-algorithms in the estimation of a targeted VIM, initially proposed and explored
-in \citep{tuglus2008targeted}. In the present work, we focus on a software package,
-\texttt{methyvim}, that implements a variant of this methodology, specifically tailored
-to differential methylation analysis.
+were administered the treatment and the methylation values a CpG site would have
+if treatment were withheld from all subjects. Although these counterfactual
+outcomes are, of course, impossible to observe, they do have statistical analogs
+that may be reliably estimated (i.e., identified) from observed data under a
+small number of untestable assumptions \citep{pearl2009causality}. We describe an
+algorithm that incorporates, in its final step, the use \emph{targeted minimum
+loss-based estimators} (\textbf{TMLE}) \citep{vdl2006targeted} of a given VIM of interest,
+though we defer rigorous and detailed descriptions of this aspect of the
+statistical methodology to work outside the scope of the present manuscript
+\citep[\citet{vdl2011targeted}, \citet{vdl2018targeted}]{vdl2006targeted}. The proposed methodology
+assesses the individual importance of a given CpG site, as a proposed measure of
+differential methylation, by utilizing state-of-the-art machine learning
+algorithms in deriving targeted estimates and robust inference of a VIM, as
+considered more broadly for biomarkers in \citet{bembom2009biomarker} and
+\citet{tuglus2011targeted}. In the present work, we focus on the \texttt{methyvim} software
+package, available through the Bioconductor project \citep[
+\citet{huber2015orchestrating}]{gentleman2004bioconductor} for the R language and environment for statistical
+computing \citep{R}, which implements a particular realization of this methodology
+specifically tailored for the analysis and identification of differentially
+methylated positions (DMPs).
 
-For a general discussion of the framework of targeted minimum loss-based
-estimation and the role this approach plays in statistical causal inference, the
-interested reader is invited to consult \citet{vdl2011targeted} and \citet{vdl2018targeted}.
-For a more general introduction to (statistical) causal inference,
-\citet{pearl2009causality} and \citet{hernan2018causal} may be of interest.
+For an extended discussion of the general framework of targeted minimum
+loss-based estimation and detailed accounts of how this approach may be brought
+to bear in developing answers to complex scientific problems through tatistical
+and causal inference, the interested reader is invited to consult
+\citet{vdl2011targeted} and \citet{vdl2018targeted}. For a more general introduction to causal
+inference, \citet{pearl2009causality} and \citet{hernan2018causal} may be of interest.
 
@@
 
-We consider an observed data structure \(O = (W_k, A, Y_j)\), where \(Y_j, j = 1, \ldots J\) represents a given CpG site of interest, \(A \in \{0, 1\}\) represents a
-binary phenotype-level treatment, and \(W_k, k = 1, \ldots K(j)\) is a matrix of
-the observed values of CpG sites in the same neighborhood as \(j\) as well as any
-potential phenotype-level confounders (e.g., sex, age). We consider having
-access to measurements on a large number \(G\) of CpG sites (e.g., \(850,000\), as
-measured by the Illumina MethylationEPIC BeadChip arrays), and as a matter of
-practicality, let \(J \leq G\), so that \(J\) indexes only those CpG sites that pass
-an arbitrary pre-screening procedure; note that equality is only obtained in the
-case that all sites pass the pre-screening procedure. Further, we let \(K(j)\)
-index the set of all CpG sites so that \(k\), the realization of \(K(j)\) for a
-given CpG site \(j\), is a set of the CpG sites that neighbor the CpG site \(j\),
-where the definition of a neighborhood is left as user-specified; thus, \(W_k\) is
-a matrix of observed methylation values for the set of CpG sites \(k\) that
-neighbor a particular target site \(j\) as well as any potential phenotype-level
-confounders, where the latter does not vary between realizations of \(K(j)\). As
-a variable importance measure, we see to estimate a target parameter (\(\psi_j\))
-that quantifies the effect of changing a treatment \(A\) on the methylation of a
-CpG site \(j\), accounting for any potential confounding from phenotype-level
-covariates and the observed methylation at the set of sites \(k\) that neighbor
-the site \(j\). After a complete run of the procedure, we have access to \(j\) VIM
-estimates \(\psi_j\), each corresponding to the degree to which a given CpG site
-\(j\) was responsive to a change in the binary treatment \(A\), after controlling
-for potential confounding. When machine learning estimators are employed in
-estimating \(\psi_j\), this produces a nonparametric variable importance measure
-of differential methylation, allowing for the identification of differentially
-methylated positions (DMPs). We note that \(\psi(j)\), for any given \(j\), is a
-function of the \emph{true} data-generating distribution; so, to formalize, we
-consider observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where we
-have \(O \sim P \in \mathcal{M}\), which is to say that the random variable \(O\) is
-governed by an unknown probability distribution \(P\), assumed only to reside in a
-nonparametric statistical model \(\mathcal{M}\) that places no restrictions on the
+We consider an observed data structure, on a single experimental subject (e.g.,
+a patient), \(O = (W, A, (Y(j) : j))\), where \((Y(j) : j = 1, \ldots J)\) is the
+set of CpG sites measured by the assay in question, \(A \in \{0, 1\}\) represents
+a binary phenotype-level treatment, and \(W\) is a vector of the phenotype-level
+baseline covariates that are potential confounders (e.g., age, sex). We consider
+having access to measurements on a large number \(J\) of CpG sites (e.g.,
+\(850,000\), as measured by the Illumina \emph{MethylationEPIC} BeadChip arrays).
+Further, let \(K: j \to S_j\) be a procedure that assigns to a given CpG site \(j\)
+a set of neighbors \(S_j\) -- that is, \(S_j\) is a collection of indices of the
+neighbors of \(j\) just as \(j\) indexes the full set of CpG sites; moreover, since
+\(Y(j)\) is the measured methylation value for a given CpG site \(j\), \(Y(S_j)\) is
+the measured methylation values at the neighbors \(S_j\) of \(j\). Note that the
+definition of a neighborhood \(\{j, S_j\}\) is left vague so as to facilitate the
+use of user-specified strategies in implementation. We consider the case of
+observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where \(O \sim P \in \mathcal{M}\), which is to say that the random variable \(O\) is governed by an
+unknown probability distribution \(P\), assumed only to reside in a nonparametric
+statistical model \(\mathcal{M}\) that places no restrictions on the
 data-generating process.
+
+\hypertarget{variable-importance-measure}{%
+\subsection{Variable Importance Measure}\label{variable-importance-measure}}
+
+With the data structure above in hand, we let the estimand of interest be a
+\(j\)-specific variable importance measure (VIM) \(\Psi_j(P)\), which is defined
+through the \emph{true} (and unknown) probability distribution \(P\). As a motivating
+example, consider the case where \(\Psi_j(P)\) is
+\begin{equation}\label{vim_param}
+  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
+    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
+\end{equation}
+where \(Y(S_j)\) is the subvector of \(Y\) that contains the measured methylation
+values of the \emph{neighboring sites} of \(j\) (but not site \(j\) itself). This
+parameter of interest is a variant of the \emph{average treatment effect}, which has
+been the subject of much attention in statistical causal inference \textbf{CITATION
+NEEDED}. As a measure of variable importance, we seek to estimate this target
+parameter (\(\Psi_j(P)\)), which quantifies the effect of changing a treatment \(A\)
+on the methylation \(Y(j)\) of a CpG site \(j\), accounting for any potential
+confounding from phenotype-level covariates \(W\) and observed methylation
+\(Y(S_j)\) at the neighboring sites \(S_j\) of \(j\). We propose a procedure that
+estimates this target parameter \(\Psi_j(P)\) across all CpG sites of interest
+\(j: 1, \ldots, J\). When data-adaptive regression procedures (i.e., machine
+learning) are employed in estimating \(\Psi_j(P)\), this produces a nonparametric
+variable importance measure of differential methylation, allowing for the
+identification of differentially methylated positions (DMPs) while avoiding many
+assumptions common in the use of standard parametric regression procedures. We
+propose estimating \(\Psi_j(P)\) via targeted minimum loss-based estimation, which
+allows for data-adaptive regression procedures to be employed in a
+straightforward manner.
+
+\hypertarget{pre-screening-procedure}{%
+\subsection{Pre-Screening Procedure}\label{pre-screening-procedure}}
+
+As a matter of practicality, we consider only estimating the target VIM
+\(\Psi_j(P)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in fact
+span the full set of assayed CpG sites. In order to determine this subset of CpG
+sites, we propose the use of a pre-screening procedure, which need be nothing
+more than a method for differential methylation analysis that is computationally
+less demanding than the method proposed here. Formally, let the pre-screening
+procedure \(\theta(j): j \to \{0, 1\}\), which is to say that \(\theta\) takes as
+input a CpG site \(j\) and returns a binary decision rule of whether to include
+CpG site \(j\) in a subset to be evaluated further or not. Details on the
+implementation of pre-screening approaches is available in the sequel.
+
+\hypertarget{reducing-neighbors-via-clustering}{%
+\subsection{Reducing Neighbors via Clustering}\label{reducing-neighbors-via-clustering}}
+
+Due to the data-adaptive nature of the regression procedures employed in
+evaluating the target VIM \(\Psi_j(P)\) via targeted minimum loss-based
+estimation, it is possible the a given CpG site \(j\) may have \emph{too many}
+neighbors \(S_j\) to be controlled for in the estimation procedure. Heuristically,
+the inclusion of too many neighbors when controlling for potential confounders
+may lead to instability in the estimates produced. In such cases, we propose and
+implement the use of a clustering technique (e.g., Partitioning Around Medoids)
+to select a \emph{representative} subset of neighbors. Formally, there is likely no
+best choice of a specific clustering algorithm, so we leave this aspect of the
+proposed algorithm as flexible for the user \citep{kleinberg2003impossibility}.
+
+\hypertarget{targeted-minimum-loss-based-estimation-and-inference}{%
+\subsection{Targeted Minimum Loss-Based Estimation and Inference}\label{targeted-minimum-loss-based-estimation-and-inference}}
+
+Given the choice of target parameter \(\Psi_j(P)\), we propose the use of targeted
+minimum loss-based estimation (TMLE) to construct and evaluate an estimator
+\(\psi_{n, j}\) of \(\Psi_j(P)\). In the case of our motivating example, where the
+target VIM is the average treatment effect, we make use of a TML estimator of
+this parameter, which has been implemented for a general case in the
+\href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package \citep{gruber2011tmle}.
+Generally speaking, a TML estimator is constructed from a few simple components:
+(1) an estimator of the propensity score \citep{rosenbaum1983central} (often denoted
+by \(g(A, W)\)), (2) an estimator of the outcome regression (often denoted \(Q(A, W)\)), and (3) a targeting step applied to the initial estimator constructed from
+the aforementioned components \citep{gruber2009targeted}. In the sequel, we describe
+a few of the key mathematical properties of TML estimators relevant for
+statistical inference, deferring extended discussion of the technical details to
+\citet{vdl2011targeted} and \citet{vdl2018targeted}; moreover, as a matter of practicality, we
+make use of the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package of
+\citet{gruber2011tmle} as a backend for computing estimators of relevant target
+parameters in our software package, so the interested user is encouraged to
+discuss the documentation of that software package for further details.
+
+As noted previously, data-adaptive regression techniques may be used in the
+construction of TML estimators. This is done by generating a stacked regression
+model from a given set of machine learning algorithms via the Super Learner
+algorithm \citep[breiman1996stacked, vdl2007super]{wolpert1992stacked}, which ensures
+that the resulting ensemble learner satisifies important optimality properties
+with respect to cross-validation \citep[\citet{dudoit2005asymptotics},
+\citet{vdl2003unified}]{vdl2004asymptotic}. Importantly, TML estimators are well suited for inference,
+having an asymptotically normal limiting distribution
+\citep[\citet{vdl2006targeted}]{tsiatis2007semiparametric} \(\sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2)\), where \(\sigma_{\text{EIF}}^2\) is
+the variance of the efficient influence function of the target parameter
+evaluated at the observed data. This allows for confidence intervals to be
+constructed and and hypothesis tests (i.e., \(H_{0, j}: \Psi_j(P) = 0\)) to be
+performed in a straightforward manner.
+
+\hypertarget{correction-for-multiple-testing}{%
+\subsection{Correction for Multiple Testing}\label{correction-for-multiple-testing}}
+
+Given that we seek to estimate \(\Psi_j(P)\) for a possibly large number of CpG
+sites \((j: 1, \ldots J)\), the need to perform corrections for multiple testing
+is clear. In order to curb the potential for false discoveries, we recommend the
+use of the Benjamini and Hochberg procedure for controlling the False Discovery
+Rate (FDR) \citep{benjamini1995controlling}; however, as the proposed procedure
+involves a pre-screening step, naive application of the Benjamini and Hochberg
+procedure (BH) is invalid -- instead, we rely on a modification of the procedure
+to control the FDR, with established theoretical guarantees when pre-screening
+is employed \citep{tuglus2009modified}. In brief, the modified marginal Benjamini and
+Hochberg procedure to control the FDR under pre-screening works by applying the
+standard BH procedure to a padded vector of p-values -- that is, letting \(J\) be
+the number of CpG sites tested and \(K\) the number of CpG sites filtered out (so
+that \(J + K = P\), where \(P\) is the original dimension of the genomic assay), the
+modified marginal BH procedure is the application of the original BH procedure
+to a vector of p-values, composed of the \(J\) p-values from performing \(J\)
+hypothesis tests (of the form \(H_{0, j}: \Psi_j(P) = 0\)) and \(K\) additional
+p-values automatically set to \(1\) (for the hypothesis tests not performed on
+account of pre-screening). This procedure is guaranteed to control the FDR at
+the same desired rate when pre-screening is performed whereas naive application
+of the BH procedure fails to do so.
 
@@
 a manner that is very nearly assumption-free, that is, within a \emph{fully
-nonparametric statistical model}. \textbf{The statistical algorithm consists in
-several major steps}:
+nonparametric statistical model}. \textbf{The statistical algorithm consists in the
+several major steps elaborated upon above and summarized below}:
 
@@
 \item
-  Pre-screening of genomic sites is used to isolate a subset of sites for
-  which there is cursory evidence of differential methylation. For the sake of
-  computational feasibility, targeted minimum loss-based estimates of VIMs are
-  computed only for this subset of sites. Currently, the available screening
-  approach adapts core routines from the
-  \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R
-  package. Future releases will support functionality from other packages
-  (e.g., \href{https://CRAN.R-project.org/package=randomForest}{\texttt{randomForest}},
-  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}}). Following the
+  \emph{Pre-screening} of genomic sites is used to isolate a subset of sites for
+  which there is cursory evidence of differential methylation. Currently, the
+  available screening approach adapts core routines from the
+  \href{http://bioconductor.org/packages/limma}{\texttt{limma}} R package. Following the
   style of the function for performing screening via \texttt{limma}, users may write
   their own screening functions and are invited to contribute such functions to
-  the core software package by opening pull requests at the GitHub repository.
+  the core software package by opening pull requests at the GitHub repository:
+  \url{https://github.com/nhejazi/methyvim}.
 \item
@@
   computed at each of the CpG sites passing the screening step. The VIMs are
-  defined in such a way that the estimated effects is of an exposure/treatment
-  on the methylation status of a target CpG site, controlling for the observed
+  defined in such a way that the estimated effects is of an binary treatment on
+  the methylation status of a target CpG site, controlling for the observed
   methylation status of the neighbors of that site. Currently, routines are
   adapted from the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package.
-  Future releases will support doubly-robust estimates of these VIMs (via the
-  \href{https://cran.r-project.org/package=drtmle}{\texttt{drtmle}} package)
-  and add parameters for continuous treatments/exposures (via the
-  \href{https://CRAN.R-project.org/package=tmle.npvi}{\texttt{tmle.npvi}} and
-  \href{https://github.com/nhejazi/txshift}{\texttt{txshift}} R packages).
 \item
-  Since pre-screening is performed prior to estimating VIMs, we make use of a
-  multiple testing correction uniquely suited to such settings. Due to the
-  multiple testing nature of the estimation problem, a variant of the Benjamini
-  and Hochberg procedure for controlling the False Discovery Rate (FDR) is
-  applied \citep{benjamini1995controlling}. Specifically, we apply a modified
-  marginal Benjamini and Hochberg step-up False Discovery Rate controlling
-  procedure for multi-stage analyses (FDR-MSA), which has established
-  theoretical guarantees to control the FDR at the same rate as if all sites
-  were tested (i.e., without screening) \citep{tuglus2009modified}.
+  Since pre-screening is performed prior to estimating VIMs, we apply the
+  modified marginal Benjamini and Hochberg step-up False Discovery Rate
+  controlling procedure for multi-stage analyses (FDR-MSA), which is
+  well-suited for avoiding false positive discoveries when testing is only
+  performed on a subset of potential targets.
 \end{enumerate}
@@
 
-Support for continuous-valued treatments or exposures is \emph{planned but not yet
-available}. Future releases will allow users to assess continuous-valued
-treatments by relying on parameters estimable through implementations available
-in the following software packages:
+\hypertarget{future-work}{%
+\paragraph{Future Work}\label{future-work}}
+
+Support for continuous-valued treatments or exposures is \emph{planned and ongoing}
+but not yet available in stable releases of the \texttt{methyvim} software package.
+Future releases will allow users to assess continuous-valued treatments by
+relying on parameters estimable through implementations available in the
+following software packages:
 
@@
 
-We refer the reader to the package vignette, ``\texttt{methyvim}: Targeted Data-Adaptive
-Estimation and Inference for Differential Methylation Analysis,'' included in any
-distribution of the software package, for further details.
+The \texttt{show} method of the \texttt{methytmle} class summarizes a selection of the above
+information for the user while masking some of the wealth of information given
+when calling the same method for \texttt{GenomicRatioSet}. All information contained in
+\texttt{GenomicRatioSet} objects is preserved in \texttt{methytmle} objects, so as to easen
+interoperability with other differential methylation software for experienced
+users. We refer the reader to the package vignette, ``\texttt{methyvim}: Targeted
+Data-Adaptive Estimation and Inference for Differential Methylation Analysis,''
+included in any distribution of the software package, for further details.
 
@@
 \begin{Highlighting}[]
-\KeywordTok{suppressMessages}\NormalTok{(}\KeywordTok{library}\NormalTok{(minfiData))}
+\KeywordTok{suppressMessages}\NormalTok{(}
+  \CommentTok{# numerous messages displayed at time of loading}
+  \KeywordTok{library}\NormalTok{(minfiData)}
+\NormalTok{)}
 \KeywordTok{data}\NormalTok{(MsetEx)}
@@
 ##                   lwr_ci       est_ate        upr_ci      var_ate
-## cg14008030 -0.11956597277 -0.0314159619  0.0567340489 0.002022705230
-## cg20253340 -0.08850636712 -0.0588661418 -0.0292259165 0.000228691940
-## cg21870274 -0.09499057246 -0.0291189817  0.0367526091 0.001129494604
-## cg17308840 -0.04626018088 -0.0071524518  0.0319552773 0.000398119136
-## cg00645010 -0.02677328296 -0.0134655543 -0.0001578256 0.000046099449
-## cg27534567  0.06745648461  0.1157118536  0.1639672225 0.000606148645
-## cg08258224  0.13650451332  0.3050884951  0.4736724770 0.007398104677
-## cg20275697 -0.30210263597 -0.1231299608  0.0558427144 0.008337988977
-## cg24373735 -0.04283533168  0.0076666975  0.0581687266 0.000663904349
-## cg12445832 -0.06082410546  0.0150395574  0.0909032203 0.001498150599
-## cg01097950 -0.04392159158  0.0658323796  0.1755863507 0.003135655504
-## cg01782097 -0.01082071508  0.0010232901  0.0128672954 0.000036516155
-##                                                      pval n_neighbors
-## cg14008030 0.48484680819856229572195616128738038241863251           0
-## cg20253340 0.00009917426050587487018107835101687896894873           0
-## cg21870274 0.38625371823607712595816110479063354432582855           2
-## cg17308840 0.71999433317147576438799205789109691977500916           2
-## cg00645010 0.04734006904146480232409288646522327326238155           2
-## cg27534567 0.00000260293555116907031197386393484016764432           1
-## cg08258224 0.00038959136906232085580809032521187873499002           1
-## cg20275697 0.17751545277429223168574878854997223243117332           0
-## cg24373735 0.76604893436462262457098404411226511001586914           0
-## cg12445832 0.69760217179702999068524604808771982789039612           0
-## cg01097950 0.23973765213124559325663653908122796565294266           0
-## cg01782097 0.86553022244577304533663664187770336866378784           1
-##            n_neighbors_control max_cor_neighbors
-## cg14008030                   0                NA
-## cg20253340                   0                NA
-## cg21870274                   1        0.94435796
-## cg17308840                   1        0.94435796
-## cg00645010                   2        0.52368097
-## cg27534567                   0        0.93629683
-## cg08258224                   0        0.93629683
-## cg20275697                   0                NA
-## cg24373735                   0                NA
-## cg12445832                   0                NA
-## cg01097950                   0                NA
-## cg01782097                   1       -0.39410834
+## cg14008030 -1.195660e-01 -0.0314159619  0.0567340489 2.022705e-03
+## cg20253340 -8.850637e-02 -0.0588661418 -0.0292259165 2.286919e-04
+## cg21870274 -9.499057e-02 -0.0291189817  0.0367526091 1.129495e-03
+## cg17308840 -4.626018e-02 -0.0071524518  0.0319552773 3.981191e-04
+## cg00645010 -2.677328e-02 -0.0134655543 -0.0001578256 4.609945e-05
+## cg27534567  6.745648e-02  0.1157118536  0.1639672225 6.061486e-04
+## cg08258224  1.365045e-01  0.3050884951  0.4736724770 7.398105e-03
+## cg20275697 -3.021026e-01 -0.1231299608  0.0558427144 8.337989e-03
+## cg24373735 -4.283533e-02  0.0076666975  0.0581687266 6.639043e-04
+## cg12445832 -6.082411e-02  0.0150395574  0.0909032203 1.498151e-03
+## cg01097950 -4.392159e-02  0.0658323796  0.1755863507 3.135656e-03
+## cg01782097 -1.082072e-02  0.0010232901  0.0128672954 3.651615e-05
+##                    pval n_neighbors n_neighbors_control max_cor_neighbors
+## cg14008030 4.848468e-01           0                   0                NA
+## cg20253340 9.917426e-05           0                   0                NA
+## cg21870274 3.862537e-01           2                   1        0.94435796
+## cg17308840 7.199943e-01           2                   1        0.94435796
+## cg00645010 4.734007e-02           2                   2        0.52368097
+## cg27534567 2.602936e-06           1                   0        0.93629683
+## cg08258224 3.895914e-04           1                   0        0.93629683
+## cg20275697 1.775155e-01           0                   0                NA
+## cg24373735 7.660489e-01           0                   0                NA
+## cg12445832 6.976022e-01           0                   0                NA
+## cg01097950 2.397377e-01           0                   0                NA
+## cg01782097 8.655302e-01           1                   1       -0.39410834
 ##  [ reached getOption("max.print") -- omitted 238 rows ]
@@
 \begin{Highlighting}[]
-\NormalTok{fdr_p <-}\StringTok{ }\KeywordTok{fdr_msa}\NormalTok{(}\DataTypeTok{pvals =}\NormalTok{ methyvim_cancer_ate}\OperatorTok{@}\NormalTok{vim}\OperatorTok{$}\NormalTok{pval,}
+\NormalTok{fdr_p <-}\StringTok{ }\KeywordTok{fdr_msa}\NormalTok{(}\DataTypeTok{pvals =} \KeywordTok{vim}\NormalTok{(methyvim_cancer_ate)}\OperatorTok{$}\NormalTok{pval,}
                  \DataTypeTok{total_obs =} \KeywordTok{nrow}\NormalTok{(methyvim_cancer_ate))}

### 2018-06-29 R/methytmle-class.R

@@ -11,12 +11,15 @@
 #' @export .methytmle
 #' @exportClass methytmle
 #'
+#' @rdname methytmle-class
+#'
 #' @examples
 #' library(methyvimData)
 #' suppressMessages(library(SummarizedExperiment))
 #' data(grsExample)
 #' # cast the GenomicRatioSet to class methytmle
 #' methy_tmle <- .methytmle(grsExample)
+#' methy_tmle
 #
 .methytmle <- methods::setClass(
   Class = "methytmle",
@@ -46,19 +49,33 @@ methods::setMethod("show", "methytmle", function(object) {
 
 ################################################################################
 
-#' @importFrom methods setGeneric setMethod
+#' Accessor for Parameter Information
+#'
+#' @param object S4 object of class \code{methytmle}.
+#'
+#' @rdname methytmle-class
+#'
+#' @keywords internal
 #
-methods::setGeneric("param", function(object) standardGeneric("param"))
-methods::setMethod("param", "methytmle", function(object) {
+param <- function(object) {
+  stopifnot(class(object) == "methytmle")
   object@param
-})
+}
 
 ################################################################################
 
-#' @importFrom methods setGeneric setMethod
+#' Accessor for Variable Importance Measure Information
+#'
+#' @param object S4 object of class \code{methytmle}.
+#'
+#' @rdname methytmle-class
+#'
+#' @keywords internal
+#'
+#' @export
 #
-methods::setGeneric("vim", function(object) standardGeneric("vim"))
-methods::setMethod("vim", "methytmle", function(object) {
+vim <- function(object) {
+  stopifnot(class(object) == "methytmle")
   object@vim
-})
+}
 

### 2018-07-06 paper/paper_BiocF1000.Rmd
manuscript update

@@
 abstract: |
-  We present `methyvim`, an R package implementing a general algorithm for the
-  nonparametric estimation of treatment effects on DNA methylation at CpG sites
-  throughout the genome, complete with straightforward statistical inference for
-  such estimates. The approach leverages variable importance measures derived
-  from statistical parameters arising in causal inference, defined in such a
-  manner that they may be used to obtain targeted estimates of the relative
-  importance of individual CpG sites with respect to a binary treatment assigned
-  at the phenotype level. The procedure implemented is computationally
-  efficient, incorporating a preliminary screening step to isolate a subset of
-  sites for which there is cursory evidence of differential methylation as well
-  as a unique multiple testing correction to control the False Discovery Rate
-  with the same rigor as if all sites were tested. This technique for analysis
-  of differentially methylated positions provides an avenue to incorporate
-  flexible state-of-the-art machine learning algorithms into the estimation of
-  differential methylation effects without the loss of interpretable statistical
-  inference.
+  We present `methyvim`, an R package implementing an algorithm for the
+  nonparametric estimation of the effects of exposures on DNA methylation at CpG
+  sites throughout the genome, complete with straightforward statistical
+  inference for such estimates. The approach leverages variable importance
+  measures derived from statistical parameters arising in causal inference,
+  defined in such a manner that they may be used to obtain targeted estimates of
+  the relative importance of individual CpG sites with respect to a binary
+  treatment assigned at the phenotype level, thereby providing a new approach to
+  identifying differentially methylated positions. The procedure implemented is
+  computationally efficient, incorporating a preliminary screening step to
+  isolate a subset of sites for which there is cursory evidence of differential
+  methylation as well as a unique multiple testing correction to control the
+  False Discovery Rate with the same rigor as would be available if all sites
+  were subjected to testing. This novel technique for analysis of differentially
+  methylated positions provides an avenue for incorporating flexible
+  state-of-the-art data-adaptive regression procedures (i.e., machine learning)
+  into the estimation of differential methylation effects without the loss of
+  interpretable statistical inference for the estimated quantity.
 keywords: DNA methylation, differential methylation, epigenetics, causal
@@
 
-## Data Structure
-
-We consider an observed data structure, on a single experimental subject (e.g.,
-a patient), $O = (W, A, (Y(j) : j))$, where $(Y(j) : j = 1, \ldots J)$ is the
-set of CpG sites measured by the assay in question, $A \in \{0, 1\}$ represents
-a binary phenotype-level treatment, and $W$ is a vector of the phenotype-level
-baseline covariates that are potential confounders (e.g., age, sex). We consider
-having access to measurements on a large number $J$ of CpG sites (e.g.,
-$850,000$, as measured by the Illumina _MethylationEPIC_ BeadChip arrays).
-Further, let $K: j \to S_j$ be a procedure that assigns to a given CpG site $j$
-a set of neighbors $S_j$ -- that is, $S_j$ is a collection of indices of the
-neighbors of $j$ just as $j$ indexes the full set of CpG sites; moreover, since
-$Y(j)$ is the measured methylation value for a given CpG site $j$, $Y(S_j)$ is
-the measured methylation values at the neighbors $S_j$ of $j$. Note that the
-definition of a neighborhood $\{j, S_j\}$ is left vague so as to facilitate the
-use of user-specified strategies in implementation. We consider the case of
-observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P \in
-\mathcal{M}$, which is to say that the random variable $O$ is governed by an
-unknown probability distribution $P$, assumed only to reside in a nonparametric
-statistical model $\mathcal{M}$ that places no restrictions on the
-data-generating process.
-
-## Variable Importance Measure
-
-With the data structure above in hand, we let the estimand of interest be a
-$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
-through the _true_ (and unknown) probability distribution $P$. As a motivating
-example, consider the case where $\Psi_j(P)$ is
-\begin{equation}\label{vim_param}
-  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
-    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
-\end{equation}
-where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
-values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
-parameter of interest is a variant of the _average treatment effect_, which has
-been the subject of much attention in statistical causal inference **CITATION
-NEEDED**. As a measure of variable importance, we seek to estimate this target
-parameter ($\Psi_j(P)$), which quantifies the effect of changing a treatment $A$
-on the methylation $Y(j)$ of a CpG site $j$, accounting for any potential
-confounding from phenotype-level covariates $W$ and observed methylation
-$Y(S_j)$ at the neighboring sites $S_j$ of $j$. We propose a procedure that
-estimates this target parameter $\Psi_j(P)$ across all CpG sites of interest
-$j: 1, \ldots, J$. When data-adaptive regression procedures (i.e., machine
-learning) are employed in estimating $\Psi_j(P)$, this produces a nonparametric
-variable importance measure of differential methylation, allowing for the
-identification of differentially methylated positions (DMPs) while avoiding many
-assumptions common in the use of standard parametric regression procedures. We
-propose estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which
-allows for data-adaptive regression procedures to be employed in a
-straightforward manner.
-
-## Pre-Screening Procedure
-
-As a matter of practicality, we consider only estimating the target VIM
-$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
-span the full set of assayed CpG sites. In order to determine this subset of CpG
-sites, we propose the use of a pre-screening procedure, which need be nothing
-more than a method for differential methylation analysis that is computationally
-less demanding than the method proposed here. Formally, let the pre-screening
-procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
-input a CpG site $j$ and returns a binary decision rule of whether to include
-CpG site $j$ in a subset to be evaluated further or not. Details on the
-implementation of pre-screening approaches is available in the sequel.
-
-## Reducing Neighbors via Clustering
-
-Due to the data-adaptive nature of the regression procedures employed in
-evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
-estimation, it is possible the a given CpG site $j$ may have _too many_
-neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
-the inclusion of too many neighbors when controlling for potential confounders
-may lead to instability in the estimates produced. In such cases, we propose and
-implement the use of a clustering technique (e.g., Partitioning Around Medoids)
-to select a _representative_ subset of neighbors. Formally, there is likely no
-best choice of a specific clustering algorithm, so we leave this aspect of the
-proposed algorithm as flexible for the user [@kleinberg2003impossibility].
-
-## Targeted Minimum Loss-Based Estimation and Inference
-
-Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
-minimum loss-based estimation (TMLE) to construct and evaluate an estimator
-$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
-target VIM is the average treatment effect, we make use of a TML estimator of
-this parameter, which has been implemented for a general case in the
-[`tmle`](https://CRAN.R-project.org/package=tmle) R package [@gruber2011tmle].
-Generally speaking, a TML estimator is constructed from a few simple components:
-(1) an estimator of the propensity score [@rosenbaum1983central] (often denoted
-by $g(A, W)$), (2) an estimator of the outcome regression (often denoted $Q(A,
-W)$), and (3) a targeting step applied to the initial estimator constructed from
-the aforementioned components [@gruber2009targeted]. In the sequel, we describe
-a few of the key mathematical properties of TML estimators relevant for
-statistical inference, deferring extended discussion of the technical details to
-@vdl2011targeted and @vdl2018targeted; moreover, as a matter of practicality, we
-make use of the [`tmle`](https://CRAN.R-project.org/package=tmle)  R package of
-@gruber2011tmle as a backend for computing estimators of relevant target
-parameters in our software package, so the interested user is encouraged to
-discuss the documentation of that software package for further details.
-
-As noted previously, data-adaptive regression techniques may be used in the
-construction of TML estimators. This is done by generating a stacked regression
-model from a given set of machine learning algorithms via the Super Learner
-algorithm [@wolpert1992stacked, breiman1996stacked, vdl2007super], which ensures
-that the resulting ensemble learner satisifies important optimality properties
-with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
-@vdl2003unified]. Importantly, TML estimators are well suited for inference,
-having an asymptotically normal limiting distribution
-[@tsiatis2007semiparametric, @vdl2006targeted] $\sqrt{n}(\psi_{n, j} -
-\Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2)$, where $\sigma_{\text{EIF}}^2$ is
-the variance of the efficient influence function of the target parameter
-evaluated at the observed data. This allows for confidence intervals to be
-constructed and and hypothesis tests (i.e., $H_{0, j}: \Psi_j(P) = 0$) to be
-performed in a straightforward manner.
-
-## Correction for Multiple Testing
-
-Given that we seek to estimate $\Psi_j(P)$ for a possibly large number of CpG
-sites $(j: 1, \ldots J)$, the need to perform corrections for multiple testing
-is clear. In order to curb the potential for false discoveries, we recommend the
-use of the Benjamini and Hochberg procedure for controlling the False Discovery
-Rate (FDR) [@benjamini1995controlling]; however, as the proposed procedure
-involves a pre-screening step, naive application of the Benjamini and Hochberg
-procedure (BH) is invalid -- instead, we rely on a modification of the procedure
-to control the FDR, with established theoretical guarantees when pre-screening
-is employed [@tuglus2009modified]. In brief, the modified marginal Benjamini and
-Hochberg procedure to control the FDR under pre-screening works by applying the
-standard BH procedure to a padded vector of p-values -- that is, letting $J$ be
-the number of CpG sites tested and $K$ the number of CpG sites filtered out (so
-that $J + K = P$, where $P$ is the original dimension of the genomic assay), the
-modified marginal BH procedure is the application of the original BH procedure
-to a vector of p-values, composed of the $J$ p-values from performing $J$
-hypothesis tests (of the form $H_{0, j}: \Psi_j(P) = 0$) and $K$ additional
-p-values automatically set to $1$ (for the hypothesis tests not performed on
-account of pre-screening). This procedure is guaranteed to control the FDR at
-the same desired rate when pre-screening is performed whereas naive application
-of the BH procedure fails to do so.
-
 ## Implementation
@@
 a manner that is very nearly assumption-free, that is, within a _fully
-nonparametric statistical model_. __The statistical algorithm consists in the
-several major steps elaborated upon above and summarized below__:
+nonparametric statistical model_. The statistical algorithm consists in the
+several major steps summarized below. Additional methodological details on the
+use of targeted minimum loss-based estimation in this problem setting is
+provided in a brief appendix (see section __Appendix__).
 
@@
 
-For the CpG sites that pass the pre-screening step, a user-specified target
+For CpG sites that pass the pre-screening step, a user-specified target
 parameter of interest is estimated independently at each site. _In all cases,
@@
 
-For discrete-valued treatments or exposures:
+Two popular target causal parameters for discrete-valued treatments or
+exposures are
 
@@
 
+<!--
 #### Future Work
@@
   respect to another variable measured in the same study.
+
+-->
 
@@
 
+# Appendix
+
+## Data Structure
+
+We consider an observed data structure, on a single experimental subject (e.g.,
+a patient), $O = (W, A, (Y(j) : j))$, where $(Y(j) : j = 1, \ldots J)$ is the
+set of CpG sites measured by the assay in question, $A \in \{0, 1\}$ represents
+a binary phenotype-level treatment, and $W$ is a vector of the phenotype-level
+baseline covariates that are potential confounders (e.g., age, sex). We consider
+having access to measurements on a large number $J$ of CpG sites (e.g.,
+$850,000$, as measured by the Illumina _MethylationEPIC_ BeadChip arrays).
+Further, let $K: j \to S_j$ be a procedure that assigns to a given CpG site $j$
+a set of neighbors $S_j$ -- that is, $S_j$ is a collection of indices of the
+neighbors of $j$ just as $j$ indexes the full set of CpG sites; moreover, since
+$Y(j)$ is the measured methylation value for a given CpG site $j$, $Y(S_j)$ is
+the measured methylation values at the neighbors $S_j$ of $j$. Note that the
+definition of a neighborhood $\{j, S_j\}$ is left vague so as to facilitate the
+use of user-specified strategies in implementation. We consider the case of
+observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P \in
+\mathcal{M}$, which is to say that the random variable $O$ is governed by an
+unknown probability distribution $P$, assumed only to reside in a nonparametric
+statistical model $\mathcal{M}$ that places no restrictions on the
+data-generating process.
+
+## Variable Importance Measure
+
+With the data structure above in hand, we let the estimand of interest be a
+$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
+through the _true_ (and unknown) probability distribution $P$. As a motivating
+example, consider the case where $\Psi_j(P)$ is
+\begin{equation}\label{vim_param}
+  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
+    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
+\end{equation}
+where $Y(S_j)$ is the subvector of $Y$ that contains the measured methylation
+values of the _neighboring sites_ of $j$ (but not site $j$ itself). This
+parameter of interest is a variant of the _average treatment effect_, which has
+been the subject of much attention in statistical causal inference
+[@holland1986statistics, @hahn1998role, @hirano2003efficient]. As a measure of
+variable importance, we seek to estimate this target parameter ($\Psi_j(P)$),
+which quantifies the effect of changing a treatment $A$ on the methylation
+$Y(j)$ of a CpG site $j$, accounting for any potential confounding from
+phenotype-level covariates $W$ and observed methylation $Y(S_j)$ at the
+neighboring sites $S_j$ of $j$. We propose a procedure that estimates this
+target parameter $\Psi_j(P)$ across all CpG sites of interest $j: 1, \ldots, J$.
+When data-adaptive regression procedures (i.e., machine learning) are employed
+in estimating $\Psi_j(P)$, this produces a nonparametric variable importance
+measure of differential methylation, allowing for the identification of
+differentially methylated positions (DMPs) while avoiding many assumptions
+common in the use of standard parametric regression procedures. We propose
+estimating $\Psi_j(P)$ via targeted minimum loss-based estimation, which allows
+for data-adaptive regression procedures to be employed in a straightforward
+manner using the Super Learner algorithm for ensemble modeling [@vdl2007super,
+@wolpert1992stacked, @breiman1996stacked, @gruber2009targeted,
+@vdl2011targeted].
+
+## Pre-Screening Procedure
+
+As a matter of practicality, we consider only estimating the target VIM
+$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
+span the full set of assayed CpG sites. In order to determine this subset of CpG
+sites, we propose the use of a pre-screening procedure, which need be nothing
+more than a method for differential methylation analysis that is computationally
+less demanding than the method proposed here. Formally, let the pre-screening
+procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
+input a CpG site $j$ and returns a binary decision rule of whether to include
+CpG site $j$ in a subset to be evaluated further or not. As an example, one
+might consider employing a classical differential methylation analysis
+procedure, such as the linear modeling approach of the
+[`limma`](https://bioconductor.org/packages/limma) R package
+[@smyth2004linear, @robinson2014statistical], using only CpG sites that pass a
+reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
+p-values) for the subsequent analysis steps in the `methyvim` pipeline. Details
+on the implementation of pre-screening approaches is provided in the sequel.
+
+## Reducing Neighbors via Clustering
+
+Due to the data-adaptive nature of the regression procedures employed in
+evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
+estimation, it is possible the a given CpG site $j$ may have _too many_
+neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
+the inclusion of too many neighbors when controlling for potential confounders
+may lead to instability in the estimates produced. In such cases, we propose and
+implement the use of a clustering technique (e.g., partitioning around medoids)
+to select a _representative_ subset of neighbors. Formally, there is likely no
+best choice of a specific clustering algorithm, so we leave this aspect of the
+proposed algorithm as flexible for the user [@kleinberg2003impossibility]. Note
+that the goal of employing a clustering procedure in `methyvim` is to obtain a
+smaller but still highly representative set of neighbors $S(j)$, so as to allow
+for estimates of $\Psi_j(P)$ to account for as much confounding from neighboring
+sites as is allowed by the available data.
+
+## Targeted Minimum Loss-Based Estimation and Statistical Inference
+
+Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
+minimum loss-based estimation (TMLE) to construct and evaluate an estimator
+$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
+target VIM is based on the average treatment effect, we make use of a TML
+estimator of this parameter, which has been implemented for a general case in
+the [`tmle`](https://CRAN.R-project.org/package=tmle) R package
+[@gruber2011tmle]; users of `methyvim` may wish to consult the documentation of
+that software package as a supplement. Generally speaking, a TML estimator is
+constructed from a few simple components: (1) an estimator of the propensity
+score [@rosenbaum1983central], often denoted by $g(A \mid W)$; (2) an estimator
+of the outcome regression, often denoted $Q(A, W)$; and (3) a targeting step
+applied to the initial estimators constructed from the aforementioned components
+[@gruber2009targeted]. While we describe a few of the key properties of TML
+estimators in the sequel, extended technical discussion is deferred to more
+comprehensive work [@vdl2011targeted, @vdl2018targeted]. In order to construct
+an estimate $\psi_{n, j}$ of $\Psi_j(P)$, it is necessary to accurately estimate
+to nuisance parameters, these being the propensity score ($g(A \mid W)$) and the
+outcome regression ($Q(A, W)$); moreover, data-adaptive regression procedures
+may be used to obtain consistent estimates of these quantities through the
+creation of a stacked regression model via the Super Learner algorithm
+[@wolpert1992stacked, @breiman1996stacked, @vdl2007super], which ensures
+that the resulting ensemble learner satisfies important optimality properties
+with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
+@vdl2003unified]. After constructing such estimates, an iterative procedure
+(i.e., the targeting step) may be used to combine these individual estimates
+into an estimate of the target parameter $\Psi_j(P)$.
+
+Importantly, TML estimators are well-suited for statistical inference, having an
+asymptotically normal limiting distribution [@tsiatis2007semiparametric,
+@vdl2006targeted], which allows for a closed-form expression of the variance
+to be derived:
+\begin{equation}\label{eqn:lim_psi}
+  \sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2),
+\end{equation}
+where $\sigma_{\text{EIF}}^2$ is the variance of the efficient influence
+function, with respect to a nonparametric statistical model, of the target
+parameter evaluated at the observed data. Such a convenience allows for
+confidence intervals and hypothesis tests to be constructed (i.e.,
+$H_{0, j}: \Psi_j(P) = 0$) in a straightforward manner.
+
+## Correction for Multiple Testing
+
+Given that we seek to estimate $\Psi_j(P)$ for a possibly large number of CpG
+sites $(j: 1, \ldots J)$, the need to perform corrections for multiple testing
+is clear. In order to curb the potential for false discoveries, we recommend the
+use of the Benjamini and Hochberg procedure for controlling the False Discovery
+Rate (FDR) [@benjamini1995controlling]; however, as the proposed procedure
+involves a pre-screening step, naive application of the Benjamini and Hochberg
+procedure (BH) is invalid -- instead, we rely on a modification of the procedure
+to control the FDR, with established theoretical guarantees when pre-screening
+is employed [@tuglus2009modified]. In brief, the modified marginal Benjamini and
+Hochberg procedure to control the FDR under pre-screening works by applying the
+standard BH procedure to a padded vector of p-values -- that is, letting $J$ be
+the number of CpG sites tested and $K$ the number of CpG sites filtered out (so
+that $J + K = P$, where $P$ is the original dimension of the genomic assay), the
+modified marginal BH procedure is the application of the original BH procedure
+to a vector of p-values, composed of the $J$ p-values from performing $J$
+hypothesis tests (of the form $H_{0, j}: \Psi_j(P) = 0$) and $K$ additional
+p-values automatically set to $1$ (for the hypothesis tests not performed on
+account of pre-screening). This procedure is guaranteed to control the FDR at
+the same desired rate when pre-screening is performed whereas naive application
+of the BH procedure fails to do so.
+
 # References

### 2018-07-06 paper/paper_BiocF1000.tex
manuscript update

@@
 \begin{abstract}
-We present \texttt{methyvim}, an R package implementing a general algorithm for the
-nonparametric estimation of treatment effects on DNA methylation at CpG sites
-throughout the genome, complete with straightforward statistical inference for
-such estimates. The approach leverages variable importance measures derived
-from statistical parameters arising in causal inference, defined in such a
-manner that they may be used to obtain targeted estimates of the relative
-importance of individual CpG sites with respect to a binary treatment assigned
-at the phenotype level. The procedure implemented is computationally
-efficient, incorporating a preliminary screening step to isolate a subset of
-sites for which there is cursory evidence of differential methylation as well
-as a unique multiple testing correction to control the False Discovery Rate
-with the same rigor as if all sites were tested. This technique for analysis
-of differentially methylated positions provides an avenue to incorporate
-flexible state-of-the-art machine learning algorithms into the estimation of
-differential methylation effects without the loss of interpretable statistical
-inference.
+We present \texttt{methyvim}, an R package implementing an algorithm for the
+nonparametric estimation of the effects of exposures on DNA methylation at CpG
+sites throughout the genome, complete with straightforward statistical
+inference for such estimates. The approach leverages variable importance
+measures derived from statistical parameters arising in causal inference,
+defined in such a manner that they may be used to obtain targeted estimates of
+the relative importance of individual CpG sites with respect to a binary
+treatment assigned at the phenotype level, thereby providing a new approach to
+identifying differentially methylated positions. The procedure implemented is
+computationally efficient, incorporating a preliminary screening step to
+isolate a subset of sites for which there is cursory evidence of differential
+methylation as well as a unique multiple testing correction to control the
+False Discovery Rate with the same rigor as would be available if all sites
+were subjected to testing. This novel technique for analysis of differentially
+methylated positions provides an avenue for incorporating flexible
+state-of-the-art data-adaptive regression procedures (i.e., machine learning)
+into the estimation of differential methylation effects without the loss of
+interpretable statistical inference for the estimated quantity.
 \end{abstract}
@@
 
-\hypertarget{data-structure}{%
-\subsection{Data Structure}\label{data-structure}}
-
-We consider an observed data structure, on a single experimental subject (e.g.,
-a patient), \(O = (W, A, (Y(j) : j))\), where \((Y(j) : j = 1, \ldots J)\) is the
-set of CpG sites measured by the assay in question, \(A \in \{0, 1\}\) represents
-a binary phenotype-level treatment, and \(W\) is a vector of the phenotype-level
-baseline covariates that are potential confounders (e.g., age, sex). We consider
-having access to measurements on a large number \(J\) of CpG sites (e.g.,
-\(850,000\), as measured by the Illumina \emph{MethylationEPIC} BeadChip arrays).
-Further, let \(K: j \to S_j\) be a procedure that assigns to a given CpG site \(j\)
-a set of neighbors \(S_j\) -- that is, \(S_j\) is a collection of indices of the
-neighbors of \(j\) just as \(j\) indexes the full set of CpG sites; moreover, since
-\(Y(j)\) is the measured methylation value for a given CpG site \(j\), \(Y(S_j)\) is
-the measured methylation values at the neighbors \(S_j\) of \(j\). Note that the
-definition of a neighborhood \(\{j, S_j\}\) is left vague so as to facilitate the
-use of user-specified strategies in implementation. We consider the case of
-observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where \(O \sim P \in \mathcal{M}\), which is to say that the random variable \(O\) is governed by an
-unknown probability distribution \(P\), assumed only to reside in a nonparametric
-statistical model \(\mathcal{M}\) that places no restrictions on the
-data-generating process.
-
-\hypertarget{variable-importance-measure}{%
-\subsection{Variable Importance Measure}\label{variable-importance-measure}}
-
-With the data structure above in hand, we let the estimand of interest be a
-\(j\)-specific variable importance measure (VIM) \(\Psi_j(P)\), which is defined
-through the \emph{true} (and unknown) probability distribution \(P\). As a motivating
-example, consider the case where \(\Psi_j(P)\) is
-\begin{equation}\label{vim_param}
-  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
-    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
-\end{equation}
-where \(Y(S_j)\) is the subvector of \(Y\) that contains the measured methylation
-values of the \emph{neighboring sites} of \(j\) (but not site \(j\) itself). This
-parameter of interest is a variant of the \emph{average treatment effect}, which has
-been the subject of much attention in statistical causal inference \textbf{CITATION
-NEEDED}. As a measure of variable importance, we seek to estimate this target
-parameter (\(\Psi_j(P)\)), which quantifies the effect of changing a treatment \(A\)
-on the methylation \(Y(j)\) of a CpG site \(j\), accounting for any potential
-confounding from phenotype-level covariates \(W\) and observed methylation
-\(Y(S_j)\) at the neighboring sites \(S_j\) of \(j\). We propose a procedure that
-estimates this target parameter \(\Psi_j(P)\) across all CpG sites of interest
-\(j: 1, \ldots, J\). When data-adaptive regression procedures (i.e., machine
-learning) are employed in estimating \(\Psi_j(P)\), this produces a nonparametric
-variable importance measure of differential methylation, allowing for the
-identification of differentially methylated positions (DMPs) while avoiding many
-assumptions common in the use of standard parametric regression procedures. We
-propose estimating \(\Psi_j(P)\) via targeted minimum loss-based estimation, which
-allows for data-adaptive regression procedures to be employed in a
-straightforward manner.
-
-\hypertarget{pre-screening-procedure}{%
-\subsection{Pre-Screening Procedure}\label{pre-screening-procedure}}
-
-As a matter of practicality, we consider only estimating the target VIM
-\(\Psi_j(P)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in fact
-span the full set of assayed CpG sites. In order to determine this subset of CpG
-sites, we propose the use of a pre-screening procedure, which need be nothing
-more than a method for differential methylation analysis that is computationally
-less demanding than the method proposed here. Formally, let the pre-screening
-procedure \(\theta(j): j \to \{0, 1\}\), which is to say that \(\theta\) takes as
-input a CpG site \(j\) and returns a binary decision rule of whether to include
-CpG site \(j\) in a subset to be evaluated further or not. Details on the
-implementation of pre-screening approaches is available in the sequel.
-
-\hypertarget{reducing-neighbors-via-clustering}{%
-\subsection{Reducing Neighbors via Clustering}\label{reducing-neighbors-via-clustering}}
-
-Due to the data-adaptive nature of the regression procedures employed in
-evaluating the target VIM \(\Psi_j(P)\) via targeted minimum loss-based
-estimation, it is possible the a given CpG site \(j\) may have \emph{too many}
-neighbors \(S_j\) to be controlled for in the estimation procedure. Heuristically,
-the inclusion of too many neighbors when controlling for potential confounders
-may lead to instability in the estimates produced. In such cases, we propose and
-implement the use of a clustering technique (e.g., Partitioning Around Medoids)
-to select a \emph{representative} subset of neighbors. Formally, there is likely no
-best choice of a specific clustering algorithm, so we leave this aspect of the
-proposed algorithm as flexible for the user \citep{kleinberg2003impossibility}.
-
-\hypertarget{targeted-minimum-loss-based-estimation-and-inference}{%
-\subsection{Targeted Minimum Loss-Based Estimation and Inference}\label{targeted-minimum-loss-based-estimation-and-inference}}
-
-Given the choice of target parameter \(\Psi_j(P)\), we propose the use of targeted
-minimum loss-based estimation (TMLE) to construct and evaluate an estimator
-\(\psi_{n, j}\) of \(\Psi_j(P)\). In the case of our motivating example, where the
-target VIM is the average treatment effect, we make use of a TML estimator of
-this parameter, which has been implemented for a general case in the
-\href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package \citep{gruber2011tmle}.
-Generally speaking, a TML estimator is constructed from a few simple components:
-(1) an estimator of the propensity score \citep{rosenbaum1983central} (often denoted
-by \(g(A, W)\)), (2) an estimator of the outcome regression (often denoted \(Q(A, W)\)), and (3) a targeting step applied to the initial estimator constructed from
-the aforementioned components \citep{gruber2009targeted}. In the sequel, we describe
-a few of the key mathematical properties of TML estimators relevant for
-statistical inference, deferring extended discussion of the technical details to
-\citet{vdl2011targeted} and \citet{vdl2018targeted}; moreover, as a matter of practicality, we
-make use of the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package of
-\citet{gruber2011tmle} as a backend for computing estimators of relevant target
-parameters in our software package, so the interested user is encouraged to
-discuss the documentation of that software package for further details.
-
-As noted previously, data-adaptive regression techniques may be used in the
-construction of TML estimators. This is done by generating a stacked regression
-model from a given set of machine learning algorithms via the Super Learner
-algorithm \citep[breiman1996stacked, vdl2007super]{wolpert1992stacked}, which ensures
-that the resulting ensemble learner satisifies important optimality properties
-with respect to cross-validation \citep[\citet{dudoit2005asymptotics},
-\citet{vdl2003unified}]{vdl2004asymptotic}. Importantly, TML estimators are well suited for inference,
-having an asymptotically normal limiting distribution
-\citep[\citet{vdl2006targeted}]{tsiatis2007semiparametric} \(\sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2)\), where \(\sigma_{\text{EIF}}^2\) is
-the variance of the efficient influence function of the target parameter
-evaluated at the observed data. This allows for confidence intervals to be
-constructed and and hypothesis tests (i.e., \(H_{0, j}: \Psi_j(P) = 0\)) to be
-performed in a straightforward manner.
-
-\hypertarget{correction-for-multiple-testing}{%
-\subsection{Correction for Multiple Testing}\label{correction-for-multiple-testing}}
-
-Given that we seek to estimate \(\Psi_j(P)\) for a possibly large number of CpG
-sites \((j: 1, \ldots J)\), the need to perform corrections for multiple testing
-is clear. In order to curb the potential for false discoveries, we recommend the
-use of the Benjamini and Hochberg procedure for controlling the False Discovery
-Rate (FDR) \citep{benjamini1995controlling}; however, as the proposed procedure
-involves a pre-screening step, naive application of the Benjamini and Hochberg
-procedure (BH) is invalid -- instead, we rely on a modification of the procedure
-to control the FDR, with established theoretical guarantees when pre-screening
-is employed \citep{tuglus2009modified}. In brief, the modified marginal Benjamini and
-Hochberg procedure to control the FDR under pre-screening works by applying the
-standard BH procedure to a padded vector of p-values -- that is, letting \(J\) be
-the number of CpG sites tested and \(K\) the number of CpG sites filtered out (so
-that \(J + K = P\), where \(P\) is the original dimension of the genomic assay), the
-modified marginal BH procedure is the application of the original BH procedure
-to a vector of p-values, composed of the \(J\) p-values from performing \(J\)
-hypothesis tests (of the form \(H_{0, j}: \Psi_j(P) = 0\)) and \(K\) additional
-p-values automatically set to \(1\) (for the hypothesis tests not performed on
-account of pre-screening). This procedure is guaranteed to control the FDR at
-the same desired rate when pre-screening is performed whereas naive application
-of the BH procedure fails to do so.
-
 \hypertarget{implementation}{%
@@
 a manner that is very nearly assumption-free, that is, within a \emph{fully
-nonparametric statistical model}. \textbf{The statistical algorithm consists in the
-several major steps elaborated upon above and summarized below}:
+nonparametric statistical model}. The statistical algorithm consists in the
+several major steps summarized below. Additional methodological details on the
+use of targeted minimum loss-based estimation in this problem setting is
+provided in a brief appendix (see section \textbf{Appendix}).
 
@@
 
-For the CpG sites that pass the pre-screening step, a user-specified target
+For CpG sites that pass the pre-screening step, a user-specified target
 parameter of interest is estimated independently at each site. \emph{In all cases,
@@
 \citet{vdl2011targeted}.
-
-\hypertarget{future-work}{%
-\paragraph{Future Work}\label{future-work}}
-
-Support for continuous-valued treatments or exposures is \emph{planned and ongoing}
-but not yet available in stable releases of the \texttt{methyvim} software package.
-Future releases will allow users to assess continuous-valued treatments by
-relying on parameters estimable through implementations available in the
-following software packages:
-
-\begin{itemize}
-\item
-  A nonparametric variable importance measure (NPVI) \citep{chambaz2012estimation}
-  (R package \texttt{tmle.npvi}). The effect of a continuous-valued exposure or
-  treatment (the observed methylation at a target CpG site) on an outcome of
-  interest is estimated, controlling for the observed methylation value at all
-  other CpG sites in the same neighborhood as the target site. This uses a
-  parameter that compares values of the treatment against a user-specified
-  reference value taken to be the null value. In particular, the implementation
-  to be provided is designed to assess the effect of differential methylation at
-  the target CpG site on an outcome of interest (e.g., survival), providing a
-  nonparametric evaluation of the impact of methylation at the target site.
-\item
-  The causal effect of shifting the value of an observed intervention, defined
-  as the counterfactual outcome under a posited shift of a continuous-value
-  treatment of interest using stochastic intervention policies
-  \citep[\citet{diaz2018stochastic}, \citet{hejazi2018txshift}]{munoz2012population} (R package
-  \texttt{txshift}). The value of an outcome of interest under an unobserved value of
-  the (continuous-valued) treatment, specified through a user-provided
-  \emph{additive shift} of the observed treatment, may be data-adaptively estimated.
-  This allows for the effect of changes/shifts in the observed methylation at a
-  CpG site on an outcome of interest to be evaluated, providing a nonparametric
-  evalution of the relative importance of a particular target CpG site with
-  respect to another variable measured in the same study.
-\end{itemize}
 
@@
 
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+\hypertarget{data-structure}{%
+\subsection{Data Structure}\label{data-structure}}
+
+We consider an observed data structure, on a single experimental subject (e.g.,
+a patient), \(O = (W, A, (Y(j) : j))\), where \((Y(j) : j = 1, \ldots J)\) is the
+set of CpG sites measured by the assay in question, \(A \in \{0, 1\}\) represents
+a binary phenotype-level treatment, and \(W\) is a vector of the phenotype-level
+baseline covariates that are potential confounders (e.g., age, sex). We consider
+having access to measurements on a large number \(J\) of CpG sites (e.g.,
+\(850,000\), as measured by the Illumina \emph{MethylationEPIC} BeadChip arrays).
+Further, let \(K: j \to S_j\) be a procedure that assigns to a given CpG site \(j\)
+a set of neighbors \(S_j\) -- that is, \(S_j\) is a collection of indices of the
+neighbors of \(j\) just as \(j\) indexes the full set of CpG sites; moreover, since
+\(Y(j)\) is the measured methylation value for a given CpG site \(j\), \(Y(S_j)\) is
+the measured methylation values at the neighbors \(S_j\) of \(j\). Note that the
+definition of a neighborhood \(\{j, S_j\}\) is left vague so as to facilitate the
+use of user-specified strategies in implementation. We consider the case of
+observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where \(O \sim P \in \mathcal{M}\), which is to say that the random variable \(O\) is governed by an
+unknown probability distribution \(P\), assumed only to reside in a nonparametric
+statistical model \(\mathcal{M}\) that places no restrictions on the
+data-generating process.
+
+\hypertarget{variable-importance-measure}{%
+\subsection{Variable Importance Measure}\label{variable-importance-measure}}
+
+With the data structure above in hand, we let the estimand of interest be a
+\(j\)-specific variable importance measure (VIM) \(\Psi_j(P)\), which is defined
+through the \emph{true} (and unknown) probability distribution \(P\). As a motivating
+example, consider the case where \(\Psi_j(P)\) is
+\begin{equation}\label{vim_param}
+  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
+    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
+\end{equation}
+where \(Y(S_j)\) is the subvector of \(Y\) that contains the measured methylation
+values of the \emph{neighboring sites} of \(j\) (but not site \(j\) itself). This
+parameter of interest is a variant of the \emph{average treatment effect}, which has
+been the subject of much attention in statistical causal inference
+\citep[\citet{hahn1998role}, \citet{hirano2003efficient}]{holland1986statistics}. As a measure of
+variable importance, we seek to estimate this target parameter (\(\Psi_j(P)\)),
+which quantifies the effect of changing a treatment \(A\) on the methylation
+\(Y(j)\) of a CpG site \(j\), accounting for any potential confounding from
+phenotype-level covariates \(W\) and observed methylation \(Y(S_j)\) at the
+neighboring sites \(S_j\) of \(j\). We propose a procedure that estimates this
+target parameter \(\Psi_j(P)\) across all CpG sites of interest \(j: 1, \ldots, J\).
+When data-adaptive regression procedures (i.e., machine learning) are employed
+in estimating \(\Psi_j(P)\), this produces a nonparametric variable importance
+measure of differential methylation, allowing for the identification of
+differentially methylated positions (DMPs) while avoiding many assumptions
+common in the use of standard parametric regression procedures. We propose
+estimating \(\Psi_j(P)\) via targeted minimum loss-based estimation, which allows
+for data-adaptive regression procedures to be employed in a straightforward
+manner using the Super Learner algorithm for ensemble modeling \citep[
+\citet{wolpert1992stacked}, \citet{breiman1996stacked}, \citet{gruber2009targeted},
+\citet{vdl2011targeted}]{vdl2007super}.
+
+\hypertarget{pre-screening-procedure}{%
+\subsection{Pre-Screening Procedure}\label{pre-screening-procedure}}
+
+As a matter of practicality, we consider only estimating the target VIM
+\(\Psi_j(P)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in fact
+span the full set of assayed CpG sites. In order to determine this subset of CpG
+sites, we propose the use of a pre-screening procedure, which need be nothing
+more than a method for differential methylation analysis that is computationally
+less demanding than the method proposed here. Formally, let the pre-screening
+procedure \(\theta(j): j \to \{0, 1\}\), which is to say that \(\theta\) takes as
+input a CpG site \(j\) and returns a binary decision rule of whether to include
+CpG site \(j\) in a subset to be evaluated further or not. As an example, one
+might consider employing a classical differential methylation analysis
+procedure, such as the linear modeling approach of the
+\href{https://bioconductor.org/packages/limma}{\texttt{limma}} R package
+\citep[\citet{robinson2014statistical}]{smyth2004linear}, using only CpG sites that pass a
+reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
+p-values) for the subsequent analysis steps in the \texttt{methyvim} pipeline. Details
+on the implementation of pre-screening approaches is provided in the sequel.
+
+\hypertarget{reducing-neighbors-via-clustering}{%
+\subsection{Reducing Neighbors via Clustering}\label{reducing-neighbors-via-clustering}}
+
+Due to the data-adaptive nature of the regression procedures employed in
+evaluating the target VIM \(\Psi_j(P)\) via targeted minimum loss-based
+estimation, it is possible the a given CpG site \(j\) may have \emph{too many}
+neighbors \(S_j\) to be controlled for in the estimation procedure. Heuristically,
+the inclusion of too many neighbors when controlling for potential confounders
+may lead to instability in the estimates produced. In such cases, we propose and
+implement the use of a clustering technique (e.g., partitioning around medoids)
+to select a \emph{representative} subset of neighbors. Formally, there is likely no
+best choice of a specific clustering algorithm, so we leave this aspect of the
+proposed algorithm as flexible for the user \citep{kleinberg2003impossibility}. Note
+that the goal of employing a clustering procedure in \texttt{methyvim} is to obtain a
+smaller but still highly representative set of neighbors \(S(j)\), so as to allow
+for estimates of \(\Psi_j(P)\) to account for as much confounding from neighboring
+sites as is allowed by the available data.
+
+\hypertarget{targeted-minimum-loss-based-estimation-and-statistical-inference}{%
+\subsection{Targeted Minimum Loss-Based Estimation and Statistical Inference}\label{targeted-minimum-loss-based-estimation-and-statistical-inference}}
+
+Given the choice of target parameter \(\Psi_j(P)\), we propose the use of targeted
+minimum loss-based estimation (TMLE) to construct and evaluate an estimator
+\(\psi_{n, j}\) of \(\Psi_j(P)\). In the case of our motivating example, where the
+target VIM is based on the average treatment effect, we make use of a TML
+estimator of this parameter, which has been implemented for a general case in
+the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package
+\citep{gruber2011tmle}; users of \texttt{methyvim} may wish to consult the documentation of
+that software package as a supplement. Generally speaking, a TML estimator is
+constructed from a few simple components: (1) an estimator of the propensity
+score \citep{rosenbaum1983central}, often denoted by \(g(A \mid W)\); (2) an estimator
+of the outcome regression, often denoted \(Q(A, W)\); and (3) a targeting step
+applied to the initial estimators constructed from the aforementioned components
+\citep{gruber2009targeted}. While we describe a few of the key properties of TML
+estimators in the sequel, extended technical discussion is deferred to more
+comprehensive work \citep[\citet{vdl2018targeted}]{vdl2011targeted}. In order to construct
+an estimate \(\psi_{n, j}\) of \(\Psi_j(P)\), it is necessary to accurately estimate
+to nuisance parameters, these being the propensity score (\(g(A \mid W)\)) and the
+outcome regression (\(Q(A, W)\)); moreover, data-adaptive regression procedures
+may be used to obtain consistent estimates of these quantities through the
+creation of a stacked regression model via the Super Learner algorithm
+\citep[\citet{breiman1996stacked}, \citet{vdl2007super}]{wolpert1992stacked}, which ensures
+that the resulting ensemble learner satisfies important optimality properties
+with respect to cross-validation \citep[\citet{dudoit2005asymptotics},
+\citet{vdl2003unified}]{vdl2004asymptotic}. After constructing such estimates, an iterative procedure
+(i.e., the targeting step) may be used to combine these individual estimates
+into an estimate of the target parameter \(\Psi_j(P)\).
+
+Importantly, TML estimators are well-suited for statistical inference, having an
+asymptotically normal limiting distribution \citep[
+\citet{vdl2006targeted}]{tsiatis2007semiparametric}, which allows for a closed-form expression of the variance
+to be derived:
+\begin{equation}\label{eqn:lim_psi}
+  \sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2),
+\end{equation}
+where \(\sigma_{\text{EIF}}^2\) is the variance of the efficient influence
+function, with respect to a nonparametric statistical model, of the target
+parameter evaluated at the observed data. Such a convenience allows for
+confidence intervals and hypothesis tests to be constructed (i.e.,
+\(H_{0, j}: \Psi_j(P) = 0\)) in a straightforward manner.
+
+\hypertarget{correction-for-multiple-testing}{%
+\subsection{Correction for Multiple Testing}\label{correction-for-multiple-testing}}
+
+Given that we seek to estimate \(\Psi_j(P)\) for a possibly large number of CpG
+sites \((j: 1, \ldots J)\), the need to perform corrections for multiple testing
+is clear. In order to curb the potential for false discoveries, we recommend the
+use of the Benjamini and Hochberg procedure for controlling the False Discovery
+Rate (FDR) \citep{benjamini1995controlling}; however, as the proposed procedure
+involves a pre-screening step, naive application of the Benjamini and Hochberg
+procedure (BH) is invalid -- instead, we rely on a modification of the procedure
+to control the FDR, with established theoretical guarantees when pre-screening
+is employed \citep{tuglus2009modified}. In brief, the modified marginal Benjamini and
+Hochberg procedure to control the FDR under pre-screening works by applying the
+standard BH procedure to a padded vector of p-values -- that is, letting \(J\) be
+the number of CpG sites tested and \(K\) the number of CpG sites filtered out (so
+that \(J + K = P\), where \(P\) is the original dimension of the genomic assay), the
+modified marginal BH procedure is the application of the original BH procedure
+to a vector of p-values, composed of the \(J\) p-values from performing \(J\)
+hypothesis tests (of the form \(H_{0, j}: \Psi_j(P) = 0\)) and \(K\) additional
+p-values automatically set to \(1\) (for the hypothesis tests not performed on
+account of pre-screening). This procedure is guaranteed to control the FDR at
+the same desired rate when pre-screening is performed whereas naive application
+of the BH procedure fails to do so.
+
 \renewcommand\refname{References}

### 2018-08-18 README.Rmd
fixes and pkgdown updates

@@
 
-> Targeted Data-Adaptive Estimation and Inference for Differential Methylation
-> Analysis
+> Targeted, Robust, and Model-free Differential Methylation Analysis
 
@@
             and {van der Laan}, Mark J},
-          title = {methyvim: Targeted and model-free differential methylation
-            analysis in R},
+          title = {methyvim: Targeted, Robust, and Model-free Differential
+            Methylation Analysis in {R}},
           journal = {}

### 2018-08-18 R/methytmle-class.R

@@ -7,6 +7,7 @@
 #' @importClassesFrom SummarizedExperiment SummarizedExperiment
 #'  RangedSummarizedExperiment
 #' @importClassesFrom S4Vectors Vector Annotated
+#' @import BiocGenerics
 #'
 #' @export .methytmle
 #' @exportClass methytmle

### 2018-08-18 R/zzz.R

@@ -2,7 +2,6 @@
   packageStartupMessage(paste0(
     "methyvim v",
     utils::packageDescription("methyvim")$Version,
-    ": Targeted Variable Importance for ",
-    "Differential Methylation Analysis"
+    ": Targeted, Robust, and Model-free Differential Methylation Analysis"
   ))
 }

### 2018-08-20 paper/paper_BiocF1000.Rmd

@@
 ---
-title: '`methyvim`: Targeted, Robust, and Model-free Analysis of Differential
-  Methylation in R'
+title: '`methyvim`: Targeted, Robust, and Model-free Differential Methylation
+  Analysis in R'
 author:
@@
 keywords: DNA methylation, differential methylation, epigenetics, causal
-  inference, statistical variable importance, machine learning, targeted minimum
-  loss-based estimation
+  inference, variable importance, machine learning, targeted loss-based
+  estimation
 bibliography: references.bib
@@
 use of user-specified strategies in implementation. We consider the case of
-observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P \in
-\mathcal{M}$, which is to say that the random variable $O$ is governed by an
-unknown probability distribution $P$, assumed only to reside in a nonparametric
-statistical model $\mathcal{M}$ that places no restrictions on the
+observing $n$ iid copies of $O$ (i.e., $O_1, \ldots, O_n$), where $O \sim P_0
+\in \mathcal{M}$, which is to say that the random variable $O$ is governed by an
+unknown probability distribution $P_0$, assumed only to reside in a
+nonparametric statistical model $\mathcal{M}$ that places no restrictions on the
 data-generating process.
@@
 With the data structure above in hand, we let the estimand of interest be a
-$j$-specific variable importance measure (VIM) $\Psi_j(P)$, which is defined
-through the _true_ (and unknown) probability distribution $P$. As a motivating
-example, consider the case where $\Psi_j(P)$ is
+$j$-specific variable importance measure (VIM) $\Psi_j(P_0)$, which is defined
+through the _true_ (and unknown) probability distribution $P_0$. As a motivating
+example, consider the case where $\Psi_j(P_0)$ is
 \begin{equation}\label{vim_param}
-  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
-    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
+  \Psi_j(P_0) = \mathbb{E}_{P_0}(\mathbb{E}_{P_0}(Y(j) \mid A = 1, W, Y(S_j)) -
+    \mathbb{E}_{P_0} \mathbb{E}_{P_0}(Y(j) \mid A = 0, W, Y(S_j)),
 \end{equation}
@@
 [@holland1986statistics, @hahn1998role, @hirano2003efficient]. As a measure of
-variable importance, we seek to estimate this target parameter ($\Psi_j(P)$),
+variable importance, we seek to estimate this target parameter ($\Psi_j(P_0)$),
 which quantifies the effect of changing a treatment $A$ on the methylation
@@
 phenotype-level covariates $W$ and observed methylation $Y(S_j)$ at the
-neighboring sites $S_j$ of $j$. We propose a procedure that estimates this
-target parameter $\Psi_j(P)$ across all CpG sites of interest $j: 1, \ldots, J$.
-When data-adaptive regression procedures (i.e., machine learning) are employed
-in estimating $\Psi_j(P)$, this produces a nonparametric variable importance
-measure of differential methylation, allowing for the identification of
-differentially methylated positions (DMPs) while avoiding many assumptions
+neighboring sites $S_j$ of $j$. Importantly, the use of such well-defined
+parameters as variable importance measures allows the construction of
+inferential procedures (e.g., confidence intervals and hypothesis tests) for the
+estimate [@vdl2006statistical]. We propose a procedure that estimates this
+target parameter $\Psi_j(P_0)$ across all CpG sites of interest $j: 1, \ldots,
+J$. When data-adaptive regression procedures (i.e., machine learning) are
+employed in estimating $\Psi_j(P_0)$, this produces a nonparametric variable
+importance measure of differential methylation, allowing for the identification
+of differentially methylated positions (DMPs) while avoiding many assumptions
 common in the use of standard parametric regression procedures. We propose
@@
 for data-adaptive regression procedures to be employed in a straightforward
-manner using the Super Learner algorithm for ensemble modeling [@vdl2007super,
-@wolpert1992stacked, @breiman1996stacked, @gruber2009targeted,
+manner using the Super Learner algorithm for the construction of ensemble models
+[@vdl2007super, @wolpert1992stacked, @breiman1996stacked, @gruber2009targeted,
 @vdl2011targeted].
@@
 As a matter of practicality, we consider only estimating the target VIM
-$\Psi_j(P)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in fact
-span the full set of assayed CpG sites. In order to determine this subset of CpG
-sites, we propose the use of a pre-screening procedure, which need be nothing
-more than a method for differential methylation analysis that is computationally
-less demanding than the method proposed here. Formally, let the pre-screening
-procedure $\theta(j): j \to \{0, 1\}$, which is to say that $\theta$ takes as
-input a CpG site $j$ and returns a binary decision rule of whether to include
-CpG site $j$ in a subset to be evaluated further or not. As an example, one
-might consider employing a classical differential methylation analysis
-procedure, such as the linear modeling approach of the
+$\Psi_j(P_0)$ at a subset of CpG sites, so that $j: 1, \ldots J$ does not, in
+fact span the full set of assayed CpG sites. In order to determine this subset
+of CpG sites, we propose the use of a pre-screening procedure, which need be
+nothing more than a method for differential methylation analysis that is
+computationally less demanding than the method proposed here. Formally, let the
+pre-screening procedure $\theta(j): j \to \{0, 1\}$, which is to say that
+$\theta$ takes as input a CpG site $j$ and returns a binary decision rule of
+whether to include CpG site $j$ in a subset to be evaluated further or not. As
+an example, one might consider employing a classical differential methylation
+analysis procedure, such as the linear modeling approach of the
 [`limma`](https://bioconductor.org/packages/limma) R package
@@
 reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
-p-values) for the subsequent analysis steps in the `methyvim` pipeline. Details
-on the implementation of pre-screening approaches is provided in the sequel.
+p-values) for the subsequent analysis steps in the `methyvim` pipeline.
 
@@
 Due to the data-adaptive nature of the regression procedures employed in
-evaluating the target VIM $\Psi_j(P)$ via targeted minimum loss-based
-estimation, it is possible the a given CpG site $j$ may have _too many_
+evaluating the target VIM $\Psi_j(P_0)$ via targeted minimum loss-based
+estimation, it is possible that a given CpG site $j$ may have _too many_
 neighbors $S_j$ to be controlled for in the estimation procedure. Heuristically,
@@
 smaller but still highly representative set of neighbors $S(j)$, so as to allow
-for estimates of $\Psi_j(P)$ to account for as much confounding from neighboring
-sites as is allowed by the available data.
+for estimates of $\Psi_j(P_0)$ to account for as much confounding from
+neighboring sites as is allowed by the available data.
 
@@
 
-Given the choice of target parameter $\Psi_j(P)$, we propose the use of targeted
-minimum loss-based estimation (TMLE) to construct and evaluate an estimator
-$\psi_{n, j}$ of $\Psi_j(P)$. In the case of our motivating example, where the
-target VIM is based on the average treatment effect, we make use of a TML
-estimator of this parameter, which has been implemented for a general case in
-the [`tmle`](https://CRAN.R-project.org/package=tmle) R package
+Given the choice of target parameter $\Psi_j(P_0)$, we propose the use of
+targeted minimum loss-based estimation (TMLE) to construct and evaluate an
+estimator $\psi_{n, j}$ of $\Psi_j(P_0)$. In the case of our motivating example,
+where the target VIM is based on the average treatment effect, we make use of a
+TML estimator of this parameter, which has been implemented for a general case
+in the [`tmle`](https://CRAN.R-project.org/package=tmle) R package
 [@gruber2011tmle]; users of `methyvim` may wish to consult the documentation of
@@
 constructed from a few simple components: (1) an estimator of the propensity
-score [@rosenbaum1983central], often denoted by $g(A \mid W)$; (2) an estimator
-of the outcome regression, often denoted $Q(A, W)$; and (3) a targeting step
-applied to the initial estimators constructed from the aforementioned components
+score [@rosenbaum1983central], often denoted $g(A \mid W)$; (2) an estimator of
+the outcome regression, often denoted $\bar{Q}(A, W)$; and (3) a targeting step
+that revises the initial estimators of the aforementioned components such that
+the resultant estimator satisfies a set of score-like equations
 [@gruber2009targeted]. While we describe a few of the key properties of TML
@@
 comprehensive work [@vdl2011targeted, @vdl2018targeted]. In order to construct
-an estimate $\psi_{n, j}$ of $\Psi_j(P)$, it is necessary to accurately estimate
-to nuisance parameters, these being the propensity score ($g(A \mid W)$) and the
-outcome regression ($Q(A, W)$); moreover, data-adaptive regression procedures
-may be used to obtain consistent estimates of these quantities through the
-creation of a stacked regression model via the Super Learner algorithm
-[@wolpert1992stacked, @breiman1996stacked, @vdl2007super], which ensures
-that the resulting ensemble learner satisfies important optimality properties
-with respect to cross-validation [@vdl2004asymptotic, @dudoit2005asymptotics,
-@vdl2003unified]. After constructing such estimates, an iterative procedure
-(i.e., the targeting step) may be used to combine these individual estimates
-into an estimate of the target parameter $\Psi_j(P)$.
+an estimate $\psi_{n, j}$ of $\Psi_j(P_0)$, it is necessary to accurately
+estimate two nuisance parameters, these being the propensity score
+($g(A \mid W)$) and the outcome regression ($\bar{Q}(A, W)$); moreover,
+data-adaptive regression procedures may be used to obtain consistent estimates
+of these quantities through the creation of a stacked regression model via the
+Super Learner algorithm [@wolpert1992stacked, @breiman1996stacked,
+@vdl2007super], which ensures that the resulting ensemble model satisfies
+important optimality properties with respect to cross-validation
+[@vdl2004asymptotic, @dudoit2005asymptotics, @vdl2003unified]. To use our
+running example, one would construct a TML estimator of the average treatment
+effect (ATE) as a variable importance measure in the following short series of
+steps:
+
+1. Construct initial estimates of the propensity score $g(A \mid W)$ and outcome
+   regression $\bar{Q}(A, W)$. As noted previously, the Super Learner algorithm
+   may be employed to construct such estimates.
+2. Update these initial estimates (e.g., iteratively) so as to solve the
+   efficient influence function estimation equation of the parameter of interest
+   relative to a nonparametric statistical model.
+3. In the efficient influence function for the ATE, $\text{EIF}(P_0)(O) =
+   H_n(Y - \bar{Q}(A, W))$, the auxiliary term $H_n$ may be denoted as
+   $H_n = \frac{\mathbb{I}(A = 1)}{g(1 \mid W)} -
+   \frac{\mathbb{I}(A = 0)}{g(0 \mid W)}$. Note that the exact form of the
+   efficient influence function and auxiliary covariate varies with the target
+   parameter of interest.
+4. The TML estimator of the parameter of interest is constructed by a procedure
+   that updates the initial estimators such that the empirical mean of the
+   efficient influence function estimating equation(s) is nearly zero (with
+   respect to an error proportional to the sample size, i.e.,
+   $\frac{1}{\sqrt{n}}$).
 
@@
 \begin{equation}\label{eqn:lim_psi}
-  \sqrt{n}(\psi_{n, j} - \Psi_j(P)) \sim N(0, \sigma_{\text{EIF}}^2),
+  \sqrt{n}(\psi_{n, j} - \Psi_j(P_0)) \sim N(0, \sigma_{\text{EIF}}^2),
 \end{equation}
@@
 confidence intervals and hypothesis tests to be constructed (i.e.,
-$H_{0, j}: \Psi_j(P) = 0$) in a straightforward manner.
+$H_{0, j}: \Psi_j(P_0) = 0$) in a straightforward manner. What's more, under
+standard regularity conditions, and with consistent estimation of the propensity
+score and outcome regression, the TML estimator is asymptotically efficient and
+achieve the lowest possible variance among a large class of estimators
+[@bickel1993efficient].
 
@@
 
-Given that we seek to estimate $\Psi_j(P)$ for a possibly large number of CpG
+Given that we seek to estimate $\Psi_j(P_0)$ for a possibly large number of CpG
 sites $(j: 1, \ldots J)$, the need to perform corrections for multiple testing
@@
 to a vector of p-values, composed of the $J$ p-values from performing $J$
-hypothesis tests (of the form $H_{0, j}: \Psi_j(P) = 0$) and $K$ additional
+hypothesis tests (of the form $H_{0, j}: \Psi_j(P_0) = 0$) and $K$ additional
 p-values automatically set to $1$ (for the hypothesis tests not performed on

### 2018-08-20 paper/paper_BiocF1000.tex

@@
 
-\title{\texttt{methyvim}: Targeted, Robust, and Model-free Analysis of Differential Methylation in R}
+\title{\texttt{methyvim}: Targeted, Robust, and Model-free Differential Methylation Analysis in R}
 
@@
 \section*{Keywords}
-DNA methylation, differential methylation, epigenetics, causal inference, statistical variable importance, machine learning, targeted minimum loss-based estimation
+DNA methylation, differential methylation, epigenetics, causal inference, variable importance, machine learning, targeted loss-based estimation
 
@@
 \begin{verbatim}
-## methyvim v1.3.1: Targeted Variable Importance for Differential Methylation Analysis
+## methyvim v1.3.1: Targeted, Robust, and Model-free Differential Methylation Analysis
 \end{verbatim}
@@
 \end{Shaded}
+
+\begin{verbatim}
+## Loading required package: nnls
+\end{verbatim}
 
@@
 
-\begin{verbatim}
-## [1] 0.1 0.3 0.6 0.8 1.0
-\end{verbatim}
-
 \begin{center}\includegraphics{paper_BiocF1000_files/figure-latex/methyvim-heatmap-1} \end{center}
@@
 use of user-specified strategies in implementation. We consider the case of
-observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where \(O \sim P \in \mathcal{M}\), which is to say that the random variable \(O\) is governed by an
-unknown probability distribution \(P\), assumed only to reside in a nonparametric
-statistical model \(\mathcal{M}\) that places no restrictions on the
+observing \(n\) iid copies of \(O\) (i.e., \(O_1, \ldots, O_n\)), where \(O \sim P_0 \in \mathcal{M}\), which is to say that the random variable \(O\) is governed by an
+unknown probability distribution \(P_0\), assumed only to reside in a
+nonparametric statistical model \(\mathcal{M}\) that places no restrictions on the
 data-generating process.
@@
 With the data structure above in hand, we let the estimand of interest be a
-\(j\)-specific variable importance measure (VIM) \(\Psi_j(P)\), which is defined
-through the \emph{true} (and unknown) probability distribution \(P\). As a motivating
-example, consider the case where \(\Psi_j(P)\) is
+\(j\)-specific variable importance measure (VIM) \(\Psi_j(P_0)\), which is defined
+through the \emph{true} (and unknown) probability distribution \(P_0\). As a motivating
+example, consider the case where \(\Psi_j(P_0)\) is
 \begin{equation}\label{vim_param}
-  \Psi_j(P) = \mathbb{E}_P(\mathbb{E}_P(Y(j) \mid A = 1, W, Y(S_j)) -
-    \mathbb{E}_P \mathbb{E}_P(Y(j) \mid A = 0, W, Y(S_j)),
+  \Psi_j(P_0) = \mathbb{E}_{P_0}(\mathbb{E}_{P_0}(Y(j) \mid A = 1, W, Y(S_j)) -
+    \mathbb{E}_{P_0} \mathbb{E}_{P_0}(Y(j) \mid A = 0, W, Y(S_j)),
 \end{equation}
@@
 \citep[\citet{hahn1998role}, \citet{hirano2003efficient}]{holland1986statistics}. As a measure of
-variable importance, we seek to estimate this target parameter (\(\Psi_j(P)\)),
+variable importance, we seek to estimate this target parameter (\(\Psi_j(P_0)\)),
 which quantifies the effect of changing a treatment \(A\) on the methylation
@@
 phenotype-level covariates \(W\) and observed methylation \(Y(S_j)\) at the
-neighboring sites \(S_j\) of \(j\). We propose a procedure that estimates this
-target parameter \(\Psi_j(P)\) across all CpG sites of interest \(j: 1, \ldots, J\).
-When data-adaptive regression procedures (i.e., machine learning) are employed
-in estimating \(\Psi_j(P)\), this produces a nonparametric variable importance
-measure of differential methylation, allowing for the identification of
-differentially methylated positions (DMPs) while avoiding many assumptions
+neighboring sites \(S_j\) of \(j\). Importantly, the use of such well-defined
+parameters as variable importance measures allows the construction of
+inferential procedures (e.g., confidence intervals and hypothesis tests) for the
+estimate \citep{vdl2006statistical}. We propose a procedure that estimates this
+target parameter \(\Psi_j(P_0)\) across all CpG sites of interest \(j: 1, \ldots, J\). When data-adaptive regression procedures (i.e., machine learning) are
+employed in estimating \(\Psi_j(P_0)\), this produces a nonparametric variable
+importance measure of differential methylation, allowing for the identification
+of differentially methylated positions (DMPs) while avoiding many assumptions
 common in the use of standard parametric regression procedures. We propose
@@
 for data-adaptive regression procedures to be employed in a straightforward
-manner using the Super Learner algorithm for ensemble modeling \citep[
-\citet{wolpert1992stacked}, \citet{breiman1996stacked}, \citet{gruber2009targeted},
+manner using the Super Learner algorithm for the construction of ensemble models
+\citep[\citet{wolpert1992stacked}, \citet{breiman1996stacked}, \citet{gruber2009targeted},
 \citet{vdl2011targeted}]{vdl2007super}.
@@
 As a matter of practicality, we consider only estimating the target VIM
-\(\Psi_j(P)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in fact
-span the full set of assayed CpG sites. In order to determine this subset of CpG
-sites, we propose the use of a pre-screening procedure, which need be nothing
-more than a method for differential methylation analysis that is computationally
-less demanding than the method proposed here. Formally, let the pre-screening
-procedure \(\theta(j): j \to \{0, 1\}\), which is to say that \(\theta\) takes as
-input a CpG site \(j\) and returns a binary decision rule of whether to include
-CpG site \(j\) in a subset to be evaluated further or not. As an example, one
-might consider employing a classical differential methylation analysis
-procedure, such as the linear modeling approach of the
+\(\Psi_j(P_0)\) at a subset of CpG sites, so that \(j: 1, \ldots J\) does not, in
+fact span the full set of assayed CpG sites. In order to determine this subset
+of CpG sites, we propose the use of a pre-screening procedure, which need be
+nothing more than a method for differential methylation analysis that is
+computationally less demanding than the method proposed here. Formally, let the
+pre-screening procedure \(\theta(j): j \to \{0, 1\}\), which is to say that
+\(\theta\) takes as input a CpG site \(j\) and returns a binary decision rule of
+whether to include CpG site \(j\) in a subset to be evaluated further or not. As
+an example, one might consider employing a classical differential methylation
+analysis procedure, such as the linear modeling approach of the
 \href{https://bioconductor.org/packages/limma}{\texttt{limma}} R package
@@
 reasonable cutoff based on the linear model (e.g., magnitude of t-statistics,
-p-values) for the subsequent analysis steps in the \texttt{methyvim} pipeline. Details
-on the implementation of pre-screening approaches is provided in the sequel.
+p-values) for the subsequent analysis steps in the \texttt{methyvim} pipeline.
 
@@
 Due to the data-adaptive nature of the regression procedures employed in
-evaluating the target VIM \(\Psi_j(P)\) via targeted minimum loss-based
-estimation, it is possible the a given CpG site \(j\) may have \emph{too many}
+evaluating the target VIM \(\Psi_j(P_0)\) via targeted minimum loss-based
+estimation, it is possible that a given CpG site \(j\) may have \emph{too many}
 neighbors \(S_j\) to be controlled for in the estimation procedure. Heuristically,
@@
 smaller but still highly representative set of neighbors \(S(j)\), so as to allow
-for estimates of \(\Psi_j(P)\) to account for as much confounding from neighboring
-sites as is allowed by the available data.
+for estimates of \(\Psi_j(P_0)\) to account for as much confounding from
+neighboring sites as is allowed by the available data.
 
@@
 
-Given the choice of target parameter \(\Psi_j(P)\), we propose the use of targeted
-minimum loss-based estimation (TMLE) to construct and evaluate an estimator
-\(\psi_{n, j}\) of \(\Psi_j(P)\). In the case of our motivating example, where the
-target VIM is based on the average treatment effect, we make use of a TML
-estimator of this parameter, which has been implemented for a general case in
-the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package
+Given the choice of target parameter \(\Psi_j(P_0)\), we propose the use of
+targeted minimum loss-based estimation (TMLE) to construct and evaluate an
+estimator \(\psi_{n, j}\) of \(\Psi_j(P_0)\). In the case of our motivating example,
+where the target VIM is based on the average treatment effect, we make use of a
+TML estimator of this parameter, which has been implemented for a general case
+in the \href{https://CRAN.R-project.org/package=tmle}{\texttt{tmle}} R package
 \citep{gruber2011tmle}; users of \texttt{methyvim} may wish to consult the documentation of
@@
 constructed from a few simple components: (1) an estimator of the propensity
-score \citep{rosenbaum1983central}, often denoted by \(g(A \mid W)\); (2) an estimator
-of the outcome regression, often denoted \(Q(A, W)\); and (3) a targeting step
-applied to the initial estimators constructed from the aforementioned components
+score \citep{rosenbaum1983central}, often denoted \(g(A \mid W)\); (2) an estimator of
+the outcome regression, often denoted \(\bar{Q}(A, W)\); and (3) a targeting step
+that revises the initial estimators of the aforementioned components such that
+the resultant estimator satisfies a set of score-like equations
 \citep{gruber2009targeted}. While we describe a few of the key properties of TML
@@
 comprehensive work \citep[\citet{vdl2018targeted}]{vdl2011targeted}. In order to construct
-an estimate \(\psi_{n, j}\) of \(\Psi_j(P)\), it is necessary to accurately estimate
-to nuisance parameters, these being the propensity score (\(g(A \mid W)\)) and the
-outcome regression (\(Q(A, W)\)); moreover, data-adaptive regression procedures
-may be used to obtain consistent estimates of these quantities through the
-creation of a stacked regression model via the Super Learner algorithm
-\citep[\citet{breiman1996stacked}, \citet{vdl2007super}]{wolpert1992stacked}, which ensures
-that the resulting ensemble learner satisfies important optimality properties
-with respect to cross-validation \citep[\citet{dudoit2005asymptotics},
-\citet{vdl2003unified}]{vdl2004asymptotic}. After constructing such estimates, an iterative procedure
-(i.e., the targeting step) may be used to combine these individual estimates
-into an estimate of the target parameter \(\Psi_j(P)\).
+an estimate \(\psi_{n, j}\) of \(\Psi_j(P_0)\), it is necessary to accurately
+estimate two nuisance parameters, these being the propensity score
+(\(g(A \mid W)\)) and the outcome regression (\(\bar{Q}(A, W)\)); moreover,
+data-adaptive regression procedures may be used to obtain consistent estimates
+of these quantities through the creation of a stacked regression model via the
+Super Learner algorithm \citep[\citet{breiman1996stacked},
+\citet{vdl2007super}]{wolpert1992stacked}, which ensures that the resulting ensemble model satisfies
+important optimality properties with respect to cross-validation
+\citep[\citet{dudoit2005asymptotics}, \citet{vdl2003unified}]{vdl2004asymptotic}. To use our
+running example, one would construct a TML estimator of the average treatment
+effect (ATE) as a variable importance measure in the following short series of
+steps:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Construct initial estimates of the propensity score \(g(A \mid W)\) and outcome
+  regression \(\bar{Q}(A, W)\). As noted previously, the Super Learner algorithm
+  may be employed to construct such estimates.
+\item
+  Update these initial estimates (e.g., iteratively) so as to solve the
+  efficient influence function estimation equation of the parameter of interest
+  relative to a nonparametric statistical model.
+\item
+  In the efficient influence function for the ATE, \(\text{EIF}(P_0)(O) = H_n(Y - \bar{Q}(A, W))\), the auxiliary term \(H_n\) may be denoted as
+  \(H_n = \frac{\mathbb{I}(A = 1)}{g(1 \mid W)} - \frac{\mathbb{I}(A = 0)}{g(0 \mid W)}\). Note that the exact form of the
+  efficient influence function and auxiliary covariate varies with the target
+  parameter of interest.
+\item
+  The TML estimator of the parameter of interest is constructed by a procedure
+  that updates the initial estimators such that the empirical mean of the
+  efficient influence function estimating equation(s) is nearly zero (with
+  respect to an error proportional to the sample size, i.e.,
+  \(\frac{1}{\sqrt{n}}\)).
+\end{enumerate}
 
@@
 confidence intervals and hypothesis tests to be constructed (i.e.,
-\(H_{0, j}: \Psi_j(P) = 0\)) in a straightforward manner.
+\(H_{0, j}: \Psi_j(P_0) = 0\)) in a straightforward manner. What's more, under
+standard regularity conditions, and with consistent estimation of the propensity
+score and outcome regression, the TML estimator is asymptotically efficient and
+achieve the lowest possible variance among a large class of estimators
+\citep{bickel1993efficient}.
 
@@
 
-Given that we seek to estimate \(\Psi_j(P)\) for a possibly large number of CpG
+Given that we seek to estimate \(\Psi_j(P_0)\) for a possibly large number of CpG
 sites \((j: 1, \ldots J)\), the need to perform corrections for multiple testing
@@
 to a vector of p-values, composed of the \(J\) p-values from performing \(J\)
-hypothesis tests (of the form \(H_{0, j}: \Psi_j(P) = 0\)) and \(K\) additional
+hypothesis tests (of the form \(H_{0, j}: \Psi_j(P_0) = 0\)) and \(K\) additional
 p-values automatically set to \(1\) (for the hypothesis tests not performed on

### 2018-08-21 README.Rmd
update DOI badge in readme and pkgdown

@@
 [![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
-[![DOI](https://zenodo.org/badge/79256902.svg)](https://zenodo.org/badge/latestdoi/79256902)
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1401298.svg)](https://doi.org/10.5281/zenodo.1401298)
 

### 2018-08-28 README.Rmd
add bioc badge + update site

--- 2018-08-18 README.Rmd
@@
 [![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
+[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/methyvim.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/methyvim)
 [![Bioc Time](http://bioconductor.org/shields/years-in-bioc/methyvim.svg)](https://bioconductor.org/packages/release/bioc/html/methyvim.html)

### 2018-09-27 README.Rmd
update citation

@@
         @article{hejazi2018methyvim,
-          doi = {},
-          url = {},
+          doi = {10.12688/f1000research.16047.1},
+          url = {https://dx.doi.org/10.12688/f1000research.16047.1},
           year = {2018},
-          month = {},
-          publisher = {},
-          volume = {},
+          publisher = {Faculty of 1000 Ltd},
+          volume = {7},
+          number = {1424},
           author = {Hejazi, Nima S and Phillips, Rachael V and Hubbard, Alan E
             and {van der Laan}, Mark J},
-          title = {methyvim: Targeted, Robust, and Model-free Differential
-            Methylation Analysis in {R}},
-          journal = {}
+          title = {{methyvim}: Targeted, robust, and model-free differential
+            methylation analysis in {R}},
+          journal = {F1000Research}
         }

### 2018-10-31 README.Rmd
update docs

@@
 framework plays in statistical causal inference, the recommended references are
-@vdl2011targeted and @vdl2018targeted. @hernan2018causal and @pearl2009causality
+@vdl2011targeted and @vdl2018targeted. @hernan2018causal and @pearl2000causality
 may be of interest to those desiring a more general introduction to statistical
@@
 ```{r bioc-installation, eval = FALSE}
-if (!("BiocManager" %in% installed.packages())) {
+if (!requireNamespace("BiocManager", quietly=TRUE)) {
   install.packages("BiocManager")

### 2019-09-14 README.Rmd
authorship updates
pkgdown fixes

@@
   rmarkdown::github_document
-bibliography: "README-refs.bib"
+bibliography: "inst/REFERENCES.bib"
 ---
@@
 
-__Author:__ [Nima Hejazi](https://nimahejazi.org)
+__Author:__ [Nima Hejazi](https://nimahejazi.org) and [Mark van der
+  Laan](https://vanderlaan-lab.org)
 
@@
 
-&copy; 2017-2018 [Nima S. Hejazi](https://nimahejazi.org)
+&copy; 2017-2019 [Nima S. Hejazi](https://nimahejazi.org)
 

### 2019-09-14 sandbox/1_setup_data.R

@@ -1,87 +1,0 @@
-# needed pkgs
-library(here)
-library(minfi)
-library(foreach)
-library(parallel)
-library(doParallel)
-n_cores <- detectCores()
-registerDoParallel(n_cores)
-
-# check system, makes sure you're on bluevelvet
-stopifnot(
-  stringr::str_split(Sys.info()["nodename"], "\\.")[[1]][1] == "bluevelvet"
-)
-
-# load sample data for playing around
-load(here("..", "data", "tmleMethyData", "epigen_grs.RData"))
-
-# methyvim args, defaults that are inputted into the main functions
-exp_var_int = 1 # binary exposure (for ATE)
-# column variables in the coldata matrix
-out_var_int = 2 # continuous outcome (for NPVI)
-window_bp = 1e3
-corr_max = 0.35
-# maximum correlation that a neighboring site can have with the target site
-obs_per_covar = 15
-# how many obs you need for every covariate you include in W (20/covariate)
-# otherwise you have insufficient data to control for all of this
-type = "Mval"
-filter = "limma"
-# one based on npvi and another for wilsons package, pass object into screening
-# appropriate function and sites that pass are the ones we analyze
-cutoff = 0.05 # defines what passes through the filter, p-value
-# limma might want to use t-stat
-# npvi use p-value
-# data adaptive w wilon might want to use rank of site w/i final table produced
-parallel = TRUE
-return_ic = FALSE # ic estimates out, slot in methytmle object
-shrink_ic = FALSE # apply limma to reduce variance in ic based estimates
-# at that point you wont care about values from tmle estimation procedure
-# ic based estimate for each site and variance of these sites is variance of
-# efficient influence curve, so each site is distributed w mean (psi)
-# then do t-test w the expectation with the ate
-# (controlling erratic spikes in the t stat based on low variance)
-family = "binomial"  # for logistic fluctuation model with scaled Y
-# or gaussian
-g_lib = c("SL.mean", "SL.glm", "SL.randomForest") #sensible defaults
-Q_lib = c("SL.mean", "SL.randomForest")
-npvi_cutoff = 0.25
-# keep a continuous (ate discretizes things) but you have to map some range of
-# the values to zero (null range with enough observations)
-# cutoff defines the quantile in the obs data that you'll use to map to zero
-# antione had long discussion about this and theres an optimal way to deciding
-# this cutoff
-npvi_conf = 0.95
-npvi_descr = list(f = identity, iter = 10, cvControl = 2, nMax = 30,
[…]

### 2019-09-14 sandbox/2_methyvim_funs.R

@@ -1,136 +1,0 @@
-# define methytmle class, core class in the package
-.methytmle <- methods::setClass(
-       Class = "methytmle",
-       slots = list(call = "call",  # so the user can remember the args they pass in
-                    screen_ind = "numeric",
-                    clusters = "numeric",
-                    g = "matrix",
-                    Q = "matrix",
-                    ic = "data.frame",
-                    vim = "data.frame"),
-       contains = "GenomicRatioSet"
-)
-
-# set up methyvim object, hacky to create something of class call
-call <- "testing123"
-class(call) <- "call" # hacking the call object
-methy_tmle <- .methytmle(catch_inputs_ate$data)
-
-#data was originally a genomic ratio set now its a methytmle object
-methy_tmle@call <- call
-
-# using LIMMA for screening
-limma_screen <- function(methytmle, var_int, type, cutoff = 0.05) {
-  # setup design matrix
-  design <- as.numeric(colData(methytmle)[, var_int])
-  design <- as.matrix(cbind(rep(1, times = length(design)), design))
-
-  # create expression object for modeling
-  if (type == "Beta") {
-    methytmle_exprs <- minfi::getBeta(methytmle)
-  } else if (type == "Mval") {
-    methytmle_exprs <- minfi::getM(methytmle)
-  }
-
-  # fit limma model and apply empirical Bayes shrinkage
-  mod_fit <- limma::lmFit(object = methytmle_exprs, design = design)
-  mod_fit <- limma::eBayes(mod_fit)
-
-  # extract indices of relevant CpG sites
-  tt_out <- limma::topTable(mod_fit, coef = 2, num = Inf, sort.by = "none")
-  indices_pass <- which(tt_out$P.Value < cutoff)
-
-  #takes in methytmle object and returns same but with screen slot filled
-
-  # add to appropriate slot in the methytmle input object
-  methytmle@screen_ind <- indices_pass
-  return(methytmle)
-}
-
-# function to cluster sites
-cluster_sites <- function(methy_tmle, window_size = 1000) {
-  gr <- SummarizedExperiment::rowRanges(methy_tmle)
-  pos <- BiocGenerics::start(IRanges::ranges(gr))
-  clusters <- bumphunter::clusterMaker(chr = GenomeInfoDb::seqnames(gr),
-                                       pos = pos,
-                                       assumeSorted = FALSE,
-                                       maxGap = window_size)
-  methy_tmle@clusters <- as.numeric(clusters)
[…]

### 2019-09-15 README.Rmd
readme cleanup

@@
 
-__Author:__ [Nima Hejazi](https://nimahejazi.org) and [Mark van der
+__Authors:__ [Nima Hejazi](https://nimahejazi.org) and [Mark van der
   Laan](https://vanderlaan-lab.org)
@@
 `methyvim` is an R package that provides facilities for differential methylation
-analysis based on _variable importance measures_ (VIMs), a class of
-statistically estimable target parameters that arise in causal inference.
+analysis based on _variable importance measures_ (VIMs), statistical target
+parameters inspired by causal inference.
 
-The statistical methodology implemented computes targeted minimum loss-based
-estimates of several well-characterized variable importance measures:
+The statistical methodology implemented computes targeted minimum loss estimates
+of several well-characterized variable importance measures:
 
@@
 
-For continous-valued treatments or exposures:
+For continuous-valued treatments or exposures (_WIP: support planned_):
 
@@
 
-_In all cases, an estimator of the target parameter is constructed via targeted
-minimum loss-based estimation._
-
 These methods allow differential methylation effects to be quantified in a
-manner that is largely free of assumptions, especially of the variety exploited
+manner that is largely assumption-free, especially of the variety exploited
 in parametric models. __The statistical algorithm consists in several major
@@
   multiple testing correction uniquely suited to such settings. Due to the
-multiple testing nature of the estimation problem, a variant of the Benjamini &
-Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
+  multiple testing nature of the estimation problem, a variant of the Benjamini
+  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
   [@benjamini1995controlling]. Specifically, we apply the modified marginal
@@
 
-For a general discussion of the framework of targeted minimum loss-based
-estimation, the many applications of this methodology, and the role the
-framework plays in statistical causal inference, the recommended references are
-@vdl2011targeted and @vdl2018targeted. @hernan2018causal and @pearl2000causality
-may be of interest to those desiring a more general introduction to statistical
-causal inference.
+For a general discussion of the framework of targeted minimum loss estimation
+and its myriad applications, the canonical references are @vdl2011targeted and
+@vdl2018targeted. @hernan2018causal and @pearl2000causality may be of interest
+to those desiring a more general introduction to statistical causal inference.
 

### 2019-10-05 README.Rmd
readme/pkgdown updates + appveyor tweak

@@
 __Authors:__ [Nima Hejazi](https://nimahejazi.org) and [Mark van der
-  Laan](https://vanderlaan-lab.org)
+  Laan](https://vanderlaan-lab.org/about)
 
@@
 To contribute, install the bleeding-edge _development version_ from GitHub via
-[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):
+[`remotes`](https://CRAN.R-project.org/package=remotes):
 
 ```{r gh-master-installation, eval = FALSE}
-devtools::install_github("nhejazi/methyvim")
+remotes::install_github("nhejazi/methyvim")
 ```
@@
 ```{r gh-develop-installation, eval = FALSE}
-devtools::install_github("nhejazi/methyvim", ref = "RELEASE_3_6")
+remotes::install_github("nhejazi/methyvim", ref = "RELEASE_3_6")
 ```
@@
 
+        @manual{hejazi2019methyvimbioc,
+          author = {Hejazi, Nima S and {van der Laan}, Mark J},
+          title = {{methyvim}: Targeted, robust, and model-free differential
+            methylation analysis},
+          doi = {10.18129/B9.bioc.methyvim},
+          url = {https://bioconductor.org/packages/methyvim},
+          note = {R package version 1.6.0}
+        }
+
 ---

### 2019-10-29 README.Rmd
catching up with tmle package changes

@@
           url = {https://bioconductor.org/packages/methyvim},
-          note = {R package version 1.6.0}
+          note = {R package version 1.8.0}
         }

### 2019-10-29 R/utils.R

@@ -45,8 +45,7 @@ set_parallel <- function(parallel = c(TRUE, FALSE),
     }
   } else if (parallel == FALSE) {
     warning(paste(
-      "Sequential evaluation is strongly discouraged.",
-      "\n Proceed with caution."
+      "Sequential evaluation over many probes may take a long time."
     ))
     future::plan(future::sequential)
   }
@@ -74,9 +73,13 @@ set_parallel <- function(parallel = c(TRUE, FALSE),
 #' @param fun A \code{function} to be wrapped in a \code{try} statement.
 #' @param ... Additional arguments passed to the previous argument \code{fun}.
 wrap_in_try <- function(fun, ...) {
-  wrapped <- function(...)
-    try({
-      fun(...)
-    }, silent = TRUE)
+  wrapped <- function(...) {
+    try(
+      {
+        fun(...)
+      },
+      silent = TRUE
+    )
+  }
   return(wrapped)
 }
